### 2014-09-17 paper.tex
added preliminary draft of abstract

\documentclass[conference]{IEEEtran}
\usepackage[pdftex]{graphicx}
\usepackage[cmex10]{amsmath}
\interdisplaylinepenalty=2500
\usepackage{algorithmic}
\usepackage{array}
\usepackage{mdwmath}
\usepackage{mdwtab}
\usepackage{eqparbox}
\usepackage[font=footnotesize]{subfig}
\usepackage{fixltx2e}
\usepackage{url}
\hyphenation{op-tical net-works semi-conduc-tor}

% additional packages and utility commands
\usepackage{minted}
% \usepackage{flushend}

% \usepackage{lineno}
% \setlength\linenumbersep{1mm}
% \linenumbers

\begin{document}

% to avoid syntax error highlighting - e.g. foo's not js ;-)
\expandafter\def\csname PY@tok@err\endcsname{}

\title{A Domain Specific Language for Functional Code Fusion of
       Intrusion Detection Algorithms in Wireless Sensor Networks}

\author{\IEEEauthorblockN{Christophe Van Ginneken, Jef Maerien, Christophe
Huygens, Danny Hughes, Wouter Joosen}%
\IEEEauthorblockA{iMinds - DistriNet - KU Leuven\\
B-3001, Leuven, Belgium\\
\{firstname.lastname\}@cs.kuleuven.be}}

\maketitle

\begin{abstract}

Introducing intrusion detection in a wireless sensor network puts a strain on
the limited resources of its nodes. Fusing the underlying algorithms' code can
alleviate this, but the manual and repetitive labor proves to an unacceptable
economic burden. Intrusion detection algorithms share common functionality and
therefore are a prime candidate for applying a domain specific language. In
this paper we propose such a domain specific language to describe these
algorithms in a formal and functional way. This allows for the automation of
the fusion of these algorithms on a functional level though code generation.
The code, produced by a prototype code generator, shows to have less impact on
e.g. the execution time of the resulting intrusion detection system and its use
of the wireless network.

\end{abstract}

\bibliographystyle{IEEEtran}
\bibliography{referenties}

\end{document}

### 2014-09-22 paper.tex
more minor changes
more abstract touch-ups + preliminary structure

@@
 
-\title{A Domain Specific Language for Functional Code Fusion of
+\title{A Domain Specific Language for Functional Code Fusion for
        Intrusion Detection Algorithms in Wireless Sensor Networks}
@@
 
-Introducing intrusion detection in a wireless sensor network puts a strain on
-the limited resources of its nodes. Fusing the underlying algorithms' code can
-alleviate this, but the manual and repetitive labor proves to an unacceptable
-economic burden. Intrusion detection algorithms share common functionality and
-therefore are a prime candidate for applying a domain specific language. In
-this paper we propose such a domain specific language to describe these
-algorithms in a formal and functional way. This allows for the automation of
-the fusion of these algorithms on a functional level though code generation.
-The code, produced by a prototype code generator, shows to have less impact on
-e.g. the execution time of the resulting intrusion detection system and its use
-of the wireless network.
+Introducing intrusion detection in wireless sensor network (WSNs) puts a strain
+on the limited resources of its nodes. Fusion of the underlying algorithms can
+alleviate this, but the manual and repetitive labor proves to be an
+unacceptable economic burden and is error-prone. Intrusion detection algorithms
+share common functionality and structure, and therefore are a prime candidate
+for the application of a domain specific language (DSL). In this paper we
+propose such a DSL to describe these algorithms in a formal and functional way.
+This allows for the automation of the fusion of these algorithms on a
+functional level though code generation. The code, produced by a prototype code
+generator implementation, shows to have less impact on e.g. the execution time
+of the resulting intrusion detection system and its use of the wireless
+network, in comparison with a basic side-by-side implementation of the same
+algorithms.
 
@@
 
-\bibliographystyle{IEEEtran}
-\bibliography{referenties}
+\section{Introduction}
+
+WSN
+
+IDS as a non-functional cost
+
+Economics of development
+
+Functional Code Fusion
+
+The remainder of this paper proceeds as follows: ...
+
+\section{Related Work}
+
+DSL for WSN
+
+DSL for IDS
+
+Code Generation for WSN
+
+Code Optimization for embedded systems/WSN
+
+Code Generation for IDS
+
+\section{Background}
+
+Theory of Loop Fusion
+
+Examples of code/loop fusion (e.g. Haskell Stream Fusion)
+
+Function Inlining (e.g. TinyOS)
+
+\section{Problem Analysis}
+
+\section{Evaluation}
+
+network usages: number of packets, number of bytes
+
+processing time
+
+image size
+
+TODO: memory usage
+
+2 algorithms
+TODO: adding 3rd ?
+
+\section{Conclusions and Future Work}
+
+theoretical gains are met in practice
+
+\section*{Acknowledgements}
+
+We would like to thank the reviewers for their thoughtful and helpful comments
+that enhanced the readability of this paper. Our gratitude and respect also
+goes out to all members of the NES task force at KU Leuven/DistriNet for
+creating the nurturing environment where these ideas could grow.
+
+% \bibliographystyle{IEEEtran}
+% \bibliography{referenties}
 

### 2014-09-22 paper.tex
added initial references

@@
 
-DSL for WSN
+DSL for WSN \cite{naumowicz2009prototyping,levis2004tinyscript}
 
-DSL for IDS
+DSL for IDS \cite{eckmann2002statl}
 
-Code Generation for WSN
+Code Generation for embedded systems/WSNs \cite{leupers2000code,marwedel2002code}
 
-Code Optimization for embedded systems/WSN
+Code Optimization for embedded systems/WSNs \cite{panda2001data,naik2001software}
 
-Code Generation for IDS
+Code Generation for IDS \cite{charitakis2003code}
 
@@
 
-Theory of Loop Fusion
+Theory of Loop Fusion \cite{darte2000complexity}
 
-Examples of code/loop fusion (e.g. Haskell Stream Fusion)
+Examples of code/loop fusion (e.g. Haskell Stream Fusion) \cite{coutts2007stream}
 
-Function Inlining (e.g. TinyOS)
+Function Inlining (e.g. TinyOS) \cite{gay2007software,gay2003nesc}
 
 \section{Problem Analysis}
+
+TODO
+
+\section{Problem Solution}
+
+TODO
 
@@
 
-TODO: memory usage
+TODO: memory usage ?
 
 2 algorithms
-TODO: adding 3rd ?
+
+TODO: add 3rd ?
 
@@
 theoretical gains are met in practice
+
+explore more domains + implement corresponding DSLs for embedded systems
+
+extract generic language as host for DSLs
 
@@
 
-% \bibliographystyle{IEEEtran}
-% \bibliography{referenties}
+\bibliographystyle{IEEEtran}
+\bibliography{literature/referenties}
 

### 2014-09-23 paper.tex
added more references
updated affiliation + acknowledgements

@@
 Huygens, Danny Hughes, Wouter Joosen}%
-\IEEEauthorblockA{iMinds - DistriNet - KU Leuven\\
-B-3001, Leuven, Belgium\\
+\IEEEauthorblockA{iMinds-DistriNet, KU Leuven\\
+3001 Leuven, Belgium\\
 \{firstname.lastname\}@cs.kuleuven.be}}
@@
 
-WSN
+WSN \cite{baggio2005wireless,werner2005monitoring,hughes2006gridstix}
 
-IDS as a non-functional cost
+Economics of development, IDS as a non-functional cost \cite{lee2002toward}
 
-Economics of development
+Energy in WSNs \cite{hughes2013energy}
 
@@
 \section{Related Work}
+
+IDS in WSN \cite{perrig2004security,mishra2004intrusion}
+
+DSL \cite{fowler2010domain,mernik2005and}
 
@@
 
-2 algorithms
+2 algorithms \cite{ganeriwal2008reputation}
 
-TODO: add 3rd ?
+TODO: add 3rd ? \cite{krontiris2009cooperative}
 
@@
 
-We would like to thank the reviewers for their thoughtful and helpful comments
+This research is partially funded by the Interuniversity Attraction Poles
+Programme Belgian State, Belgian Science Policy, and by the Research Fund KU
+Leuven. This research is partially funded by the EU FP7 project NESSoS. We
+would like to thank the reviewers for their thoughtful and helpful comments
 that enhanced the readability of this paper. Our gratitude and respect also

### 2014-09-29 paper.tex
some last-minute details

@@
 share common functionality and structure, and therefore are a prime candidate
-for the application of a domain specific language (DSL). In this paper we
-propose such a DSL to describe these algorithms in a formal and functional way.
-This allows for the automation of the fusion of these algorithms on a
-functional level though code generation. The code, produced by a prototype code
-generator implementation, shows to have less impact on e.g. the execution time
-of the resulting intrusion detection system and its use of the wireless
-network, in comparison with a basic side-by-side implementation of the same
-algorithms.
+for the application of a domain specific language (DSL). This paper propose
+such a DSL to describe these algorithms in a formal and functional way. This
+allows for the automation of the fusion of these algorithms on a functional
+level though code generation. The more optimally fused code, produced by a
+prototype code generator implementation, shows to have less impact on e.g. the
+execution time of the resulting intrusion detection system and its use of the
+wireless network, in comparison with a basic side-by-side implementation of the
+same algorithms.
 

### 2014-09-29 paper.tex
rewrote title and abstract

--- 2014-09-23 paper.tex
@@
 
-\title{A Domain Specific Language for Functional Code Fusion for
-       Intrusion Detection Algorithms in Wireless Sensor Networks}
+\title{Efficiently Detecting Intrusions\\in the Internet of Things}
 
@@
 
-Introducing intrusion detection in wireless sensor network (WSNs) puts a strain
-on the limited resources of its nodes. Fusion of the underlying algorithms can
-alleviate this, but the manual and repetitive labor proves to be an
-unacceptable economic burden and is error-prone. Intrusion detection algorithms
-share common functionality and structure, and therefore are a prime candidate
-for the application of a domain specific language (DSL). In this paper we
-propose such a DSL to describe these algorithms in a formal and functional way.
-This allows for the automation of the fusion of these algorithms on a
-functional level though code generation. The code, produced by a prototype code
-generator implementation, shows to have less impact on e.g. the execution time
-of the resulting intrusion detection system and its use of the wireless
-network, in comparison with a basic side-by-side implementation of the same
-algorithms.
+Supporting multiple intrusion detection (ID) algorithms imposes significant
+overhead on Internet of Things (IoT) devices. Fusion of the underlying
+algorithms can alleviate this, yet, proves to be time-consuming, repetitive and
+error-prone. To address this problem we introduce id-foo, a framework for the
+development of efficient intrusion detection systems (IDS). It consists of a
+domain specific language (DSL) that allows describing ID algorithms in a formal
+and functional way. The id-foo code generator then fuses these algorithms on a
+functional level and produces more optimally organised source code. A
+side-by-side comparison shows that id-foo-based code significantly reduces
+memory footprint, message passing overhead and execution time in comparison to
+a straightforward and sequential implementation of the same algorithms.
 

### 2014-10-01 paper.tex
added structure in comments
first very rough draft for introduction

@@
 \usepackage{url}
-\hyphenation{op-tical net-works semi-conduc-tor}
+\hyphenation{}
 
@@
 \usepackage{minted}
-% \usepackage{flushend}
+\usepackage{flushend}
 
-% \usepackage{lineno}
-% \setlength\linenumbersep{1mm}
-% \linenumbers
+\usepackage{lineno}
+\setlength\linenumbersep{1mm}
+\linenumbers
 
@@
 
-WSN \cite{baggio2005wireless,werner2005monitoring,hughes2006gridstix}
+% context
 
-Economics of development, IDS as a non-functional cost \cite{lee2002toward}
+The promise of the Internet of Things (IoT) is exciting. Soon my car will
+notify the central heating system at home that I will be late, due to
+unforeseen heavy traffic. It now can turn up the heat a bit later, thus
+lowering my energy bills and adding to a greener world. Wonderful, I'm in.
 
-Energy in WSNs \cite{hughes2013energy}
+But what if the energy company is capable of doing the same? What if they
+decide to turn up the heat while I'm gone, making my home nice and cosy but
+also cranking up my bill and their profit. No thank you, let me out.
 
-Functional Code Fusion
+Along with all the wonderful possibilities, the IoT also holds an enormous
+threat: by opening our homes and lives to all these interconnected devices, we
+also open them to everyone who is technologically savvy enough to break these
+new virtual locks, only now most of the time without a trace.
 
-The remainder of this paper proceeds as follows: ...
+In classic (wired) networks, members of the network are protected by a firewall
+that cleanses all traffic before it actually enters the local network.
+Firewalls are primarily focusing on the actual network traffic, prohibiting
+access to certain destinations and/or services. Attackers though often use
+complicated strategies that abuse flaws in the actual services themselves. The
+firewall can't see these, they pass the wall of fire unnoticed and deliver
+their payload to the system under attack.
+
+At that point it becomes important to know that something went wrong and if
+possible how it went wrong. This information will at least enable us to learn
+from the incident and prepare for future situations or correctly assess the
+impact and mitigate it. This is where intrusion detection steps in. An
+intrusion detection (ID) probe monitors all traffic that passes the firewall
+and looks for known patterns. To do this, it typically gets a copied feed of
+all traffic from the firewall. Often the intrusion detection system (IDS) also
+has a feedback link to the firewall, enabling it to lock out the evildoer based
+on its origin, but most of the time post-mortem.
+
+In a wireless network however, it is not possible to have this single point in
+the network that can oversee all traffic. Every node is uniquely addressable,
+without the need to pass by any other node. So if the security perimeter
+diminishes and every device stands on its own, every device will have to
+implement its own lines of defence.
+
+There is one upside to this situation though: with protection and ID present on
+every device, the services running on those devices can take advantage of these
+security layers. If the IDS detects a problem, the service can immediately
+terminate the communication with the attacker, before it actually experiences
+the consequences. Just like a firewall, the localised IDS actually can be a
+layer between the hostile outside world and the software running on the device.
+
+This integration will be a key point in development of security layers for IoT
+devices. More than in a classic network IDS, the algorithms that try to
+discover the attacks operate with a large degree of uncertainty. They also
+often have to collaborate with other nearby devices in the network, to unmask
+the culprit. This collaboration also increases the level of uncertainty,
+because no other device can be fully trusted. In the end the IDS on an IoT
+device will be able to provide information about the trust it puts in other
+nodes in the network, but it will be up to some application-level policy to
+decide what to do with this information.
+
+% gap analysis
+
+Current developments in this area focus on programming frameworks to structure
+the implementation of different algorithms \cite{todo}, but don't offer a way
+to optimise the usage of resources. They also don't support systematic reuse to
+create different configurations for different nodes or evolution in time.
+\cite{todo}.
+
+% problem -> requirements for solution
+
+The IoT is a very heterogeneous environment, both with respect to hardware and
+software. Being able to address this in a transparent and automated way is a
+key success factor for any kind of development in this area. For a
+non-functional and generic layer, such as ID, this aspect becomes even more
+paramount.
+
+A solution that supports the integration of multiple algorithms and targets IoT
+devices needs to consider the impact on the resources of these devices. These
+devices not only need to survive using a limited battery, the production cost
+should also taken into consideration. Different algorithms should be combined
+as to avoid cumulating their impact.
+
+% my approach (redundant -> contributions)
+
+% This paper introduces id-foo, a framework that allows describing ID algorithms
+% in a formal and functional way using a domain specific language (DSL). It also
+% provides a code generator (CG) that employes functional code fusing (FCF) to
+% produce source code that organises the source code in a more optimal way. This
+% way it lowers the impact on memory usage, execution time and network usage.
+
+% scientific contribution
+
+The first contribution of this paper is a framework that enables the formal
+description of ID algorithms on a functional level. Given these descriptions an
+automated code generation process produces the corresponding source code, for a
+given platform and configuration policy.
+
+The second contribution is the application of functional code fusion (FCF)
+during code generation. FCF identifies common data and functions and organises
+the code in such a way that redundant iterations, tests and computations are
+eliminated. Our results show substantially improved memory usage, execution
+time and usage of the wireless network. All of these components add to the
+reduction of energy and a longer battery life and therefore autonomy of the
+device.
+
+The final contribution is the unified integration of the generated IDS with the
+actual software. All information from the different algorithms is centralised
+and made available to the applications using an event-driven application
+programming interface (API).
+
+The remainder of this paper proceeds as follows: Section \ref{background}
+details some of the ideas behind FCF and identifies the different aspects of
+the problem space. Section \ref{design} describes the design we applied to
+construct the framework, its language and code generator. Section
+\ref{evaluation} evaluates an implementation and application of the domain
+specific language (DSL) and code generator, using a set of ID algorithms found
+in literature. Section \ref{related} explores related work in the field of ID
+DSLs and code generation for the IoT. Finally, section \ref{conclusion}
+summarises out findings, draws conclusions and identifies topics for future
+work.
+
+\section{Background}
+\label{background}
+
+Theory of Loop Fusion \cite{darte2000complexity}
+
+Examples of code/loop fusion (e.g. Haskell Stream Fusion) \cite{coutts2007stream}
+
+Function Inlining (e.g. TinyOS) \cite{gay2007software,gay2003nesc}
+
+TODO
+
+\section{Design}
+\label{design}
+
+TODO
+
+\section{Evaluation}
+\label{evaluation}
+
+TODO
 
 \section{Related Work}
+\label{related}
 
@@
 
-\section{Background}
+\section{Conclusion and Future Work}
+\label{conclusion}
 
-Theory of Loop Fusion \cite{darte2000complexity}
+theoretical gains are met in practice
 
-Examples of code/loop fusion (e.g. Haskell Stream Fusion) \cite{coutts2007stream}
+explore more domains + implement corresponding DSLs for embedded systems
 
-Function Inlining (e.g. TinyOS) \cite{gay2007software,gay2003nesc}
+extract generic language as host for DSLs
 
-\section{Problem Analysis}
-
-TODO
-
-\section{Problem Solution}
-
-TODO
-
-\section{Evaluation}
+\section{Implementation and Evaluation}
 
@@
 TODO: add 3rd ? \cite{krontiris2009cooperative}
-
-\section{Conclusions and Future Work}
-
-theoretical gains are met in practice
-
-explore more domains + implement corresponding DSLs for embedded systems
-
-extract generic language as host for DSLs
 

### 2014-10-03 paper.tex
some touch-ups

@@
 development of efficient intrusion detection systems (IDS). It consists of a
-domain specific language (DSL) that allows describing ID algorithms in a formal
-and functional way. The id-foo code generator then fuses these algorithms on a
-functional level and produces more optimally organised source code. A
-side-by-side comparison shows that id-foo-based code significantly reduces
-memory footprint, message passing overhead and execution time in comparison to
-a straightforward and sequential implementation of the same algorithms.
+domain specific language (DSL) that allows formally describing ID algorithms on
+a functional level. The id-foo code generator then fuses these algorithms and
+produces more optimally organised source code. A side-by-side comparison shows
+that generated code reduces memory footprint, message passing overhead and
+execution time in comparison to a straightforward sequential implementation of
+the same algorithms.
 
@@
 
-The promise of the Internet of Things (IoT) is exciting. Soon my car will
-notify the central heating system at home that I will be late, due to
-unforeseen heavy traffic. It now can turn up the heat a bit later, thus
-lowering my energy bills and adding to a greener world. Wonderful, I'm in.
+The promise of the Internet of Things (IoT) is exciting: my car will notify the
+central heating system at my smart home \cite{aldrich2003smart} that I will be
+late. It now can turn up the heat later, thus lowering my energy bills and
+adding to a greener world. Wonderful, I'm in.
 
-But what if the energy company is capable of doing the same? What if they
-decide to turn up the heat while I'm gone, making my home nice and cosy but
-also cranking up my bill and their profit. No thank you, let me out.
+But what if the energy company can do the same? What if they are able to turn
+up the heat while I'm gone, making my home nice and cosy but also cranking up
+my bill and their profit. No thank you, let me out.
 
-Along with all the wonderful possibilities, the IoT also holds an enormous
+Along with all the wonderful possibilities, the IoT clearly holds an enormous
 threat: by opening our homes and lives to all these interconnected devices, we
-also open them to everyone who is technologically savvy enough to break these
-new virtual locks, only now most of the time without a trace.
+also open them to everyone who is able to break these new, virtual locks, often
+without leaving a trace.
 
-In classic (wired) networks, members of the network are protected by a firewall
-that cleanses all traffic before it actually enters the local network.
-Firewalls are primarily focusing on the actual network traffic, prohibiting
-access to certain destinations and/or services. Attackers though often use
-complicated strategies that abuse flaws in the actual services themselves. The
-firewall can't see these, they pass the wall of fire unnoticed and deliver
-their payload to the system under attack.
+In classic networks, members are protected by a firewall that inspects all
+traffic before it's passed on into the network. Firewalls focus on individual
+network packets, prohibiting access to certain destinations and/or services.
+But attackers often use strategies that abuse flaws in one or more services
+themselves. The firewall can't distinguish these from legitimate packets, they
+pass unnoticed and deliver their payload.
 
-At that point it becomes important to know that something went wrong and if
-possible how it went wrong. This information will at least enable us to learn
-from the incident and prepare for future situations or correctly assess the
-impact and mitigate it. This is where intrusion detection steps in. An
-intrusion detection (ID) probe monitors all traffic that passes the firewall
-and looks for known patterns. To do this, it typically gets a copied feed of
-all traffic from the firewall. Often the intrusion detection system (IDS) also
-has a feedback link to the firewall, enabling it to lock out the evildoer based
-on its origin, but most of the time post-mortem.
+Now it becomes important to know that something went wrong and, if possible,
+how it went wrong. This information will at least enable us to learn from the
+incident and prepare for future situations or correctly assess the impact and
+mitigate it. This is where intrusion detection steps in. An intrusion detection
+(ID) probe monitors all traffic that passes the firewall and looks for known
+patterns. To do this, it typically gets a copied feed of all traffic from the
+firewall. Often the intrusion detection system (IDS) also has a feedback link
+to the firewall, enabling it to lock out the evildoer based on its origin, but
+most of the time this can't stop the initial attack.
 
-In a wireless network however, it is not possible to have this single point in
-the network that can oversee all traffic. Every node is uniquely addressable,
-without the need to pass by any other node. So if the security perimeter
-diminishes and every device stands on its own, every device will have to
-implement its own lines of defence.
+In a mobile ad hoc network (MANET) however, it is not possible to have this
+single point in the network that can oversee all traffic
+\cite{mishra2004intrusion}. Every node is individually approachable, without
+the need to pass by any other node. So if the security perimeter diminishes and
+every device stands on its own, every device will have to implement its own
+lines of defence.
 
@@
 terminate the communication with the attacker, before it actually experiences
-the consequences. Just like a firewall, the localised IDS actually can be a
-layer between the hostile outside world and the software running on the device.
+the consequences. Just like a firewall, the IDS can be a layer between the
+hostile network and the software on the device.
 
@@
 often have to collaborate with other nearby devices in the network, to unmask
-the culprit. This collaboration also increases the level of uncertainty,
-because no other device can be fully trusted. In the end the IDS on an IoT
-device will be able to provide information about the trust it puts in other
+the culprit \cite{marchang2008collaborative,krontiris2009cooperative}. This
+collaboration increases the level of uncertainty, because no other device can
+be fully trusted. In the end the IDS on an IoT device will be able to provide
+information about the trust \cite{ganeriwal2008reputation} it puts in other
 nodes in the network, but it will be up to some application-level policy to
@@
 Current developments in this area focus on programming frameworks to structure
-the implementation of different algorithms \cite{todo}, but don't offer a way
-to optimise the usage of resources. They also don't support systematic reuse to
-create different configurations for different nodes or evolution in time.
-\cite{todo}.
+the implementation of different algorithms \cite{valero2012di} and offer the
+required basic functional components to implement actual algorithms
+\cite{krontiris2008lidea}, but don't offer a way to optimise the usage of
+resources. They also don't support systematic reuse to create different
+configurations for different nodes or evolution in time.
 
@@
 key success factor for any kind of development in this area. For a
-non-functional and generic layer, such as ID, this aspect becomes even more
-paramount.
+non-functional layer, such as ID, this aspect becomes even more paramount.
 
@@
 devices not only need to survive using a limited battery, the production cost
-should also taken into consideration. Different algorithms should be combined
-as to avoid cumulating their impact.
+should also taken into consideration. Different algorithms should be merged to
+avoid accumulating their impact.
 
@@
 the problem space. Section \ref{design} describes the design we applied to
-construct the framework, its language and code generator. Section
-\ref{evaluation} evaluates an implementation and application of the domain
-specific language (DSL) and code generator, using a set of ID algorithms found
-in literature. Section \ref{related} explores related work in the field of ID
-DSLs and code generation for the IoT. Finally, section \ref{conclusion}
-summarises out findings, draws conclusions and identifies topics for future
-work.
+construct the framework, its domain specific language (DSL) and code generator.
+Section \ref{evaluation} evaluates an implementation and application of the DSL
+and code generator, using a set of ID algorithms found in literature. Section
+\ref{related} explores related work in the field of ID DSLs and code generation
+for the IoT. Finally, section \ref{conclusion} summarises out findings, draws
+conclusions and identifies topics for future work.
 

### 2014-10-03 paper.tex
reduced text a bit

--- 2014-09-29 paper.tex
@@
 development of efficient intrusion detection systems (IDS). It consists of a
-domain specific language (DSL) that allows describing ID algorithms in a formal
-and functional way. The id-foo code generator then fuses these algorithms on a
-functional level and produces more optimally organised source code. A
-side-by-side comparison shows that id-foo-based code significantly reduces
-memory footprint, message passing overhead and execution time in comparison to
-a straightforward and sequential implementation of the same algorithms.
+domain specific language (DSL) that allows formally describing ID algorithms on
+a functional level. The id-foo code generator then fuses these algorithms and
+produces more optimally organised source code. Integration of applications with
+this code further enables reacting on threats as they occur, not post-mortem. A
+side-by-side comparison shows that generated code reduces memory footprint,
+message passing overhead and execution time in comparison to a straightforward
+sequential implementation of the same algorithms.
 
@@
 
-WSN \cite{baggio2005wireless,werner2005monitoring,hughes2006gridstix}
+% context
 
-Economics of development, IDS as a non-functional cost \cite{lee2002toward}
+% IoT + threat
 
-Energy in WSNs \cite{hughes2013energy}
+The promise of the Internet of Things (IoT) is exciting: my smart car will
+notify the my smart home \cite{aldrich2003smart} that I will be late. It now
+can turn up the heat later, thus lowering my energy bills and adding to a
+greener world. Wonderful, I'm in. But what if the energy company can do the
+same? What if they are able to turn up the heat while I'm gone, making my home
+nice and cosy but also cranking up my bill and their profit. No thank you, let
+me out.
 
-Functional Code Fusion
+Along with all the wonderful possibilities, the IoT clearly holds an enormous
+threat: by opening our homes and lives to all these interconnected devices, we
+also open them to everyone who is able to break these new, virtual locks, often
+without leaving a trace.
 
-The remainder of this paper proceeds as follows: ...
+% classic networks: fw and central ids
+
+In classic networks, firewalls protect the outer perimeter of the network,
+filtering unwanted packets. But attacks on flaws in services can pass
+unnoticed. This is where intrusion detection (ID) steps in: an ID probe
+monitors all traffic that passes through the firewall, looking for patterns and
+optionally alerting the firewall if a malicious pattern is detected.
+
+% manet: local ids + opportunity of integration
+
+In mobile ad hoc networks (MANET), which make up an important part of the IoT,
+it is not possible to have this single point in the network that can oversee
+all traffic \cite{mishra2004intrusion}. Every node is individually
+approachable. If this outer security perimeter diminishes, every device has to
+implement its own lines of defence. The services running on those devices can
+take advantage of these local security layers. If the IDS detects a problem,
+the service can immediately terminate the communication with the attacker,
+before it's actually abused.
+
+% manet: IDS algorithms -> coorporation, trust
+
+More than in a classic network IDS, ID algorithms for MANETS operate with a
+large degree of uncertainty. They often have to collaborate with other devices
+in the network \cite{marchang2008collaborative,krontiris2009cooperative}, which
+increases the level of uncertainty, because no other device can fully be
+trusted. An IDS on an IoT device will be able to provide the services with
+information about the reputation \cite{ganeriwal2008reputation} of other nodes
+in the network.
+
+% manet: limited resources
+
+Resources are another important factor: IoT devices typically have limited
+batteries, little processing power and memory is restricted to the bare minimum.
+
+% gap analysis
+
+Current developments in ID focus on programming frameworks to structure the
+implementation of different algorithms \cite{valero2012di} and offer the
+required basic functional components to implement actual algorithms
+\cite{krontiris2008lidea}, but don't offer a way to optimise the usage of
+resources. They also don't support systematic reuse to create different
+configurations for different nodes or evolution in time.
+
+% problem -> requirements for solution
+
+The IoT is a heterogeneous environment, both in hardware and software.
+Addressing this in a transparent and automated way is a key success factor for
+any kind of development in this area. For a non-functional layer, such as ID,
+this aspect becomes even more paramount.
+
+A solution that supports the integration of multiple algorithms on IoT devices
+needs to consider the impact on their resources. Different algorithms should be
+merged to avoid accumulating their impact.
+
+% my approach merged into scientific contribution
+
+% 1. framework: language + code generator
+
+The first contribution of this paper is a framework that enables the formal
+description of ID algorithms on a functional level. An automated code
+generation process produces corresponding source code, for a given platform and
+configuration.
+
+% 2. FCF to optimise
+
+The second contribution introduces functional code fusion (FCF) as a generation
+paradigm. FCF identifies common data and functions, and organises code in such
+a way that redundant iterations, tests and computations are eliminated. Our
+results show substantially improved memory usage, execution time and usage of
+the wireless radio, reducing energy consumption.
+
+% 3. integration
+
+The final contribution is the integration of the generated IDS with the actual
+software. Information from the different algorithms is unified and made
+available to the applications, that can use it to actively protect themselves.
+
+% structure
+
+The remainder of this paper proceeds as follows: section \ref{background}
+details some of the ideas behind FCF and identifies the different aspects of
+the problem space. Section \ref{design} describes the design we applied to
+construct the framework, its domain specific language (DSL) and code generator.
+Section \ref{evaluation} evaluates an implementation and application of the
+framework. Section \ref{related} explores related work in the field of DSLs for
+ID and code generation for the IoT. Finally, section \ref{conclusion}
+summarises our findings, draws conclusions and identifies topics for future
+work.
+
+\section{Background}
+\label{background}
+
+Theory of Loop Fusion \cite{darte2000complexity}
+
+Examples of code/loop fusion (e.g. Haskell Stream Fusion) \cite{coutts2007stream}
+
+Function Inlining (e.g. TinyOS) \cite{gay2007software,gay2003nesc}
+
+TODO
+
+\section{Design}
+\label{design}
+
+TODO
+
+\section{Evaluation}
+\label{evaluation}
+
+TODO
 
 \section{Related Work}
+\label{related}
 

### 2014-10-06 paper.tex
roughly updated title/abstract/title after meeting

@@
 
+% until we settle on the final name ;-)
+\usepackage{xspace}
+\newcommand{\NAME}{id-foo\xspace}
+
 \begin{document}
@@
 
-\title{Efficiently Detecting Intrusions\\in the Internet of Things}
+\title{
+Introducing the \NAME framework\\
+for efficient Intrusion Detection\\
+in the Internet of Things
+}
 
@@
 algorithms can alleviate this, yet, proves to be time-consuming, repetitive and
-error-prone. To address this problem we introduce id-foo, a framework for the
+error-prone. To address this problem we introduce \NAME, a framework for the
 development of efficient intrusion detection systems (IDS). It consists of a
 domain specific language (DSL) that allows formally describing ID algorithms on
-a functional level. The id-foo code generator then fuses these algorithms and
-produces more optimally organised source code. Integration of applications with
-this code further enables reacting on threats as they occur, not post-mortem. A
-side-by-side comparison shows that generated code reduces memory footprint,
-message passing overhead and execution time in comparison to a straightforward
-sequential implementation of the same algorithms.
+a functional level. A code generator then fuses these algorithms and produces
+more optimally organised source code. This way, \NAME addresses the
+heterogeneity of both the IoT and ID, as well as the limited resources of IoT
+devices. A side-by-side comparison shows that generated code reduces memory
+footprint, message passing overhead and execution time in comparison to a
+straightforward sequential implementation of the same algorithms.
 
@@
 
-Along with all the wonderful possibilities, the IoT clearly holds an enormous
+Along with all the wonderful possibilities, the IoT clearly holds a significant
 threat: by opening our homes and lives to all these interconnected devices, we
@@
 
-% manet: local ids + opportunity of integration
+% manet: local ids -> localize -> limited resources
 
@@
 it is not possible to have this single point in the network that can oversee
-all traffic \cite{mishra2004intrusion}. Every node is individually
-approachable. If this outer security perimeter diminishes, every device has to
-implement its own lines of defence. The services running on those devices can
-take advantage of these local security layers. If the IDS detects a problem,
-the service can immediately terminate the communication with the attacker,
-before it's actually abused.
-
-% manet: IDS algorithms -> coorporation, trust
-
-More than in a classic network IDS, ID algorithms for MANETS operate with a
-large degree of uncertainty. They often have to collaborate with other devices
-in the network \cite{marchang2008collaborative,krontiris2009cooperative}, which
-increases the level of uncertainty, because no other device can fully be
-trusted. An IDS on an IoT device will be able to provide the services with
-information about the reputation \cite{ganeriwal2008reputation} of other nodes
-in the network.
-
-% manet: limited resources
-
-Resources are another important factor: IoT devices typically have limited
-batteries, little processing power and memory is restricted to the bare minimum.
+all traffic \cite{mishra2004intrusion}. If the outer security perimeter
+diminishes, every device has to implement its own lines of defence.
+Localisation of the IDS requires resources, and resources are another important
+factor to consider: IoT devices typically have limited batteries, little
+processing power and memory is restricted to the bare minimum.
 
@@
 
-% my approach merged into scientific contribution
+% my approach merged into ...
+% scientific contribution
 
@@
 generation process produces corresponding source code, for a given platform and
-configuration.
+configuration. The way, the framework addresses the heterogeneous nature of
+both the IoT as a hardware and software platform, and that of the different ID
+algorithms.
 
@@
 the wireless radio, reducing energy consumption.
-
-% 3. integration
-
-The final contribution is the integration of the generated IDS with the actual
-software. Information from the different algorithms is unified and made
-available to the applications, that can use it to actively protect themselves.
 
@@
 
-\section{Evaluation}
+\section{Implementation and Evaluation}
 \label{evaluation}
 
-TODO
+network usages: number of packets, number of bytes
+
+processing time
+
+image size
+
+TODO: memory usage ?
+
+2 algorithms \cite{ganeriwal2008reputation}
+
+TODO: add 3rd ? \cite{krontiris2009cooperative}
 
@@
 
+% manet: IDS algorithms -> coorporation, trust
+
+More than in a classic network IDS, ID algorithms for MANETS operate with a
+large degree of uncertainty. They often have to collaborate with other devices
+in the network \cite{marchang2008collaborative,krontiris2009cooperative}, which
+increases the level of uncertainty, because no other device can fully be
+trusted. An IDS on an IoT device will be able to provide the services with
+information about the reputation \cite{ganeriwal2008reputation} of other nodes
+in the network.
+
 \section{Conclusion and Future Work}
@@
 
-\section{Implementation and Evaluation}
+Integration of applications with generated IDS allows actively and dynamically
+responding to changes in the reputation of other nodes.
 
-network usages: number of packets, number of bytes
-
-processing time
-
-image size
-
-TODO: memory usage ?
-
-2 algorithms \cite{ganeriwal2008reputation}
-
-TODO: add 3rd ? \cite{krontiris2009cooperative}
+The services running on those devices can take advantage of these local
+security layers. If the IDS detects a problem, the service can immediately
+terminate the communication with the attacker, before it's actually abused.
 

### 2014-10-06 push.sh
added push all-in-one script

@@ -1,0 +1,4 @@
+#!/bin/bash
+
+git push private master
+git push dnet master

### 2014-10-06 paper.tex
made title casing more consistent
augmenting the boringness-factor ;-)

@@
 \title{
-Introducing the \NAME framework\\
-for efficient Intrusion Detection\\
+Introducing the \NAME Framework\\
+for Efficient Intrusion Detection\\
 in the Internet of Things
@@
 
-The promise of the Internet of Things (IoT) is exciting: my smart car will
-notify the my smart home \cite{aldrich2003smart} that I will be late. It now
-can turn up the heat later, thus lowering my energy bills and adding to a
-greener world. Wonderful, I'm in. But what if the energy company can do the
-same? What if they are able to turn up the heat while I'm gone, making my home
-nice and cosy but also cranking up my bill and their profit. No thank you, let
-me out.
-
-Along with all the wonderful possibilities, the IoT clearly holds a significant
-threat: by opening our homes and lives to all these interconnected devices, we
-also open them to everyone who is able to break these new, virtual locks, often
-without leaving a trace.
+Along with its potential, the IoT also presents a significant threat: by
+opening our smart homes \cite{aldrich2003smart} and personal data to all these
+interconnected devices, we also open them to everyone who is able to break
+these new, virtual locks, often without leaving a trace.
 
@@
 
-In classic networks, firewalls protect the outer perimeter of the network,
-filtering unwanted packets. But attacks on flaws in services can pass
-unnoticed. This is where intrusion detection (ID) steps in: an ID probe
-monitors all traffic that passes through the firewall, looking for patterns and
-optionally alerting the firewall if a malicious pattern is detected.
+In classic networks, firewalls focus on the outer perimeter of the network,
+filtering unwanted packets and protecting the entire internal network. But
+attacks on flaws in services can pass unnoticed. This is where intrusion
+detection (ID) steps in: an ID probe monitors all traffic that passes through
+the firewall, looking for patterns and optionally alerting the firewall if a
+malicious pattern is detected \cite{denning1987intrusion}.
 
@@
 it is not possible to have this single point in the network that can oversee
-all traffic \cite{mishra2004intrusion}. If the outer security perimeter
-diminishes, every device has to implement its own lines of defence.
-Localisation of the IDS requires resources, and resources are another important
-factor to consider: IoT devices typically have limited batteries, little
-processing power and memory is restricted to the bare minimum.
+all traffic \cite{zhang2000intrusion, mishra2004intrusion}. If the outer
+security perimeter diminishes, every device has to implement its own lines of
+defence. Making the IDS a local service on each device requires resources,
+which are another important constraint: IoT devices typically have limited
+batteries, little processing power and memory is restricted to the bare minimum.
 
@@
 generation process produces corresponding source code, for a given platform and
-configuration. The way, the framework addresses the heterogeneous nature of
-both the IoT as a hardware and software platform, and that of the different ID
-algorithms.
+configuration. This way, the framework addresses the heterogeneous nature of
+both the IoT, as a hardware and software platform, and ID algorithms.
 

### 2014-10-06 paper.tex
started describing the background of FCF with algo

@@
 \interdisplaylinepenalty=2500
-\usepackage{algorithmic}
+% \usepackage{algorithmic}
 \usepackage{array}
@@
 
+% while preparing, linenumbers come in handy
 \usepackage{lineno}
@@
 \newcommand{\NAME}{id-foo\xspace}
+
+% http://tex.stackexchange.com/questions/299/how-to-get-long-texttt-sections-to-break
+\newcommand*\justify{%
+  \fontdimen2\font=0.4em% interword space
+  \fontdimen3\font=0.2em% interword stretch
+  \fontdimen4\font=0.1em% interword shrink
+  \fontdimen7\font=0.1em% extra space
+  \hyphenchar\font=`\-% allowing hyphenation
+}
+
+% http://en.wikibooks.org/wiki/LaTeX/Customizing_LaTeX
+\newcommand{\ttt}[1]{\texttt{\justify{#1}}}
+
+\usepackage{algpseudocode}
+% custom Let command
+\newcommand*\Let[2]{\State #1 $\gets$ #2}
+\renewcommand{\algorithmicforall}{\textbf{for each}}
+\let\ForEach\ForAll
 
@@
 
+ID algorithms targeting MANETs share a common design pattern as illustrated in
+pseudocode listing \ref{alg:id-algo-pattern}. It consists of two functions:
+\ttt{process\_message} and \ttt{do\_housekeeping}. The former function
+represents the fact that ID algorithms inspect the network traffic that is
+captured by the wireless radio and process it, looking for malicious activity.
+The latter allows for algorithms to perform periodical tasks, checking their
+internal information and optionally act upon it. Some algorithms will implement
+both functions, some don't, but the abstract structure is applicable for a
+significant number of algorithms.
+
+\begin{listing}
+\begin{algorithmic}[1]
+  \Require{nodes, global storage for information about nodes}
+  \Function{process\_message}{$msg$}
+    \ForEach{$byte \in msg$} \label{alg:id-algo-pattern-loop1}
+     \State \dots \Comment{analyse byte-sequences}
+    \EndFor
+    \Let{$nodes_x$}{value}  \Comment{optionally update node info}
+    \State \Call{send}{$nodes_y$, ``info''} \Comment{optionally exchange info} \label{alg:id-algo-pattern-send}
+  \EndFunction
+  \State
+  \Function{do\_housekeeping}{}
+    \ForEach{$node \in nodes$} \label{alg:id-algo-pattern-loop2}
+      \If{$node > \dots$} \Comment{validate recorded value}
+        \State \dots \Comment{take actions}
+      \EndIf
+    \EndFor
+  \EndFunction
+  \State
+\end{algorithmic}
+\caption{ID Algorithm Pattern\label{alg:id-algo-pattern}}
+\end{listing}
+
+The parts of interest here the iterations at lines
+\ref{alg:id-algo-pattern-loop1} and \ref{alg:id-algo-pattern-loop2}, and the
+interaction with the network at line \ref{alg:id-algo-pattern-send}.
+
 Theory of Loop Fusion \cite{darte2000complexity}

### 2014-10-12 paper.tex
IEEE only accepts figure and table floats
added first draft background

@@
 
-ID algorithms targeting MANETs share a common design pattern as illustrated in
-pseudocode listing \ref{alg:id-algo-pattern}. It consists of two functions:
-\ttt{process\_message} and \ttt{do\_housekeeping}. The former function
-represents the fact that ID algorithms inspect the network traffic that is
-captured by the wireless radio and process it, looking for malicious activity.
-The latter allows for algorithms to perform periodical tasks, checking their
-internal information and optionally act upon it. Some algorithms will implement
-both functions, some don't, but the abstract structure is applicable for a
-significant number of algorithms.
+The underlying problem addressed in this paper is of an economic nature: adding
+ID doesn't add functionality, so its impact should be reduced as much as
+possible.
 
-\begin{listing}
+In literature, many ID algorithm are described, but hardly any are implemented.
+If implementations would be available, one would typically apply a pattern as
+shown in listing \ref{alg:id-algo-application}.
+
+\begin{figure}[H]
+\begin{algorithmic}[1]
+  \Let{$msg$}{$network$} \Comment{all observed messages}
+  \ForEach{$algorithm \in algorithms$}
+    \State \Call{algorithm.progress\_message}{msg}
+  \EndFor
+  \State \dots
+  \Comment{at a given interval}
+  \ForEach{$algorithm \in algorithms$}
+    \State \Call{algorithm.do\_housekeeping}{}
+  \EndFor
+\end{algorithmic}
+\captionof{listing}{Sequential implementation of several ID algorithms\label{alg:id-algo-application}}
+\end{figure}
+
+The implementations would be linked and called in a sequential way, when new
+messages are observed on the network and at a regular interval to allow them to
+do internal housekeeping, processing the information extracted from the network
+traffic, often in a statistical way.
+
+This is a common design pattern of ID algorithms targetting MANETs. The
+implementation of this pattern also shares common structures, as illustrated in
+listing \ref{alg:id-algo-pattern}.
+
+\begin{figure}[H]
 \begin{algorithmic}[1]
@@
     \Let{$nodes_x$}{value}  \Comment{optionally update node info}
-    \State \Call{send}{$nodes_y$, ``info''} \Comment{optionally exchange info} \label{alg:id-algo-pattern-send}
+    \State \Call{send}{$nodes_y$, ``info''} \Comment{optionally exchange info} \label{alg:id-algo-pattern-send1}
   \EndFunction
@@
   \Function{do\_housekeeping}{}
-    \ForEach{$node \in nodes$} \label{alg:id-algo-pattern-loop2}
+    \ForEach{$node \in nodes$} \label{alg:id-algo-pattern-loop2} \label{alg:id-algo-pattern-common-data}
       \If{$node > \dots$} \Comment{validate recorded value}
         \State \dots \Comment{take actions}
+        \State \Call{send}{$node$, ``info''} \Comment{e.g. communicate} \label{alg:id-algo-pattern-send2}
       \EndIf
@@
 \end{algorithmic}
-\caption{ID Algorithm Pattern\label{alg:id-algo-pattern}}
-\end{listing}
+\captionof{listing}{ID Algorithm Pattern\label{alg:id-algo-pattern}}
+\end{figure}
 
-The parts of interest here the iterations at lines
-\ref{alg:id-algo-pattern-loop1} and \ref{alg:id-algo-pattern-loop2}, and the
-interaction with the network at line \ref{alg:id-algo-pattern-send}.
+The parts of interest here are the loops at lines
+\ref{alg:id-algo-pattern-loop1} and \ref{alg:id-algo-pattern-loop2}, the common
+\emph{nodes} data on line \ref{alg:id-algo-pattern-common-data} and the
+interaction with the network at lines \ref{alg:id-algo-pattern-send1} and
+\ref{alg:id-algo-pattern-send2}.
 
-Theory of Loop Fusion \cite{darte2000complexity}
+\subsection{Loops and Common Data}
 
-Examples of code/loop fusion (e.g. Haskell Stream Fusion) \cite{coutts2007stream}
+Given several ID algorithms, that all perform the same parsing loop, over the
+same byte stream, performing comparable operations to find similar patterns,
+one would clearly benefit from combining these loops, avoiding redundant
+reevaluation of the same byte patterns.
 
-Function Inlining (e.g. TinyOS) \cite{gay2007software,gay2003nesc}
+The same goes for common data. Each algorithm not only has to parse the same
+message, it also has to store information about nodes within its own scope,
+resulting in a lot of common data that is duplicated in memory.
 
-TODO
+The only way to solve this, would be to change the provided implementation,
+pulling up the loops over the message and the nodes to a level above the loop
+over the algorithms. But this comes at a development cost. This cost typically
+isn't a one-time cost, e.g. in case a new version of the algorithm is released,
+or simply when a different set of algorithms is selected. Finally, changing, or
+simply implementing the algorithms from scratch, also holds the risk of making
+mistakes.
+
+\subsection{Network Usage}
+
+But, as said before, there are hardly any implementations readily available.
+Probably the main reason for this is the heterogeneity of the IoT. A prime
+example is the \ttt{SEND} function. Different software stacks will have
+different application programming interfaces (APIs) for accessing the network.
+Simple, macro-like solutions can overcome this, but the dispersion of the calls
+to access the network is causing the wireless radio to be on a lot more that
+one would hope for.
+
+A valid solution would try to combine messages in a single, reducing the access
+to the network and even sharing some network overhead. The impact to implement
+this, again would weigh to much on
 

### 2014-10-13 paper.tex
completed draft background

@@
 messages are observed on the network and at a regular interval to allow them to
-do internal housekeeping, processing the information extracted from the network
-traffic, often in a statistical way.
+do internal housekeeping. This includes processing the information extracted
+from the network traffic, often in a statistical way.
 
-This is a common design pattern of ID algorithms targetting MANETs. The
-implementation of this pattern also shares common structures, as illustrated in
+This duality is a common design pattern of ID algorithms targeting MANETs. The
+internals of the pattern reveal more common structures, as illustrated in
 listing \ref{alg:id-algo-pattern}.
@@
 
-Given several ID algorithms, that all perform the same parsing loop, over the
+Given several ID algorithms, that all execute the same parsing loop, over the
 same byte stream, performing comparable operations to find similar patterns,
@@
 message, it also has to store information about nodes within its own scope,
-resulting in a lot of common data that is duplicated in memory.
+resulting in a lot of duplicated information in memory.
 
@@
 isn't a one-time cost, e.g. in case a new version of the algorithm is released,
-or simply when a different set of algorithms is selected. Finally, changing, or
-simply implementing the algorithms from scratch, also holds the risk of making
-mistakes.
+or simply when a different set of algorithms is selected due to a changing risk
+analysis. Finally, changing, or simply implementing the algorithms from
+scratch, also holds the risk of making mistakes.
 
@@
 
-But, as said before, there are hardly any implementations readily available.
-Probably the main reason for this is the heterogeneity of the IoT. A prime
-example is the \ttt{SEND} function. Different software stacks will have
-different application programming interfaces (APIs) for accessing the network.
-Simple, macro-like solutions can overcome this, but the dispersion of the calls
-to access the network is causing the wireless radio to be on a lot more that
-one would hope for.
+But there are hardly any implementations readily available. Probably the main
+reason for this is the heterogeneity of the IoT. A prime example is the
+\ttt{SEND} function. Different software stacks will have different application
+programming interfaces (APIs) for accessing the network. Simple, macro-like
+solutions can overcome this, but the dispersion of the calls to access the
+network is causing the wireless radio to be on a lot more that one would hope
+for.
 
-A valid solution would try to combine messages in a single, reducing the access
-to the network and even sharing some network overhead. The impact to implement
-this, again would weigh to much on
+A valid solution would try to combine messages in a single instance, reducing
+the access to the network and even sharing some network overhead. The impact to
+implement this manually will again outweigh the gain. An option would be to use
+a software library, that provides a unified API for accessing the network and
+collects individual calls and combines them in a single network message.
+
+Besides the fact that all researchers would have to use the same library in a
+consistent way, they would limit their operational scope to a single language,
+serving only a fraction of the heterogeneous IoT.
+
+\subsection{Situation Summary}
+
+Starting from scratch allows optimising the implementation of an IDS, but
+doesn't scale. Reusing existing implementations also doesn't help, due to the
+far from optimal sequential execution of the algorithms and the redundant
+storage of common data. Software libraries can solve many of the issues, but
+only cover parts of the entire spectrum, due to their tie-in with languages and
+frameworks.
 
@@
 \label{design}
-
-TODO
 

### 2014-10-14 paper.tex
added subsection on the DSL component

@@
 % additional packages and utility commands
+\newcommand{\TODO}{\textbf{\color{red}TODO}}
+
 \usepackage{minted}
@@
 \let\ForEach\ForAll
+
+\usepackage{listings}
 
@@
 \begin{figure}[H]
+\captionof{lstlisting}{Sequential implementation of ID algorithms\label{alg:id-algo-application}}
 \begin{algorithmic}[1]
@@
 \end{algorithmic}
-\captionof{listing}{Sequential implementation of several ID algorithms\label{alg:id-algo-application}}
 \end{figure}
@@
 \begin{figure}[H]
+\captionof{lstlisting}{ID Algorithm Pattern\label{alg:id-algo-pattern}}
 \begin{algorithmic}[1]
@@
 \end{algorithmic}
-\captionof{listing}{ID Algorithm Pattern\label{alg:id-algo-pattern}}
 \end{figure}
@@
 \label{design}
+
+To bridge the gap between the research world and the heterogeneous IoT where
+developers want to implement an actual IDS, while still respecting the limited
+resources they have at their disposal, we introduce \NAME, a framework
+consisting of a DSL, code generator and software library.
+
+The DSL enables researchers to formally describe their ID algorithms, without
+focusing on any specific language or platform. Given such formal descriptions
+of ID algorithms, a code generator is now able to combine these, while avoiding
+the sequential patterns/problems presented in listings
+\ref{alg:id-algo-application} and \ref{alg:id-algo-pattern}. It can also target
+different languages and platforms, while reusing a software library that
+provides e.g. message parsing, network access wrapping,\dots
+
+\subsection{DSL}
+
+The DSL component of \NAME has two major goals and one minor: (1) allow the
+description of ID algorithms in a formal way and (2) avoid that those
+descriptions contain \emph{non-fusible} logic and/or data. Besides these (3)
+the usability of the language is also taken into account.
+
+The first goal is covered by the construction of the language itself. Given a
+description of an ID algorithm using the DSL, the algorithm can be
+unambiguously interpreted. The DSL borrows a lot of its syntax from C-like
+languages, dropping only a small subset of its capabilities, but replacing
+those with other higher-level constructs taken from object-oriented (OO)
+languages, aspect-orientation,\dots
+
+The most important omission from the C syntax are basic loops. Loops are very
+generic and make it hard to identify if they deal with the same logic and/or
+data. Fusing these loops across different algorithms is often non-trivial. In
+\NAME's DSL, other constructs have been introduced to lift the definition of
+these loops to a more functional level: scheduling and scoping of the execution
+of functions, reacting to events and pattern matching are three examples that
+allow describing the actual meaning of loops in stead of simply the loop
+itself. It sometimes requires rethinking the reason behind a technical loop,
+but often results in a much clearer description.
+
+To deal with common data, the domain provides the notion of \emph{nodes}. As
+most, if not all, ID algorithms targeting the IoT require some notion of the
+other nodes or devices in the vicinity, the DSL makes this the primary
+information carrier. All functional constructs, mentioned above, typically
+operate on the collection of known nodes, or a functionally defined subset of
+that. Each algorithm also has the possibility to \emph{extend} nodes, adding
+its own properties needed for its execution.
+
+Based on a C-like syntax the language is also very accessible. To further
+nurture the usability, syntactic sugar from OO languages is also adopted.
+Because not everything should be implemented in such a high-level language, it
+also provides a way to \emph{import} external functionality. In the first place
+this will include a standard library of functions that can be called, but it
+can include any custom built software component. Finally it tries to limit the
+need for type information and performs as much type inference as possible. As
+it aims to be language that describes algorithms on a functional level it seeks
+to safeguard its users from these \emph{technicalities}. The fact that the
+opportunities of such a formal description include generation of code in a
+variety of languages, each with possibly different types, also results the need
+for a more functional approach to types.
+
+Listing \ref{lst:heartbeat} illustrates some of the features mentioned above,
+using a small but representative example of an algorithm: a heartbeat. Using a
+heartbeat, nodes can broadcast a message at a regular interval, allowing other
+nodes to monitor its availability. Missing heartbeats could indicate problems,
+such as physical capturing of a certain node.
+
+\lstinputlisting[
+  float,
+  basicstyle=\small\ttfamily,
+  label=lst:heartbeat,
+  caption=Heartbeat in \NAME
+]{src/heartbeat.foo}
+
+\subsection{Code Generator}
+
+\TODO
+
+- semantic model \cite{fowler2010domain}
+- model transformations
+- code model + syntactic mapping using coding pattern-like instructions
+
+\subsection{Software Library}
+
+\TODO
+
+- incoming message parsing
+- outgoing message grouping
+- execution of schedules
+- standard library: high-level conceptual functions (e.g. now(), sha1,...)
 

### 2014-10-15 paper.tex
added previous evaluation

@@
 
-network usages: number of packets, number of bytes
+To evaluate the prototype we chose to use a system based on the Atmel
+ATMEGA1284p micro-controller \cite{datasheet:atmega1284p} and the Digi XBee S2
+ZigBee module \cite{manual:xbee}. To drive the hardware a minimalistic library
+was used, wrapping the technical calls to the components in more easy to use
+functions.
 
-processing time
+The reason for this at first sight unusual setup is simple: relying on basic
+hardware and software allows us to validate that the generator is capable of
+generating code even for the most basic environment available. Any step up,
+both in hardware or software, would offer better and higher abstractions that
+would make it easier to generate code for that situation. It shows that the
+requirements of the generator towards the platform are minimal, don't rely on
+advanced frameworks or operating systems and can therefore be applied in any
+environment, targeting any platform.
 
-image size
+For the evaluation we constructed a small WSN consisting of three nodes: an
+end-device, a router and a coordinator. The end-device and the router were
+completely autonomous nodes, while the coordinator consisted of an XBee module,
+directly connected to a computer via a USB break-out board, allowing serial
+access via a terminal.
 
-TODO: memory usage ?
+We started from a basic application that measures light intensity and reports
+it to the coordinator. The application was written both manually and generated.
+On top of this baseline we added two intrusion detection related algorithms: a
+heartbeat, that allows nodes to validate each other's continuous presence, and
+a reputation-building algorithm that checks if a parent node is cooperative and
+actually forwards messages with a destination further down the
+network\cite{ganeriwal2008reputation}.
 
-2 algorithms \cite{ganeriwal2008reputation}
+Three criteria were evaluated: the image size of the resulting compiled code,
+the network usage in number of frames and bytes and the time required to
+perform once cycle of the event loop. These metrics were collected in four
+situations: without ID algorithms, with a heartbeat, with reputation-tracking
+and with both algorithms implemented.
 
-TODO: add 3rd ? \cite{krontiris2009cooperative}
+In case of the manual implementation, both algorithms were constructed as
+standalone modules and sequentially called from the base-application's event
+loop. The code generator was provided with FOO-lang descriptions of the
+algorithms and generated all four cases.
+
+Before reviewing the results, it is very important to put these results in a
+proper perspective. Comparing manually written code to generated code is no
+exact science. One can always argue that the quality of the two code bases is
+not comparable and that the manual code always can be improved.
+
+Still it might be interesting to look at the numbers, given that both code
+bases are constructed with the same intentions and practices. We believe we
+have achieved this and have taken honest decisions while constructing the
+implementations, making them comparable. Numerical analysis of the
+implementation offers us a preliminary estimation of the effect of our proposed
+solution.
+
+Tables \ref{tbl:manual} and \ref{tbl:generated} show the collected data
+respectively for the manual and the generated implementation. The base case is
+presented in absolute values, while the implementations with the added
+algorithms are presented as relative values.
+
+\begin{table}[H]
+  \centering
+  \begin{tabular}{lrrrr}
+  \hline
+      & base & heartbeat & reputation & both\\
+  \hline
+  size (bytes) & 10500 & 148\% & 127\% & 175\%\\
+  frames & 20 & 255\% & 160\% & 315\%\\
+  bytes & 476 & 406\% & 181\% & 487\%\\
+  time ($\mu$s) & 48 & 196\% & 183\% & 310\%\\
+  \hline
+  \end{tabular}
+  \caption{Results for the manual implementation.}
+  \label{tbl:manual}
+\end{table}
+
+\begin{table}[H]
+  \centering
+  \begin{tabular}{lrrrr}
+  \hline
+         & base & heartbeat & reputation & both\\
+  \hline
+  size (bytes) & 10496 & 175\% & 156\% & 200\%\\
+  frames & 20 & 245\% & 160\% & 275\%\\
+  bytes & 476 & 399\% & 186\% & 454\%\\
+  time ($\mu$s) & 48 & 252\% & 252\% & 288\%\\
+  \hline
+  \end{tabular}
+  \caption{Results for the generated implementation.}
+  \label{tbl:generated}
+\end{table}
+
+From these raw results, we learn that adding algorithms manually to the
+base-application, results in a cumulative increase. The impact of both
+algorithms is simply the sum of the impact of each algorithm by itself. In the
+case of the time required to execute one cycle of the event loop, the total
+time is even a bit higher than simply the sum.
+
+In the case of the generated implementation, this no longer holds: although
+that even the individual increases due to adding a single algorithm are higher
+than in the manual case, the result for the implementation of both algorithms
+is less than the sum of both. Here we clearly see the effect of reusing the
+framework that comes with FOO-lang. The same goes for all other metrics. Maybe
+most remarkable is the effect on the time of one event loop cycle: both
+algorithms by itself add 150\% with respect to the base case, but when
+combined, the impact hardly increases. Here we learn that the introduced
+framework is responsible for the initial overhead, but it clearly pays for
+itself when adding more algorithms.
+
+Table \ref{tbl:summary} compares the two situations by subtracting the manual
+case from the generated case, showing the impact of the code generation. A
+relative value for the entire implementation with both algorithms is also
+presented.
+
+\begin{table}[H]
+  \centering
+  \begin{tabular}{lrrrrr}
+  \hline
+                & base & heartbeat & reputation & both  & result \\
+  \hline
+  size (bytes)  & -4    & 2822     & 3070       & 2664  & 115\%  \\
+  frames        & 0     & -2       & 0          & -8    & 87\%   \\
+  bytes         & 0     & -36      & 24         & -156  & 93\%   \\
+  time ($\mu$s) & 0     & 27       & 33         & -11   & 93\%   \\
+  \hline
+  \end{tabular}
+  \caption{Comparison of both implementations.}
+  \label{tbl:summary}
+\end{table}
+
+When comparing the results of both implementations we first can conclude that
+the generator produces exactly the same code for the base case. We couldn't
+measure any real differences between both implementations.
+
+Secondly, we see that the framework that comes with the generated code adds to
+the size of the resulting image. But the cost is almost constant and is
+relatively lower when algorithms are combined. With roughly 3KB of additional
+overhead, or 15\% in this simple case, this impact is affordable.
+
+From a functional point of view, the generated code lives up to the
+expectactions: both the number of frames as the bytes sent benefit from well
+organized code and reuse of a common framework.
+
+Finally we see that both algorithms by itself add an equal amount of time to
+the processing of a single event loop cycle, but when combined, the event loop
+is about 7\% faster compared to the manual case.
+
+Without any effort to optimize the framework code, nor leveraging knowledge
+about the specific algorithms, this very basic generated code shows that the
+proposed principles are not only theoretically feasible but actually result in
+interesting improvements.
 

### 2014-10-15 paper.tex
evaluation updates to match new scope/wording

@@
 
-For the evaluation we constructed a small WSN consisting of three nodes: an
-end-device, a router and a coordinator. The end-device and the router were
-completely autonomous nodes, while the coordinator consisted of an XBee module,
-directly connected to a computer via a USB break-out board, allowing serial
-access via a terminal.
+For the evaluation we constructed a small meshed network consisting of three
+devices, with only one device able to talk to both other devices, requiring
+routing of messages through this device.
 
-We started from a basic application that measures light intensity and reports
-it to the coordinator. The application was written both manually and generated.
-On top of this baseline we added two intrusion detection related algorithms: a
-heartbeat, that allows nodes to validate each other's continuous presence, and
-a reputation-building algorithm that checks if a parent node is cooperative and
-actually forwards messages with a destination further down the
+We started from a basic application that measures light intensity. The
+application was written both manually and generated. On top of this baseline we
+added two intrusion detection related algorithms: a heartbeat, that allows
+nodes to validate each other's continuous presence, and a reputation-building
+algorithm that checks if a parent node is cooperative and actually forwards
+messages with a destination further down the
 network\cite{ganeriwal2008reputation}.
@@
 standalone modules and sequentially called from the base-application's event
-loop. The code generator was provided with FOO-lang descriptions of the
-algorithms and generated all four cases.
+loop. The code generator was provided with \NAME descriptions of the algorithms
+and generated all four cases.
 
@@
 exact science. One can always argue that the quality of the two code bases is
-not comparable and that the manual code always can be improved.
-
-Still it might be interesting to look at the numbers, given that both code
-bases are constructed with the same intentions and practices. We believe we
-have achieved this and have taken honest decisions while constructing the
-implementations, making them comparable. Numerical analysis of the
-implementation offers us a preliminary estimation of the effect of our proposed
-solution.
+not comparable and that the manual code always can be improved. Still it might
+be interesting to look at the numbers, given that both code bases are
+constructed with the same intentions and practices. We believe we have achieved
+this and have taken honest decisions while constructing the implementations,
+making them comparable. Numerical analysis of the implementation offers us a
+preliminary estimation of the effect of our proposed solution.
 
@@
 is less than the sum of both. Here we clearly see the effect of reusing the
-framework that comes with FOO-lang. The same goes for all other metrics. Maybe
-most remarkable is the effect on the time of one event loop cycle: both
+software library that comes with \NAME. The same goes for all other metrics.
+Maybe most remarkable is the effect on the time of one event loop cycle: both
 algorithms by itself add 150\% with respect to the base case, but when
 combined, the impact hardly increases. Here we learn that the introduced
-framework is responsible for the initial overhead, but it clearly pays for
-itself when adding more algorithms.
+software library is responsible for the initial overhead, but it clearly pays
+for itself when adding more algorithms.
 
@@
 
-Secondly, we see that the framework that comes with the generated code adds to
-the size of the resulting image. But the cost is almost constant and is
+Secondly, we see that the software library that comes with the generated code
+adds to the size of the resulting image. But the cost is almost constant and is
 relatively lower when algorithms are combined. With roughly 3KB of additional
@@
 From a functional point of view, the generated code lives up to the
-expectactions: both the number of frames as the bytes sent benefit from well
-organized code and reuse of a common framework.
+expectations: both the number of frames as the bytes sent benefit from well
+organised code and reuse of a common software library.
 
@@
 
-Without any effort to optimize the framework code, nor leveraging knowledge
+Without any effort to optimise the software library, nor leveraging knowledge
 about the specific algorithms, this very basic generated code shows that the

### 2014-10-15 paper.tex
setup code generator and semantic model

@@
 \usepackage{listings}
+
+\usepackage{tikz}
 
@@
 
+To convert the DSL based ID algorithms into production-ready code, we propose
+the use of a code generator. Although many approaches here are applicable, like
+template based text expansion, XSLT transformations or even using the
+CodeDom\cite{dollard2004code}, we believe that for the level of functional
+transformations that need to be performed, a richer model is required.
+
+A semantic model \cite{fowler2010domain}, tied closely to the DSL itself, can
+fulfil this role. Semantic transformations can now be applied to create a
+code-oriented model, comparable to the idea of CodeDom. From there on,
+syntactic transformations an emit actual source code. Figure
+\ref{fig:code-generation} shows the overall procedure.
+
+\begin{figure}[H]
+\begin{tikzpicture}[
+    nonterminal/.style={
+% The shape:
+rectangle,
+% The size:
+      minimum size=6mm,
+% The border:
+      very thick,
+      draw=red!50!black!50,
+      % The filling:
+top color=white,
+bottom color=red!50!black!20, % and something else at the bottom % Font
+font=\itshape
+    }]
+  \node [nonterminal] {TODO};
+\end{tikzpicture}
+\caption{Overview code generation process}
+\label{fig:code-generation}
+\end{figure}
+
+\subsection{Semantic Model}
+
 \TODO
 
-- semantic model \cite{fowler2010domain}
-- model transformations
-- code model + syntactic mapping using coding pattern-like instructions
+Figure \ref{fig:meta-model} shows the meta-model of the the semantic model.
+
+\begin{figure}[H]
+\begin{tikzpicture}[
+    nonterminal/.style={
+% The shape:
+rectangle,
+% The size:
+      minimum size=6mm,
+% The border:
+      very thick,
+      draw=red!50!black!50,
+      % The filling:
+top color=white,
+bottom color=red!50!black!20, % and something else at the bottom % Font
+font=\itshape
+    }]
+  \node [nonterminal] {TODO};
+\end{tikzpicture}
+\caption{Semantic Model Meta-Model}
+\label{fig:meta-model}
+\end{figure}
 

### 2014-10-15 paper.tex
added draft figures

@@
 \usepackage{tikz}
+\usetikzlibrary{positioning}
+\usetikzlibrary{calc}
+\usetikzlibrary{graphs}
+\usetikzlibrary{trees}
 
@@
 
+\tikzset{
+  resource/.style={
+    rectangle,
+    minimum size=7mm,
+    thick, draw=black,
+    top color=white,
+    font=\ttfamily
+  },
+  representation/.style={
+    rectangle, minimum size=6mm, rounded corners=3mm,
+    minimum size=7mm,
+    thick, draw=black,
+    top color=white,
+    font=\itshape
+  }
+}
+
 \begin{figure}[H]
 \begin{tikzpicture}[
-    nonterminal/.style={
-% The shape:
-rectangle,
-% The size:
-      minimum size=6mm,
-% The border:
-      very thick,
-      draw=red!50!black!50,
-      % The filling:
-top color=white,
-bottom color=red!50!black!20, % and something else at the bottom % Font
-font=\itshape
-    }]
-  \node [nonterminal] {TODO};
+  node distance=8mm,
+  ,>=stealth,thick,black!50,text=black,
+  every new ->/.style={shorten >=1pt},
+]
+  \node (dsl)  [resource]                    {DSL};
+  \node (sm)   [representation,right=of dsl] {Semantic\\Model};
+  \node (cm)   [representation,right=of sm]  {Code\\Model};
+  \node (code) [resource,right=of cm]        {Code};
+
+  \draw [->] (dsl) -> (sm)  node [midway,above=10pt] {parse};
+  \draw [->] (sm)  -> (cm)   node [midway,above=10pt,align=center] {semantic\\transform};
+  \draw [->] (cm)  -> (code) node [midway,above=10pt] {emit code};
+
+  \draw [rounded corners,->]
+        ($ (sm.east) - (5mm,3.5mm) $)
+        -- ++(0,-.5)
+        -| ($ (sm.west) + (5mm,-3.5mm) $)
+           node [near start, below=1pt] {type inference};
+
+  \draw [rounded corners,->]
+        ($ (cm.east) - (5mm,3.5mm) $)
+        -- ++(0,-.5)
+        -| ($ (cm.west) + (5mm,-3.5mm) $)
+           node [near start, below=1pt,align=center] {syntactic\\transform};
+
 \end{tikzpicture}
@@
 
+\tikzset{
+  class/.style={
+    rectangle,
+    minimum size=1cm,
+    thick, draw=black,
+    text centered,
+    text width=1.75cm,
+  }
+}
+
 \begin{figure}[H]
 \begin{tikzpicture}[
-    nonterminal/.style={
-% The shape:
-rectangle,
-% The size:
-      minimum size=6mm,
-% The border:
-      very thick,
-      draw=red!50!black!50,
-      % The filling:
-top color=white,
-bottom color=red!50!black!20, % and something else at the bottom % Font
-font=\itshape
-    }]
-  \node [nonterminal] {TODO};
+  edge from parent fork down,
+  sibling distance=5mm, level distance=20mm, node distance=8mm
+  ]
+  \node (model) [class] {Model}
+      child {node (module ) [class] {Module}
+        child {node (strategy) [class,align=center,xshift=-9mm] {Execution\\Strategy}
+          child {node(every) [class,xshift=-9mm] {Every}}
+          child {node (when) [class,xshift=9mm] {When}}
+        }
+        child {node [class,xshift=9mm] {function}}
+      };
+  \node (constant) [class,right=of model] {Constant};
+  \node (scope)    [class,left=of strategy] {Scope};
 \end{tikzpicture}
-\caption{Semantic Model Meta-Model}
+\caption{Partial Semantic Model Meta-Model}
 \label{fig:meta-model}

### 2014-10-15 paper.tex
finally found solution for UML diagram

@@
 \usetikzlibrary{trees}
+\usetikzlibrary{arrows}
 
@@
     text width=1.75cm,
-  }
+  },
+  subclass/.style={thick,draw=black,>=open triangle 60,<-},
+  aggregate/.style={thick,draw=black,>=diamond,<-}
 }
@@
   \node (model) [class] {Model}
-      child {node (module ) [class] {Module}
-        child {node (strategy) [class,align=center,xshift=-9mm] {Execution\\Strategy}
-          child {node(every) [class,xshift=-9mm] {Every}}
-          child {node (when) [class,xshift=9mm] {When}}
+      child {node (module ) [class] {Module} edge from parent [aggregate]
+        child {node (strategy) [class,align=center,xshift=-9mm] {Execution\\Strategy} edge from parent [aggregate]
+          child {node(every) [class,xshift=-9mm] {Every} edge from parent [subclass]}
+          child {node (when) [class,xshift=9mm] {When} edge from parent [subclass]}
         }
-        child {node [class,xshift=9mm] {function}}
+        child {node [class,xshift=9mm] {function} edge from parent [aggregate]}
       };
@@
   \node (scope)    [class,left=of strategy] {Scope};
+
+  \draw [thick,>=latex,->] (strategy) -- (scope);
+  \draw [aggregate] (model) -- (constant);
 \end{tikzpicture}

### 2014-10-16 paper.tex
completed design up to semantic model

--- 2014-10-15 paper.tex
@@
 \usepackage{tikz}
+\usetikzlibrary{positioning}
+\usetikzlibrary{calc}
+\usetikzlibrary{graphs}
+\usetikzlibrary{trees}
+\usetikzlibrary{arrows}
 
@@
 
-\begin{figure}[H]
+\tikzset{
+  resource/.style={
+    rectangle,
+    minimum width=1cm,
+    minimum height=8mm,
+    thick, draw=black,
+    top color=white,
+    font=\ttfamily
+  },
+  representation/.style={
+    rectangle, rounded corners=3mm,
+    minimum width=2cm,
+    minimum height=8mm,
+    thick, draw=black,
+    top color=white,
+    font=\itshape
+  }
+}
+
+\begin{figure}[ht]
+  \centering
+  \scalebox{.85}{
 \begin{tikzpicture}[
-    nonterminal/.style={
-% The shape:
-rectangle,
-% The size:
-      minimum size=6mm,
-% The border:
-      very thick,
-      draw=red!50!black!50,
-      % The filling:
-top color=white,
-bottom color=red!50!black!20, % and something else at the bottom % Font
-font=\itshape
-    }]
-  \node [nonterminal] {TODO};
+  node distance=8mm,
+  ,>=stealth,thick,black!50,text=black,
+  every new ->/.style={shorten >=1pt}
+]
+  \node (dsl)  [resource]                                 {DSL};
+  \node (sm)   [representation,right=of dsl,align=center] {Semantic\\Model};
+  \node (cm)   [representation,right=of sm,align=center]  {Code\\Model};
+  \node (code) [resource,right=of cm]        {Code};
+
+  \draw [->] (dsl) -> (sm)   node [midway,above=10pt]              {parse};
+  \draw [->] (sm)  -> (cm)   node [midway,above=10pt,align=center] {semantic\\transform};
+  \draw [->] (cm)  -> (code) node [midway,above=10pt]              {emit code};
+
+  \draw [rounded corners,->]
+        ($ (sm.east) - (5mm,4.5mm) $)
+        -- ++(0,-.5)
+        -| ($ (sm.west) + (5mm,-4.5mm) $)
+           node [near start, below=1pt] {type inference};
+
+  \draw [rounded corners,->]
+        ($ (cm.east) - (5mm,4.5mm) $)
+        -- ++(0,-.5)
+        -| ($ (cm.west) + (5mm,-4.5mm) $)
+           node [near start, below=1pt,align=center] {syntactic\\transform};
+
 \end{tikzpicture}
+}
 \caption{Overview code generation process}
@@
 
+The DSL is parsed and imported into a semantic model. Based on the minimal type
+information present in the model, all remaining types are inferred. Through a
+series of semantic transformations, the model evolves from the functional to a
+more technical one, a code model. This code model can be considered as a rich
+AST with an extensive set of language constructs. Syntactic transformations now
+lower this level of expressiveness according to the desired platform and
+language. In the end the code model can be directly emitted to the target
+language.
+
 \subsection{Semantic Model}
 
-\TODO
+The semantic model is the core of the solution. It offers a way to describe ID
+algorithms in a way that allows semantic reinterpretation. This way, the
+functionality of different algorithms can be reorganised to allow for more
+optimal execution and memory usage. Hence the name \NAME of the framework:
+intrusion detection functionality organisation optimisation.
 
-Figure \ref{fig:meta-model} shows the meta-model of the the semantic model.
+Figure \ref{fig:meta-model} shows the central part of the semantic model as
+implemented for \NAME.
 
-\begin{figure}[H]
+\tikzset{
+  class/.style={
+    rectangle,
+    minimum size=1cm,
+    thick, draw=black,
+    text centered,
+    text width=1.75cm,
+  },
+  subclass/.style={thick,draw=black,>=open triangle 60,<-},
+  aggregate/.style={thick,draw=black,>=open diamond,<-},
+  composite/.style={thick,draw=black,>=diamond,<-},
+  dependency/.style={thick,>=latex,->}
+}
+
+\begin{figure}[ht]
+  \centering
+  \scalebox{.85}{
 \begin{tikzpicture}[
-    nonterminal/.style={
-% The shape:
-rectangle,
-% The size:
-      minimum size=6mm,
-% The border:
-      very thick,
-      draw=red!50!black!50,
-      % The filling:
-top color=white,
-bottom color=red!50!black!20, % and something else at the bottom % Font
-font=\itshape
-    }]
-  \node [nonterminal] {TODO};
+  edge from parent fork down,
+  sibling distance=5mm, level distance=20mm, node distance=6mm
+  ]
+  \node (model) [class] {Model}
+      child {node (module) [class] {Module} edge from parent [composite]
+        child {node (strategy) [class,align=center,xshift=-9mm] {Execution\\Strategy} edge from parent [composite]
+          child {node(every) [class,xshift=-9mm] {Every} edge from parent [subclass]}
+          child {node (when) [class,xshift=9mm] {When} edge from parent [subclass]}
+        }
+        child {node (function) [class,xshift=9mm] {Function} edge from parent [composite]}
+      };
+  \node (constant)  [class,right=of module]   {Constant};
+  \node (scope)     [class,left=of strategy]  {Scope};
+  \node (statement) [class,right=of function] {Statement};
+  \node (nodes)     [class,left=of module]    {Nodes\\Domain};
+
+  \draw [aggregate]  (module)   -- (nodes);
+  \draw [composite]  (module)   -- (constant);
+  \draw [dependency] (strategy) -- (scope);
+  \draw [composite]  (function) -- (statement);
 \end{tikzpicture}
-\caption{Semantic Model Meta-Model}
+}
+\caption{Partial Semantic Model for \NAME}
 \label{fig:meta-model}
 \end{figure}
+
+The concept driving this model is the structure starting at the \emph{Model} up
+to the \emph{Execution Strategy}. This structure surrounds the actual
+\emph{Function}s and underlying \emph{Statement}s and allows for interpretation
+and fusion of the algorithms.
+
+\subsection{Function Code Fusion}
+
+\TODO
 

### 2014-10-16 paper.tex
UK -> US

@@
 a functional level. A code generator then fuses these algorithms and produces
-more optimally organised source code. This way, \NAME addresses the
+more optimally organized source code. This way, \NAME addresses the
 heterogeneity of both the IoT and ID, as well as the limited resources of IoT
@@
 security perimeter diminishes, every device has to implement its own lines of
-defence. Making the IDS a local service on each device requires resources,
+defense. Making the IDS a local service on each device requires resources,
 which are another important constraint: IoT devices typically have limited
@@
 required basic functional components to implement actual algorithms
-\cite{krontiris2008lidea}, but don't offer a way to optimise the usage of
+\cite{krontiris2008lidea}, but don't offer a way to optimize the usage of
 resources. They also don't support systematic reuse to create different
@@
 The second contribution introduces functional code fusion (FCF) as a generation
-paradigm. FCF identifies common data and functions, and organises code in such
+paradigm. FCF identifies common data and functions, and organizes code in such
 a way that redundant iterations, tests and computations are eliminated. Our
@@
 ID and code generation for the IoT. Finally, section \ref{conclusion}
-summarises our findings, draws conclusions and identifies topics for future
+summarizes our findings, draws conclusions and identifies topics for future
 work.
@@
     \ForEach{$byte \in msg$} \label{alg:id-algo-pattern-loop1}
-     \State \dots \Comment{analyse byte-sequences}
+     \State \dots \Comment{analyze byte-sequences}
     \EndFor
@@
 
-Starting from scratch allows optimising the implementation of an IDS, but
+Starting from scratch allows optimizing the implementation of an IDS, but
 doesn't scale. Reusing existing implementations also doesn't help, due to the
@@
 A semantic model \cite{fowler2010domain}, tied closely to the DSL itself, can
-fulfil this role. Semantic transformations can now be applied to create a
+fulfill this role. Semantic transformations can now be applied to create a
 code-oriented model, comparable to the idea of CodeDom. From there on,
@@
 more technical one, a code model. This code model can be considered as a rich
-AST with an extensive set of language constructs. Syntactic transformations now
-lower this level of expressiveness according to the desired platform and
-language. In the end the code model can be directly emitted to the target
-language.
+abstract syntax tree (AST) with an extensive set of language constructs.
+Syntactic transformations now lower this level of expressiveness according to
+the desired platform and language. In the end the code model can be directly
+emitted to the target language.
 
@@
 algorithms in a way that allows semantic reinterpretation. This way, the
-functionality of different algorithms can be reorganised to allow for more
+functionality of different algorithms can be reorganized to allow for more
 optimal execution and memory usage. Hence the name \NAME of the framework:
-intrusion detection functionality organisation optimisation.
+intrusion detection functionality organization optimization.
 
@@
 expectations: both the number of frames as the bytes sent benefit from well
-organised code and reuse of a common software library.
+organized code and reuse of a common software library.
 
@@
 
-Without any effort to optimise the software library, nor leveraging knowledge
+Without any effort to optimize the software library, nor leveraging knowledge
 about the specific algorithms, this very basic generated code shows that the

### 2014-10-16 paper.tex
added design FCF & Code Model

@@
 
-\subsection{Function Code Fusion}
+\subsection{Functional Code Fusion}
 
-\TODO
+Thanks to the functional description of events and the effects that are
+triggered by them, the code generator is able to identify the same actions
+across different algorithms. This way it is possible to extract the parsing of
+incoming messages, group actions on the same sets of data or scheduled function
+executions.
+
+FCF typically looks for patterns in the semantic model and performs model
+transformations to change these patterns to a more optimized organization.
+Semantic transformations are performed on the semantic model itself and end
+with a transformation of the semantic model into a code model, which encodes
+the semantics into technical code constructs.
+
+\subsection{Code Model}
+
+The code model can be compared to an AST for a very expressive language. It
+contains constructs found in certain languages, but not in others. Therefore a
+second round of model transformations needs to be performed. These syntactic
+transformations morph the unsupported constructs into comparable constructs
+that are supported by the targeted platform and language.
+
+Depending on the target platform, some of the constructions used to e.g. parse
+incoming messages, schedule tasks,\dots may not be transformed into
+self-containing code. The generator then uses a software library to perform
+these tasks. It will look for patterns that indicate any of these actions and
+transform them into calls into the \NAME software library.
 

### 2014-10-17 paper.tex
started rewrite, added old id algo classification

@@
 \title{
-Introducing the \NAME Framework\\
-for Efficient Intrusion Detection\\
+The \NAME Framework:\\
+Efficient Intrusion Detection\\
 in the Internet of Things
@@
 
-Along with its potential, the IoT also presents a significant threat: by
-opening our smart homes \cite{aldrich2003smart} and personal data to all these
+The IoT presents a great potential to positively influence our daily work and
+life: domotics, assisted living, e-health and enhanced learning are only a few
+examples\cite{atzori2010internet}. But with this potential the IoT also
+presents a significant threat: by opening our smart homes
+\cite{aldrich2003smart} and personal data\cite{weber2010internet} to all these
 interconnected devices, we also open them to everyone who is able to break
-these new, virtual locks, often without leaving a trace.
+these new, virtual locks. IoT devices typically have limited batteries, little
+processing power and memory is restricted to the bare minimum. These properties
+make it hard to add any additional service, including security.
 
@@
 the firewall, looking for patterns and optionally alerting the firewall if a
-malicious pattern is detected \cite{denning1987intrusion}.
+malicious pattern is detected\cite{denning1987intrusion}.
 
@@
 
-In mobile ad hoc networks (MANET), which make up an important part of the IoT,
-it is not possible to have this single point in the network that can oversee
-all traffic \cite{zhang2000intrusion, mishra2004intrusion}. If the outer
-security perimeter diminishes, every device has to implement its own lines of
-defense. Making the IDS a local service on each device requires resources,
-which are another important constraint: IoT devices typically have limited
-batteries, little processing power and memory is restricted to the bare minimum.
+In wireless networks of resource constraint devices, which make up an important
+part of the IoT, it is not possible to have this single point in the network
+that can oversee all traffic \cite{zhang2000intrusion, mishra2004intrusion}. If
+the outer security perimeter diminishes, every device has to implement its own
+lines of defense. Making the IDS a local service on each device requires
+resources, which are already very constraint.
 
@@
 Addressing this in a transparent and automated way is a key success factor for
-any kind of development in this area. For a non-functional layer, such as ID,
-this aspect becomes even more paramount.
-
-A solution that supports the integration of multiple algorithms on IoT devices
-needs to consider the impact on their resources. Different algorithms should be
-merged to avoid accumulating their impact.
+any kind of development in this area. A solution that supports the integration
+of multiple algorithms on IoT devices needs to consider the impact on their
+resources. Different algorithms should be merged to avoid accumulating their
+impact.
 
@@
 
-The remainder of this paper proceeds as follows: section \ref{background}
-details some of the ideas behind FCF and identifies the different aspects of
-the problem space. Section \ref{design} describes the design we applied to
-construct the framework, its domain specific language (DSL) and code generator.
-Section \ref{evaluation} evaluates an implementation and application of the
-framework. Section \ref{related} explores related work in the field of DSLs for
-ID and code generation for the IoT. Finally, section \ref{conclusion}
-summarizes our findings, draws conclusions and identifies topics for future
-work.
+The remainder of this paper proceeds as follows: section \ref{landscape} takes
+a look at ID algorithms and classifies them. Section \ref{analysis} looks for
+patterns in the implementation of these classes of algorithms, details some of
+the ideas behind FCF and identifies the different aspects of the problem space.
+Section \ref{design} describes the design we applied to construct the
+framework, its domain specific language (DSL) and code generator. Section
+\ref{evaluation} evaluates an implementation and application of the framework.
+Section \ref{related} explores related work in the field of DSLs for ID and
+code generation for the IoT. Finally, section \ref{conclusion} summarizes our
+findings, draws conclusions and identifies topics for future work.
 
-\section{Background}
-\label{background}
+\section{Landscape of ID Algorithms}
+\label{landscape}
 
-The underlying problem addressed in this paper is of an economic nature: adding
-ID doesn't add functionality, so its impact should be reduced as much as
-possible.
+Research into ID in wireless networks typically focuses on two major topics
+that complement the architecture: the devices as single entities and the
+network as a group of such devices. Both are important because the group cannot
+make a decision without members that detect malicious behavior and a
+group-based decision is often needed because nodes can easily miss out on
+certain events that could indicate intrusions, due to their wireless and
+not-always-on nature.
+
+\subsection{Detecting Intrusions}
+\label{subsection:detecting}
+
+There are different ways to construct a taxonomy for intrusion detection, but
+common themes do appear. According to \cite{mishra2004intrusion} and
+\cite{ioannis2007towards} there are three major categories: anomaly detection,
+signature or misuse detection, and specification-based detection. In
+\cite{alrajeh2013intrusion} the authors mostly agree with this topology, but
+add the notion of hybrid intrusion detection systems and cross layer intrusion
+detection systems as recent advances in research. Although these additional
+categories offer very interesting prospects, the authors have to admit that the
+impact on the resource-constrained sensor nodes might still be too high.
+
+The meshed network topology of WSNs and its routing protocols have produced
+many related attacks and each of these attacks has been the focus of many
+research papers presenting algorithms to detect them. In the following
+paragraphs we take a look at some popular ones.
+
+In a wormhole attack, an artificial low-latency link is created between two
+distant nodes in the network. This causes the routing protocol to divert more
+traffic through this link, offering the attacker an abundance of packets to
+inspect. In \cite{maheshwari2007detecting} an algorithm is proposed to allow
+nodes to determine if a wormhole is actively present. In essence, the algorithm
+relies on the exchange of neighbour lists to allow nodes to determine if
+unexpected substructures are present in the graph representing the network.
+
+Another attack that is related to the wormhole is the sinkhole. When launching
+a sinkhole attack, as described in \cite{krontiris2008launching}, the attacker
+first needs to compromise an existing node. Being in control of the node, the
+attacker can make his captured node look more \emph{attractive} to the
+surrounding nodes and thus draw more traffic to itself. The sinkhole attack is
+typically a foundation for other routing level attacks, such as selective
+forwarding, modification of packets or even dropping them altogether. In its
+simplest form, the sinkhole offers the attacker with lots of packets to analyze
+and use to gather information about the network and its functionality.
+
+Detecting sinkholes is very hard and is often only possible based on other
+attacks that are supported by the sinkhole. When the sinkhole is used to
+implement selective forwarding, \cite{ngai2006intruder} proposes an algorithm
+that allows a base station, the aggregating master node to which all nodes
+typically send their information, to observe missing data from nodes in the
+same area in a statistical way.
+
+More typical attacks such as flooding, the sybil attack, rushing\dots are
+amongst others presented in \cite{wood2002denial} and \cite{djenouri2005survey}.
+
+\subsection{Cooperative Decision Making}
+\label{subsection:coorperative}
+
+As in many other situations, a group is often stronger than the sum of its
+individual components. This is surely the case for WSNs. Because not all nodes
+are always actively participating in the network, some nodes might miss clues
+that would otherwise lead them to detect intrusions. It is hardly impossible
+for a single node to detect a complex attack by itself. Therefore a second
+important research topic consists of cooperative algorithms to combine
+information from single nodes into a group-based decision about alleged
+intrusions.
+
+In \cite{krontiris2009cooperative}, the authors first present a theoretical
+foundation to analyze cooperative algorithms. Given these foundations they
+continue to present an algorithm consisting of 5 phases. It essentially
+implements a voting system based on the Guy Fawkes protocol
+\cite{anderson1998new} to identify an intruder. It enables the exchange of
+suspected intruder information and allows distributed authentication of those
+\emph{votes}. Based on these authenticated votes, a distributed decision can be
+made about the commonly identified intruder.
+
+A distributed, cooperative algorithm can sometimes be implemented locally. This
+is illustrated by the reputation-based detection algorithm introduced in
+\cite{ganeriwal2008reputation}. Using this algorithm, nodes can exchange
+information about the reputation of other nodes. Combining this information
+allows them to decide about their trust in a given node, based on more than
+their own, often partial, observations.
+
+\subsection{Software Attestation}
+\label{subsection:attestation}
+
+A research topic that exists in parallel to the detection and cooperation
+duality is software attestation. Software attestation offers algorithms for
+exchanging information about the software that is running on a node, and aims
+to identify nodes that can no longer prove that their content is unaltered.
+Examples of evolving algorithms to implement this functionality include SWATT
+\cite{seshadri2004swatt}, ICE/SCUBA \cite{seshadri2006scuba} and SAKE
+\cite{seshadri2008sake}.
+
+Software attestation is hard and many details can cause an algorithm to fail.
+Interesting is the discussion surrounding this topic that was started by
+\cite{castelluccia2009difficulty}, in response to the previously mentioned
+papers. In \cite{perrig2010refutation} the authors of the original papers
+counter many of the objections made to their work, but the general feeling is
+that even in the best conditions, there is always a way to circumvent even the
+most ingenious algorithm to perform software-based attestation.
+
+\section{Analysis of ID Algorithms}
+\label{analysis}
+
+\TODO
 

### 2014-10-17 paper.tex
clean up of classification of ID algorithms

@@
 \interdisplaylinepenalty=2500
-% \usepackage{algorithmic}
 \usepackage{array}
@@
 % additional packages and utility commands
+
 \newcommand{\TODO}{\textbf{\color{red}TODO}}
 
-\usepackage{minted}
 \usepackage{flushend}
@@
 \title{
-The \NAME Framework:\\
+\NAME: A framework for \\
 Efficient Intrusion Detection\\
@@
 \cite{aldrich2003smart} and personal data\cite{weber2010internet} to all these
-interconnected devices, we also open them to everyone who is able to break
-these new, virtual locks. IoT devices typically have limited batteries, little
-processing power and memory is restricted to the bare minimum. These properties
-make it hard to add any additional service, including security.
+interconnected devices, we also open them to everyone who is able to break the
+virtual locks that protect them. IoT devices typically have limited batteries,
+little processing power and memory is restricted to the bare minimum. These
+properties make it hard to add any additional service, including security.
 
@@
 
-In wireless networks of resource constraint devices, which make up an important
-part of the IoT, it is not possible to have this single point in the network
-that can oversee all traffic \cite{zhang2000intrusion, mishra2004intrusion}. If
-the outer security perimeter diminishes, every device has to implement its own
-lines of defense. Making the IDS a local service on each device requires
-resources, which are already very constraint.
+In wireless networks of resource constraint devices, which make up an
+significant part of the IoT, it is not possible to have this single point in
+the network that can oversee all traffic\cite{zhang2000intrusion,
+mishra2004intrusion}. If the outer security perimeter diminishes, every device
+has to implement its own lines of defense. But making the IDS a local service
+on each device requires local resources, which are the scarce commodity to IoT
+devices.
 
@@
 Current developments in ID focus on programming frameworks to structure the
-implementation of different algorithms \cite{valero2012di} and offer the
-required basic functional components to implement actual algorithms
-\cite{krontiris2008lidea}, but don't offer a way to optimize the usage of
-resources. They also don't support systematic reuse to create different
-configurations for different nodes or evolution in time.
+implementation of algorithms\cite{valero2012di} and offer the required basic
+functional components to implement them\cite{krontiris2008lidea}. But they
+don't offer a way to optimize the usage of resources. Nor do they support
+systematic reuse to create different configurations for different devices.
 
@@
 of multiple algorithms on IoT devices needs to consider the impact on their
-resources. Different algorithms should be merged to avoid accumulating their
-impact.
+resources. Different algorithms should be merged somehow to avoid accumulating
+their impact.
 
@@
 configuration. This way, the framework addresses the heterogeneous nature of
-both the IoT, as a hardware and software platform, and ID algorithms.
+both the IoT and ID algorithms.
 
@@
 The second contribution introduces functional code fusion (FCF) as a generation
-paradigm. FCF identifies common data and functions, and organizes code in such
-a way that redundant iterations, tests and computations are eliminated. Our
-results show substantially improved memory usage, execution time and usage of
-the wireless radio, reducing energy consumption.
+paradigm to solve the problem of hard to reuse and combine ID algorithms. FCF
+identifies common data and functions, and organizes code in such a way that
+redundant iterations, tests and computations are eliminated. Our results show
+substantially improved memory usage, execution time and usage of the wireless
+radio, reducing energy consumption.
 
@@
 
-The remainder of this paper proceeds as follows: section \ref{landscape} takes
-a look at ID algorithms and classifies them. Section \ref{analysis} looks for
-patterns in the implementation of these classes of algorithms, details some of
-the ideas behind FCF and identifies the different aspects of the problem space.
-Section \ref{design} describes the design we applied to construct the
+The remainder of this paper proceeds as follows: section \ref{classification}
+takes a look at ID algorithms and classifies them. Section \ref{pattern} looks
+for patterns in the implementation of these classes of algorithms, details some
+of the ideas behind FCF and identifies the different aspects of the problem
+space. Section \ref{design} describes the design we applied to construct the
 framework, its domain specific language (DSL) and code generator. Section
@@
 
-\section{Landscape of ID Algorithms}
-\label{landscape}
+\section{A Classification of ID Algorithms}
+\label{classification}
 
@@
 \subsection{Detecting Intrusions}
-\label{subsection:detecting}
+\label{detection}
 
@@
 
-The meshed network topology of WSNs and its routing protocols have produced
-many related attacks and each of these attacks has been the focus of many
-research papers presenting algorithms to detect them. In the following
-paragraphs we take a look at some popular ones.
+\subsection{Cooperative Decision Making}
+\label{coorperative}
 
-In a wormhole attack, an artificial low-latency link is created between two
-distant nodes in the network. This causes the routing protocol to divert more
-traffic through this link, offering the attacker an abundance of packets to
-inspect. In \cite{maheshwari2007detecting} an algorithm is proposed to allow
-nodes to determine if a wormhole is actively present. In essence, the algorithm
-relies on the exchange of neighbour lists to allow nodes to determine if
-unexpected substructures are present in the graph representing the network.
-
-Another attack that is related to the wormhole is the sinkhole. When launching
-a sinkhole attack, as described in \cite{krontiris2008launching}, the attacker
-first needs to compromise an existing node. Being in control of the node, the
-attacker can make his captured node look more \emph{attractive} to the
-surrounding nodes and thus draw more traffic to itself. The sinkhole attack is
-typically a foundation for other routing level attacks, such as selective
-forwarding, modification of packets or even dropping them altogether. In its
-simplest form, the sinkhole offers the attacker with lots of packets to analyze
-and use to gather information about the network and its functionality.
-
-Detecting sinkholes is very hard and is often only possible based on other
-attacks that are supported by the sinkhole. When the sinkhole is used to
-implement selective forwarding, \cite{ngai2006intruder} proposes an algorithm
-that allows a base station, the aggregating master node to which all nodes
-typically send their information, to observe missing data from nodes in the
-same area in a statistical way.
-
-More typical attacks such as flooding, the sybil attack, rushing\dots are
-amongst others presented in \cite{wood2002denial} and \cite{djenouri2005survey}.
-
-\subsection{Cooperative Decision Making}
-\label{subsection:coorperative}
+These taxonomies focus on the act of detecting intrusions itself, the local
+algorithm so to speak. In a highly distributed environment, the orthogonal
+categorization, dealing with the distributed algorithm, is maybe even more
+important: how do algorithms deal with the information coming from the group
+they are part of?
 
 As in many other situations, a group is often stronger than the sum of its
-individual components. This is surely the case for WSNs. Because not all nodes
-are always actively participating in the network, some nodes might miss clues
-that would otherwise lead them to detect intrusions. It is hardly impossible
-for a single node to detect a complex attack by itself. Therefore a second
-important research topic consists of cooperative algorithms to combine
-information from single nodes into a group-based decision about alleged
-intrusions.
+individual components. Because not all devices are always actively
+participating in the network, some might miss clues that would otherwise lead
+them to detect intrusions. It is hardly impossible for a single device to
+detect a complex attack by itself. Therefore a second important research topic
+consists of cooperative algorithms to combine information from single
+identities into a group-based decision about alleged
+intrusions\cite{zhang2000intrusion}.
+
+A distributed, cooperative algorithm can sometimes be implemented locally. This
+is illustrated by the reputation-based detection algorithm introduced in
+\cite{ganeriwal2008reputation}, based on the same principle of a watchdog also
+found in \cite{mishra2004intrusion} Using this algorithm, nodes can exchange
+information about the reputation of other nodes. Combining this information
+allows them to decide about their trust in a given node, based on more than
+their own, often partial, observations alone.
 
@@
 continue to present an algorithm consisting of 5 phases. It essentially
-implements a voting system based on the Guy Fawkes protocol
-\cite{anderson1998new} to identify an intruder. It enables the exchange of
-suspected intruder information and allows distributed authentication of those
-\emph{votes}. Based on these authenticated votes, a distributed decision can be
-made about the commonly identified intruder.
-
-A distributed, cooperative algorithm can sometimes be implemented locally. This
-is illustrated by the reputation-based detection algorithm introduced in
-\cite{ganeriwal2008reputation}. Using this algorithm, nodes can exchange
-information about the reputation of other nodes. Combining this information
-allows them to decide about their trust in a given node, based on more than
-their own, often partial, observations.
+implements a voting system based on the Guy Fawkes
+protocol\cite{anderson1998new} to identify an intruder. It enables the exchange
+of suspected intruder information and allows distributed authentication of
+those \emph{votes}. Based on these authenticated votes, a distributed decision
+can be made about the commonly identified intruder.
 
@@
 to identify nodes that can no longer prove that their content is unaltered.
-Examples of evolving algorithms to implement this functionality include SWATT
-\cite{seshadri2004swatt}, ICE/SCUBA \cite{seshadri2006scuba} and SAKE
-\cite{seshadri2008sake}.
+Examples of a family of algorithms to implement this functionality include
+SWATT\cite{seshadri2004swatt}, ICE/SCUBA\cite{seshadri2006scuba} and
+SAKE\cite{seshadri2008sake}.
 
@@
 
-\section{Analysis of ID Algorithms}
-\label{analysis}
+\section{A Design Pattern for ID Algorithms}
+\label{pattern}
 

### 2014-10-17 paper.tex
fix float issues after removing too much
a bit of rephrasing

@@
 % additional packages and utility commands
+
+\usepackage{float}
 
@@
 
-These taxonomies focus on the act of detecting intrusions itself, the local
-algorithm so to speak. In a highly distributed environment, the orthogonal
-categorization, dealing with the distributed algorithm, is maybe even more
-important: how do algorithms deal with the information coming from the group
-they are part of?
+These taxonomies focus on the act of detecting intrusions themselves. In a
+highly distributed environment, an orthogonal categorization, dealing with the
+distributed nature of the network, may be even more important.
 
@@
 A research topic that exists in parallel to the detection and cooperation
-duality is software attestation. Software attestation offers algorithms for
-exchanging information about the software that is running on a node, and aims
-to identify nodes that can no longer prove that their content is unaltered.
-Examples of a family of algorithms to implement this functionality include
+duality is software attestation. It offers algorithms for exchanging
+information about the software that is running on a device, and aims to
+identify devices that can no longer prove that their content is unaltered. A
+family of algorithms to implement this functionality are
 SWATT\cite{seshadri2004swatt}, ICE/SCUBA\cite{seshadri2006scuba} and

### 2014-10-17 paper.tex
added intro for design pattern for ID algorithms

@@
 
+Given this classification, we now introduce a design pattern, common to these
+algorithms. It will allow us to define a strategy for optimizing the
+combination of these algorithms.
+
+From the classification and survey of several algorithms, we can extract two
+activities common to most algorithms: the inspection of network traffic and the
+evaluation of thresholds.
+
+The former is rather trivial. Network traffic is the carrier along which
+malicious payloads reach the device or where patterns of communication can be
+observed when other devices in the vicinity are affected by such payloads.
+
+The latter is more complex. An algorithm based on signature or misuse detection
+can sometimes undeniably be sure if an attack is real or not. But when dealing
+with anomaly detection this line might not be that crisp. Especially in
+distributed situations where even a neighbouring device can not be trusted,
+detecting intrusion is no absolute science. In this case one has to rely on
+statistics and thresholds to decide when to drop trust in a given device.
+
+Simply checking a threshold after updating a counter, is also not always a good
+strategy. A simple example is that of a watchdog\cite{mishra2004intrusion},
+observing the actions of other devices. There are valid reasons why a devices
+might miss its theoretical window in which certain actions are expected. Simply
+reacting on such basic threshold might result in false positives. A way to
+overcome this is to decouple the detection of problems and the decision what to
+do about it.
+
+This decoupling of detection and decision is the core of the design pattern for
+ID algorithms. Listing \ref{alg:id-algo-pattern} presents the design pattern in
+pseudo-code.
+
+\begin{figure}[!ht]
+\captionof{lstlisting}{ID Algorithm Pattern\label{alg:id-algo-pattern}}
+\begin{algorithmic}[1]
+  \Require{nodes, global storage for information about nodes}
+  \Function{process\_message}{$msg$}
+    \ForEach{$byte \in msg$} \label{alg:id-algo-pattern-loop1}
+     \State \dots \Comment{analyze byte-sequences}
+    \EndFor
+    \Let{$nodes_x$}{value}  \Comment{optionally update node info}
+    \State \Call{send}{$nodes_y$, ``info''} \Comment{optionally exchange info} \label{alg:id-algo-pattern-send1}
+  \EndFunction
+  \State
+  \Function{do\_housekeeping}{}
+    \ForEach{$node \in nodes$} \label{alg:id-algo-pattern-loop2} \label{alg:id-algo-pattern-common-data}
+      \If{$node > \dots$} \Comment{validate recorded value}
+        \State \dots \Comment{take actions}
+        \State \Call{send}{$node$, ``info''} \Comment{e.g. communicate} \label{alg:id-algo-pattern-send2}
+      \EndIf
+    \EndFor
+  \EndFunction
+  \State
+\end{algorithmic}
+\end{figure}
+
 \TODO
@@
 
-\begin{figure}[H]
+\begin{figure}[!ht]
 \captionof{lstlisting}{Sequential implementation of ID algorithms\label{alg:id-algo-application}}
@@
 listing \ref{alg:id-algo-pattern}.
-
-\begin{figure}[H]
-\captionof{lstlisting}{ID Algorithm Pattern\label{alg:id-algo-pattern}}
-\begin{algorithmic}[1]
-  \Require{nodes, global storage for information about nodes}
-  \Function{process\_message}{$msg$}
-    \ForEach{$byte \in msg$} \label{alg:id-algo-pattern-loop1}
-     \State \dots \Comment{analyze byte-sequences}
-    \EndFor
-    \Let{$nodes_x$}{value}  \Comment{optionally update node info}
-    \State \Call{send}{$nodes_y$, ``info''} \Comment{optionally exchange info} \label{alg:id-algo-pattern-send1}
-  \EndFunction
-  \State
-  \Function{do\_housekeeping}{}
-    \ForEach{$node \in nodes$} \label{alg:id-algo-pattern-loop2} \label{alg:id-algo-pattern-common-data}
-      \If{$node > \dots$} \Comment{validate recorded value}
-        \State \dots \Comment{take actions}
-        \State \Call{send}{$node$, ``info''} \Comment{e.g. communicate} \label{alg:id-algo-pattern-send2}
-      \EndIf
-    \EndFor
-  \EndFunction
-  \State
-\end{algorithmic}
-\end{figure}
 

### 2014-10-17 paper.tex
rewrote pattern analysis

@@
 
-\TODO
-
-In literature, many ID algorithm are described, but hardly any are implemented.
-If implementations would be available, one would typically apply a pattern as
-shown in listing \ref{alg:id-algo-application}.
+We now extend the scope of the design pattern, to include its application. When
+implementing different algorithms, one will typically encounter a pattern such
+as describes in listing \ref{alg:id-algo-application}.
 
@@
 
-The implementations would be linked and called in a sequential way, when new
-messages are observed on the network and at a regular interval to allow them to
-do internal housekeeping. This includes processing the information extracted
-from the network traffic, often in a statistical way.
+The different implementations would be linked and called in a sequential way,
+when new messages are observed on the network and at a regular interval to
+allow them to do internal housekeeping.
 
-This duality is a common design pattern of ID algorithms targeting MANETs. The
-internals of the pattern reveal more common structures, as illustrated in
-listing \ref{alg:id-algo-pattern}.
-
-The parts of interest here are the loops at lines
-\ref{alg:id-algo-pattern-loop1} and \ref{alg:id-algo-pattern-loop2}, the common
-\emph{nodes} data on line \ref{alg:id-algo-pattern-common-data} and the
-interaction with the network at lines \ref{alg:id-algo-pattern-send1} and
-\ref{alg:id-algo-pattern-send2}.
+Given these patterns, we can now identify the problem areas that tend to result
+in suboptimal code when combining multiple ID algorithms.
 
@@
 
-Given several ID algorithms, that all execute the same parsing loop, over the
-same byte stream, performing comparable operations to find similar patterns,
-one would clearly benefit from combining these loops, avoiding redundant
-reevaluation of the same byte patterns.
+Two issues arise from the nested loops that originate from the combination of
+the ID algorithm patterns and its application pattern. First, each algorithm,
+by itself, will parse each observed message on the network, processing the same
+byte stream, performing comparable operation to find similar patterns.
 
-The same goes for common data. Each algorithm not only has to parse the same
-message, it also has to store information about nodes within its own scope,
-resulting in a lot of duplicated information in memory.
+Second, there is common data: Each algorithm also has to store information
+about nodes within its own scope, resulting in a lot of duplicated information
+in memory.
 
-The only way to solve this, would be to change the provided implementation,
-pulling up the loops over the message and the nodes to a level above the loop
-over the algorithms. But this comes at a development cost. This cost typically
-isn't a one-time cost, e.g. in case a new version of the algorithm is released,
-or simply when a different set of algorithms is selected due to a changing risk
-analysis. Finally, changing, or simply implementing the algorithms from
-scratch, also holds the risk of making mistakes.
+Reimplementing the algorithms to eliminate these issues is not a valid option.
+This development cost typically isn't a one-time cost, e.g. in case a new
+version of the algorithm is released, or simply when a different set of
+algorithms is selected due to a changing risk analysis. Finally, changing, or
+simply implementing the algorithms from scratch, also holds the risk of making
+mistakes.
 
@@
 
-But there are hardly any implementations readily available. Probably the main
-reason for this is the heterogeneity of the IoT. A prime example is the
-\ttt{SEND} function. Different software stacks will have different application
-programming interfaces (APIs) for accessing the network. Simple, macro-like
-solutions can overcome this, but the dispersion of the calls to access the
-network is causing the wireless radio to be on a lot more that one would hope
-for.
+A third issue is hidden in the use of the network. The calls to the \ttt{SEND}
+function are scattered around the algorithm, causing the wireless radio to be
+accessed multiple times, keeping it active and energy consuming.
 
-A valid solution would try to combine messages in a single instance, reducing
-the access to the network and even sharing some network overhead. The impact to
-implement this manually will again outweigh the gain. An option would be to use
-a software library, that provides a unified API for accessing the network and
-collects individual calls and combines them in a single network message.
+The heterogeneity of the IoT devices, results in many different software
+stacks, each with their own network API. This also explains why there are very
+few implementations readily available to developers, their is no common
+platform to target.
 
-Besides the fact that all researchers would have to use the same library in a
-consistent way, they would limit their operational scope to a single language,
-serving only a fraction of the heterogeneous IoT.
-
-\subsection{Situation Summary}
-
-Starting from scratch allows optimizing the implementation of an IDS, but
-doesn't scale. Reusing existing implementations also doesn't help, due to the
-far from optimal sequential execution of the algorithms and the redundant
-storage of common data. Software libraries can solve many of the issues, but
-only cover parts of the entire spectrum, due to their tie-in with languages and
-frameworks.
+Even using a unified software library would not only require a consistent
+implementation of it, but would also limit the operational scope of the
+research implementation, still restricting it to a single language.
 

### 2014-10-17 paper.tex
disabled flushend, causing issues on last page
started rewrite design + added remaining TODO’s

@@
 
-\usepackage{flushend}
+% \usepackage{flushend}
 
@@
 of ID algorithms, a code generator is now able to combine these, while avoiding
-the sequential patterns/problems presented in listings
-\ref{alg:id-algo-application} and \ref{alg:id-algo-pattern}. It can also target
-different languages and platforms, while reusing a software library that
-provides e.g. message parsing, network access wrapping,\dots
+the problems arising from the sequential patterns/problems presented in
+listings \ref{alg:id-algo-application} and \ref{alg:id-algo-pattern}. It can
+also target different languages and platforms, while reusing a software library
+that provides e.g. message parsing, network access wrapping,\dots
 
 \subsection{DSL}
+
+\TODO
 
@@
 
+\TODO
+
 IDS in WSN \cite{perrig2004security,mishra2004intrusion}

### 2014-10-20 paper.tex
some more rewriting of background

@@
 
+These two topics lead to the first two classes: detecting intrusions and
+cooperative decision making. We also identify software attestation as a third
+class. Although it can be considered as a special case of a combination of both
+other classes, its focus is not on the detection of intrusions, but on the
+effect of an intrusion on the system itself.
+
 \subsection{Detecting Intrusions}
@@
 categories offer very interesting prospects, the authors have to admit that the
-impact on the resource-constrained sensor nodes might still be too high.
+impact on the resource-constrained devices might still be too much. We
+therefore focus on the three categories agreed upon by most sources.
+
+\begin{LaTeXdescription}
+
+  \item[Anomaly detection] considers a baseline profile and evaluates all
+  activity according to this profile. Aberrant behavior is flagged as a
+  possible intrusion, but this is no exact science. The baseline profile can
+  evolve over time, especially in wireless and mobile environments. The flagged
+  incidents should therefore also be handled with care and no absolute
+  decisions can be made from single events.
+
+  \item[Signature or misuse detection] starts with known patterns of attacks
+  and tries to find evidence in the actions of and communication with a device.
+  Because many attacks use valid ways to communicate with services on the
+  devices, this can also produce false positives and offers no black and white
+  certainty.
+
+  \item[Specification-based detection] uses constraints that describe the
+  correct behavior of the system. Although that this approach can detect
+  previously unknown attacks, it can sometimes not detect specific attacks that
+  can be detected by anomaly or misuse detection.
+
+\end{LaTeXdescription}
 
@@
 highly distributed environment, an orthogonal categorization, dealing with the
-distributed nature of the network, may be even more important.
+distributed nature of the network and its ways of communicating, may be even
+more important.
 
@@
 identities into a group-based decision about alleged
-intrusions\cite{zhang2000intrusion}.
+intrusions\cite{zhang2000intrusion}. We identify two ways by which this
+cooperative decision making can be constructed and focus on the way
+participants exchange information: broadcasting and interactive.
 
-A distributed, cooperative algorithm can sometimes be implemented locally. This
-is illustrated by the reputation-based detection algorithm introduced in
-\cite{ganeriwal2008reputation}, based on the same principle of a watchdog also
-found in \cite{mishra2004intrusion} Using this algorithm, nodes can exchange
-information about the reputation of other nodes. Combining this information
+\begin{LaTeXdescription}
+
+  \item[Broadcasting] can sometimes allow a distributed, cooperative algorithm
+  to be implemented locally. This is illustrated by the reputation-based
+  detection algorithm introduced in \cite{ganeriwal2008reputation}, based on
+  the same principle of a watchdog also found in \cite{mishra2004intrusion}
+  Using this algorithm, nodes can exchange information about the reputation of
+  other nodes, using only broadcasted messages. Combining this information
   allows them to decide about their trust in a given node, based on more than
@@
 
-In \cite{krontiris2009cooperative}, the authors first present a theoretical
-foundation to analyze cooperative algorithms. Given these foundations they
-continue to present an algorithm consisting of 5 phases. It essentially
-implements a voting system based on the Guy Fawkes
-protocol\cite{anderson1998new} to identify an intruder. It enables the exchange
-of suspected intruder information and allows distributed authentication of
-those \emph{votes}. Based on these authenticated votes, a distributed decision
-can be made about the commonly identified intruder.
+  \item[Interactive] communication allows algorithms to exchange information to
+  reach a consensus. In \cite{krontiris2009cooperative}, the authors first
+  present a theoretical foundation to analyze cooperative algorithms. Given
+  these foundations they continue to present an algorithm consisting of 5
+  phases. It essentially implements a voting system based on the Guy Fawkes
+  protocol\cite{anderson1998new} to identify an intruder. It enables the
+  exchange of suspected intruder information and allows distributed
+  authentication of those \emph{votes}. Based on these authenticated votes, a
+  distributed decision can be made about the commonly identified intruder.
+
+\end{LaTeXdescription}
 
@@
 
-Software attestation is hard and many details can cause an algorithm to fail.
-Interesting is the discussion surrounding this topic that was started by
-\cite{castelluccia2009difficulty}, in response to the previously mentioned
-papers. In \cite{perrig2010refutation} the authors of the original papers
-counter many of the objections made to their work, but the general feeling is
-that even in the best conditions, there is always a way to circumvent even the
-most ingenious algorithm to perform software-based attestation.
+Software attestation is hard and many details can cause an algorithm to fail. A
+discussion surrounding this topic, started by \cite{castelluccia2009difficulty}
+in response to the previously mentioned papers, illustrates this. In
+\cite{perrig2010refutation} the authors of the original papers counter many of
+the objections made to their work, but the general feeling is that even in the
+best conditions, there is always a way to circumvent the most ingenious
+algorithm to perform software-based attestation.
+
+Within the context of this paper, we will not consider software attestation as
+a special class and exclude it from our scope, only focusing on detecting
+intrusions and distributed decision making.
+
+Figure \ref{fig:classification} gives a visual overview of the classification
+presented above.
+
+\begin{figure}[!ht]
+  \centering
+  \scalebox{.85}{
+\begin{tikzpicture}[
+  sibling distance=20mm,node distance=40mm
+  ]
+  \node (detection)             [align=center] {Detecting\\Intrusions}
+    child {node (anomaly)       [align=center] {Anomaly}}
+    child {node (signature)     [align=center] {Signature\\Misuse}}
+    child {node (specification) [align=center] {Specification}};
+  \node (decision)              [align=center,right=of detection] {Decision\\Making}
+    child {node (broadcasting)  [align=center] {Broadcasting}}
+    child {node (interactive)   [align=center] {Interactive}};
+\end{tikzpicture}
+}
+  \caption{Classification of ID Algorithms}
+  \label{fig:classification}
+\end{figure}
 
@@
 
-Given this classification, we now introduce a design pattern, common to these
-algorithms. It will allow us to define a strategy for optimizing the
-combination of these algorithms.
+Given this classification, we now introduce a design pattern that identifies
+common aspects of these algorithms. It will allow us to define a strategy for
+optimizing the combination of these algorithms. From the classification and
+survey of several algorithms, we extract two activities common to most
+algorithms: the inspection of network traffic and the evaluation of thresholds.
 
-From the classification and survey of several algorithms, we can extract two
-activities common to most algorithms: the inspection of network traffic and the
-evaluation of thresholds.
+The former is trivial: network traffic is the carrier along which malicious
+payloads reach the device or where patterns of communication can be observed
+when other devices in the vicinity are affected by such payloads.
 
-The former is rather trivial. Network traffic is the carrier along which
-malicious payloads reach the device or where patterns of communication can be
-observed when other devices in the vicinity are affected by such payloads.
-
-The latter is more complex. An algorithm based on signature or misuse detection
-can sometimes undeniably be sure if an attack is real or not. But when dealing
-with anomaly detection this line might not be that crisp. Especially in
-distributed situations where even a neighbouring device can not be trusted,
+The latter is more complex: an algorithm based on signature or misuse detection
+maybe sometimes can undeniably decide if an attack is real or not. But when
+dealing with anomaly detection this line surely isn't that crisp. Especially in
+distributed situations where even a neighboring device can not be trusted,
 detecting intrusion is no absolute science. In this case one has to rely on
@@
 
-Simply checking a threshold after updating a counter, is also not always a good
-strategy. A simple example is that of a watchdog\cite{mishra2004intrusion},
-observing the actions of other devices. There are valid reasons why a devices
-might miss its theoretical window in which certain actions are expected. Simply
-reacting on such basic threshold might result in false positives. A way to
-overcome this is to decouple the detection of problems and the decision what to
-do about it.
+Simply checking a threshold after updating some counter, is also not always a
+good strategy. A simple example is that of a
+watchdog\cite{mishra2004intrusion}, observing the actions of other devices.
+There are valid reasons why a devices might miss the theoretical window in
+which certain actions are expected. Simply reacting on such basic threshold
+might result in false positives. A way to overcome this is to decouple the
+detection of problems and the decision what to do about them.
 
@@
 
-We now extend the scope of the design pattern, to include its application. When
-implementing different algorithms, one will typically encounter a pattern such
-as describes in listing \ref{alg:id-algo-application}.
+We now extend the scope of the design pattern, to include its application in
+the context of multiple algorithms. When implementing different algorithms, one
+will typically encounter a pattern such as describes in listing
+\ref{alg:id-algo-application}.
 
@@
 when new messages are observed on the network and at a regular interval to
-allow them to do internal housekeeping.
+allow them to do internal housekeeping, such as threshold evaluation.
 
@@
 
-Reimplementing the algorithms to eliminate these issues is not a valid option.
-This development cost typically isn't a one-time cost, e.g. in case a new
-version of the algorithm is released, or simply when a different set of
+Reimplementing the algorithms manually to eliminate these issues is not a valid
+option. This development cost typically isn't a one-time cost, e.g. in case a
+new version of the algorithm is released, or simply when a different set of
 algorithms is selected due to a changing risk analysis. Finally, changing, or
@@
 
-\subsection{Network Usage}
+\subsection{Network Usage and Heterogeneity}
 
@@
 stacks, each with their own network API. This also explains why there are very
-few implementations readily available to developers, their is no common
-platform to target.
+few implementations readily available to developers: there is no common
+platform to target for researchers.
 
@@
 
-Listing \ref{lst:heartbeat} illustrates some of the features mentioned above,
-using a small but representative example of an algorithm: a heartbeat. Using a
-heartbeat, nodes can broadcast a message at a regular interval, allowing other
-nodes to monitor its availability. Missing heartbeats could indicate problems,
-such as physical capturing of a certain node.
+Listing \ref{lst:watchdog} illustrates some of the features mentioned above,
+using a small but representative example of an algorithm: a
+watchdog\cite{mishra2004intrusion}. Using a heartbeat, nodes can broadcast a
+message at a regular interval, allowing other nodes to monitor its
+availability. Missing heartbeats could indicate problems, such as physical
+capturing of a certain node.
 
@@
   basicstyle=\small\ttfamily,
-  label=lst:heartbeat,
+  label=lst:watchdog,
   caption=Heartbeat in \NAME
-]{src/heartbeat.foo}
+]{src/watchdog.foo}
 

### 2014-10-21 paper.tex
added foo-lang syntax highlighting

@@
 capturing of a certain node.
+\lstdefinelanguage{foo-lang}{
+  emph={module,const,from,import,extend,with,after,do,function,@every,having},
+  emphstyle={\textbf},
+  morecomment=[l]{//},
+  moredelim=[is][]{@@}{@@},       % temp solution to allow keywords without emph
+  moredelim=[is][\emph]{!!}{!!}   % temp solution to hightlight #atoms
+}
 
 \lstinputlisting[
+  language=foo-lang,
   float,
-  basicstyle=\small\ttfamily,
+  basicstyle=\footnotesize\ttfamily\color{black},
   label=lst:watchdog,
-  caption=Heartbeat in \NAME
+  caption=Watchdog in \NAME
 ]{src/watchdog.foo}

### 2014-10-21 paper.tex
completed draft rewrite of background/design

@@
   these foundations they continue to present an algorithm consisting of 5
-  phases. It essentially implements a voting system based on the Guy Fawkes
+  phases. It essentially implements a voting system-based on the Guy Fawkes
   protocol\cite{anderson1998new} to identify an intruder. It enables the
@@
 of ID algorithms, a code generator is now able to combine these, while avoiding
-the problems arising from the sequential patterns/problems presented in
-listings \ref{alg:id-algo-application} and \ref{alg:id-algo-pattern}. It can
-also target different languages and platforms, while reusing a software library
-that provides e.g. message parsing, network access wrapping,\dots
+the problems arising from the patterns/problems presented in listings
+\ref{alg:id-algo-application} and \ref{alg:id-algo-pattern}. It can also target
+different languages and platforms, while reusing a software library that
+provides e.g. message parsing, network access wrapping,\dots
 
-\subsection{DSL}
+\subsection{The Meaning of Loops}
 
-\TODO
+From the design patterns we learned that loops are a major contributor to the
+problem of algorithm fusibility. Loops in program code are technical constructs
+and in fact often hide the actual intended functionality.
 
-The DSL component of \NAME has two major goals and one minor: (1) allow the
-description of ID algorithms in a formal way and (2) avoid that those
-descriptions contain \emph{non-fusible} logic and/or data. Besides these (3)
-the usability of the language is also taken into account.
+To illustrate this, we take the example of parsing network messages. Each
+algorithm deals with parsing in the same way: loop over the bytes in the
+payload and look for a pattern that identifies what the algorithm is looking
+for. This is in fact a technical implementation choice. The underlying
+functional goal is to react when such pattern is present in a message. The loop
+has \emph{activated} this functionality, but the real goal is to
+\emph{passively} react on an event, not \emph{actively} look for it.
 
-The first goal is covered by the construction of the language itself. Given a
-description of an ID algorithm using the DSL, the algorithm can be
-unambiguously interpreted. The DSL borrows a lot of its syntax from C-like
-languages, dropping only a small subset of its capabilities, but replacing
-those with other higher-level constructs taken from object-oriented (OO)
-languages, aspect-orientation,\dots
+This is how \NAME removes the need for loops: by introducing event handling,
+constructions with loops are transformed to their functional meaning, allowing
+much easier identification and fusion of common functionalities.
 
-The most important omission from the C syntax are basic loops. Loops are very
-generic and make it hard to identify if they deal with the same logic and/or
-data. Fusing these loops across different algorithms is often non-trivial. In
-\NAME's DSL, other constructs have been introduced to lift the definition of
-these loops to a more functional level: scheduling and scoping of the execution
-of functions, reacting to events and pattern matching are three examples that
-allow describing the actual meaning of loops in stead of simply the loop
-itself. It sometimes requires rethinking the reason behind a technical loop,
-but often results in a much clearer description.
+The same goes for actively polling some variable until it reaches a certain
+threshold. This also can be transformed to an event, eliminating the need for a
+technical loop-construction.
 
-To deal with common data, the domain provides the notion of \emph{nodes}. As
-most, if not all, ID algorithms targeting the IoT require some notion of the
-other nodes or devices in the vicinity, the DSL makes this the primary
-information carrier. All functional constructs, mentioned above, typically
-operate on the collection of known nodes, or a functionally defined subset of
-that. Each algorithm also has the possibility to \emph{extend} nodes, adding
-its own properties needed for its execution.
+\subsection{A DSL for Nodes}
 
-Based on a C-like syntax the language is also very accessible. To further
-nurture the usability, syntactic sugar from OO languages is also adopted.
-Because not everything should be implemented in such a high-level language, it
-also provides a way to \emph{import} external functionality. In the first place
-this will include a standard library of functions that can be called, but it
-can include any custom built software component. Finally it tries to limit the
-need for type information and performs as much type inference as possible. As
-it aims to be language that describes algorithms on a functional level it seeks
-to safeguard its users from these \emph{technicalities}. The fact that the
-opportunities of such a formal description include generation of code in a
-variety of languages, each with possibly different types, also results the need
-for a more functional approach to types.
+The second problem that could be observed from the patterns, was that of common
+data. All algorithms deal with information about neighboring devices. To
+centralize this information \NAME provides a unified \texttt{nodes} concept as
+part of its DSL and exposes it as an object to the implementor.
+
+Nodes can be extended with properties only available to the algorithm that
+defines them. This way only one instance for every known devices will be
+present in memory. Finally, nodes can define a scope for some function to be
+executed in. This way, nodes also offer a way to iterate all known devices,
+again without the need for a loop.
+
+\subsection{Unified Messaging}
+
+Sending and receiving of network messages is tightly integrated with the
+previously mentioned concepts of event handling and the nodes domain. An
+underlying software framework will handle all interaction with the network,
+triggering functions by means of events when new messages are parsed and
+exposing an API to send messages through the nodes domain. This approach gives
+the framework complete control over the way messages are sent, allowing
+optimized marshaling, grouping,\dots
 
@@
 using a small but representative example of an algorithm: a
-watchdog\cite{mishra2004intrusion}. Using a heartbeat, nodes can broadcast a
+watchdog\cite{mishra2004intrusion}. Like a heartbeat, nodes can broadcast a
 message at a regular interval, allowing other nodes to monitor its
 availability. Missing heartbeats could indicate problems, such as physical
-capturing of a certain node.
+capturing of a device.
+
 \lstdefinelanguage{foo-lang}{
@@
 
-To convert the DSL based ID algorithms into production-ready code, we propose
+To convert the DSL-based ID algorithms into production-ready code, we propose
 the use of a code generator. Although many approaches here are applicable, like
-template based text expansion, XSLT transformations or even using the
+template-based text expansion, XSLT transformations or even using the
 CodeDom\cite{dollard2004code}, we believe that for the level of functional
@@
 
-\TODO
-
-- incoming message parsing
-- outgoing message grouping
-- execution of schedules
-- standard library: high-level conceptual functions (e.g. now(), sha1,...)
+We already touched upon the underlying software library when discussing the
+unified messaging framework: incoming messages are parsed, outgoing messages
+grouped. Besides these examples, other common functionality also isn't
+generated. The execution of functions based on a schedule or the implementation
+of higher level conceptual functions, such as \texttt{now} and \texttt{sha1},
+are calls into the standard software library that supports the code generator.
 

### 2014-10-22 paper.tex
externalised Tikz node styling

@@
 \usetikzlibrary{arrows}
+
+\include{tikz}
 
@@
 
-\tikzset{
-  resource/.style={
-    rectangle,
-    minimum width=1cm,
-    minimum height=8mm,
-    thick, draw=black,
-    top color=white,
-    font=\ttfamily
-  },
-  representation/.style={
-    rectangle, rounded corners=3mm,
-    minimum width=2cm,
-    minimum height=8mm,
-    thick, draw=black,
-    top color=white,
-    font=\itshape
-  }
-}
-
 \begin{figure}[ht]
@@
 
-  \draw [->] (dsl) -> (sm)   node [midway,above=10pt]              {parse};
-  \draw [->] (sm)  -> (cm)   node [midway,above=10pt,align=center] {semantic\\transform};
-  \draw [->] (cm)  -> (code) node [midway,above=10pt]              {emit code};
+  \draw [->] (dsl) -> (sm)   node [midway,above=15pt]              {parse};
+  \draw [->] (sm)  -> (cm)   node [midway,above=15pt,align=center] {semantic\\transform};
+  \draw [->] (cm)  -> (code) node [midway,above=15pt]              {emit code};
 
   \draw [rounded corners,->]
-        ($ (sm.east) - (5mm,4.5mm) $)
+        ($ (sm.east) - (5mm,5mm) $)
         -- ++(0,-.5)
-        -| ($ (sm.west) + (5mm,-4.5mm) $)
+        -| ($ (sm.west) + (5mm,-5mm) $)
            node [near start, below=1pt] {type inference};
@@
   \draw [rounded corners,->]
-        ($ (cm.east) - (5mm,4.5mm) $)
+        ($ (cm.east) - (5mm,5mm) $)
         -- ++(0,-.5)
-        -| ($ (cm.west) + (5mm,-4.5mm) $)
+        -| ($ (cm.west) + (5mm,-5mm) $)
            node [near start, below=1pt,align=center] {syntactic\\transform};
@@
 implemented for \NAME.
-
-\tikzset{
-  class/.style={
-    rectangle,
-    minimum size=1cm,
-    thick, draw=black,
-    text centered,
-    text width=1.75cm,
-  },
-  subclass/.style={thick,draw=black,>=open triangle 60,<-},
-  aggregate/.style={thick,draw=black,>=open diamond,<-},
-  composite/.style={thick,draw=black,>=diamond,<-},
-  dependency/.style={thick,>=latex,->}
-}
 

### 2014-10-22 tikz.tex
externalised Tikz node styling

%!TEX root=paper.tex

% basics

\tikzset{
  box/.style            = { rectangle, minimum width=2cm, minimum height=1cm },
  border/.style         = { thick, draw=black },
  round/.style          = { rounded corners=3mm },
}

% system

\tikzset{
  artifact/.style       = { box, font=\ttfamily },
  support/.style        = { box, font=\itshape },
  user/.style           = { artifact, fill=black!15!white },
  dom/.style            = { support,  fill=black!30!white },
  platform/.style       = { support,  fill=black!55!white, text=white },
  native/.style         = { support,  fill=black!70!white, text=white },
  generated/.style      = { artifact, border },
  resource/.style       = { artifact, border },
  representation/.style = { support, border, round }
}

% UML

\tikzset{
  class/.style={
    box,
    border,
    text centered,
    text width=1.75cm
  },
  relation/.style       = { thick, draw=black, <-},
  subclass/.style       = { relation, >=open triangle 60 },
  aggregate/.style      = { relation, >=open diamond },
  composite/.style      = { relation, >=diamond },
  dependency/.style     = { relation, >=latex, -> }
}

### 2014-10-22 paper.tex
fixed heartbeat/watchdog issues + listings order

@@
 the problems arising from the patterns/problems presented in listings
-\ref{alg:id-algo-application} and \ref{alg:id-algo-pattern}. It can also target
+\ref{alg:id-algo-pattern} and \ref{alg:id-algo-application}. It can also target
 different languages and platforms, while reusing a software library that
@@
 perform once cycle of the event loop. These metrics were collected in four
-situations: without ID algorithms, with a heartbeat, with reputation-tracking
+situations: without ID algorithms, with a watchdog, with reputation-tracking
 and with both algorithms implemented.
@@
   \hline
-      & base & heartbeat & reputation & both\\
+      & base & watchdog & reputation & both\\
   \hline
@@
   \hline
-         & base & heartbeat & reputation & both\\
+         & base & watchdog & reputation & both\\
   \hline
@@
   \hline
-                & base & heartbeat & reputation & both  & result \\
+                & base & watchdog & reputation & both  & result \\
   \hline

### 2014-10-22 paper.tex
added three more citations wrt DSL for IDS
improved old evaluation text a bit

@@
 
-For the evaluation we constructed a small meshed network consisting of three
-devices, with only one device able to talk to both other devices, requiring
-routing of messages through this device.
+For the evaluation we constructed a small meshed network. In this network, not
+all devices have direct access to one another, requiring routing through shared
+neighboring devices.
 
 We started from a basic application that measures light intensity. The
-application was written both manually and generated. On top of this baseline we
-added two intrusion detection related algorithms: a heartbeat, that allows
-nodes to validate each other's continuous presence, and a reputation-building
-algorithm that checks if a parent node is cooperative and actually forwards
-messages with a destination further down the
-network\cite{ganeriwal2008reputation}.
+application was constructed as an includable piece of code. This way it could
+be integrated in an manual implementation and merged into the generated
+solution. On top of this baseline we added two intrusion detection related
+algorithms: a watchdog\cite{mishra2004intrusion}, that allows nodes to validate
+each other's continuous presence, and a reputation-building
+algorithm\cite{ganeriwal2008reputation}, that checks if a parent node is
+cooperative and actually forwards messages with a destination further down the
+network.
+
+\subsection{Evaluation}
 
@@
 
-DSL for IDS \cite{eckmann2002statl}
+DSL for IDS \cite{sekar1999high} \cite{uppuluri2001experiences} \cite{eckmann2002statl} \cite{vigna2003designing}
 

### 2014-10-22 paper.tex
added architecture diagram to evaluation

@@
 
+In this section we discuss a prototype implementation and evaluate its
+performance. We first present the architecture of this implementation and
+consider the impact it has from a development point of view. Next we use this
+implementation to quantify its merits using a side-by-side comparison of
+generated code versus a manual implementation.
+
+\subsection{Architecture}
+
+Figure \ref{fig:architecture} shows the architecture of the implementation.
+
+\begin{figure}[ht]
+  \centering
+  \scalebox{.85}{
+\begin{tikzpicture}[
+  node distance=10mm and 20mm,
+  ,>=stealth,thick,black!50,text=black,
+  every new ->/.style={shorten >=1pt},
+]
+  \node (dsl)      [user,align=center]                                     {DSL};
+  \node (code)     [generated,right=of dsl]                                {code};
+  \node (image)    [generated,right=of code]                               {image};
+
+  \node (nodes)    [dom,above right=of dsl,xshift=-20mm]                   {Nodes};
+  \node (platform) [platform,below right=of dsl,align=center,xshift=-20mm] {Platform \&\\Language};
+  \node (lib)      [native,below right=of code,align=center,xshift=-20mm]  {Software\\Library};
+
+  \draw [->] (dsl)  -> (code)  node [midway,above=2pt] {generation};
+  \draw [->] (code) -> (image) node [midway,above=2pt] {compilation};
+
+  \draw [->,dashed] (nodes.south)    -> ++(0,-15mm);
+  \draw [->,dashed] (platform.north) -> ++(0,15mm);
+  \draw [->,dashed] (lib.north)      -> ++(0,15mm);
+\end{tikzpicture}
+}
+  \caption{Architecture of Prototype Implementation}
+  \label{fig:architecture}
+\end{figure}
+
+\TODO
+
+\subsection{Implementation}
+
 To evaluate the prototype we chose to use a system based on the Atmel

### 2014-10-22 paper.tex
completed draft text for architecture evaluation

@@
 
-Figure \ref{fig:architecture} shows the architecture of the implementation.
+Figure \ref{fig:architecture} shows the architecture of the implementation. It
+illustrates the generation of the DSL into compilable code, using the Nodes
+domain information, a platform and language description and the native software
+library. The applied grayscale indicates the components' platform dependency:
+the DSL is platform and language independent and deals with subsystems, such as
+the network, in an abstract way. The Nodes domain implements some of these
+abstracted subsystems, but still in a platform and language independent way.
+The platform and language descriptions define the actual target platform and
+language and convert the abstract constructions in actually supported code. The
+software library is written in native code targeting the platform and language.
+Finally, the code and the image are artifacts of the automated generation and
+compilation process. They aren't a product of manual work and therefore aren't
+directly dependent on the platform.
 
@@
 
-\TODO
+This hierarchy is important from a development effort point of view. The level
+of dependency on the platform also dictates the level of effort to implement
+changes to the platform. For example: changing the language will require a
+complete rewrite of the software library and a partial rewrite of the language
+description of the generator. Although this is a one-time effort, it is
+important in relation to the scalability of the framework.
 
@@
 
-\subsection{Evaluation}
+\subsection{Experimental Results}
 

### 2014-10-22 paper.tex
minor changes to the original evaluation text

@@
 respectively for the manual and the generated implementation. The base case is
-presented in absolute values, while the implementations with the added
-algorithms are presented as relative values.
+presented in absolute values, while the implementations with the ID algorithms
+added are presented as relative values.
 
@@
 the generator produces exactly the same code for the base case. We couldn't
-measure any real differences between both implementations.
+measure any real differences between both implementations. It's important to
+validate this aspect to ensure that both implementations are structurally
+comparable.
 

### 2014-10-23 paper.tex
changed formatting of subsection titles
marked conclusion and future work as TODO
minor changes while reviewing before starting ;-)

@@
 
-\subsection{Detecting Intrusions}
+\subsection*{Detecting Intrusions}
 \label{detection}
@@
 
-\subsection{Cooperative Decision Making}
+\subsection*{Cooperative Decision Making}
 \label{coorperative}
@@
 
-\subsection{Software Attestation}
+\subsection*{Software Attestation}
 \label{subsection:attestation}
@@
 
-Figure \ref{fig:classification} gives a visual overview of the classification
+Figure \ref{fig:classification} visually summarizes the dual classification
 presented above.
@@
 
-\begin{figure}[!ht]
+\begin{figure}[b]
 \captionof{lstlisting}{ID Algorithm Pattern\label{alg:id-algo-pattern}}
@@
 
-\subsection{Loops and Common Data}
+\subsection*{Loops and Common Data}
 
@@
 
-\subsection{Network Usage and Heterogeneity}
+\subsection*{Network Usage and Heterogeneity}
 
@@
 
-\subsection{The Meaning of Loops}
+\subsection*{The Meaning of Loops}
 
@@
 
-\subsection{A DSL for Nodes}
+\subsection*{A DSL for Nodes}
 
@@
 
-\subsection{Unified Messaging}
+\subsection*{Unified Messaging}
 
@@
 
-\subsection{Code Generator}
+\subsection*{Code Generator}
 
@@
 
-\subsection{Semantic Model}
+\subsection*{Semantic Model}
 
@@
 
-\subsection{Functional Code Fusion}
+\subsection*{Functional Code Fusion}
 
@@
 
-\subsection{Code Model}
+\subsection*{Code Model}
 
@@
 
-\subsection{Software Library}
+\subsection*{Software Library}
 
@@
 
-\subsection{Architecture}
+\subsection*{Architecture}
 
@@
 This hierarchy is important from a development effort point of view. The level
-of dependency on the platform also dictates the level of effort to implement
-changes to the platform. For example: changing the language will require a
-complete rewrite of the software library and a partial rewrite of the language
-description of the generator. Although this is a one-time effort, it is
-important in relation to the scalability of the framework.
+of dependency on the targeted platform also dictates the level of effort to
+implement changes to that platform. For example: to emit a different language
+will require a complete rewrite of the software library and a partial rewrite
+of the language description of the generator. Although this is a reusable,
+one-time effort, it is important in relation to the scalability of the
+framework.
 
-\subsection{Implementation}
+\subsection*{Implementation}
 
@@
 
-\subsection{Experimental Results}
+\subsection*{Experimental Results}
 
@@
 
+\TODO
+
 theoretical gains are met in practice

### 2014-10-23 paper.tex
added related work subsection on DSLs

@@
 \include{tikz}
+
+\usepackage[autostyle]{csquotes}
 
@@
 
-\TODO
+Section \ref{classification} introduced some ID
+algorithms\cite{ganeriwal2008reputation,mishra2004intrusion,krontiris2009cooperative}
+and presented a classification\cite{mishra2004intrusion,ioannis2007towards,alrajeh2013intrusion}
+based on their detection approach and interaction style.
 
-IDS in WSN \cite{perrig2004security,mishra2004intrusion}
+This section now focuses on related work in the field of domain specific
+languages and code generation, both in relation to wireless networks of
+resource constraint devices. These two components are the core of the \NAME
+framework and the solution it proposes. We also look at the more general scope,
+how these concepts are applied in other disciplines and evaluate their
+relevance in the scope of intrusion detection in the IoT and the proposed \NAME
+framework.
 
-DSL \cite{fowler2010domain,mernik2005and}
+\subsection{Domain Specific Languages}
 
-DSL for WSN \cite{naumowicz2009prototyping,levis2004tinyscript}
+Using a DSL, and especially an external DSL, as a foundation for any solution
+requires special care. It's important to remember that a DSL is merely a
+\emph{thin veneer over a (semantic) model}\cite{fowler2010domain}. Its sole
+purpose should be the easing of filling out the model. In the case of \NAME the
+reasons to implement a DSL are valid: the key goal is to remove any technical
+restrictions from a formal description of ID algorithms, allowing domain
+experts to produce these specifications\cite{mernik2005and}.
 
-DSL for IDS \cite{sekar1999high} \cite{uppuluri2001experiences} \cite{eckmann2002statl} \cite{vigna2003designing}
+Zooming in on wireless networks of resource constraint devices, we see that
+DSLs are indeed introduced \enquote{to shorten the development cycles
+dramatically during deployments, to reduce the dependency on hardware and
+software engineers, and to lead to a wider adoption outside the computer
+science field}\cite{sadilek2008domain,naumowicz2009prototyping}.
+
+TinyScript\cite{levis2004tinyscript} is an example of a language for wireless
+sensor networks. It's a complete programing language that aims to ease the
+description of algorithms in general. The Basic-like syntax does enable more
+productive development. But this approach doesn't address the underlying
+resource problems inherently connected to the context.
+
+SEAL\cite{elsts2013seal} is a novel complete programming language that focuses
+on novice programmers. It is a clear fit for \emph{sense-and-send} type of
+applications, and doesn't pretend to be the best choice for any type of
+application. It does however share the idea of event to overcome the technical
+nature of loops.
+
+Shifting the scope to intrusion detection, we learn that DSLs are predominantly
+used to describe patterns of events found in network
+packets\cite{sekar1999high,roesch1999snort} in a concise way.
+
+BMSL\cite{uppuluri2001experiences} is a DSL that allows the specification of
+patterns and actions. Patterns can contain events extracted from log files,
+system calls and network packets.
+
+The STAT framework with its STATL
+language\cite{eckmann2002statl,vigna2003designing} are not intrusion detection
+specific, but through extensions they allow to model attack scenarios using
+state machines. This is an interesting approach that shares the same ideas as
+our event-driven concept. The framework generates C++.
+
+\subsection{Code Generation}
+
+Code generation can be applied for a number reasons. In combination with a DSL
+it can optimize a development process, reducing the amount of code that needs
+to be constructed or it can be used to systematically create better code.
+
+Embedded systems are a prime example of an environment where code needs to be
+of a high quality.
 

### 2014-10-23 paper.tex
added subsection on code generation

@@
 Embedded systems are a prime example of an environment where code needs to be
-of a high quality.
+of a high quality. Focus on size and performance are a logical choice when
+dealing with limited storage and processing power. Different techniques are
+described in literature and implemented in compilers, starting at the lowest
+level\cite{leupers2000code, marwedel2002code}.
 
-Code Generation for embedded systems/WSNs \cite{leupers2000code,marwedel2002code}
+Optimizing for power efficiency however proves to be concurrent to size and
+performance optimization. Different strategies are needed. Data and memory
+optimizations \cite{panda2001data} can result in interesting improvements on a
+lower level. But to really preserve energy higher-level strategies are
+required, targeting architecture, radio handling, efficient
+software,\dots\cite{naik2001software}. Most of these are beyond the control of
+the compiler and show that a higher level of abstraction is required to address
+them.
 
-Code Optimization for embedded systems/WSNs \cite{panda2001data,naik2001software}
+DSLs are a solution here, but other paradigms are possible. A mixed framework
+consisting of a virtual machine (VM) and native code
+extension\cite{sadilek2007energy}, combined with energy-aware compilation,
+seems a promising approach. By offloading the optimization to a VM and only
+compiling parts and not the whole, the solution allows updating at a very
+granular scale, thus reducing the update cost, while maintaining flexibility
+and lowering the development effort. As stated in the introduction, \NAME
+focuses on the reduction of the development effort, while preserving
+flexibility.
 
-Code Generation for IDS \cite{charitakis2003code}
-
-% manet: IDS algorithms -> coorporation, trust
-
-More than in a classic network IDS, ID algorithms for MANETS operate with a
-large degree of uncertainty. They often have to collaborate with other devices
-in the network \cite{marchang2008collaborative,krontiris2009cooperative}, which
-increases the level of uncertainty, because no other device can fully be
-trusted. An IDS on an IoT device will be able to provide the services with
-information about the reputation \cite{ganeriwal2008reputation} of other nodes
-in the network.
+Finally, applying code generation to IDS, typically tries to generate native
+code to avoid the interpretation of the description of patterns.
+I2S\cite{charitakis2003code} is an example that generates code for
+SNORT\cite{roesch1999snort} rules. To optimise this highly efficient code, it
+often targets a specific platform.
 

### 2014-10-23 paper.tex
added software dev strategy classification diagram

@@
 
+\usepackage{tikz-3dplot}
+
 \include{tikz}
@@
 
-The remainder of this paper proceeds as follows: section \ref{classification}
-takes a look at ID algorithms and classifies them. Section \ref{pattern} looks
-for patterns in the implementation of these classes of algorithms, details some
-of the ideas behind FCF and identifies the different aspects of the problem
-space. Section \ref{design} describes the design we applied to construct the
-framework, its domain specific language (DSL) and code generator. Section
-\ref{evaluation} evaluates an implementation and application of the framework.
-Section \ref{related} explores related work in the field of DSLs for ID and
-code generation for the IoT. Finally, section \ref{conclusion} summarizes our
-findings, draws conclusions and identifies topics for future work.
+The remainder of this paper proceeds as follows: section \ref{approach}
+classifies our approach as a development strategy for wireless devices, section
+\ref{classification} takes a look at ID algorithms and classifies them. Section
+\ref{pattern} looks for patterns in the implementation of these classes of
+algorithms, details some of the ideas behind FCF and identifies the different
+aspects of the problem space. Section \ref{design} describes the design we
+applied to construct the framework, its domain specific language (DSL) and code
+generator. Section \ref{evaluation} evaluates an implementation and application
+of the framework. Section \ref{related} explores related work in the field of
+DSLs for ID and code generation for the IoT. Finally, section \ref{conclusion}
+summarizes our findings, draws conclusions and identifies topics for future
+work.
+
+\section{Classification of Code Generation as a Development Strategy}
+\label{approach}
+
+Figure \ref{fig:approach} shows a classification of several development
+strategies that can be adopted for projects involving wireless devices.
+
+\begin{figure}[ht]
+\tdplotsetmaincoords{110}{-10}
+\begin{tikzpicture}[scale=1,tdplot_main_coords]
+  % axis
+  \draw[very thick,->] (0,0,0) -- (7,0,0) node[anchor=north east,xshift=8mm]{$flexibility$};
+  \draw[very thick,->] (0,0,0) -- (0,7,0) node[anchor=north west]{$update$};
+  \draw[very thick,->] (0,0,0) -- (0,0,4) node[anchor=south]{$development$};
+  % points
+  \coordinate (O)        at (0,0,0);
+  \coordinate (ideal)    at (6,0,0);
+  \coordinate (vm)       at (2,2,2);
+  \coordinate (modular)  at (4,4,4);
+  \coordinate (reconfig) at (4,2,4);
+  \coordinate (full)     at (6,6,4);
+  \coordinate (cg)       at (6,4.5,3);
+  % "lines"
+  \draw[dashed] (2,0,0)    -- (2,2,0) -- (0,2,0);
+  \draw[dashed] (4,0,0)    -- (4,4,0) -- (0,4,0);
+  \draw[dashed] (6,0,0)    -- (6,6,0) -- (0,6,0);
+  \draw[dashed] (vm)       -- (2,2,0);
+  \draw[dashed] (modular)  -- (4,4,0);
+  \draw[dashed] (full)     -- (6,6,0);
+  \draw[dashed] (reconfig) -- (4,2,0);
+  \draw[dashed] (cg)       -- (6,4.5,0);
+  % "balls"
+  % 1. ideal
+  \tdplottransformmainscreen{6}{0}{0}
+  \shadedraw[tdplot_screen_coords, ball color=lightgray,draw=lightgray]
+    (\tdplotresx,\tdplotresy) circle (0.20)
+    node[above,xshift=+5pt,yshift=+5pt,text=gray]{\footnotesize ideal};
+  % 2. VM
+  \tdplottransformmainscreen{2}{2}{2}
+  \shadedraw[tdplot_screen_coords, ball color=lightgray,draw=lightgray]
+    (\tdplotresx,\tdplotresy) circle (0.20)
+    node[above,yshift=+7pt,text=gray]{\footnotesize virtual machine};
+  % 3. modular
+  \tdplottransformmainscreen{4}{4}{4}
+  \shadedraw[tdplot_screen_coords, ball color=lightgray,draw=lightgray]
+    (\tdplotresx,\tdplotresy) circle (0.20)
+    node[left,xshift=-5pt,text=gray]{\footnotesize modular native image};
+  % 4. reconfig
+  \tdplottransformmainscreen{4}{2}{4}
+  \shadedraw[tdplot_screen_coords, ball color=lightgray,draw=lightgray]
+    (\tdplotresx,\tdplotresy) circle (0.20)
+    node[above,yshift=+7pt,text=gray]{\footnotesize reconfigurable virtual machine};
+  % 5. full
+  \tdplottransformmainscreen{6}{6}{4}
+  \shadedraw[tdplot_screen_coords, ball color=lightgray,draw=lightgray]
+    (\tdplotresx,\tdplotresy) circle (0.20)
+    node[right,xshift=+5pt,text=gray]{\footnotesize full native image};
+  % 5. cg
+  \tdplottransformmainscreen{6}{4.5}{3}
+  \shadedraw[tdplot_screen_coords, ball color=gray,draw=black]
+    (\tdplotresx,\tdplotresy) circle (0.20)
+    node[right,xshift=+5pt,text=black,align=center]{\footnotesize code generation};
+\end{tikzpicture}
+\caption{Classification of Development Strategies for Wireless Devices (based on \cite{sadilek2007energy})}
+\label{fig:approach}
+\end{figure}
+
+\TODO
 

### 2014-10-23 paper.tex
completed draft of id-foo classification

--- 2014-10-23 paper.tex
@@
 
-The remainder of this paper proceeds as follows: section \ref{classification}
-takes a look at ID algorithms and classifies them. Section \ref{pattern} looks
-for patterns in the implementation of these classes of algorithms, details some
-of the ideas behind FCF and identifies the different aspects of the problem
-space. Section \ref{design} describes the design we applied to construct the
-framework, its domain specific language (DSL) and code generator. Section
-\ref{evaluation} evaluates an implementation and application of the framework.
-Section \ref{related} explores related work in the field of DSLs for ID and
-code generation for the IoT. Finally, section \ref{conclusion} summarizes our
-findings, draws conclusions and identifies topics for future work.
+The remainder of this paper proceeds as follows: section \ref{positioning}
+positions our approach as a development strategy for wireless devices, section
+\ref{classification} takes a look at ID algorithms and classifies them. Section
+\ref{pattern} looks for patterns in the implementation of these classes of
+algorithms, details some of the ideas behind FCF and identifies the different
+aspects of the problem space. Section \ref{design} describes the design we
+applied to construct the framework, its domain specific language (DSL) and code
+generator. Section \ref{evaluation} evaluates an implementation and application
+of the framework. Section \ref{related} explores related work in the field of
+DSLs for ID and code generation for the IoT. Finally, section \ref{conclusion}
+summarizes our findings, draws conclusions and identifies topics for future
+work.
+
+\section{\NAME as a Development Strategy}
+\label{positioning}
+
+Figure \ref{fig:approach} shows a classification of several development
+strategies that can be adopted for projects involving wireless devices.
+
+\newcommand{\ball}[5][above,yshift=+5pt]{
+\tdplottransformmainscreen{#2}{#3}{#4}
+\shadedraw[tdplot_screen_coords, ball color=lightgray,draw=lightgray,shading angle=-90]
+  (\tdplotresx,\tdplotresy) circle (0.20)
+  node[text=gray,font=\footnotesize,#1]{#5};
+}
+
+\begin{figure}[ht]
+  \centering
+  \scalebox{.85}{
+\tdplotsetmaincoords{110}{-10}
+\begin{tikzpicture}[scale=1,tdplot_main_coords]
+  % axis
+  \draw[very thick,->] (0,0,0) -- (7,0,0) node[anchor=north east,xshift=10mm]{$flexibility$};
+  \draw[very thick,->] (0,0,0) -- (0,7,0) node[anchor=north west]{$update cost$};
+  \draw[very thick,->] (0,0,0) -- (0,0,4) node[anchor=south]{$development$};
+  % points
+  \coordinate (O)        at (0,0,0);
+  \coordinate (ideal)    at (6,0,0);
+  \coordinate (vm)       at (2,2,2);
+  \coordinate (modular)  at (4,4,4);
+  \coordinate (reconfig) at (4,2,4);
+  \coordinate (full)     at (6,6,4);
+  \coordinate (cg)       at (6,4,2.5);
+  % "lines"
+  \draw[dashed] (2,0,0)    -- (2,2,0) -- (0,2,0);
+  \draw[dashed] (4,0,0)    -- (4,4,0) -- (0,4,0);
+  \draw[dashed] (6,0,0)    -- (6,6,0) -- (0,6,0);
+  \draw[dashed] (vm)       -- (2,2,0);
+  \draw[dashed] (modular)  -- (4,4,0);
+  \draw[dashed] (full)     -- (6,6,0);
+  \draw[dashed] (reconfig) -- (4,2,0);
+  \draw[dashed] (cg)       -- (6,4,0);
+  % "balls"
+  \ball{6}{0}{0}{ideal}
+  \ball{2}{2}{2}{virtual machine}
+  \ball[left,xshift=-5pt]{4}{4}{4}{\shortstack{modular\\native image}}
+  \ball{4}{2}{4}{\shortstack{reconfigurable\\virtual machine}}
+  \ball[right,xshift=+5pt,yshift=+5pt]{6}{6}{4}{full native image}
+  %\ball[right,xshift=+5pt,text=black]{6}{4.5}{3}{code generation};
+  % 5. cg
+  \tdplottransformmainscreen{6}{4}{2.5}
+  \shadedraw[tdplot_screen_coords, ball color=gray,draw=black]
+    (\tdplotresx,\tdplotresy) circle (0.20)
+    node[right,xshift=+5pt,text=black,align=center,font=\footnotesize]{code generation\\+ functional code fusion};
+\end{tikzpicture}
+}
+\caption{Classification of Development Strategies for Wireless Devices (based on \cite{sadilek2007energy})}
+\label{fig:approach}
+\end{figure}
+
+Three properties are used to classify the strategies\cite{sadilek2007energy}:
+flexibility, update cost and development effort. The flexibility is a measure
+for the number of applications that can be deployed on a device without
+physical access. The update cost is the energy required to deploy a new version
+of an application on a node. The development effort covers the time to write,
+test and deploy an application. The \emph{ideal} situation is that were one
+ends up with a maximal flexibility and no update nor development cost.
+
+If we consider the trade off between flexibility and update cost, we can
+distinguish two extremes: virtual machines and full native images. Virtual
+machines allow for very small updates due to their concise byte code, but at
+the same time are restricted by the limitations of their instruction set and
+predefined capabilities. A full native image on the other hand has unlimited
+flexibility, but requires a large transfer of the image over the network.
+
+Two intermediate alternatives are possible: modular native image and
+reconfigurable virtual machines. With modular native images, only modules are
+changed, limiting the update size, but the framework to support it is static
+and limits the possibilities. Reconfigurable virtual machines take this same
+approach, but apply it to virtual machines, allowing to \emph{upgrade} the
+instruction set of the virtual machine.
+
+Choosing a development strategy, with this trade off take into account, will
+also determine the development cost. Developing for a virtual machine will be
+less demanding than targeting a native image.
+
+With \NAME we opt to start from the full native image, allowing full
+flexibility, but try to reduce both the update cost and the development cost.
+Firstly, by applying code generation and functional code fusion, we optimize
+the code organisation, which has a positive effect on the code size and hence
+also on the update cost. Secondly, by enabling reuse on a functional level and
+automating the writing phase, the development cost is also lowered.
 

### 2014-10-23 paper.tex
fixed a few details

@@
   \tdplottransformmainscreen{6}{4}{2.5}
-  \shadedraw[tdplot_screen_coords, ball color=gray,draw=black]
+  \shadedraw[tdplot_screen_coords, ball color=gray,draw=black,shading angle=-90]
     (\tdplotresx,\tdplotresy) circle (0.20)
@@
 and lowering the development effort. As stated in the introduction, \NAME
-focuses on the reduction of the development effort, while preserving
-flexibility.
+focuses on the reduction of the development effort and update cost, while
+preserving flexibility.
 

### 2014-10-28 paper.tex
processed review remarks

@@
 \include{tikz}
+\include{foo-lang}
 
@@
 
-Supporting multiple intrusion detection (ID) algorithms imposes significant
-overhead on Internet of Things (IoT) devices. Fusion of the underlying
-algorithms can alleviate this, yet, proves to be time-consuming, repetitive and
+When intrusion prevention fails, intrusion detection serves as a second layer
+of defense. It looks for patterns, anomalies,\dots that can indicate malicious
+behavior. But supporting these intrusion detection algorithms imposes
+significant overhead on resource-restricted devices such as those targeting the
+Internet of Things. Fusion of the algorithms, to optimize the use of the shared
+resources, can alleviate this, yet, proves to be time-consuming, repetitive and
 error-prone. To address this problem we introduce \NAME, a framework for the
-development of efficient intrusion detection systems (IDS). It consists of a
-domain specific language (DSL) that allows formally describing ID algorithms on
-a functional level. A code generator then fuses these algorithms and produces
-more optimally organized source code. This way, \NAME addresses the
-heterogeneity of both the IoT and ID, as well as the limited resources of IoT
-devices. A side-by-side comparison shows that generated code reduces memory
+development of efficient intrusion detection systems. It consists of a domain
+specific language, to formally describing intrusion detection algorithms on a
+functional level, and a code generator to produce more optimally organized
+source code. A side-by-side comparison shows that generated code reduces memory
 footprint, message passing overhead and execution time in comparison to a
@@
 
-The IoT presents a great potential to positively influence our daily work and
-life: domotics, assisted living, e-health and enhanced learning are only a few
-examples\cite{atzori2010internet}. But with this potential the IoT also
-presents a significant threat: by opening our smart homes
-\cite{aldrich2003smart} and personal data\cite{weber2010internet} to all these
-interconnected devices, we also open them to everyone who is able to break the
-virtual locks that protect them. IoT devices typically have limited batteries,
-little processing power and memory is restricted to the bare minimum. These
-properties make it hard to add any additional service, including security.
+The internet of things (IoT) presents a great potential to positively influence
+our daily work and life: domotics, assisted living, e-health and enhanced
+learning are only a few examples\cite{atzori2010internet}. But with this
+potential the IoT also presents a significant threat: by opening our smart
+homes and personal data to all these interconnected devices, we also open them
+to everyone who is able to break the virtual locks that protect them. IoT
+devices typically have limited batteries, little processing power and memory is
+restricted to the bare minimum. These properties make it hard to add any
+additional service, including security.
 
@@
 significant part of the IoT, it is not possible to have this single point in
-the network that can oversee all traffic\cite{zhang2000intrusion,
-mishra2004intrusion}. If the outer security perimeter diminishes, every device
-has to implement its own lines of defense. But making the IDS a local service
-on each device requires local resources, which are the scarce commodity to IoT
-devices.
+the network that can oversee all traffic\cite{mishra2004intrusion}. If the
+outer security perimeter diminishes, every device has to implement its own
+lines of defense. But making the IDS a local service on each device requires
+local resources, which are the scarce commodity to IoT devices.
 
@@
 implementation of algorithms\cite{valero2012di} and offer the required basic
-functional components to implement them\cite{krontiris2008lidea}. But they
-don't offer a way to optimize the usage of resources. Nor do they support
-systematic reuse to create different configurations for different devices.
-
-% problem -> requirements for solution
-
-The IoT is a heterogeneous environment, both in hardware and software.
+functional components to implement them\cite{krontiris2008lidea}. The first
+aspect that is missing, is that they don't offer a way to optimize the usage of
+resources. A solution that supports the integration of multiple algorithms on
+IoT devices needs to consider the impact on their resources and avoid
+accumulating their impact.
+
+Secondly, they don't support systematic reuse to create different
+configurations for different devices, both dealing with hard- and software.
 Addressing this in a transparent and automated way is a key success factor for
-any kind of development in this area. A solution that supports the integration
-of multiple algorithms on IoT devices needs to consider the impact on their
-resources. Different algorithms should be merged somehow to avoid accumulating
-their impact.
-
-% my approach merged into ...
+any kind of development in this area.
+
 % scientific contribution
@@
 identifies common data and functions, and organizes code in such a way that
-redundant iterations, tests and computations are eliminated. Our results show
+redundant iterations, tests and computations are eliminated. Results show
 substantially improved memory usage, execution time and usage of the wireless
@@
 
-The remainder of this paper proceeds as follows: section \ref{positioning}
-positions our approach as a development strategy for wireless devices, section
-\ref{classification} takes a look at ID algorithms and classifies them. Section
-\ref{pattern} looks for patterns in the implementation of these classes of
-algorithms, details some of the ideas behind FCF and identifies the different
-aspects of the problem space. Section \ref{design} describes the design we
-applied to construct the framework, its domain specific language (DSL) and code
-generator. Section \ref{evaluation} evaluates an implementation and application
-of the framework. Section \ref{related} explores related work in the field of
-DSLs for ID and code generation for the IoT. Finally, section \ref{conclusion}
-summarizes our findings, draws conclusions and identifies topics for future
-work.
-
-\section{\NAME as a Development Strategy}
+The remainder of this paper proceeds as follows: section \ref{classification}
+takes a look at ID algorithms and classifies them. Section \ref{pattern} looks
+for patterns in the implementation of these classes of algorithms, details some
+of the ideas behind FCF and identifies the different aspects of the problem
+space. Section \ref{design} describes the design we applied to construct the
+framework, its domain specific language (DSL) and code generator. Section
+\ref{evaluation} evaluates an implementation and application of the framework.
+Section \ref{related} explores related work in the field of DSLs for ID and
+code generation for the IoT. Finally, section \ref{conclusion} summarizes our
+findings, draws conclusions and identifies topics for future work.
+
+\section{A Classification of ID Algorithms}
+\label{classification}
+
+Research into ID in wireless networks typically focuses on two major topics
+that complement the architecture: the devices as single entities and the
+network as a group of such devices. Both are important because the group cannot
+make a decision without members that detect malicious behavior and a
+group-based decision is often needed because nodes can easily miss out on
+certain events that could indicate intrusions, due to their wireless and
+not-always-on nature.
+
+These two topics lead to the first two classes: detecting intrusions and
+cooperative decision making. We also identify software attestation as a third
+class. Although it can be considered as a special case of a combination of both
+other classes, its focus is not on the detection of intrusions, but on the
+effect of an intrusion on the system itself.
+
+\subsection*{Detecting Intrusions}
+\label{detection}
+
+There are different ways to construct a taxonomy for intrusion detection, but
+common themes do appear. According to \cite{mishra2004intrusion} and
+\cite{ioannis2007towards} there are three major categories: anomaly detection,
+signature or misuse detection, and specification-based detection. In
+\cite{alrajeh2013intrusion} the authors mostly agree with this topology, but
+add the notion of hybrid intrusion detection systems and cross layer intrusion
+detection systems as recent advances in research. Although these additional
+categories offer very interesting prospects, the authors have to admit that the
+impact on the resource-constrained devices might still be too much. We
+therefore focus on the three categories agreed upon by most sources.
+
+\begin{LaTeXdescription}
+
+  \item[Anomaly detection] considers a baseline profile and evaluates all
+  activity according to this profile. Aberrant behavior is flagged as a
+  possible intrusion, but this is no exact science. The baseline profile can
+  evolve over time, especially in wireless and mobile environments. The flagged
+  incidents should therefore also be handled with care and no absolute
+  decisions can be made from single events.
+
+  \item[Signature or misuse detection] starts with known patterns of attacks
+  and tries to find evidence in the actions of and communication with a device.
+  Because many attacks use valid ways to communicate with services on the
+  devices, this can also produce false positives and offers no black and white
+  certainty.
+
+  \item[Specification-based detection] uses constraints that describe the
+  correct behavior of the system. Although that this approach can detect
+  previously unknown attacks, it can sometimes not detect specific attacks that
+  can be detected by anomaly or misuse detection.
+
+\end{LaTeXdescription}
+
+\subsection*{Cooperative Decision Making}
+\label{coorperative}
+
+These taxonomies focus on the act of detecting intrusions themselves. In a
+highly distributed environment, an orthogonal categorization, dealing with the
+distributed nature of the network and its ways of communicating, may be even
+more important.
+
+As in many other situations, a group is often stronger than the sum of its
+individual components. Because not all devices are always actively
+participating in the network, some might miss clues that would otherwise lead
+them to detect intrusions. It is hardly impossible for a single device to
+detect a complex attack by itself. Therefore a second important research topic
+consists of cooperative algorithms to combine information from single
+identities into a group-based decision about alleged intrusions. We identify
+two ways by which this cooperative decision making can be constructed and focus
+on the way participants exchange information: broadcasting and interactive.
+
+\begin{LaTeXdescription}
+
+  \item[Broadcasting] can sometimes allow a distributed, cooperative algorithm
+  to be implemented locally. This is illustrated by the reputation-based
+  detection algorithm introduced in \cite{ganeriwal2008reputation}, based on
+  the same principle of a watchdog also found in \cite{mishra2004intrusion}.
+  Using this algorithm, nodes can exchange information about the reputation of
+  other nodes, solely with broadcasted messages. Combining this information
+  allows them to decide about their trust in a given node, based on more than
+  their own, often partial, observations alone.
+
+  \item[Interactive] communication allows algorithms to exchange information to
+  reach a consensus. In \cite{krontiris2009cooperative}, the authors first
+  present a theoretical foundation to analyze cooperative algorithms. Given
+  these foundations they continue to present an algorithm consisting of 5
+  phases. It essentially implements a voting system-based on the Guy Fawkes
+  protocol\cite{anderson1998new} to identify an intruder. It enables the
+  exchange of suspected intruder information and allows distributed
+  authentication of those \emph{votes}. Based on these authenticated votes, a
+  distributed decision can be made about the jointly identified intruder.
+
+\end{LaTeXdescription}
+
+\subsection*{Software Attestation}
+\label{subsection:attestation}
+
+A research topic that exists in parallel to the detection and cooperation
+duality is software attestation. It offers algorithms for exchanging
+information about the software that is running on a device, and aims to
+identify devices that can no longer prove that their content is unaltered. A
+good example is SAKE\cite{seshadri2008sake}.
+
+Software attestation is hard and many details can cause an algorithm to fail.
+The general feeling is that even in the best conditions, there is always a way
+to circumvent the most ingenious algorithm to perform software-based
+attestation.
+
+Within the context of this paper, we will not consider software attestation as
+a special class and exclude it from our scope, only focusing on detecting
+intrusions and distributed decision making. Figure \ref{fig:classification}
+visually summarizes this dual classification.
+
+\begin{figure}[ht]
+  \centering
+  \scalebox{.85}{
+\begin{tikzpicture}[
+  sibling distance=20mm,node distance=40mm
+  ]
+  \node (detection)             [align=center] {Detecting\\Intrusions}
+    child {node (anomaly)       [align=center] {Anomaly}}
+    child {node (signature)     [align=center] {Signature\\Misuse}}
+    child {node (specification) [align=center] {Specification}};
+  \node (decision)              [align=center,right=of detection] {Decision\\Making}
+    child {node (broadcasting)  [align=center] {Broadcasting}}
+    child {node (interactive)   [align=center] {Interactive}};
+\end{tikzpicture}
+}
+  \caption{Classification of ID Algorithms}
+  \label{fig:classification}
+\end{figure}
+
+\section{A Design Pattern for ID Algorithms}
+\label{pattern}
+
+Given this classification, we now introduce a design pattern that identifies
+common aspects of these algorithms. It will allow us to define a strategy for
+optimizing the combination of these algorithms. From the classification and
+survey of several algorithms, we extract two activities common to most
+algorithms: the inspection of network traffic and the evaluation of thresholds.
+
+The former is trivial: network traffic is the carrier along which malicious
+payloads reach the device or where patterns of communication can be observed
+when other devices in the vicinity are affected by such payloads.
+
+The latter is more complex: an algorithm based on signature or misuse detection
+can maybe undeniably decide if an attack is real or not. But when dealing with
+anomaly detection this line surely isn't that crisp. Especially in distributed
+situations where even a neighboring device can not be trusted, detecting
+intrusion is no absolute science. In this case one has to rely on statistics
+and thresholds to decide when to drop trust in a given device.
+
+But simply checking a threshold after updating some counter, is not a good
+strategy. A simple example is that of a watchdog\cite{mishra2004intrusion},
+observing the actions of other devices. There are valid reasons why a devices
+might miss the theoretical window in which certain actions are expected. Simply
+reacting on such basic threshold might result in false positives. A way to
+overcome this is to decouple the detection of problems and the decision what to
+do about them.
+
+This decoupling of detection and decision is the core of the design pattern for
+ID algorithms. Listing \ref{alg:id-algo-pattern} presents the design pattern in
+pseudo-code.
+
+\begin{figure}[ht]
+\captionof{lstlisting}{ID Algorithm Pattern\label{alg:id-algo-pattern}}
+\begin{algorithmic}[1]
+  \Require{nodes, global storage for information about nodes}
+  \Function{process\_message}{$msg$}
+    \ForEach{$byte \in msg$} \label{alg:id-algo-pattern-loop1}
+     \State \dots \Comment{analyze byte-sequences}
+    \EndFor
+    \Let{$nodes_x$}{value}  \Comment{optionally update node info}
+    \State \Call{send}{$nodes_y$, ``info''} \Comment{optionally exchange info} \label{alg:id-algo-pattern-send1}
+  \EndFunction
+  \State
+  \Function{do\_housekeeping}{}
+    \ForEach{$node \in nodes$} \label{alg:id-algo-pattern-loop2} \label{alg:id-algo-pattern-common-data}
+      \If{$node > \dots$} \Comment{validate recorded value}
+        \State \dots \Comment{take actions}
+        \State \Call{send}{$node$, ``info''} \Comment{e.g. communicate} \label{alg:id-algo-pattern-send2}
+      \EndIf
+    \EndFor
+  \EndFunction
+  \State
+\end{algorithmic}
+\end{figure}
+
+We now extend the scope of the design pattern, to include its application in
+the context of multiple algorithms. When implementing different algorithms, one
+will typically encounter a pattern such as describes in listing
+\ref{alg:id-algo-application}.
+
+\begin{figure}[ht]
+\captionof{lstlisting}{Application pattern of multiple ID algorithms\label{alg:id-algo-application}}
+\begin{algorithmic}[1]
+  \Let{$msg$}{$network$} \Comment{all observed messages}
+  \ForEach{$algorithm \in algorithms$}
+    \State \Call{algorithm.progress\_message}{msg}
+  \EndFor
+  \State \dots
+  \Comment{at a given interval}
+  \ForEach{$algorithm \in algorithms$}
+    \State \Call{algorithm.do\_housekeeping}{}
+  \EndFor
+\end{algorithmic}
+\end{figure}
+
+The different implementations would be linked and called in a sequential way,
+when new messages are observed on the network and at a regular interval to
+allow them to do internal housekeeping, such as threshold evaluation.
+
+\subsection*{Inherent Problems}
+\label{pattern-problems}
+
+Given these patterns, we can now identify the problem areas that tend to result
+in suboptimal code when combining multiple ID algorithms. Two issues arise from
+the nested loops that originate from the combination of the ID algorithm
+pattern and the application pattern: loops and common data. A third issue is
+the use of the network.
+
+\begin{LaTeXdescription}
+
+  \item[Loops] by themselves are not a problem, but when different loops
+  perform the same operations in sequence, an overhead might be incurred. An
+  example from the ID algorithm pattern is the handling of messages: each
+  algorithm parses each observed message on the network, processing the same
+  byte stream, performing comparable operations to find similar patterns.
+
+  \item[Common data] is cross-cutting the different algorithms, that have to
+  store information about nodes within its own scope. This results in a lot of
+  duplicated information in memory that also needs to be maintained.
+
+  \item[The network] allows algorithms to operate on a distributed scale. But
+  calls to the \ttt{SEND} function are scattered around the algorithms, causing
+  the wireless radio to be accessed multiple times, keeping it active and
+  energy consuming more than strictly needed.
+
+\end{LaTeXdescription}
+
+Reimplementing the algorithms manually to eliminate these issues is not a valid
+option. This development cost typically isn't a one-time cost, e.g. in case a
+new version of the algorithm is released, or simply when a different set of
+algorithms is selected due to a changing risk analysis. Also, changing, or
+simply implementing the algorithms from scratch, holds the risk of making
+mistakes.
+
+The heterogeneity of IoT devices results in many different software stacks,
+each with e.g. their own network API. This also explains why there are very few
+implementations readily available to developers: there is no common platform to
+target for researchers. Even using a unified software library would not only
+require a consistent implementation of it, but would also limit the operational
+scope of the research implementation, still restricting it to a single language.
+
+\section{Design}
+\label{design}
+
+To bridge the gap between the research world and the heterogeneous IoT where
+developers want to implement an actual IDS, while still respecting the limited
+resources they have at their disposal, we introduce \NAME, a framework
+consisting of a DSL, code generator and software library.
+
+The DSL enables researchers to formally describe their ID algorithms, without
+focusing on any specific language or platform. Given such formal descriptions
+of ID algorithms, a code generator is now able to combine these, while avoiding
+the problems arising from the patterns/problems presented in listings
+\ref{alg:id-algo-pattern} and \ref{alg:id-algo-application}. It can also target
+different languages and platforms, while reusing a software library that
+provides e.g. message parsing, network access wrapping,\dots
+
+Choosing for a DSL and code generator is choosing a development strategy. This
+implies several choices with respect to flexibility, update cost and
+development effort. We will first position \NAME and motivate its design with
+respect to these trade offs (\ref{positioning}). Second, we will look at the
+principal design components of the DSL and how it deals with loops and common
+data (\ref{dsl-design}). Third, the structure of the code generator will be
+discussed (\ref{code-generator-design}). Finally, fourth, the supporting
+software library is introduced (\ref{software-lib-design}).
+
+\subsection{\NAME as a Development Strategy}
 \label{positioning}
 
-Figure \ref{fig:approach} shows a classification of several development
+Figure \ref{fig:positioning} shows a classification of several development
 strategies that can be adopted for projects involving wireless devices.
@@
   \draw[very thick,->] (0,0,0) -- (7,0,0) node[anchor=north east,xshift=10mm]{$flexibility$};
-  \draw[very thick,->] (0,0,0) -- (0,7,0) node[anchor=north west]{$update cost$};
+  \draw[very thick,->] (0,0,0) -- (0,7,0) node[anchor=north west]{$update\ cost$};
   \draw[very thick,->] (0,0,0) -- (0,0,4) node[anchor=south]{$development$};
@@
     node[right,xshift=+5pt,text=black,align=center,font=\footnotesize]{code generation\\+ functional code fusion};
+  % "box around our solution" ;-)
+  \tdplottransformmainscreen{0}{0}{0}
+  \draw[tdplot_screen_coords] (4.95,1) rectangle (8.5,0.25);
+
 \end{tikzpicture}
@@
 \caption{Classification of Development Strategies for Wireless Devices (based on \cite{sadilek2007energy})}
-\label{fig:approach}
+\label{fig:positioning}
 \end{figure}
@@
 Three properties are used to classify the strategies\cite{sadilek2007energy}:
-flexibility, update cost and development effort. The flexibility is a measure
-for the number of applications that can be deployed on a device without
-physical access. The update cost is the energy required to deploy a new version
-of an application on a node. The development effort covers the time to write,
-test and deploy an application. The \emph{ideal} situation is that were one
-ends up with a maximal flexibility and no update nor development cost.
+flexibility, update cost and development effort.
+
+\begin{LaTeXdescription}
+
+  \item[Flexibility] is a measure for the number of applications that can be
+  deployed on a device without physical access.
+
+  \item[Update cost] is equivalent to the energy required to deploy a new
+  version of an application on a device.
+
+  \item[Development effort] covers the time to write, test and deploy an
+  application.
+
+\end{LaTeXdescription}
+
+The \emph{ideal} situation is that were one ends up with a maximal flexibility
+and no update nor development cost.
 
 If we consider the trade off between flexibility and update cost, we can
-distinguish two extremes: virtual machines and full native images. Virtual
-machines allow for very small updates due to their concise byte code, but at
-the same time are restricted by the limitations of their instruction set and
-predefined capabilities. A full native image on the other hand has unlimited
-flexibility, but requires a large transfer of the image over the network.
-
-Two intermediate alternatives are possible: modular native image and
-reconfigurable virtual machines. With modular native images, only modules are
-changed, limiting the update size, but the framework to support it is static
-and limits the possibilities. Reconfigurable virtual machines take this same
-approach, but apply it to virtual machines, allowing to \emph{upgrade} the
-instruction set of the virtual machine.
-
-Choosing a development strategy, with this trade off take into account, will
-also determine the development cost. Developing for a virtual machine will be
-less demanding than targeting a native image.
+distinguish two extremes: \emph{virtual machines} and \emph{full native
+images}. Virtual machines allow for very small updates due to their concise
+byte code, but at the same time are restricted by the limitations of their
+instruction set and predefined capabilities. A full native image on the other
+hand has unlimited flexibility, but requires transfer of a larger image over
+the network.
+
+Two intermediate alternatives are possible: \emph{modular native image} and
+\emph{reconfigurable virtual machines}. With modular native images, only
+modules are changed, limiting the update size, but the framework to support it
+is static and limits the possibilities. Reconfigurable virtual machines take
+this same approach, but apply it to virtual machines, allowing to
+\emph{upgrade} the instruction set of the virtual machine.
+
+Choosing a development strategy also determines the development cost: e.g.
+developing for a virtual machine will be less demanding than targeting a native
+image.
 
@@
 Firstly, by applying code generation and functional code fusion, we optimize
-the code organisation, which has a positive effect on the code size and hence
+the code organization, which has a positive effect on the code size and hence
 also on the update cost. Secondly, by enabling reuse on a functional level and
@@
 
-\section{A Classification of ID Algorithms}
-\label{classification}
-
-Research into ID in wireless networks typically focuses on two major topics
-that complement the architecture: the devices as single entities and the
-network as a group of such devices. Both are important because the group cannot
-make a decision without members that detect malicious behavior and a
-group-based decision is often needed because nodes can easily miss out on
-certain events that could indicate intrusions, due to their wireless and
-not-always-on nature.
-
-These two topics lead to the first two classes: detecting intrusions and
-cooperative decision making. We also identify software attestation as a third
-class. Although it can be considered as a special case of a combination of both
-other classes, its focus is not on the detection of intrusions, but on the
-effect of an intrusion on the system itself.
-
-\subsection*{Detecting Intrusions}
-\label{detection}
-
-There are different ways to construct a taxonomy for intrusion detection, but
-common themes do appear. According to \cite{mishra2004intrusion} and
-\cite{ioannis2007towards} there are three major categories: anomaly detection,
-signature or misuse detection, and specification-based detection. In
-\cite{alrajeh2013intrusion} the authors mostly agree with this topology, but
-add the notion of hybrid intrusion detection systems and cross layer intrusion
-detection systems as recent advances in research. Although these additional
-categories offer very interesting prospects, the authors have to admit that the
-impact on the resource-constrained devices might still be too much. We
-therefore focus on the three categories agreed upon by most sources.
-
-\begin{LaTeXdescription}
-
-  \item[Anomaly detection] considers a baseline profile and evaluates all
-  activity according to this profile. Aberrant behavior is flagged as a
-  possible intrusion, but this is no exact science. The baseline profile can
-  evolve over time, especially in wireless and mobile environments. The flagged
-  incidents should therefore also be handled with care and no absolute
-  decisions can be made from single events.
-
-  \item[Signature or misuse detection] starts with known patterns of attacks
-  and tries to find evidence in the actions of and communication with a device.
-  Because many attacks use valid ways to communicate with services on the
-  devices, this can also produce false positives and offers no black and white
-  certainty.
-
-  \item[Specification-based detection] uses constraints that describe the
-  correct behavior of the system. Although that this approach can detect
-  previously unknown attacks, it can sometimes not detect specific attacks that
-  can be detected by anomaly or misuse detection.
-
-\end{LaTeXdescription}
-
-\subsection*{Cooperative Decision Making}
-\label{coorperative}
-
-These taxonomies focus on the act of detecting intrusions themselves. In a
-highly distributed environment, an orthogonal categorization, dealing with the
-distributed nature of the network and its ways of communicating, may be even
-more important.
-
-As in many other situations, a group is often stronger than the sum of its
-individual components. Because not all devices are always actively
-participating in the network, some might miss clues that would otherwise lead
-them to detect intrusions. It is hardly impossible for a single device to
-detect a complex attack by itself. Therefore a second important research topic
-consists of cooperative algorithms to combine information from single
-identities into a group-based decision about alleged
-intrusions\cite{zhang2000intrusion}. We identify two ways by which this
-cooperative decision making can be constructed and focus on the way
-participants exchange information: broadcasting and interactive.
-
-\begin{LaTeXdescription}
-
-  \item[Broadcasting] can sometimes allow a distributed, cooperative algorithm
-  to be implemented locally. This is illustrated by the reputation-based
-  detection algorithm introduced in \cite{ganeriwal2008reputation}, based on
-  the same principle of a watchdog also found in \cite{mishra2004intrusion}
-  Using this algorithm, nodes can exchange information about the reputation of
-  other nodes, using only broadcasted messages. Combining this information
-  allows them to decide about their trust in a given node, based on more than
-  their own, often partial, observations alone.
-
-  \item[Interactive] communication allows algorithms to exchange information to
-  reach a consensus. In \cite{krontiris2009cooperative}, the authors first
-  present a theoretical foundation to analyze cooperative algorithms. Given
-  these foundations they continue to present an algorithm consisting of 5
-  phases. It essentially implements a voting system-based on the Guy Fawkes
-  protocol\cite{anderson1998new} to identify an intruder. It enables the
-  exchange of suspected intruder information and allows distributed
-  authentication of those \emph{votes}. Based on these authenticated votes, a
-  distributed decision can be made about the commonly identified intruder.
-
-\end{LaTeXdescription}
-
-\subsection*{Software Attestation}
-\label{subsection:attestation}
-
-A research topic that exists in parallel to the detection and cooperation
-duality is software attestation. It offers algorithms for exchanging
-information about the software that is running on a device, and aims to
-identify devices that can no longer prove that their content is unaltered. A
-family of algorithms to implement this functionality are
-SWATT\cite{seshadri2004swatt}, ICE/SCUBA\cite{seshadri2006scuba} and
-SAKE\cite{seshadri2008sake}.
-
-Software attestation is hard and many details can cause an algorithm to fail. A
-discussion surrounding this topic, started by \cite{castelluccia2009difficulty}
-in response to the previously mentioned papers, illustrates this. In
-\cite{perrig2010refutation} the authors of the original papers counter many of
-the objections made to their work, but the general feeling is that even in the
-best conditions, there is always a way to circumvent the most ingenious
-algorithm to perform software-based attestation.
-
-Within the context of this paper, we will not consider software attestation as
-a special class and exclude it from our scope, only focusing on detecting
-intrusions and distributed decision making.
-
-Figure \ref{fig:classification} visually summarizes the dual classification
-presented above.
-
-\begin{figure}[!ht]
-  \centering
-  \scalebox{.85}{
-\begin{tikzpicture}[
-  sibling distance=20mm,node distance=40mm
-  ]
-  \node (detection)             [align=center] {Detecting\\Intrusions}
-    child {node (anomaly)       [align=center] {Anomaly}}
-    child {node (signature)     [align=center] {Signature\\Misuse}}
-    child {node (specification) [align=center] {Specification}};
-  \node (decision)              [align=center,right=of detection] {Decision\\Making}
-    child {node (broadcasting)  [align=center] {Broadcasting}}
-    child {node (interactive)   [align=center] {Interactive}};
-\end{tikzpicture}
-}
-  \caption{Classification of ID Algorithms}
-  \label{fig:classification}
-\end{figure}
-
-\section{A Design Pattern for ID Algorithms}
-\label{pattern}
-
-Given this classification, we now introduce a design pattern that identifies
-common aspects of these algorithms. It will allow us to define a strategy for
-optimizing the combination of these algorithms. From the classification and
-survey of several algorithms, we extract two activities common to most
-algorithms: the inspection of network traffic and the evaluation of thresholds.
-
-The former is trivial: network traffic is the carrier along which malicious
-payloads reach the device or where patterns of communication can be observed
-when other devices in the vicinity are affected by such payloads.
-
-The latter is more complex: an algorithm based on signature or misuse detection
-maybe sometimes can undeniably decide if an attack is real or not. But when
-dealing with anomaly detection this line surely isn't that crisp. Especially in
-distributed situations where even a neighboring device can not be trusted,
-detecting intrusion is no absolute science. In this case one has to rely on
-statistics and thresholds to decide when to drop trust in a given device.
-
-Simply checking a threshold after updating some counter, is also not always a
-good strategy. A simple example is that of a
-watchdog\cite{mishra2004intrusion}, observing the actions of other devices.
-There are valid reasons why a devices might miss the theoretical window in
-which certain actions are expected. Simply reacting on such basic threshold
-might result in false positives. A way to overcome this is to decouple the
-detection of problems and the decision what to do about them.
-
-This decoupling of detection and decision is the core of the design pattern for
-ID algorithms. Listing \ref{alg:id-algo-pattern} presents the design pattern in
-pseudo-code.
-
-\begin{figure}[b]
-\captionof{lstlisting}{ID Algorithm Pattern\label{alg:id-algo-pattern}}
-\begin{algorithmic}[1]
-  \Require{nodes, global storage for information about nodes}
-  \Function{process\_message}{$msg$}
-    \ForEach{$byte \in msg$} \label{alg:id-algo-pattern-loop1}
-     \State \dots \Comment{analyze byte-sequences}
-    \EndFor
-    \Let{$nodes_x$}{value}  \Comment{optionally update node info}
-    \State \Call{send}{$nodes_y$, ``info''} \Comment{optionally exchange info} \label{alg:id-algo-pattern-send1}
-  \EndFunction
-  \State
-  \Function{do\_housekeeping}{}
-    \ForEach{$node \in nodes$} \label{alg:id-algo-pattern-loop2} \label{alg:id-algo-pattern-common-data}
-      \If{$node > \dots$} \Comment{validate recorded value}
-        \State \dots \Comment{take actions}
-        \State \Call{send}{$node$, ``info''} \Comment{e.g. communicate} \label{alg:id-algo-pattern-send2}
-      \EndIf
-    \EndFor
-  \EndFunction
-  \State
-\end{algorithmic}
-\end{figure}
-
-We now extend the scope of the design pattern, to include its application in
-the context of multiple algorithms. When implementing different algorithms, one
-will typically encounter a pattern such as describes in listing
-\ref{alg:id-algo-application}.
-
-\begin{figure}[!ht]
-\captionof{lstlisting}{Sequential implementation of ID algorithms\label{alg:id-algo-application}}
-\begin{algorithmic}[1]
-  \Let{$msg$}{$network$} \Comment{all observed messages}
-  \ForEach{$algorithm \in algorithms$}
-    \State \Call{algorithm.progress\_message}{msg}
-  \EndFor
-  \State \dots
-  \Comment{at a given interval}
-  \ForEach{$algorithm \in algorithms$}
-    \State \Call{algorithm.do\_housekeeping}{}
-  \EndFor
-\end{algorithmic}
-\end{figure}
-
-The different implementations would be linked and called in a sequential way,
-when new messages are observed on the network and at a regular interval to
-allow them to do internal housekeeping, such as threshold evaluation.
-
-Given these patterns, we can now identify the problem areas that tend to result
-in suboptimal code when combining multiple ID algorithms.
-
-\subsection*{Loops and Common Data}
-
-Two issues arise from the nested loops that originate from the combination of
-the ID algorithm patterns and its application pattern. First, each algorithm,
-by itself, will parse each observed message on the network, processing the same
-byte stream, performing comparable operation to find similar patterns.
-
-Second, there is common data: Each algorithm also has to store information
-about nodes within its own scope, resulting in a lot of duplicated information
-in memory.
-
-Reimplementing the algorithms manually to eliminate these issues is not a valid
-option. This development cost typically isn't a one-time cost, e.g. in case a
-new version of the algorithm is released, or simply when a different set of
-algorithms is selected due to a changing risk analysis. Finally, changing, or
-simply implementing the algorithms from scratch, also holds the risk of making
-mistakes.
-
-\subsection*{Network Usage and Heterogeneity}
-
-A third issue is hidden in the use of the network. The calls to the \ttt{SEND}
-function are scattered around the algorithm, causing the wireless radio to be
-accessed multiple times, keeping it active and energy consuming.
-
-The heterogeneity of the IoT devices, results in many different software
-stacks, each with their own network API. This also explains why there are very
-few implementations readily available to developers: there is no common
-platform to target for researchers.
-
-Even using a unified software library would not only require a consistent
-implementation of it, but would also limit the operational scope of the
-research implementation, still restricting it to a single language.
-
-\section{Design}
-\label{design}
-
-To bridge the gap between the research world and the heterogeneous IoT where
-developers want to implement an actual IDS, while still respecting the limited
-resources they have at their disposal, we introduce \NAME, a framework
-consisting of a DSL, code generator and software library.
-
-The DSL enables researchers to formally describe their ID algorithms, without
-focusing on any specific language or platform. Given such formal descriptions
-of ID algorithms, a code generator is now able to combine these, while avoiding
-the problems arising from the patterns/problems presented in listings
-\ref{alg:id-algo-pattern} and \ref{alg:id-algo-application}. It can also target
-different languages and platforms, while reusing a software library that
-provides e.g. message parsing, network access wrapping,\dots
-
-\subsection*{The Meaning of Loops}
-
-From the design patterns we learned that loops are a major contributor to the
-problem of algorithm fusibility. Loops in program code are technical constructs
-and in fact often hide the actual intended functionality.
+\subsection{Domain Specific Language}
+\label{dsl-design}
+
+Using a DSL, and especially an external DSL, as a foundation for any solution
+requires special care. It's important to remember that a DSL is merely a
+\emph{thin veneer over a (semantic) model}\cite{fowler2010domain}. Its sole
+purpose should be the easing of filling out the model. In the case of \NAME
+this easing focuses on avoiding constructions that result in problems as
+suggested in \ref{pattern-problems} concerning loops, common data and the use
+of the network.
+
+\subsubsection*{Loops}
+
+Loops are a major contributor to the problem of algorithm fusibility. Loops in
+program code are technical constructs and in fact often hide the actual
+intended functionality.
 
@@
 has \emph{activated} this functionality, but the real goal is to
-\emph{passively} react on an event, not \emph{actively} look for it.
-
-This is how \NAME removes the need for loops: by introducing event handling,
-constructions with loops are transformed to their functional meaning, allowing
-much easier identification and fusion of common functionalities.
+\emph{passively} react on an event, not \emph{actively} look for it. Using
+event handling, constructions with loops are transformed to their functional
+meaning, allowing identification and fusion of common functionalities.
 
@@
 
-\subsection*{A DSL for Nodes}
-
-The second problem that could be observed from the patterns, was that of common
-data. All algorithms deal with information about neighboring devices. To
-centralize this information \NAME provides a unified \texttt{nodes} concept as
-part of its DSL and exposes it as an object to the implementor.
+\subsubsection*{The Nodes Domain}
+
+All algorithms deal with information about neighboring devices. To centralize
+this information \NAME provides a unified \ttt{nodes} concept as part of its
+DSL and exposes it as an object to the implementor.
 
 Nodes can be extended with properties only available to the algorithm that
-defines them. This way only one instance for every known devices will be
-present in memory. Finally, nodes can define a scope for some function to be
-executed in. This way, nodes also offer a way to iterate all known devices,
-again without the need for a loop.
-
-\subsection*{Unified Messaging}
+defines them. This way only one instance for every known device will be present
+in memory. Finally, nodes can define a scope for some function to be executed
+in. This way, nodes also offer a way to iterate all known devices, again
+without the need for a loop.
+
+\subsubsection*{Unified Messaging}
+\label{dsl-unified-msg}
+
+Access to the network that is scattered around an algorithm, causes the
+wireless radio to be more activated than needed. Delaying the sending of
+messages and grouping them, can lower this overhead.
 
@@
 
+\subsubsection*{An Example}
+
 Listing \ref{lst:watchdog} illustrates some of the features mentioned above,
@@
 capturing of a device.
-
-\lstdefinelanguage{foo-lang}{
-  emph={module,const,from,import,extend,with,after,do,function,@every,having},
-  emphstyle={\textbf},
-  morecomment=[l]{//},
-  moredelim=[is][]{@@}{@@},       % temp solution to allow keywords without emph
-  moredelim=[is][\emph]{!!}{!!}   % temp solution to hightlight #atoms
-}
 
@@
 
-\subsection*{Code Generator}
+\subsection{Code Generator}
+\label{code-generator-design}
 
@@
 
-A semantic model \cite{fowler2010domain}, tied closely to the DSL itself, can
+A semantic model\cite{fowler2010domain}, tied closely to the DSL itself, can
 fulfill this role. Semantic transformations can now be applied to create a
@@
 \begin{tikzpicture}[
-  node distance=8mm,
+  node distance=7mm,
   ,>=stealth,thick,black!50,text=black,
@@
 
-\subsection*{Semantic Model}
+\subsubsection*{Semantic Model}
 
@@
   ]
-  \node (model) [class] {Model}
-      child {node (module) [class] {Module} edge from parent [composite]
-        child {node (strategy) [class,align=center,xshift=-9mm] {Execution\\Strategy} edge from parent [composite]
-          child {node(every) [class,xshift=-9mm] {Every} edge from parent [subclass]}
-          child {node (when) [class,xshift=9mm] {When} edge from parent [subclass]}
+  \node (model) [class,fill=black!10!white] {Model}
+      child {node (module) [class,fill=black!10!white] {Module} edge from parent [composite]
+        child {node (strategy) [class,align=center,xshift=-10.5mm,fill=black!10!white] {Execution\\Strategy} edge from parent [composite]
+          child {node(every) [class,xshift=-9mm,fill=black!10!white] {Every} edge from parent [subclass]}
+          child {node (when) [class,xshift=9mm,fill=black!10!white] {When} edge from parent [subclass]}
         }
-        child {node (function) [class,xshift=9mm] {Function} edge from parent [composite]}
+        child {node (function) [class,xshift=10.5mm] {Function} edge from parent [composite]}
       };
@@
   \node (statement) [class,right=of function] {Statement};
-  \node (nodes)     [class,left=of module]    {Nodes\\Domain};
+  \node (nodes)     [class,left=of module,xshift=-13mm]    {Nodes\\Domain};
 
@@
   \draw [composite]  (function) -- (statement);
+
+  \draw [composite]  (nodes)    -- (scope);
+  \draw [dependency] (strategy) -- (function);
 \end{tikzpicture}
 }
-\caption{Partial Semantic Model for \NAME}
+\caption{Partial Semantic Model for \NAME. Shaded classes indicate the \emph{event/scheduling} construction around functions to allow interpretation.}
 \label{fig:meta-model}
@@
 The concept driving this model is the structure starting at the \emph{Model} up
-to the \emph{Execution Strategy}. This structure surrounds the actual
+to the \emph{Execution Strategy}. This structure encompasses the actual
 \emph{Function}s and underlying \emph{Statement}s and allows for interpretation
@@
 
-\subsection*{Functional Code Fusion}
+\subsubsection*{Functional Code Fusion (FCF)}
 
@@
 
-\subsection*{Code Model}
+\subsubsection*{Code Model}
 
@@
 
-\subsection*{Software Library}
-
-We already touched upon the underlying software library when discussing the
-unified messaging framework: incoming messages are parsed, outgoing messages
-grouped. Besides these examples, other common functionality also isn't
-generated. The execution of functions based on a schedule or the implementation
-of higher level conceptual functions, such as \texttt{now} and \texttt{sha1},
-are calls into the standard software library that supports the code generator.
+\subsection{Software Library}
+\label{software-lib-design}
+
+In \ref{pattern-problems} the possibility to use a software library to offload
+common implementation aspects was already proposed. \ref{dsl-unified-msg}
+illustrated this approach with an example use for unified messaging: incoming
+messages are parsed, outgoing messages grouped.
+
+Other common functionality also isn't fully generated. The execution of
+functions based on a schedule or the implementation of higher level conceptual
+functions, such as \ttt{now} and \ttt{sha1}, are calls into the standard
+software library that supports the code generator.
+
+Optimizing such algorithms is not within the scope of \NAME. It also gives
+freedom of choice to the users of the framework to select appropriate
+implementations.
+
+\subsection{Design Summary}
+
+As a framework \NAME tries to reduce inefficiencies in execution, memory usage
+and network access. To achieve this, it tries to avoid constructions that are
+hard to identify and reorganize, such as loops and common data. These
+restrictions are enforced through its DSL.
+
+The DSL is platform and language independent, which allows the source code
+generator to target different configurations without any change to the formal
+descriptions of the algorithms.
+
+Finally, a supporting software library allows to separate functionality that is
+out of scope for the generator.
 
@@
 software library is written in native code targeting the platform and language.
-Finally, the code and the image are artifacts of the automated generation and
+Finally, the code and image are artifacts of the automated generation and
 compilation process. They aren't a product of manual work and therefore aren't
@@
 one-time effort, it is important in relation to the scalability of the
-framework.
+framework itself.
 
@@
 ATMEGA1284p micro-controller \cite{datasheet:atmega1284p} and the Digi XBee S2
-ZigBee module \cite{manual:xbee}. To drive the hardware a minimalistic library
-was used, wrapping the technical calls to the components in more easy to use
-functions.
-
-The reason for this at first sight unusual setup is simple: relying on basic
-hardware and software allows us to validate that the generator is capable of
-generating code even for the most basic environment available. Any step up,
-both in hardware or software, would offer better and higher abstractions that
-would make it easier to generate code for that situation. It shows that the
-requirements of the generator towards the platform are minimal, don't rely on
-advanced frameworks or operating systems and can therefore be applied in any
-environment, targeting any platform.
+ZigBee module \cite{manual:xbee}. To drive the hardware a simple library was
+used, wrapping technical calls to the components in more easy to use functions.
+
+Relying on basic hardware and software allows us to validate that the generator
+is capable of generating code even for the most basic environment available.
+Any step up, both in hardware or software, would offer better and higher
+abstractions that would make it easier to generate code for that situation. It
+shows that the requirements of the generator towards the platform are minimal,
+don't rely on advanced frameworks or operating systems and can therefore be
+applied in any environment, targeting any platform.
 
@@
 application was constructed as an includable piece of code. This way it could
-be integrated in an manual implementation and merged into the generated
+be integrated in a manual implementation and merged into the generated
 solution. On top of this baseline we added two intrusion detection related
 algorithms: a watchdog\cite{mishra2004intrusion}, that allows nodes to validate
-each other's continuous presence, and a reputation-building
+each other's presence, and a reputation-based
 algorithm\cite{ganeriwal2008reputation}, that checks if a parent node is
-cooperative and actually forwards messages with a destination further down the
-network.
+cooperative and actually forwards messages further upstream.
 
@@
 and generated all four cases.
-
-Before reviewing the results, it is very important to put these results in a
-proper perspective. Comparing manually written code to generated code is no
-exact science. One can always argue that the quality of the two code bases is
-not comparable and that the manual code always can be improved. Still it might
-be interesting to look at the numbers, given that both code bases are
-constructed with the same intentions and practices. We believe we have achieved
-this and have taken honest decisions while constructing the implementations,
-making them comparable. Numerical analysis of the implementation offers us a
-preliminary estimation of the effect of our proposed solution.
 
@@
 
-From these raw results, we learn that adding algorithms manually to the
-base-application, results in a cumulative increase. The impact of both
-algorithms is simply the sum of the impact of each algorithm by itself. In the
-case of the time required to execute one cycle of the event loop, the total
-time is even a bit higher than simply the sum.
-
-In the case of the generated implementation, this no longer holds: although
-that even the individual increases due to adding a single algorithm are higher
-than in the manual case, the result for the implementation of both algorithms
-is less than the sum of both. Here we clearly see the effect of reusing the
-software library that comes with \NAME. The same goes for all other metrics.
-Maybe most remarkable is the effect on the time of one event loop cycle: both
-algorithms by itself add 150\% with respect to the base case, but when
-combined, the impact hardly increases. Here we learn that the introduced
-software library is responsible for the initial overhead, but it clearly pays
-for itself when adding more algorithms.
+These results show that adding algorithms manually to the base-application has
+a cumulative effect. The impact of both algorithms is simply the sum of the
+impact of each algorithm by itself. In case of the execution time of one cycle
+of the event loop, the total is even a bit higher.
+
+In case of the generated implementation, this effect is no longer observed:
+individual algorithms do have a larger impact by themselves, but the
+combination of both algorithms is less than the sum of both.
+
+Most remarkable is the effect on the time of one event loop cycle: both
+algorithms by itself add 152\%. Adding a second algorithm only augments the
+impact to 188\%. We see the effect of the generic handling of common
+functionality, like message parsing, event scheduling,\dots It comes with an
+overhead for a single algorithm, but clearly pays for itself when adding more
+algorithms.
 
@@
   \end{tabular}
-  \caption{Comparison of both implementations.}
+  \caption{Comparison of both manual and generated implementations.}
   \label{tbl:summary}
@@
 
-When comparing the results of both implementations we first can conclude that
-the generator produces exactly the same code for the base case. We couldn't
-measure any real differences between both implementations. It's important to
-validate this aspect to ensure that both implementations are structurally
-comparable.
+The first observation deals with the base case: both the manual and the
+generated implementations perform exactly the same. There is no fundamental
+structural difference in the overall architecture of the code.
 
@@
 relatively lower when algorithms are combined. With roughly 3KB of additional
-overhead, or 15\% in this simple case, this impact is affordable.
-
-From a functional point of view, the generated code lives up to the
-expectations: both the number of frames as the bytes sent benefit from well
-organized code and reuse of a common software library.
+overhead, or 15\% in this simple case, this impact limited.
+
+Thirdly, when turning to the goal to reduce the network usage, we see that both
+the number of frames as well as the amount of bytes sent have dropped by
+respectively 13 and 7\%.
 
@@
 is about 7\% faster compared to the manual case.
-
-Without any effort to optimize the software library, nor leveraging knowledge
-about the specific algorithms, this very basic generated code shows that the
-proposed principles are not only theoretically feasible but actually result in
-interesting improvements.
 
@@
 
-Using a DSL, and especially an external DSL, as a foundation for any solution
-requires special care. It's important to remember that a DSL is merely a
-\emph{thin veneer over a (semantic) model}\cite{fowler2010domain}. Its sole
-purpose should be the easing of filling out the model. In the case of \NAME the
-reasons to implement a DSL are valid: the key goal is to remove any technical
-restrictions from a formal description of ID algorithms, allowing domain
-experts to produce these specifications\cite{mernik2005and}.
-
-Zooming in on wireless networks of resource constraint devices, we see that
-DSLs are indeed introduced \enquote{to shorten the development cycles
-dramatically during deployments, to reduce the dependency on hardware and
-software engineers, and to lead to a wider adoption outside the computer
-science field}\cite{sadilek2008domain,naumowicz2009prototyping}.
+Within the scope of wireless networks of resource constraint devices, we see
+DSLs being introduced \enquote{to shorten the development cycles dramatically
+during deployments, to reduce the dependency on hardware and software
+engineers, and to lead to a wider adoption outside the computer science
+field}\cite{sadilek2008domain,naumowicz2009prototyping}.
 
@@
 
-Code generation can be applied for a number reasons. In combination with a DSL
-it can optimize a development process, reducing the amount of code that needs
-to be constructed or it can be used to systematically create better code.
+Code generation, in combination with a DSL, can optimize a development process
+by reducing the amount of code that needs to be constructed or systematically
+improving the code quality.
 
@@
 granular scale, thus reducing the update cost, while maintaining flexibility
-and lowering the development effort. As stated in the introduction, \NAME
+and lowering the development effort. As explained in \ref{positioning}, \NAME
 focuses on the reduction of the development effort and update cost, while
@@
 I2S\cite{charitakis2003code} is an example that generates code for
-SNORT\cite{roesch1999snort} rules. To optimise this highly efficient code, it
+SNORT\cite{roesch1999snort} rules. To optimize this highly efficient code, it
 often targets a specific platform.
@@
 would like to thank the reviewers for their thoughtful and helpful comments
-that enhanced the readability of this paper. Our gratitude and respect also
-goes out to all members of the NES task force at KU Leuven/DistriNet for
-creating the nurturing environment where these ideas could grow.
+that enhanced the readability of this paper.
 

### 2014-10-28 paper.tex
added draft conclusion

@@
 
-\TODO
-
-theoretical gains are met in practice
-
-explore more domains + implement corresponding DSLs for embedded systems
-
-extract generic language as host for DSLs
-
-Integration of applications with generated IDS allows actively and dynamically
-responding to changes in the reputation of other nodes.
-
-The services running on those devices can take advantage of these local
-security layers. If the IDS detects a problem, the service can immediately
-terminate the communication with the attacker, before it's actually abused.
+Wireless resource constraint devices require that software respects their
+energy restrictions. Efficiently adding multiple intrusion detection algorithm
+requires fusion of these algorithms on a functional level, to reduce execution
+inefficiencies, redundant memory usage and superfluous network access.
+
+In this paper we introduced \NAME, a framework consisting of a DSL, a source
+code generator and accompanying software library. Using the DSL, intrusion
+detection algorithms can be formally described in such way that, using
+functional code fusion during source code generation, it is possible to
+interpret their functional meaning and merge common functionality and shared
+data. The software library hosts common functionality, such as network message
+parsing, sent message aggregation, event scheduling,\dots as a framework for
+the generator to apply.
+
+Initial experiments confirm the theoretical positive impact on the use of
+resources. The platform independent DSL enables generation of source code
+targeting different platforms and languages, which allows high degrees of reuse
+in heterogenous environments and different configurations.
+
+Near future research will focus on the integration of applications with the
+generated intrusion detection systems. We see opportunities to allow
+application to actively and dynamically respond to intrusions as they are
+detected and avoid malicious payload to harm them.
+
+A continued effort is targeting the reduction of non-platform independent parts
+in the implementation. This way it is possible to add more platforms and
+languages with less effort.
+
+In the longer run, we want to investigate the possibility to extend the
+functional code fusion paradigm to other domains beyond intrusion detection.
 

### 2014-10-29 paper.tex
improved academic writing style removed: 74 PASSIVE VOICE 7 TYPOGRAPHY ERRORS 34 WEASEL WORDS 7 STRUNK AND WHITE 6 ABBR (courtesy of https://github.com/devd/Academic-Writing-Check)

@@
 The internet of things (IoT) presents a great potential to positively influence
-our daily work and life: domotics, assisted living, e-health and enhanced
-learning are only a few examples\cite{atzori2010internet}. But with this
-potential the IoT also presents a significant threat: by opening our smart
-homes and personal data to all these interconnected devices, we also open them
-to everyone who is able to break the virtual locks that protect them. IoT
-devices typically have limited batteries, little processing power and memory is
-restricted to the bare minimum. These properties make it hard to add any
-additional service, including security.
+our daily work and life through domotics, assisted living, e-health and
+enhanced learning\cite{atzori2010internet}. But with this potential the IoT
+also presents a significant threat: by opening our smart homes and personal
+data to all these interconnected devices, we also open them to everyone who is
+able to break the virtual locks that protect them. IoT devices typically have
+limited batteries, processing power and memory. These properties make it hard
+to add any additional service, including security.
 
@@
 detection (ID) steps in: an ID probe monitors all traffic that passes through
-the firewall, looking for patterns and optionally alerting the firewall if a
-malicious pattern is detected\cite{denning1987intrusion}.
+the firewall, looking for patterns and optionally, after detecting a malicious
+patter, alerting the firewall\cite{denning1987intrusion}.
 
@@
 configurations for different devices, both dealing with hard- and software.
-Addressing this in a transparent and automated way is a key success factor for
-any kind of development in this area.
+Addressing this in a transparent and automated way is important for any kind of
+development in this area.
 
@@
 paradigm to solve the problem of hard to reuse and combine ID algorithms. FCF
-identifies common data and functions, and organizes code in such a way that
-redundant iterations, tests and computations are eliminated. Results show
-substantially improved memory usage, execution time and usage of the wireless
-radio, reducing energy consumption.
+identifies common data and functions, and organizes code to eliminate redundant
+iterations, tests and computations. Results show improved memory usage,
+execution time and usage of the wireless radio, reducing energy consumption.
 
@@
 Section \ref{related} explores related work in the field of DSLs for ID and
-code generation for the IoT. Finally, section \ref{conclusion} summarizes our
+code generation for the IoT\@. Finally, section \ref{conclusion} summarizes our
 findings, draws conclusions and identifies topics for future work.
@@
 cooperative decision making. We also identify software attestation as a third
-class. Although it can be considered as a special case of a combination of both
-other classes, its focus is not on the detection of intrusions, but on the
-effect of an intrusion on the system itself.
+class. Although it is a combination of both other classes, its focus is not on
+the detection of intrusions, but on the effect of an intrusion on the system
+itself.
 
@@
 signature or misuse detection, and specification-based detection. In
-\cite{alrajeh2013intrusion} the authors mostly agree with this topology, but
-add the notion of hybrid intrusion detection systems and cross layer intrusion
+\cite{alrajeh2013intrusion} the authors agree with this topology and add the
+notion of hybrid intrusion detection systems and cross layer intrusion
 detection systems as recent advances in research. Although these additional
-categories offer very interesting prospects, the authors have to admit that the
-impact on the resource-constrained devices might still be too much. We
-therefore focus on the three categories agreed upon by most sources.
+categories seem promising, the authors admit that the impact on
+resource-constrained devices will be too much. We therefore focus on the three
+categories agreed upon by most sources.
 
@@
   \item[Anomaly detection] considers a baseline profile and evaluates all
-  activity according to this profile. Aberrant behavior is flagged as a
+  activity according to this profile. It tries to flag aberrant behavior as a
   possible intrusion, but this is no exact science. The baseline profile can
-  evolve over time, especially in wireless and mobile environments. The flagged
-  incidents should therefore also be handled with care and no absolute
-  decisions can be made from single events.
+  evolve over time, especially in wireless and mobile environments. Single
+  events should therefore not result in absolute decisions.
 
@@
   and tries to find evidence in the actions of and communication with a device.
-  Because many attacks use valid ways to communicate with services on the
+  Because attacks often use valid ways to communicate with services on the
   devices, this can also produce false positives and offers no black and white
@@
   previously unknown attacks, it can sometimes not detect specific attacks that
-  can be detected by anomaly or misuse detection.
+  anomaly or misuse detection can.
 
@@
 
-As in many other situations, a group is often stronger than the sum of its
-individual components. Because not all devices are always actively
-participating in the network, some might miss clues that would otherwise lead
-them to detect intrusions. It is hardly impossible for a single device to
-detect a complex attack by itself. Therefore a second important research topic
-consists of cooperative algorithms to combine information from single
-identities into a group-based decision about alleged intrusions. We identify
-two ways by which this cooperative decision making can be constructed and focus
-on the way participants exchange information: broadcasting and interactive.
+A group is often stronger than the sum of its individual components. Because
+not all devices are always actively participating in the network, they can miss
+clues that would otherwise lead them to detect intrusions. It is hardly
+impossible for a single device to detect a complex attack by itself. Therefore
+a second important research topic consists of cooperative algorithms to combine
+information from single identities into a group-based decision about alleged
+intrusions. We identify two ways to construct cooperative decision making and
+focus on the way participants exchange information: broadcasting and
+interactive.
 
@@
 
-  \item[Broadcasting] can sometimes allow a distributed, cooperative algorithm
-  to be implemented locally. This is illustrated by the reputation-based
-  detection algorithm introduced in \cite{ganeriwal2008reputation}, based on
-  the same principle of a watchdog also found in \cite{mishra2004intrusion}.
-  Using this algorithm, nodes can exchange information about the reputation of
-  other nodes, solely with broadcasted messages. Combining this information
-  allows them to decide about their trust in a given node, based on more than
-  their own, often partial, observations alone.
+  \item[Broadcasting] can sometimes allow to implement a distributed,
+  cooperative algorithm locally. The reputation-based detection algorithm
+  introduced in \cite{ganeriwal2008reputation} illustrates this, based on the
+  same principle of a watchdog also found in \cite{mishra2004intrusion}. Using
+  this algorithm, nodes can exchange information about the reputation of other
+  nodes, solely with broadcasts. Combining this information allows them to
+  decide about their trust in a given node, based on more than their own, often
+  partial, observations alone.
 
@@
   exchange of suspected intruder information and allows distributed
-  authentication of those \emph{votes}. Based on these authenticated votes, a
-  distributed decision can be made about the jointly identified intruder.
+  authentication of those \emph{votes}. Based on these authenticated votes, the
+  participants can make a distributed decision about the jointly identified
+  intruder.
 
@@
 duality is software attestation. It offers algorithms for exchanging
-information about the software that is running on a device, and aims to
-identify devices that can no longer prove that their content is unaltered. A
+information about the software that is running on a device and allows to
+identify devices with modified content, that no longer can pass attestation. A
 good example is SAKE\cite{seshadri2008sake}.
 
-Software attestation is hard and many details can cause an algorithm to fail.
-The general feeling is that even in the best conditions, there is always a way
-to circumvent the most ingenious algorithm to perform software-based
-attestation.
+Software attestation is hard and details can cause an algorithm to fail. The
+general feeling is that even in the best conditions, there is always a way to
+circumvent the most ingenious algorithm to perform software-based attestation.
 
@@
 common aspects of these algorithms. It will allow us to define a strategy for
-optimizing the combination of these algorithms. From the classification and
-survey of several algorithms, we extract two activities common to most
-algorithms: the inspection of network traffic and the evaluation of thresholds.
+optimizing the combination of these algorithms. From the classification and a
+survey of algorithms, we extract two activities common to most algorithms: the
+inspection of network traffic and the evaluation of thresholds.
 
 The former is trivial: network traffic is the carrier along which malicious
-payloads reach the device or where patterns of communication can be observed
-when other devices in the vicinity are affected by such payloads.
+payloads reach a device directly. Indirectly a device can also detect aberrant
+patterns of communication by other devices, as a result of a successful attack.
 
@@
 can maybe undeniably decide if an attack is real or not. But when dealing with
-anomaly detection this line surely isn't that crisp. Especially in distributed
-situations where even a neighboring device can not be trusted, detecting
-intrusion is no absolute science. In this case one has to rely on statistics
-and thresholds to decide when to drop trust in a given device.
+anomaly detection this line surely isn't that crisp.
+
+Especially in distributed situations it is impossible to fully trust any
+neighboring device. In this case one has to rely on statistics and thresholds
+to decide when to drop trust in a given device.
 
@@
 observing the actions of other devices. There are valid reasons why a devices
-might miss the theoretical window in which certain actions are expected. Simply
-reacting on such basic threshold might result in false positives. A way to
-overcome this is to decouple the detection of problems and the decision what to
-do about them.
+can miss a theoretical window in which one expects activity. Simply reacting on
+such basic threshold can result in false positives. A way to overcome this is
+to decouple the detection of problems and the decision what to do about them.
 
@@
         \State \dots \Comment{take actions}
-        \State \Call{send}{$node$, ``info''} \Comment{e.g. communicate} \label{alg:id-algo-pattern-send2}
+        \State \Call{send}{$node$, ``info''} \Comment{communicate} \label{alg:id-algo-pattern-send2}
       \EndIf
@@
 
-The different implementations would be linked and called in a sequential way,
-when new messages are observed on the network and at a regular interval to
-allow them to do internal housekeeping, such as threshold evaluation.
+The pattern calls the different, linked implementations in a sequential way in
+two situations: first, when the device observes new messages on the network
+and, secondly, at a regular interval to allow for internal housekeeping, such
+as threshold evaluation.
 
@@
 
-  \item[Loops] by themselves are not a problem, but when different loops
-  perform the same operations in sequence, an overhead might be incurred. An
-  example from the ID algorithm pattern is the handling of messages: each
-  algorithm parses each observed message on the network, processing the same
-  byte stream, performing comparable operations to find similar patterns.
+  \item[Loops] by themselves are not a problem, but different loops, who
+  perform the same operations in sequence, can incur an overhead. An example
+  from the ID algorithm pattern is the handling of messages: each algorithm
+  parses each observed message on the network, processing the same byte stream,
+  performing comparable operations to find similar patterns.
 
@@
   store information about nodes within its own scope. This results in a lot of
-  duplicated information in memory that also needs to be maintained.
+  duplicated information in memory.
 
   \item[The network] allows algorithms to operate on a distributed scale. But
-  calls to the \ttt{SEND} function are scattered around the algorithms, causing
-  the wireless radio to be accessed multiple times, keeping it active and
-  energy consuming more than strictly needed.
+  scattered calls to the \ttt{SEND} function cause multiple accesses to the
+  wireless radio, keeping it active and energy consuming more than strictly
+  needed.
 
@@
 Reimplementing the algorithms manually to eliminate these issues is not a valid
-option. This development cost typically isn't a one-time cost, e.g. in case a
-new version of the algorithm is released, or simply when a different set of
-algorithms is selected due to a changing risk analysis. Also, changing, or
+option. This development cost typically isn't a one-time cost. For example in
+case of a release of a new version of the algorithm, or simply a different
+selection of algorithms due to a changing risk analysis. Also, changing, or
 simply implementing the algorithms from scratch, holds the risk of making
@@
 
-The heterogeneity of IoT devices results in many different software stacks,
-each with e.g. their own network API. This also explains why there are very few
+The heterogeneity of IoT devices results in different software stacks, each
+with for example their own network API\@. This also explains why there are no
 implementations readily available to developers: there is no common platform to
@@
 To bridge the gap between the research world and the heterogeneous IoT where
-developers want to implement an actual IDS, while still respecting the limited
+developers want to implement an IDS, while still respecting the limited
 resources they have at their disposal, we introduce \NAME, a framework
@@
 different languages and platforms, while reusing a software library that
-provides e.g. message parsing, network access wrapping,\dots
+provides for example message parsing, network access wrapping,\dots
 
 Choosing for a DSL and code generator is choosing a development strategy. This
-implies several choices with respect to flexibility, update cost and
-development effort. We will first position \NAME and motivate its design with
-respect to these trade offs (\ref{positioning}). Second, we will look at the
-principal design components of the DSL and how it deals with loops and common
-data (\ref{dsl-design}). Third, the structure of the code generator will be
-discussed (\ref{code-generator-design}). Finally, fourth, the supporting
-software library is introduced (\ref{software-lib-design}).
+implies choices with respect to flexibility, update cost and development
+effort. We will first position \NAME and motivate its design with respect to
+these trade offs (\ref{positioning}). Second, we will look at the principal
+design components of the DSL and how it deals with loops and common data
+(\ref{dsl-design}). Third, we discuss the structure of the code generator
+(\ref{code-generator-design}) and finally, we introduce the supporting software
+library (\ref{software-lib-design}).
 
@@
 
-Figure \ref{fig:positioning} shows a classification of several development
-strategies that can be adopted for projects involving wireless devices.
+Figure \ref{fig:positioning} shows a classification of possible development
+strategies for projects involving wireless devices.
 
@@
   % axis
-  \draw[very thick,->] (0,0,0) -- (7,0,0) node[anchor=north east,xshift=10mm]{$flexibility$};
-  \draw[very thick,->] (0,0,0) -- (0,7,0) node[anchor=north west]{$update\ cost$};
-  \draw[very thick,->] (0,0,0) -- (0,0,4) node[anchor=south]{$development$};
+  \draw[thick,->] (0,0,0) -- (7,0,0) node[anchor=north east,xshift=10mm]{$flexibility$};
+  \draw[thick,->] (0,0,0) -- (0,7,0) node[anchor=north west]{$update\ cost$};
+  \draw[thick,->] (0,0,0) -- (0,0,4) node[anchor=south]{$development$};
   % points
@@
 
-Three properties are used to classify the strategies\cite{sadilek2007energy}:
-flexibility, update cost and development effort.
+Three properties classify the strategies\cite{sadilek2007energy}: flexibility,
+update cost and development effort.
 
@@
 
-  \item[Flexibility] is a measure for the number of applications that can be
-  deployed on a device without physical access.
+  \item[Flexibility] is a measure for the number of applications one can deploy
+  on a device without physical access.
 
@@
 distinguish two extremes: \emph{virtual machines} and \emph{full native
-images}. Virtual machines allow for very small updates due to their concise
-byte code, but at the same time are restricted by the limitations of their
-instruction set and predefined capabilities. A full native image on the other
-hand has unlimited flexibility, but requires transfer of a larger image over
-the network.
+images}. Virtual machines allow for small updates due to their concise byte
+code, but their instruction set and predefined capabilities restrict them at
+the same time. A full native image on the other hand has unlimited flexibility,
+but requires transfer of a larger image over the network.
 
 Two intermediate alternatives are possible: \emph{modular native image} and
-\emph{reconfigurable virtual machines}. With modular native images, only
-modules are changed, limiting the update size, but the framework to support it
-is static and limits the possibilities. Reconfigurable virtual machines take
-this same approach, but apply it to virtual machines, allowing to
-\emph{upgrade} the instruction set of the virtual machine.
-
-Choosing a development strategy also determines the development cost: e.g.
+\emph{reconfigurable virtual machines}. Modular native images allow changes on
+module level, limiting the update size, but the framework to support it is
+static and limits the possibilities. Reconfigurable virtual machines take this
+same approach, but apply it to virtual machines, allowing to \emph{upgrade} the
+instruction set of the virtual machine.
+
+Choosing a development strategy also determines the development cost: for example
 developing for a virtual machine will be less demanding than targeting a native
@@
 Loops are a major contributor to the problem of algorithm fusibility. Loops in
-program code are technical constructs and in fact often hide the actual
-intended functionality.
+program code are technical constructs and in fact often hide the intended
+functionality.
 
@@
 has \emph{activated} this functionality, but the real goal is to
-\emph{passively} react on an event, not \emph{actively} look for it. Using
-event handling, constructions with loops are transformed to their functional
-meaning, allowing identification and fusion of common functionalities.
+\emph{passively} react on an event, not \emph{actively} look for it. Event
+handling can transform constructions with loops to their functional meaning,
+allowing identification and fusion of common functionality.
 
 The same goes for actively polling some variable until it reaches a certain
-threshold. This also can be transformed to an event, eliminating the need for a
-technical loop-construction.
+threshold. When transformed to a reaction to an event, it eliminates the need
+for a technical loop-construction.
 
@@
 this information \NAME provides a unified \ttt{nodes} concept as part of its
-DSL and exposes it as an object to the implementor.
-
-Nodes can be extended with properties only available to the algorithm that
-defines them. This way only one instance for every known device will be present
-in memory. Finally, nodes can define a scope for some function to be executed
-in. This way, nodes also offer a way to iterate all known devices, again
-without the need for a loop.
+DSL and exposes it as an object to the implementer.
+
+Algorithms can extend nodes with their own properties. This way only one
+instance for every known device will be present in memory. Finally, nodes can
+define a scope to execute a function in. This way, nodes also offer a way to
+iterate all known devices, again without the need for a loop.
 
@@
 
-Access to the network that is scattered around an algorithm, causes the
-wireless radio to be more activated than needed. Delaying the sending of
-messages and grouping them, can lower this overhead.
+Access to the network, scattered around an algorithm, causes the wireless radio
+to be more active than needed. Delaying the sending of messages and grouping
+them, can lower this overhead.
 
@@
 underlying software framework will handle all interaction with the network,
-triggering functions by means of events when new messages are parsed and
-exposing an API to send messages through the nodes domain. This approach gives
-the framework complete control over the way messages are sent, allowing
-optimized marshaling, grouping,\dots
+triggering functions by means of events when parsing new messages and exposing
+an API to send messages through the nodes domain. This approach gives the
+framework complete control over the messages, allowing optimized marshaling,
+grouping,\dots
 
@@
 To convert the DSL-based ID algorithms into production-ready code, we propose
-the use of a code generator. Although many approaches here are applicable, like
-template-based text expansion, XSLT transformations or even using the
-CodeDom\cite{dollard2004code}, we believe that for the level of functional
-transformations that need to be performed, a richer model is required.
+the use of a code generator. Although different approaches here are applicable,
+like template-based text expansion, XSLT transformations or even using the
+CodeDom\cite{dollard2004code}, we believe that the functional transformations
+require a richer model.
 
 A semantic model\cite{fowler2010domain}, tied closely to the DSL itself, can
-fulfill this role. Semantic transformations can now be applied to create a
+fulfill this role. Semantic transformations evolve the model into a
 code-oriented model, comparable to the idea of CodeDom. From there on,
-syntactic transformations an emit actual source code. Figure
-\ref{fig:code-generation} shows the overall procedure.
+syntactic transformations an emit source code. Figure \ref{fig:code-generation}
+shows the overall procedure.
 
@@
 
-The DSL is parsed and imported into a semantic model. Based on the minimal type
-information present in the model, all remaining types are inferred. Through a
-series of semantic transformations, the model evolves from the functional to a
-more technical one, a code model. This code model can be considered as a rich
-abstract syntax tree (AST) with an extensive set of language constructs.
+After parsing the DSL, it's imported into a semantic model. The minimal type
+information present in the initial model allows inferring of the remaining
+types. Through a series of semantic transformations, the model evolves from the
+functional to a more technical code model. This code model is comparable to a
+rich abstract syntax tree (AST) with an extensive set of language constructs.
 Syntactic transformations now lower this level of expressiveness according to
@@
 The semantic model is the core of the solution. It offers a way to describe ID
-algorithms in a way that allows semantic reinterpretation. This way, the
-functionality of different algorithms can be reorganized to allow for more
-optimal execution and memory usage. Hence the name \NAME of the framework:
-intrusion detection functionality organization optimization.
+algorithms in a way that allows semantic reinterpretation. This way,
+reorganizing the functionality of different algorithms allows for more optimal
+execution and memory usage. Hence the name \NAME of the framework: intrusion
+detection functionality organization optimization.
 
 Figure \ref{fig:meta-model} shows the central part of the semantic model as
-implemented for \NAME.
+implemented for \NAME\@.
 
@@
 }
-\caption{Partial Semantic Model for \NAME. Shaded classes indicate the \emph{event/scheduling} construction around functions to allow interpretation.}
+\caption{Partial Semantic Model for \NAME\@. Shaded classes indicate the \emph{event/scheduling} construction around functions to allow interpretation.}
 \label{fig:meta-model}
@@
 The concept driving this model is the structure starting at the \emph{Model} up
-to the \emph{Execution Strategy}. This structure encompasses the actual
+to the \emph{Execution Strategy}. This structure encompasses the
 \emph{Function}s and underlying \emph{Statement}s and allows for interpretation
@@
 
-Thanks to the functional description of events and the effects that are
-triggered by them, the code generator is able to identify the same actions
-across different algorithms. This way it is possible to extract the parsing of
-incoming messages, group actions on the same sets of data or scheduled function
+Thanks to the functional description of events and the effects triggered by
+them, the code generator is able to identify the same actions across different
+algorithms. This way it is possible to extract the parsing of incoming
+messages, group actions on the same sets of data or scheduled function
 executions.
@@
 FCF typically looks for patterns in the semantic model and performs model
-transformations to change these patterns to a more optimized organization.
-Semantic transformations are performed on the semantic model itself and end
-with a transformation of the semantic model into a code model, which encodes
-the semantics into technical code constructs.
+transformations to change these patterns to a more optimized organization. It
+first performs semantic transformations on the semantic model itself and ends
+then transforms the semantic model into a code model, encoding the semantics
+into technical code constructs.
 
@@
 
-The code model can be compared to an AST for a very expressive language. It
-contains constructs found in certain languages, but not in others. Therefore a
-second round of model transformations needs to be performed. These syntactic
-transformations morph the unsupported constructs into comparable constructs
-that are supported by the targeted platform and language.
-
-Depending on the target platform, some of the constructions used to e.g. parse
-incoming messages, schedule tasks,\dots may not be transformed into
-self-containing code. The generator then uses a software library to perform
-these tasks. It will look for patterns that indicate any of these actions and
-transform them into calls into the \NAME software library.
+The code model can is comparable to an AST for an expressive language. It
+contains constructs found in certain languages, but not in others. Therefore,
+syntactic transformations morph the unsupported constructs into comparable
+constructs supported by the targeted platform and language.
+
+Depending on the target platform, some constructions aren't transformed into
+self-containing code. The generator uses a software library for example in case
+of parsing incoming messages and scheduling of tasks. It will look for patterns
+that indicate any of these actions and transform them into calls into the \NAME
+software library.
 
@@
 common implementation aspects was already proposed. \ref{dsl-unified-msg}
-illustrated this approach with an example use for unified messaging: incoming
-messages are parsed, outgoing messages grouped.
+illustrated this approach with an example use for unified messaging: parsing of
+incoming messages and grouping of outgoing messages.
 
@@
 
-Optimizing such algorithms is not within the scope of \NAME. It also gives
+Optimizing such algorithms is not within the scope of \NAME\@. It also gives
 freedom of choice to the users of the framework to select appropriate
@@
 and network access. To achieve this, it tries to avoid constructions that are
-hard to identify and reorganize, such as loops and common data. These
-restrictions are enforced through its DSL.
+hard to identify and reorganize, such as loops and common data. The DSL
+enforces these restrictions.
 
@@
 abstracted subsystems, but still in a platform and language independent way.
-The platform and language descriptions define the actual target platform and
-language and convert the abstract constructions in actually supported code. The
-software library is written in native code targeting the platform and language.
-Finally, the code and image are artifacts of the automated generation and
-compilation process. They aren't a product of manual work and therefore aren't
-directly dependent on the platform.
+The platform and language descriptions define the target platform and language
+and convert the abstract constructions in actually supported code. The software
+library is standard native code, targeting the platform and language. Finally,
+the code and image are artifacts of the automated generation and compilation
+process. They aren't a product of manual work and therefore aren't directly
+dependent on the platform.
 
@@
 ATMEGA1284p micro-controller \cite{datasheet:atmega1284p} and the Digi XBee S2
-ZigBee module \cite{manual:xbee}. To drive the hardware a simple library was
-used, wrapping technical calls to the components in more easy to use functions.
+ZigBee module \cite{manual:xbee}. A basic library wrapped the technical calls
+to the hardware components in more easy to use functions.
 
@@
 shows that the requirements of the generator towards the platform are minimal,
-don't rely on advanced frameworks or operating systems and can therefore be
-applied in any environment, targeting any platform.
+don't rely on advanced frameworks or operating systems and therefore are
+applicable in any environment, targeting any platform.
 
@@
 We started from a basic application that measures light intensity. The
-application was constructed as an includable piece of code. This way it could
-be integrated in a manual implementation and merged into the generated
-solution. On top of this baseline we added two intrusion detection related
-algorithms: a watchdog\cite{mishra2004intrusion}, that allows nodes to validate
-each other's presence, and a reputation-based
-algorithm\cite{ganeriwal2008reputation}, that checks if a parent node is
-cooperative and actually forwards messages further upstream.
+application was an includable piece of code, which allowed integration in a
+manual implementation and simple merging when generation. On top of this
+baseline we added two intrusion detection related algorithms: a
+watchdog\cite{mishra2004intrusion}, that allows nodes to validate each other's
+presence, and a reputation-based algorithm\cite{ganeriwal2008reputation}, that
+checks if a parent node is cooperative and actually forwards messages further
+upstream.
 
@@
 
-Three criteria were evaluated: the image size of the resulting compiled code,
-the network usage in number of frames and bytes and the time required to
-perform once cycle of the event loop. These metrics were collected in four
+We evaluated three criteria: the image size of the resulting compiled code, the
+network usage in number of frames and bytes and the time required to perform
+once cycle of the event loop. To do this, we collected these metrics for four
 situations: without ID algorithms, with a watchdog, with reputation-tracking
@@
 
-In case of the manual implementation, both algorithms were constructed as
-standalone modules and sequentially called from the base-application's event
-loop. The code generator was provided with \NAME descriptions of the algorithms
-and generated all four cases.
-
-Tables \ref{tbl:manual} and \ref{tbl:generated} show the collected data
-respectively for the manual and the generated implementation. The base case is
-presented in absolute values, while the implementations with the ID algorithms
-added are presented as relative values.
+The manual implementation implemented both algorithms as standalone modules and
+sequentially called them from the base-application's event loop. Given the
+\NAME descriptions of the algorithms, the code generator generated all four
+cases.
+
+Tables \ref{tbl:manual} and \ref{tbl:generated} show the collected data for the
+manual and the generated implementation. They present the base case in absolute
+values and the implementations with the ID algorithms relative to that base
+case.
 
@@
 functionality, like message parsing, event scheduling,\dots It comes with an
-overhead for a single algorithm, but clearly pays for itself when adding more
+overhead for a single algorithm, but pays for itself when adding more
 algorithms.
@@
 Secondly, we see that the software library that comes with the generated code
-adds to the size of the resulting image. But the cost is almost constant and is
-relatively lower when algorithms are combined. With roughly 3KB of additional
-overhead, or 15\% in this simple case, this impact limited.
+adds to the size of the resulting image. But the cost is close to constant and
+is lower for both algorithms together. With roughly 3KB of additional overhead,
+or 15\% in this simple case, this impact limited.
 
 Thirdly, when turning to the goal to reduce the network usage, we see that both
-the number of frames as well as the amount of bytes sent have dropped by
-respectively 13 and 7\%.
+the number of frames as well as the amount of bytes sent have dropped by 13 and
+7\%.
 
@@
 
-This section now focuses on related work in the field of domain specific
-languages and code generation, both in relation to wireless networks of
-resource constraint devices. These two components are the core of the \NAME
-framework and the solution it proposes. We also look at the more general scope,
-how these concepts are applied in other disciplines and evaluate their
-relevance in the scope of intrusion detection in the IoT and the proposed \NAME
-framework.
+This section focuses on related work in the field of domain specific languages
+and code generation, both in relation to wireless networks of resource
+constraint devices. These two components are the core of the \NAME framework
+and the solution it proposes. We also look at the more general scope: the
+application in other disciplines and their relevance in the scope of intrusion
+detection in the IoT and the proposed \NAME framework.
 
@@
 Within the scope of wireless networks of resource constraint devices, we see
-DSLs being introduced \enquote{to shorten the development cycles dramatically
-during deployments, to reduce the dependency on hardware and software
-engineers, and to lead to a wider adoption outside the computer science
-field}\cite{sadilek2008domain,naumowicz2009prototyping}.
+the introduction of DSLs \enquote{to shorten the development cycles
+dramatically during deployments, to reduce the dependency on hardware and
+software engineers, and to lead to a wider adoption outside the computer
+science field}\cite{sadilek2008domain,naumowicz2009prototyping}.
 
@@
 specific, but through extensions they allow to model attack scenarios using
-state machines. This is an interesting approach that shares the same ideas as
-our event-driven concept. The framework generates C++.
+state machines. Our event-driven approach shares this same ideas.
 
@@
 Code generation, in combination with a DSL, can optimize a development process
-by reducing the amount of code that needs to be constructed or systematically
-improving the code quality.
+by reducing the size of the code base or systematically improving the code
+quality.
 
@@
 of a high quality. Focus on size and performance are a logical choice when
-dealing with limited storage and processing power. Different techniques are
-described in literature and implemented in compilers, starting at the lowest
-level\cite{leupers2000code, marwedel2002code}.
+dealing with limited storage and processing power. Compilers implement
+different techniques to address this at the lowest level\cite{leupers2000code,
+marwedel2002code}.
 
 Optimizing for power efficiency however proves to be concurrent to size and
-performance optimization. Different strategies are needed. Data and memory
-optimizations \cite{panda2001data} can result in interesting improvements on a
-lower level. But to really preserve energy higher-level strategies are
-required, targeting architecture, radio handling, efficient
-software,\dots\cite{naik2001software}. Most of these are beyond the control of
-the compiler and show that a higher level of abstraction is required to address
-them.
+performance optimization, and requires different strategies. Data and memory
+optimizations \cite{panda2001data} can result in improvements on a lower level.
+But really preserving energy requires higher-level strategies, targeting
+architecture, radio handling, efficient software,\dots\cite{naik2001software}.
+Most of these are beyond the control of the compiler and require a higher level
+of abstraction to address them.
 
@@
 seems a promising approach. By offloading the optimization to a VM and only
-compiling parts and not the whole, the solution allows updating at a very
-granular scale, thus reducing the update cost, while maintaining flexibility
-and lowering the development effort. As explained in \ref{positioning}, \NAME
+compiling parts and not the whole, the solution allows updating at a granular
+scale, thus reducing the update cost, while maintaining flexibility and
+lowering the development effort. As explained in \ref{positioning}, \NAME
 focuses on the reduction of the development effort and update cost, while
@@
 
-In this paper we introduced \NAME, a framework consisting of a DSL, a source
-code generator and accompanying software library. Using the DSL, intrusion
-detection algorithms can be formally described in such way that, using
-functional code fusion during source code generation, it is possible to
-interpret their functional meaning and merge common functionality and shared
-data. The software library hosts common functionality, such as network message
-parsing, sent message aggregation, event scheduling,\dots as a framework for
-the generator to apply.
+We introduced \NAME, a framework consisting of a DSL, a source code generator
+and accompanying software library. Using the DSL, intrusion detection
+algorithms can be formally described in such way that, using functional code
+fusion during source code generation, it is possible to interpret their
+functional meaning and merge common functionality and shared data. The software
+library hosts common functionality, such as network message parsing, sent
+message aggregation, event scheduling,\dots as a framework for the generator to
+apply.
 
@@
 Near future research will focus on the integration of applications with the
-generated intrusion detection systems. We see opportunities to allow
-application to actively and dynamically respond to intrusions as they are
-detected and avoid malicious payload to harm them.
+generated intrusion detection systems. We see opportunities for applications
+who detect intrusion, to actively and dynamically respond and avoid malicious
+payloads to harm them.
 
@@
 Programme Belgian State, Belgian Science Policy, and by the Research Fund KU
-Leuven. This research is partially funded by the EU FP7 project NESSoS. We
+Leuven. This research is partially funded by the EU FP7 project NESSoS\@. We
 would like to thank the reviewers for their thoughtful and helpful comments

### 2014-11-05 paper.tex
processing of review comments

@@
 
-% additional packages and utility commands
-
-\usepackage{float}
-
-\newcommand{\TODO}{\textbf{\color{red}TODO}}
-
-% \usepackage{flushend}
-
-% while preparing, linenumbers come in handy
-\usepackage{lineno}
-\setlength\linenumbersep{1mm}
-\linenumbers
-
-% until we settle on the final name ;-)
-\usepackage{xspace}
-\newcommand{\NAME}{id-foo\xspace}
-
-% http://tex.stackexchange.com/questions/299/how-to-get-long-texttt-sections-to-break
-\newcommand*\justify{%
-  \fontdimen2\font=0.4em% interword space
-  \fontdimen3\font=0.2em% interword stretch
-  \fontdimen4\font=0.1em% interword shrink
-  \fontdimen7\font=0.1em% extra space
-  \hyphenchar\font=`\-% allowing hyphenation
-}
-
-% http://en.wikibooks.org/wiki/LaTeX/Customizing_LaTeX
-\newcommand{\ttt}[1]{\texttt{\justify{#1}}}
-
-\usepackage{algpseudocode}
-% custom Let command
-\newcommand*\Let[2]{\State #1 $\gets$ #2}
-\renewcommand{\algorithmicforall}{\textbf{for each}}
-\let\ForEach\ForAll
-
-\usepackage{listings}
-
-\usepackage{tikz}
-\usetikzlibrary{positioning}
-\usetikzlibrary{calc}
-\usetikzlibrary{graphs}
-\usetikzlibrary{trees}
-\usetikzlibrary{arrows}
-
-\usepackage{tikz-3dplot}
-
-\include{tikz}
-\include{foo-lang}
-
-\usepackage[autostyle]{csquotes}
+\input{packages}
 
 \begin{document}
-
-% to avoid syntax error highlighting - e.g. foo's not js ;-)
-\expandafter\def\csname PY@tok@err\endcsname{}
 
@@
 
-\author{\IEEEauthorblockN{Christophe Van Ginneken, Jef Maerien, Christophe
-Huygens, Danny Hughes, Wouter Joosen}%
+\author{\IEEEauthorblockN{Christophe Van Ginneken, Jef Maerien, Danny Hughes, 
+Christophe Huygens, Wouter Joosen}%
 \IEEEauthorblockA{iMinds-DistriNet, KU Leuven\\
@@
 When intrusion prevention fails, intrusion detection serves as a second layer
-of defense. It looks for patterns, anomalies,\dots that can indicate malicious
-behavior. But supporting these intrusion detection algorithms imposes
-significant overhead on resource-restricted devices such as those targeting the
-Internet of Things. Fusion of the algorithms, to optimize the use of the shared
-resources, can alleviate this, yet, proves to be time-consuming, repetitive and
-error-prone. To address this problem we introduce \NAME, a framework for the
-development of efficient intrusion detection systems. It consists of a domain
-specific language, to formally describing intrusion detection algorithms on a
-functional level, and a code generator to produce more optimally organized
+of defense. It looks for patterns \& anomalies that can indicate malicious
+behavior. But supporting an adequate set of intrusion detection algorithms
+imposes significant overhead on resource-restricted devices such as those
+targeting the Internet of Things. Fusion of the algorithms, to optimize sharing
+of resources, can alleviate this, yet, proves to be time-consuming, repetitive
+and error-prone. To address this problem we introduce \NAME, a framework for
+the development of efficient intrusion detection systems. It consists of a
+domain specific language, to formally describe intrusion detection algorithms
+on a functional level, and a code generator to produce more optimally organized
 source code. A side-by-side comparison shows that generated code reduces memory
@@
 
-The internet of things (IoT) presents a great potential to positively influence
+The internet of things (IoT) holds a great potential to positively influence
 our daily work and life through domotics, assisted living, e-health and
-enhanced learning\cite{atzori2010internet}. But with this potential the IoT
+enhanced learning \cite{atzori2010internet}. But with this potential the IoT
 also presents a significant threat: by opening our smart homes and personal
 data to all these interconnected devices, we also open them to everyone who is
-able to break the virtual locks that protect them. IoT devices typically have
-limited batteries, processing power and memory. These properties make it hard
-to add any additional service, including security.
+able to break the virtual locks (i.e.\ passwords, encryption keys,\dots) that
+protect them. IoT devices typically have limited batteries, processing power
+and memory. These properties make it hard to add any additional service,
+including security.
 
@@
 the firewall, looking for patterns and optionally, after detecting a malicious
-patter, alerting the firewall\cite{denning1987intrusion}.
-
-% manet: local ids -> localize -> limited resources
+pattern, alerting the firewall \cite{denning1987intrusion}.
+
+% local ids -> localize -> limited resources
 
@@
 significant part of the IoT, it is not possible to have this single point in
-the network that can oversee all traffic\cite{mishra2004intrusion}. If the
-outer security perimeter diminishes, every device has to implement its own
-lines of defense. But making the IDS a local service on each device requires
-local resources, which are the scarce commodity to IoT devices.
+the network that can oversee all traffic \cite{mishra2004intrusion}. Every
+device has to implement its own lines of defense. But making the intrusion
+detection system (IDS) a local service on each device requires local resources,
+which are a scarce commodity for IoT devices.
+
+Attackers have access to a large set of possible attack vectors
+\cite{aschenbruck2012security}, ranging from the wireless network, over the
+medium access control and routing, up to the device itself. Each level presents
+different opportunities to manipulate data, eavesdrop, perform a form of denial
+of service or access the network in a legitimate way. For each of these attacks
+an algorithm needs to look for patterns or anomalies. An IDS for IoT devices
+therefore requires a large number of implemented algorithms for an adequate
+coverage.
 
@@
 Current developments in ID focus on programming frameworks to structure the
-implementation of algorithms\cite{valero2012di} and offer the required basic
-functional components to implement them\cite{krontiris2008lidea}. The first
+implementation of algorithms \cite{valero2012di} and offer the required basic
+functional components to implement them \cite{krontiris2008lidea}. The first
 aspect that is missing, is that they don't offer a way to optimize the usage of
@@
 
-Secondly, they don't support systematic reuse to create different
-configurations for different devices, both dealing with hard- and software.
-Addressing this in a transparent and automated way is important for any kind of
+A second omission is support for systematic reuse to create different
+configurations for different devices, both in hard- and software. Addressing
+this in a transparent and automated way is important for any kind of
 development in this area.
@@
 The remainder of this paper proceeds as follows: section \ref{classification}
-takes a look at ID algorithms and classifies them. Section \ref{pattern} looks
-for patterns in the implementation of these classes of algorithms, details some
-of the ideas behind FCF and identifies the different aspects of the problem
-space. Section \ref{design} describes the design we applied to construct the
+analyses and classifies ID algorithms. Section \ref{pattern} looks for patterns
+in the implementation of these classes and details the ideas behind FCF\@.
+Section \ref{design} describes the design we applied to construct the
 framework, its domain specific language (DSL) and code generator. Section
-\ref{evaluation} evaluates an implementation and application of the framework.
-Section \ref{related} explores related work in the field of DSLs for ID and
-code generation for the IoT\@. Finally, section \ref{conclusion} summarizes our
+\ref{evaluation} evaluates an implementation of the framework. Section
+\ref{related} explores related work in the field of DSLs for ID and code
+generation for the IoT\@. Finally, section \ref{conclusion} summarizes our
 findings, draws conclusions and identifies topics for future work.
@@
 
-Research into ID in wireless networks typically focuses on two major topics
-that complement the architecture: the devices as single entities and the
-network as a group of such devices. Both are important because the group cannot
-make a decision without members that detect malicious behavior and a
-group-based decision is often needed because nodes can easily miss out on
-certain events that could indicate intrusions, due to their wireless and
-not-always-on nature.
-
-These two topics lead to the first two classes: detecting intrusions and
-cooperative decision making. We also identify software attestation as a third
-class. Although it is a combination of both other classes, its focus is not on
-the detection of intrusions, but on the effect of an intrusion on the system
-itself.
-
-\subsection*{Detecting Intrusions}
+Two aspects, currently missing from ID frameworks, that \NAME wants to cover
+are optimization of resources and support for systematic reuse. This section
+uses a classification to define the scope of \NAME, detailing which types of
+algorithms are good candidates for the proposed approach.
+
+Research into ID in wireless networks typically focuses on two major topics:
+the devices as single entities and the network as a group of such devices. Both
+are important: a group cannot make a decision without members that detect
+malicious behavior and a group-based decision is often needed because nodes in
+the network can miss certain events that could reveal intrusions. In a wireless
+network, where not all participants are within each others' radio range, nodes
+typically only have a partial view on the entire network. This prohibits them
+to observe potentially crucial events in distant parts of the network. The fact
+that resource constraint devices typically apply duty cycling to lower their
+energy consumption, further increases this problem, introducing periods of time
+where they even don't see local events.
+
+This leads to two complementary classes of ID activities: local intrusion
+detection and cooperative decision making.
+
+\subsection{Local Intrusion Detection}
 \label{detection}
@@
 signature or misuse detection, and specification-based detection. In
-\cite{alrajeh2013intrusion} the authors agree with this topology and add the
+\cite{alrajeh2013intrusion} the authors agree with this typology and add the
 notion of hybrid intrusion detection systems and cross layer intrusion
@@
   activity according to this profile. It tries to flag aberrant behavior as a
-  possible intrusion, but this is no exact science. The baseline profile can
-  evolve over time, especially in wireless and mobile environments. Single
-  events should therefore not result in absolute decisions.
-
-  \item[Signature or misuse detection] starts with known patterns of attacks
-  and tries to find evidence in the actions of and communication with a device.
-  Because attacks often use valid ways to communicate with services on the
-  devices, this can also produce false positives and offers no black and white
-  certainty.
+  possible intrusion. The baseline profile can evolve over time, especially in
+  wireless and mobile environments. Single events should therefore not result
+  in absolute decisions.
+
+  \item[Signature or misuse detection] starts with known patterns of attacks.
+  It tries to find evidence for intrusions in the actions of and communication
+  with a device. Because attacks often use valid ways to communicate with
+  services on the devices, this approach can produce false positives.
 
@@
 
-\subsection*{Cooperative Decision Making}
+\subsection{Cooperative Decision Making}
 \label{coorperative}
@@
 These taxonomies focus on the act of detecting intrusions themselves. In a
-highly distributed environment, an orthogonal categorization, dealing with the
-distributed nature of the network and its ways of communicating, may be even
-more important.
-
-A group is often stronger than the sum of its individual components. Because
-not all devices are always actively participating in the network, they can miss
-clues that would otherwise lead them to detect intrusions. It is hardly
-impossible for a single device to detect a complex attack by itself. Therefore
-a second important research topic consists of cooperative algorithms to combine
-information from single identities into a group-based decision about alleged
-intrusions. We identify two ways to construct cooperative decision making and
-focus on the way participants exchange information: broadcasting and
+distributed environment, communication between participants may be even more
+important.
+
+Because not all devices are always actively participating in the network, they
+can miss clues that would otherwise lead them to detect intrusions. It is
+hardly impossible for a single device to detect a complex attack by itself.
+Therefore a second important research topic consists of cooperative algorithms
+to combine information from single identities into a group-based decision about
+alleged intrusions. We identify two ways to construct cooperative decision
+making and focus on the way participants exchange information: broadcasting and
 interactive.
@@
   phases. It essentially implements a voting system-based on the Guy Fawkes
-  protocol\cite{anderson1998new} to identify an intruder. It enables the
+  protocol \cite{anderson1998new} to identify an intruder. It enables the
   exchange of suspected intruder information and allows distributed
@@
 
-\subsection*{Software Attestation}
-\label{subsection:attestation}
-
-A research topic that exists in parallel to the detection and cooperation
-duality is software attestation. It offers algorithms for exchanging
-information about the software that is running on a device and allows to
-identify devices with modified content, that no longer can pass attestation. A
-good example is SAKE\cite{seshadri2008sake}.
-
-Software attestation is hard and details can cause an algorithm to fail. The
-general feeling is that even in the best conditions, there is always a way to
-circumvent the most ingenious algorithm to perform software-based attestation.
-
-Within the context of this paper, we will not consider software attestation as
-a special class and exclude it from our scope, only focusing on detecting
-intrusions and distributed decision making. Figure \ref{fig:classification}
-visually summarizes this dual classification.
+\subsection{Properties and Scope of \NAME}
+\label{scope}
+
+Algorithms that fit this classification are typically platform independent.
+They analyze network messages, looking for patterns that can reveal intrusions
+and deal with the same information: other nodes in the network. To be able to
+formally describe ID algorithms and target the heterogeneous IoT, \NAME
+requires ID algorithms to expose these properties.
+
+A class of ID algorithms that does not have these properties is software
+attestation \cite{seshadri2008sake}. It enables exchanging information about
+the software that is running on a device and identifying devices with modified
+content. These algorithms are highly platform dependent and have no data in
+common with other algorithms, as they look for the effects of an intrusion, not
+the intruder itself.
+
+Figure \ref{fig:classification} visually summarizes the dual classification and
+scope of \NAME.
 
@@
 Given this classification, we now introduce a design pattern that identifies
-common aspects of these algorithms. It will allow us to define a strategy for
-optimizing the combination of these algorithms. From the classification and a
-survey of algorithms, we extract two activities common to most algorithms: the
-inspection of network traffic and the evaluation of thresholds.
-
-The former is trivial: network traffic is the carrier along which malicious
-payloads reach a device directly. Indirectly a device can also detect aberrant
-patterns of communication by other devices, as a result of a successful attack.
-
-The latter is more complex: an algorithm based on signature or misuse detection
-can maybe undeniably decide if an attack is real or not. But when dealing with
-anomaly detection this line surely isn't that crisp.
-
-Especially in distributed situations it is impossible to fully trust any
-neighboring device. In this case one has to rely on statistics and thresholds
-to decide when to drop trust in a given device.
+common functional aspects of these algorithms. It will allow us to define a
+strategy for optimizing the combination of these algorithms. From the
+classification and a survey of algorithms, we extract two activities common to
+most algorithms: the inspection of network traffic and the evaluation of
+thresholds.
+
+Network traffic is the carrier along which malicious payloads reach a device
+directly. Indirectly a device can also detect aberrant patterns of
+communication by other devices, as a result of a successful attack.
+
+Not all algorithms can decide if an attack is real or not, based on a single
+event. Signature and misuse detection can in some cases, but anomaly detection
+typically deals with some level of uncertainty. Also, in distributed situations
+it is impossible to fully trust any neighboring device. In these situations one
+has to rely on statistics and thresholds to decide when to drop trust in a
+given device.
 
 But simply checking a threshold after updating some counter, is not a good
-strategy. A simple example is that of a watchdog\cite{mishra2004intrusion},
-observing the actions of other devices. There are valid reasons why a devices
-can miss a theoretical window in which one expects activity. Simply reacting on
-such basic threshold can result in false positives. A way to overcome this is
-to decouple the detection of problems and the decision what to do about them.
+strategy. A simple example is that of a watchdog \cite{mishra2004intrusion},
+observing the actions of other devices. There are valid reasons why a device
+can miss a theoretical window in which one expects activity, i.e.\ the wireless
+network being the most predominantly unreliable component. Directly reacting on
+a basic threshold can result in false positives. A way to overcome this is to
+decouple the detection of problems and the decision what to do about them.
 
@@
 We now extend the scope of the design pattern, to include its application in
-the context of multiple algorithms. When implementing different algorithms, one
-will typically encounter a pattern such as describes in listing
-\ref{alg:id-algo-application}.
+the context of multiple algorithms. When implementing different algorithms, a
+pattern such as described in listing \ref{alg:id-algo-application} will
+typically appear.
 
@@
 
-\subsection*{Inherent Problems}
+\subsection{Inherent Problems}
 \label{pattern-problems}
 
-Given these patterns, we can now identify the problem areas that tend to result
+These patterns can now help in identifing the problem areas that tend to result
 in suboptimal code when combining multiple ID algorithms. Two issues arise from
@@
 
-  \item[Loops] by themselves are not a problem, but different loops, who
+  \item[Loops] by themselves are not a problem, but different loops, that
   perform the same operations in sequence, can incur an overhead. An example
@@
 
-  \item[Common data] is cross-cutting the different algorithms, that have to
-  store information about nodes within its own scope. This results in a lot of
+  \item[Common data] cross-cuts the different algorithms, that have to store
+  information about nodes within its own scope. This results in a lot of
   duplicated information in memory.
@@
 
-Reimplementing the algorithms manually to eliminate these issues is not a valid
-option. This development cost typically isn't a one-time cost. For example in
-case of a release of a new version of the algorithm, or simply a different
-selection of algorithms due to a changing risk analysis. Also, changing, or
-simply implementing the algorithms from scratch, holds the risk of making
-mistakes.
-
-The heterogeneity of IoT devices results in different software stacks, each
-with for example their own network API\@. This also explains why there are no
-implementations readily available to developers: there is no common platform to
-target for researchers. Even using a unified software library would not only
-require a consistent implementation of it, but would also limit the operational
-scope of the research implementation, still restricting it to a single language.
+\bigskip
+Two operational aspects are orthogonal to these problems and amplify them:
+
+\begin{LaTeXdescription}
+
+  \item[Reimplementing] the algorithms manually to eliminate these issues is
+  not a one-time cost. For example due to the release of a new version of an
+  algorithm, or simply a reselection of algorithms due to a changing risk
+  analysis. Also, changing, or simply implementing the algorithms from scratch,
+  holds the risk of making mistakes.
+
+  \item[The heterogeneity of IoT devices] results in different software stacks,
+  each with for example their own network API\@. This also explains why there
+  are no implementations readily available to developers: there is no single
+  platform to target for researchers.
+
+\end{LaTeXdescription}
 
@@
 
-To bridge the gap between the research world and the heterogeneous IoT where
-developers want to implement an IDS, while still respecting the limited
-resources they have at their disposal, we introduce \NAME, a framework
-consisting of a DSL, code generator and software library.
-
-The DSL enables researchers to formally describe their ID algorithms, without
-focusing on any specific language or platform. Given such formal descriptions
-of ID algorithms, a code generator is now able to combine these, while avoiding
-the problems arising from the patterns/problems presented in listings
-\ref{alg:id-algo-pattern} and \ref{alg:id-algo-application}. It can also target
-different languages and platforms, while reusing a software library that
-provides for example message parsing, network access wrapping,\dots
-
-Choosing for a DSL and code generator is choosing a development strategy. This
-implies choices with respect to flexibility, update cost and development
-effort. We will first position \NAME and motivate its design with respect to
-these trade offs (\ref{positioning}). Second, we will look at the principal
-design components of the DSL and how it deals with loops and common data
-(\ref{dsl-design}). Third, we discuss the structure of the code generator
-(\ref{code-generator-design}) and finally, we introduce the supporting software
-library (\ref{software-lib-design}).
+The previous sections identified loops, common data and the network as three
+important code-level problems. These problems are inherent to ID algorithms.
+But at the same time they conflict with their target environment, the IoT\@.
+\NAME provides a framework to efficiently implement an IDS for the IoT and
+offers a solution for the identified problems. It consists of a DSL, source
+code generator and software library. Figure \ref{fig:design} visualises the
+overall design.
+
+\begin{figure}[ht]
+  \centering
+  \scalebox{.85}{
+\begin{tikzpicture}[
+  node distance=7mm,
+  ,>=stealth,thick,black!50,text=black,
+  every new ->/.style={shorten >=1pt}
+]
+  \node (dsl)  [resource]                                 {DSL};
+  \node (cg)   [representation,right=of dsl,align=center] {Code\\Generator};
+  \node (code) [resource,right=of cg]                     {Code};
+
+  \draw [->] (dsl) -> (cg)   node [midway,above=25pt]              {formal};
+  \draw [->] (cg)  -> (code) node [midway,above=15pt,align=center] {platform\\ + language};
+
+\end{tikzpicture}
+}
+\caption{Overall Design}
+\label{fig:design}
+\end{figure}
+
+The DSL enables researchers to formally describe ID algorithms, without
+focusing on any specific language or platform. A code generator can combine
+such formal descriptions, avoiding the identified problems concerning loops,
+common data and network access. It can also target different languages and
+platforms, while reusing a software library that provides services including
+parsing, network access wrapping,\dots
+
+A DSL and code generator defines a development strategy. This implies choices
+with respect to flexibility, update cost and development effort. We will first
+position \NAME and motivate its design with respect to these trade offs.
+Second, we will look at the principal design components of the DSL and how it
+deals with loops and common data. Third, we discuss the structure of the code
+generator and introduce its software library, that offers a way to deal with
+the network access issue.
 
@@
 
-Three properties classify the strategies\cite{sadilek2007energy}: flexibility,
-update cost and development effort.
+Three properties can classify development strategies \cite{sadilek2007energy}:
+flexibility, update cost and development effort.
 
@@
 
-Choosing a development strategy also determines the development cost: for example
-developing for a virtual machine will be less demanding than targeting a native
-image.
-
-With \NAME we opt to start from the full native image, allowing full
-flexibility, but try to reduce both the update cost and the development cost.
-Firstly, by applying code generation and functional code fusion, we optimize
-the code organization, which has a positive effect on the code size and hence
-also on the update cost. Secondly, by enabling reuse on a functional level and
-automating the writing phase, the development cost is also lowered.
-
-\subsection{Domain Specific Language}
+\NAME starts from the full native image, allowing full flexibility, and aims to
+reduce both the update and development cost. Firstly, by applying functional
+code fusion, \NAME optimizes the code organization, which has a positive effect
+on the code size and hence also on the update cost. Secondly, by enabling reuse
+on a functional level and automating the writing phase, the development cost is
+also lowered.
+
+\subsection{\NAME as a Domain Specific Language}
 \label{dsl-design}
 
-Using a DSL, and especially an external DSL, as a foundation for any solution
-requires special care. It's important to remember that a DSL is merely a
-\emph{thin veneer over a (semantic) model}\cite{fowler2010domain}. Its sole
-purpose should be the easing of filling out the model. In the case of \NAME
-this easing focuses on avoiding constructions that result in problems as
-suggested in \ref{pattern-problems} concerning loops, common data and the use
-of the network.
-
-\subsubsection*{Loops}
+It's important to remember that a DSL is merely a \emph{thin veneer over a
+(semantic) model} \cite{fowler2010domain}. Its sole purpose should be
+populating the model. \NAME's DSL and model focuses on avoiding constructions
+that result in problems concerning loops, common data and the use of the
+network. To do this, \NAME uses an event-based approach to defining
+functionality.
+
+\subsubsection{Loops}
 
@@
 
-To illustrate this, we take the example of parsing network messages. Each
-algorithm deals with parsing in the same way: loop over the bytes in the
-payload and look for a pattern that identifies what the algorithm is looking
-for. This is in fact a technical implementation choice. The underlying
-functional goal is to react when such pattern is present in a message. The loop
-has \emph{activated} this functionality, but the real goal is to
-\emph{passively} react on an event, not \emph{actively} look for it. Event
-handling can transform constructions with loops to their functional meaning,
-allowing identification and fusion of common functionality.
+The example of parsing network messages illustrates this. Each algorithm deals
+with parsing in the same way: loop over the bytes in the payload and look for a
+pattern that identifies what the algorithm is looking for. This is in fact a
+technical implementation choice. The underlying functional goal is to react
+when such a pattern is present in a message. The loop has \emph{activated} this
+functionality, but the real goal is to \emph{passively} react on an event, not
+\emph{actively} look for it. Event handling therefore can transform
+constructions with loops to their functional meaning, allowing identification
+and fusion of common functionality.
 
@@
 
-\subsubsection*{The Nodes Domain}
+\subsubsection{The Nodes Domain}
 
@@
 
-\subsubsection*{Unified Messaging}
+\subsubsection{Unified Messaging}
 \label{dsl-unified-msg}
@@
 an API to send messages through the nodes domain. This approach gives the
-framework complete control over the messages, allowing optimized marshaling,
+framework complete control over communication, allowing optimized marshaling,
 grouping,\dots
 
-\subsubsection*{An Example}
+\subsubsection{An Example}
 
@@
 using a small but representative example of an algorithm: a
-watchdog\cite{mishra2004intrusion}. Like a heartbeat, nodes can broadcast a
+watchdog \cite{mishra2004intrusion}. Like a heartbeat, nodes can broadcast a
 message at a regular interval, allowing other nodes to monitor its
@@
 
-\subsection{Code Generator}
+\subsection{\NAME as a Code Generator}
 \label{code-generator-design}
 
-To convert the DSL-based ID algorithms into production-ready code, we propose
-the use of a code generator. Although different approaches here are applicable,
-like template-based text expansion, XSLT transformations or even using the
-CodeDom\cite{dollard2004code}, we believe that the functional transformations
-require a richer model.
-
-A semantic model\cite{fowler2010domain}, tied closely to the DSL itself, can
+To convert the DSL-based ID algorithms into production-ready code, we apply
+code generation techniques. Although different approaches here are applicable,
+like template-based text expansion, XSLT transformations or CodeDom
+\cite{dollard2004code}, we believe that the functional transformations require
+a richer model.
+
+A semantic model \cite{fowler2010domain}, tied closely to the DSL itself, can
 fulfill this role. Semantic transformations evolve the model into a
@@
 syntactic transformations an emit source code. Figure \ref{fig:code-generation}
-shows the overall procedure.
+shows the process that \NAME implements.
 
@@
 }
-\caption{Overview code generation process}
+\caption{Overview of \NAME's code generation process}
 \label{fig:code-generation}
@@
 
-\subsubsection*{Semantic Model}
+\subsubsection{Semantic Model}
 
@@
 Figure \ref{fig:meta-model} shows the central part of the semantic model as
-implemented for \NAME\@.
+implemented for \NAME.
 
@@
 }
-\caption{Partial Semantic Model for \NAME\@. Shaded classes indicate the \emph{event/scheduling} construction around functions to allow interpretation.}
+\caption{Partial Semantic Model for \NAME. Shaded classes indicate the \emph{event/scheduling} construction around functions to allow interpretation.}
 \label{fig:meta-model}
@@
 
-\subsubsection*{Functional Code Fusion (FCF)}
+\subsubsection{Functional Code Fusion (FCF)}
 
@@
 
-\subsubsection*{Code Model}
-
-The code model can is comparable to an AST for an expressive language. It
-contains constructs found in certain languages, but not in others. Therefore,
-syntactic transformations morph the unsupported constructs into comparable
+\subsubsection{Code Model}
+
+The code model is an AST for a virtual, expressive language. It contains
+constructs and ideas found in different languages and programming paradigms. No
+single existing programming language supports all features as is. Syntactic
+transformations therefore morph the unsupported constructs into comparable
 constructs supported by the targeted platform and language.
@@
 Depending on the target platform, some constructions aren't transformed into
-self-containing code. The generator uses a software library for example in case
-of parsing incoming messages and scheduling of tasks. It will look for patterns
-that indicate any of these actions and transform them into calls into the \NAME
-software library.
-
-\subsection{Software Library}
+actually corresponding code. In selected cases, the generator uses a software
+library. For example when it needs to parse incoming messages or schedule
+tasks. During the generation process, it looks for patterns that indicate any
+of these actions and transforms them into calls to the \NAME software library.
+
+\subsection{\NAME as a Software Library}
 \label{software-lib-design}
 
-In \ref{pattern-problems} the possibility to use a software library to offload
-common implementation aspects was already proposed. \ref{dsl-unified-msg}
-illustrated this approach with an example use for unified messaging: parsing of
+Section \ref{pattern-problems} introduced the option to use a software library
+to offload common implementation aspects, thus avoiding small discrepancies in
+code and suboptimal use of shared resources. Section \ref{dsl-unified-msg}
+illustrated this approach with a case for unified messaging: parsing of
 incoming messages and grouping of outgoing messages.
@@
 
-Optimizing such algorithms is not within the scope of \NAME\@. It also gives
+Optimizing such algorithms is not within the scope of \NAME. It also gives
 freedom of choice to the users of the framework to select appropriate
-implementations.
-
-\subsection{Design Summary}
-
-As a framework \NAME tries to reduce inefficiencies in execution, memory usage
-and network access. To achieve this, it tries to avoid constructions that are
-hard to identify and reorganize, such as loops and common data. The DSL
-enforces these restrictions.
-
-The DSL is platform and language independent, which allows the source code
+implementations for this standard functionality.
+
+\subsection{\NAME as a Framework}
+
+\NAME reduces inefficiencies in execution, memory usage and network access, due
+to the selection of separate ID algorithms. To achieve this, it avoids
+constructions that are hard to identify and reorganize, such as loops and
+common data. A DSL, focussing on event-handling, enforces these restrictions
+and enables formally describing ID algorithms.
+
+The DSL is also platform and language independent. This allows the source code
 generator to target different configurations without any change to the formal
@@
 
-Finally, a supporting software library allows to separate functionality that is
-out of scope for the generator.
+A supporting software library allows separation of standard functionality that
+is out of scope for the generator. Optimization of the implementation of this
+library for the specific platform and target language is possible.
 
@@
 
-In this section we discuss a prototype implementation and evaluate its
+This section discusses a prototype implementation and evaluates its
 performance. We first present the architecture of this implementation and
@@
 
-\subsection*{Architecture}
+\subsection{Architecture}
 
@@
 of dependency on the targeted platform also dictates the level of effort to
-implement changes to that platform. For example: to emit a different language
-will require a complete rewrite of the software library and a partial rewrite
-of the language description of the generator. Although this is a reusable,
-one-time effort, it is important in relation to the scalability of the
-framework itself.
-
-\subsection*{Implementation}
+implement changes to that platform. For example: to generate a different
+language will require a complete rewrite of the software library and a partial
+rewrite of the language description of the generator. Although this is a
+reusable, one-time effort, it is important in relation to the scalability of
+the framework itself.
+
+\subsection{Implementation}
 
@@
 baseline we added two intrusion detection related algorithms: a
-watchdog\cite{mishra2004intrusion}, that allows nodes to validate each other's
-presence, and a reputation-based algorithm\cite{ganeriwal2008reputation}, that
+watchdog \cite{mishra2004intrusion}, that allows nodes to validate each other's
+presence, and a reputation-based algorithm \cite{ganeriwal2008reputation}, that
 checks if a parent node is cooperative and actually forwards messages further
@@
 
-\subsection*{Experimental Results}
+\subsection{Experimental Results}
 
@@
 Section \ref{classification} introduced some ID
-algorithms\cite{ganeriwal2008reputation,mishra2004intrusion,krontiris2009cooperative}
-and presented a classification\cite{mishra2004intrusion,ioannis2007towards,alrajeh2013intrusion}
+algorithms \cite{ganeriwal2008reputation,mishra2004intrusion,krontiris2009cooperative}
+and presented a classification \cite{mishra2004intrusion,ioannis2007towards,alrajeh2013intrusion}
 based on their detection approach and interaction style.
@@
 software engineers, and to lead to a wider adoption outside the computer
-science field}\cite{sadilek2008domain,naumowicz2009prototyping}.
-
-TinyScript\cite{levis2004tinyscript} is an example of a language for wireless
+science field} \cite{sadilek2008domain,naumowicz2009prototyping}.
+
+TinyScript \cite{levis2004tinyscript} is an example of a language for wireless
 sensor networks. It's a complete programing language that aims to ease the
@@
 
-SEAL\cite{elsts2013seal} is a novel complete programming language that focuses
+SEAL \cite{elsts2013seal} is a novel complete programming language that focuses
 on novice programmers. It is a clear fit for \emph{sense-and-send} type of
@@
 used to describe patterns of events found in network
-packets\cite{sekar1999high,roesch1999snort} in a concise way.
-
-BMSL\cite{uppuluri2001experiences} is a DSL that allows the specification of
+packets \cite{sekar1999high,roesch1999snort} in a concise way.
+
+BMSL \cite{uppuluri2001experiences} is a DSL that allows the specification of
 patterns and actions. Patterns can contain events extracted from log files,
@@
 The STAT framework with its STATL
-language\cite{eckmann2002statl,vigna2003designing} are not intrusion detection
+language \cite{eckmann2002statl,vigna2003designing} are not intrusion detection
 specific, but through extensions they allow to model attack scenarios using
@@
 dealing with limited storage and processing power. Compilers implement
-different techniques to address this at the lowest level\cite{leupers2000code,
+different techniques to address this at the lowest level \cite{leupers2000code,
 marwedel2002code}.
@@
 But really preserving energy requires higher-level strategies, targeting
-architecture, radio handling, efficient software,\dots\cite{naik2001software}.
+architecture, radio handling, efficient software,\dots \cite{naik2001software}.
 Most of these are beyond the control of the compiler and require a higher level
@@
 consisting of a virtual machine (VM) and native code
-extension\cite{sadilek2007energy}, combined with energy-aware compilation,
+extension \cite{sadilek2007energy}, combined with energy-aware compilation,
 seems a promising approach. By offloading the optimization to a VM and only
@@
 code to avoid the interpretation of the description of patterns.
-I2S\cite{charitakis2003code} is an example that generates code for
-SNORT\cite{roesch1999snort} rules. To optimize this highly efficient code, it
+I2S \cite{charitakis2003code} is an example that generates code for
+SNORT \cite{roesch1999snort} rules. To optimize this highly efficient code, it
 often targets a specific platform.
@@
 
-Wireless resource constraint devices require that software respects their
-energy restrictions. Efficiently adding multiple intrusion detection algorithm
-requires fusion of these algorithms on a functional level, to reduce execution
-inefficiencies, redundant memory usage and superfluous network access.
-
-We introduced \NAME, a framework consisting of a DSL, a source code generator
-and accompanying software library. Using the DSL, intrusion detection
-algorithms can be formally described in such way that, using functional code
-fusion during source code generation, it is possible to interpret their
-functional meaning and merge common functionality and shared data. The software
+Adequate intrusion detection for IoT devices requires multiple algorithms to
+operate at the same time. Recent literature describes different algorithms, but
+hardly any are readily available for direct application when implementing an
+IDS for IoT devices. Reselection of algorithms, due to changing requirements or
+security threats, requires a flexible way to incorporate the different
+algorithms. This eliminates manual implementation and optimization.
+
+Wireless resource constraint devices, such as IoT devices, require that
+software respects their restrictions. Efficiently adding multiple intrusion
+detection algorithm requires fusion of these algorithms on a functional level,
+to reduce execution inefficiencies, redundant memory usage and superfluous
+network access.
+
+\NAME is a framework consisting of a DSL, a source code generator and
+accompanying software library. Using the DSL, intrusion detection algorithms
+can be formally described in terms of functionality. During source code
+generation, functional code fusion interprets the functional meaning of the
+algorithms and merges common functionality and shared data. The software
 library hosts common functionality, such as network message parsing, sent
-message aggregation, event scheduling,\dots as a framework for the generator to
-apply.
-
-Initial experiments confirm the theoretical positive impact on the use of
-resources. The platform independent DSL enables generation of source code
-targeting different platforms and languages, which allows high degrees of reuse
-in heterogenous environments and different configurations.
-
-Near future research will focus on the integration of applications with the
+message aggregation, event scheduling,\dots This allows the source code
+generator to focus on the optimization of the organization of the functionality.
+
+Initial experiments confirm the theoretical positive impact on resource usage.
+The platform independent DSL enables generation of source code targeting
+different platforms and languages, which allows high degrees of reuse in
+heterogenous environments and different configurations.
+
+Future research will focus on the integration of applications with the
 generated intrusion detection systems. We see opportunities for applications
@@
 
-\section*{Acknowledgements}
+\section{Acknowledgements}
 

### 2014-11-12 paper.tex
code model -> syntactic model

@@
 fulfill this role. Semantic transformations evolve the model into a
-code-oriented model, comparable to the idea of CodeDom. From there on,
+syntax-oriented model, comparable to the idea of CodeDom. From there on,
 syntactic transformations an emit source code. Figure \ref{fig:code-generation}
@@
   \node (sm)   [representation,right=of dsl,align=center] {Semantic\\Model};
-  \node (cm)   [representation,right=of sm,align=center]  {Code\\Model};
+  \node (cm)   [representation,right=of sm,align=center]  {Syntactic\\Model};
   \node (code) [resource,right=of cm]        {Code};
@@
 types. Through a series of semantic transformations, the model evolves from the
-functional to a more technical code model. This code model is comparable to a
+functional to a more technical syntactic model. This model is comparable to a
 rich abstract syntax tree (AST) with an extensive set of language constructs.
 Syntactic transformations now lower this level of expressiveness according to
-the desired platform and language. In the end the code model can be directly
-emitted to the target language.
+the desired platform and language. In the end the syntactic model can be
+directly emitted to the target language.
 
@@
 first performs semantic transformations on the semantic model itself and ends
-then transforms the semantic model into a code model, encoding the semantics
-into technical code constructs.
-
-\subsubsection{Code Model}
-
-The code model is an AST for a virtual, expressive language. It contains
+then transforms the semantic model into a syntactic model, encoding the
+semantics into technical syntax constructs.
+
+\subsubsection{Syntactic Model}
+
+The syntactic model is an AST for a virtual, expressive language. It contains
 constructs and ideas found in different languages and programming paradigms. No

### 2014-11-12 paper.tex
replaced three tables with an all-in-one

@@
 
+\begin{table*}[b]
+  \centering
+  \bgroup
+  \def\arraystretch{1.3}
+  \begin{tabular}{llrrrrrrr}
+\toprule
+              &            & base  & \multicolumn{2}{c}{watchdog} & \multicolumn{2}{c}{reputation} & \multicolumn{2}{c}{both} \\
+\midrule
+size (bytes)  & manual     & 10466 & 15530 &  +48\% & 13306 &  +27\% & 18334 &  +75\%\\
+              & generated  & 10466 & 18352 &  +75\% & 16376 &  +56\% & 20998 & +101\%\\
+\cline{2-9}
+              & difference &     0 &  2822 &  +18\% &  3070 &  +23\% &  2664 &  +15\%\\
+\hline
+packets sent  & manual     &    20 &    51 & +155\% &    32 &  +60\% &    63 & +215\%\\
+              & generated  &    20 &    49 & +145\% &    32 &  +60\% &    55 & +175\%\\
+\cline{2-9}
+              & difference &     0 &    -2 &   -4\% &     0 &   0\% &    -8 &   -13\%\\
+\hline
+bytes sent    & manual     &   476 &  1933 & +306\% &   860 &  +81\% &  2317 & +387\%\\
+              & generated  &   476 &  1897 & +299\% &   884 &  +86\% &  2161 & +354\%\\
+\cline{2-9}
+              & difference &     0 &   -36 &   -2\% &    24 &   +3\% &  -156 &   -7\%\\
+\hline
+execution
+time ($\mu$s) & manual     &    48 &    94 &  +96\% &    88 &  +83\% &   149 & +210\%\\
+              & generated  &    48 &   121 & +152\% &   121 & +152\% &   138 & +188\%\\
+\cline{2-9}
+              & difference &     0 &    27 &  +29\% &    33 &  +38\% &   -11 &   -7\%\\
+\bottomrule
+  \end{tabular}
+  \egroup
+  \caption{Results and Side-by-Side Comparison.}
+  \label{tbl:results}
+\end{table*}
+
 \subsection{Architecture}
@@
 
-Tables \ref{tbl:manual} and \ref{tbl:generated} show the collected data for the
-manual and the generated implementation. They present the base case in absolute
-values and the implementations with the ID algorithms relative to that base
-case.
-
-\begin{table}[H]
-  \centering
-  \begin{tabular}{lrrrr}
-  \hline
-      & base & watchdog & reputation & both\\
-  \hline
-  size (bytes) & 10500 & 148\% & 127\% & 175\%\\
-  frames & 20 & 255\% & 160\% & 315\%\\
-  bytes & 476 & 406\% & 181\% & 487\%\\
-  time ($\mu$s) & 48 & 196\% & 183\% & 310\%\\
-  \hline
-  \end{tabular}
-  \caption{Results for the manual implementation.}
-  \label{tbl:manual}
-\end{table}
-
-\begin{table}[H]
-  \centering
-  \begin{tabular}{lrrrr}
-  \hline
-         & base & watchdog & reputation & both\\
-  \hline
-  size (bytes) & 10496 & 175\% & 156\% & 200\%\\
-  frames & 20 & 245\% & 160\% & 275\%\\
-  bytes & 476 & 399\% & 186\% & 454\%\\
-  time ($\mu$s) & 48 & 252\% & 252\% & 288\%\\
-  \hline
-  \end{tabular}
-  \caption{Results for the generated implementation.}
-  \label{tbl:generated}
-\end{table}
+Table \ref{tbl:results} shows the collected data for the manual and the
+generated implementation, as well as a comparison in absolute values and
+relative.
 
@@
 
-Table \ref{tbl:summary} compares the two situations by subtracting the manual
-case from the generated case, showing the impact of the code generation. A
-relative value for the entire implementation with both algorithms is also
-presented.
-
-\begin{table}[H]
-  \centering
-  \begin{tabular}{lrrrrr}
-  \hline
-                & base & watchdog & reputation & both  & result \\
-  \hline
-  size (bytes)  & -4    & 2822     & 3070       & 2664  & 115\%  \\
-  frames        & 0     & -2       & 0          & -8    & 87\%   \\
-  bytes         & 0     & -36      & 24         & -156  & 93\%   \\
-  time ($\mu$s) & 0     & 27       & 33         & -11   & 93\%   \\
-  \hline
-  \end{tabular}
-  \caption{Comparison of both manual and generated implementations.}
-  \label{tbl:summary}
-\end{table}
+Table \ref{tbl:results} also compares the two situations by subtracting the
+manual case from the generated case, showing the impact of the code generation.
+A relative value for this difference, to the manual case, is also included.
 

### 2014-11-12 paper.tex
improved table layout

@@
 
+% table goes here, because a spanning table is placed on the next page, relative
+% to where it's defined.
 \begin{table*}[b]
   \centering
-  \bgroup
-  \def\arraystretch{1.3}
   \begin{tabular}{llrrrrrrr}
@@
 \midrule
+image
 size (bytes)  & manual     & 10466 & 15530 &  +48\% & 13306 &  +27\% & 18334 &  +75\%\\
               & generated  & 10466 & 18352 &  +75\% & 16376 &  +56\% & 20998 & +101\%\\
-\cline{2-9}
+\cmidrule{2-9}
               & difference &     0 &  2822 &  +18\% &  3070 &  +23\% &  2664 &  +15\%\\
-\hline
-packets sent  & manual     &    20 &    51 & +155\% &    32 &  +60\% &    63 & +215\%\\
+\midrule
+\multirow{2}{*}{\pbox{\textwidth}{total packets sent\\in a window of 90s}}
+              & manual     &    20 &    51 & +155\% &    32 &  +60\% &    63 & +215\%\\
               & generated  &    20 &    49 & +145\% &    32 &  +60\% &    55 & +175\%\\
-\cline{2-9}
+\cmidrule{2-9}
               & difference &     0 &    -2 &   -4\% &     0 &   0\% &    -8 &   -13\%\\
-\hline
-bytes sent    & manual     &   476 &  1933 & +306\% &   860 &  +81\% &  2317 & +387\%\\
+\midrule
+\multirow{2}{*}{\pbox{\textwidth}{total bytes sent\\in a window of 90s}}
+              & manual     &   476 &  1933 & +306\% &   860 &  +81\% &  2317 & +387\%\\
               & generated  &   476 &  1897 & +299\% &   884 &  +86\% &  2161 & +354\%\\
-\cline{2-9}
+\cmidrule{2-9}
               & difference &     0 &   -36 &   -2\% &    24 &   +3\% &  -156 &   -7\%\\
-\hline
-execution
-time ($\mu$s) & manual     &    48 &    94 &  +96\% &    88 &  +83\% &   149 & +210\%\\
+\midrule
+\multirow{2}{*}{\pbox{\textwidth}{event loop\\average duration ($\mu$s)}}
+              & manual     &    48 &    94 &  +96\% &    88 &  +83\% &   149 & +210\%\\
               & generated  &    48 &   121 & +152\% &   121 & +152\% &   138 & +188\%\\
-\cline{2-9}
+\cmidrule{2-9}
               & difference &     0 &    27 &  +29\% &    33 &  +38\% &   -11 &   -7\%\\
@@
   \end{tabular}
-  \egroup
   \caption{Results and Side-by-Side Comparison.}

### 2014-11-12 paper.tex
added some analytical results

@@
 is lower for both algorithms together. With roughly 3KB of additional overhead,
-or 15\% in this simple case, this impact limited.
+or 15\% in this simple case, the impact is small. Taking into account the
+available memory on the ATMega1284p, this is merely 2.3\% of the available
+resources.
 
@@
 is about 7\% faster compared to the manual case.
+
+\subsection{Analytical Results}
+
+Looking at the generated source code is another way to validate the framework.
+Reusing common data allows elimination of for example the network address
+within the scope of each individual algorithm. Analysis of the source code with
+both algorithms implemented, shows that this reduces the memory footprint for
+one node with 2 bytes from 27 bytes to 25 bytes. A ZigBee network address
+indeed consists of 16 bits. So this results in a reduction of 7.4\%. Using
+ZigBee's 64 bits unique addresses, this reduction can easily become 20.5\%.
+
+Finally comparing the amount of code versus the size of the formal description
+of IDS algorithms, can be an indication of the development effort. Implementing
+both algorithms manually results in 637 lines of clean and documented C code.
+The corresponding \NAME implementation results in 156 lines of DSL code.
 

### 2014-11-13 paper.tex
another round of improvements

@@
 behavior. But supporting an adequate set of intrusion detection algorithms
-imposes significant overhead on resource-restricted devices such as those
-targeting the Internet of Things. Fusion of the algorithms, to optimize sharing
-of resources, can alleviate this, yet, proves to be time-consuming, repetitive
-and error-prone. To address this problem we introduce \NAME, a framework for
-the development of efficient intrusion detection systems. It consists of a
-domain specific language, to formally describe intrusion detection algorithms
-on a functional level, and a code generator to produce more optimally organized
-source code. A side-by-side comparison shows that generated code reduces memory
-footprint, message passing overhead and execution time in comparison to a
-straightforward sequential implementation of the same algorithms.
+imposes significant overhead on resource-constraint devices such as those in
+the Internet of Things. Fusion of the algorithms, to optimize sharing of
+resources, can alleviate this, yet, proves to be time-consuming, repetitive and
+error-prone. To address this problem \NAME provides a framework for the
+development of efficient intrusion detection systems. It consists of a domain
+specific language, to formally describe intrusion detection algorithms on a
+functional level, and a code generator to produce more optimally organized
+source code. A side-by-side comparison shows that generated code is smaller and
+reduces message passing overhead, execution time and memory footprint in
+comparison to a straightforward sequential implementation of the same
+algorithms.
 
@@
 data to all these interconnected devices, we also open them to everyone who is
-able to break the virtual locks (i.e.\ passwords, encryption keys,\dots) that
-protect them. IoT devices typically have limited batteries, processing power
-and memory. These properties make it hard to add any additional service,
-including security.
+able to break the virtual locks that protect them (i.e.\ passwords, encryption
+keys,\dots). IoT devices typically have limited batteries, processing power and
+memory. These properties make it hard to add any additional service, including
+security.
 
@@
 attacks on flaws in services can pass unnoticed. This is where intrusion
-detection (ID) steps in: an ID probe monitors all traffic that passes through
-the firewall, looking for patterns and optionally, after detecting a malicious
-pattern, alerting the firewall \cite{denning1987intrusion}.
+detection (ID) comes into play: an intrusion detection systems (IDS) monitors
+all traffic that passes through the firewall, looking for patterns of malicious
+activity, and optionally, after detecting such pattern, alerting the firewall,
+allowing it to take corrective actions \cite{denning1987intrusion}.
 
@@
 
-In wireless networks of resource constraint devices, which make up an
+In wireless networks of resource-constraint devices, which make up a
 significant part of the IoT, it is not possible to have this single point in
 the network that can oversee all traffic \cite{mishra2004intrusion}. Every
-device has to implement its own lines of defense. But making the intrusion
-detection system (IDS) a local service on each device requires local resources,
-which are a scarce commodity for IoT devices.
+device has to implement its own lines of defense. But making the IDS a local
+service on each device requires local resources, which are a scarce commodity
+for IoT devices.
 
@@
 different opportunities to manipulate data, eavesdrop, perform a form of denial
-of service or access the network in a legitimate way. For each of these attacks
-an algorithm needs to look for patterns or anomalies. An IDS for IoT devices
-therefore requires a large number of implemented algorithms for an adequate
-coverage.
+of service or even access the network in a legitimate way. For each of these
+attacks an algorithm needs to look for patterns or anomalies. An IDS for IoT
+devices therefore requires a large number of implemented algorithms for an
+adequate coverage.
 
@@
 resources. A solution that supports the integration of multiple algorithms on
-IoT devices needs to consider the impact on their resources and avoid
-accumulating their impact.
+IoT devices needs to avoid accumulating their impact on the available resources.
 
@@
 
-The first contribution of this paper is a framework that enables the formal
-description of ID algorithms on a functional level. An automated code
-generation process produces corresponding source code, for a given platform and
-configuration. This way, the framework addresses the heterogeneous nature of
-both the IoT and ID algorithms.
+The first contribution of this paper is a framework called \NAME. It enables
+the formal description of ID algorithms on a functional level and provides a
+code generator to produces corresponding source code for a given platform and
+configuration. This way, \NAME addresses the heterogeneous nature of both the
+IoT and ID algorithms.
 
@@
 
-The second contribution introduces functional code fusion (FCF) as a generation
-paradigm to solve the problem of hard to reuse and combine ID algorithms. FCF
-identifies common data and functions, and organizes code to eliminate redundant
-iterations, tests and computations. Results show improved memory usage,
-execution time and usage of the wireless radio, reducing energy consumption.
+The second contribution introduces functional code fusion (FCF) as a source
+code generation paradigm to solve the problem of hard to reuse and combine ID
+algorithms. FCF identifies common data and functions, and organizes code to
+eliminate redundant iterations, tests and computations. Results show improved
+image size, usage of the wireless radio, execution time and memory usage.
 
@@
 analyses and classifies ID algorithms. Section \ref{pattern} looks for patterns
-in the implementation of these classes and details the ideas behind FCF\@.
-Section \ref{design} describes the design we applied to construct the
-framework, its domain specific language (DSL) and code generator. Section
-\ref{evaluation} evaluates an implementation of the framework. Section
-\ref{related} explores related work in the field of DSLs for ID and code
-generation for the IoT\@. Finally, section \ref{conclusion} summarizes our
-findings, draws conclusions and identifies topics for future work.
+in the implementation of these classes. Section \ref{problems} identifies the
+problems arising from these patterns. Section \ref{design} describes the design
+we applied to construct \NAME, its domain specific language (DSL) and code
+generator. Section \ref{evaluation} discusses and evaluates an implementation.
+Section \ref{related} explores related work in the field of DSLs and code
+generation. Finally, section \ref{conclusion} summarizes our findings, draws
+conclusions and identifies topics for future work.
 
@@
 
-Two aspects, currently missing from ID frameworks, that \NAME wants to cover
+Two aspects, currently missing from ID frameworks, that \NAME wants to address,
 are optimization of resources and support for systematic reuse. This section
@@
 are important: a group cannot make a decision without members that detect
-malicious behavior and a group-based decision is often needed because nodes in
-the network can miss certain events that could reveal intrusions. In a wireless
-network, where not all participants are within each others' radio range, nodes
-typically only have a partial view on the entire network. This prohibits them
-to observe potentially crucial events in distant parts of the network. The fact
-that resource constraint devices typically apply duty cycling to lower their
-energy consumption, further increases this problem, introducing periods of time
-where they even don't see local events.
+malicious behavior and a group-based decision is often needed because devices
+in the network can miss certain events that could reveal intrusions. In a
+wireless network, where not all participants are within each others' radio
+range, wireless devices typically only have a partial view on the entire
+network. This prohibits them to observe potentially crucial events in distant
+parts of the network. The fact that resource-constraint devices typically apply
+duty cycling to lower their energy consumption, further increases this problem,
+introducing periods of time where they even don't observe local events.
 
@@
 Because not all devices are always actively participating in the network, they
-can miss clues that would otherwise lead them to detect intrusions. It is
+can miss events that would otherwise lead them to detect intrusions. It is
 hardly impossible for a single device to detect a complex attack by itself.
@@
 alleged intrusions. We identify two ways to construct cooperative decision
-making and focus on the way participants exchange information: broadcasting and
-interactive.
+making and focus on the way participants exchange information: broadcast and
+interactive communication.
 
@@
 
-  \item[Broadcasting] can sometimes allow to implement a distributed,
+  \item[Broadcast communication] can allow implementing a distributed,
   cooperative algorithm locally. The reputation-based detection algorithm
-  introduced in \cite{ganeriwal2008reputation} illustrates this, based on the
+  introduced in \cite{ganeriwal2008reputation} illustrates this. It uses the
   same principle of a watchdog also found in \cite{mishra2004intrusion}. Using
-  this algorithm, nodes can exchange information about the reputation of other
-  nodes, solely with broadcasts. Combining this information allows them to
-  decide about their trust in a given node, based on more than their own, often
-  partial, observations alone.
-
-  \item[Interactive] communication allows algorithms to exchange information to
+  this algorithm, devices can exchange information about the reputation of
+  other devices, solely using one-way broadcast messages. Combining this
+  information allows them to decide about their trust in a given device, based
+  on more than their own, often partial, observations.
+
+  \item[Interactive communication] allows algorithms to exchange information to
   reach a consensus. In \cite{krontiris2009cooperative}, the authors first
   present a theoretical foundation to analyze cooperative algorithms. Given
-  these foundations they continue to present an algorithm consisting of 5
-  phases. It essentially implements a voting system-based on the Guy Fawkes
-  protocol \cite{anderson1998new} to identify an intruder. It enables the
-  exchange of suspected intruder information and allows distributed
+  these foundations they continue to present an algorithm consisting of five
+  phases, which essentially implement a voting system based on the Guy Fawkes
+  protocol \cite{anderson1998new}, to identify an intruder. It enables the
+  exchange of suspected intruder information and allows for distributed
   authentication of those \emph{votes}. Based on these authenticated votes, the
@@
 They analyze network messages, looking for patterns that can reveal intrusions
-and deal with the same information: other nodes in the network. To be able to
-formally describe ID algorithms and target the heterogeneous IoT, \NAME
+and all deal with the same information: other devices in the network. To be
+able to formally describe ID algorithms and target the heterogeneous IoT, \NAME
 requires ID algorithms to expose these properties.
@@
 \begin{algorithmic}[1]
-  \Require{nodes, global storage for information about nodes}
+  \Require{devices: global storage for device information}
   \Function{process\_message}{$msg$}
@@
     \EndFor
-    \Let{$nodes_x$}{value}  \Comment{optionally update node info}
-    \State \Call{send}{$nodes_y$, ``info''} \Comment{optionally exchange info} \label{alg:id-algo-pattern-send1}
+    \Let{$devices_x$}{value}  \Comment{optionally update device info}
+    \State \Call{send}{$device_y$, ``info''} \Comment{optionally exchange info} \label{alg:id-algo-pattern-send1}
   \EndFunction
@@
   \Function{do\_housekeeping}{}
-    \ForEach{$node \in nodes$} \label{alg:id-algo-pattern-loop2} \label{alg:id-algo-pattern-common-data}
-      \If{$node > \dots$} \Comment{validate recorded value}
+    \ForEach{$device \in devices$} \label{alg:id-algo-pattern-loop2} \label{alg:id-algo-pattern-common-data}
+      \If{$device > \dots$} \Comment{validate recorded value}
         \State \dots \Comment{take actions}
-        \State \Call{send}{$node$, ``info''} \Comment{communicate} \label{alg:id-algo-pattern-send2}
+        \State \Call{send}{$device$, ``info''} \Comment{communicate} \label{alg:id-algo-pattern-send2}
       \EndIf
@@
 
-We now extend the scope of the design pattern, to include its application in
-the context of multiple algorithms. When implementing different algorithms, a
-pattern such as described in listing \ref{alg:id-algo-application} will
+Applying the pattern in the context of multiple algorithms, extends its scope
+and a pattern such as described in listing \ref{alg:id-algo-application} will
 typically appear.
@@
 
-The pattern calls the different, linked implementations in a sequential way in
-two situations: first, when the device observes new messages on the network
-and, secondly, at a regular interval to allow for internal housekeeping, such
-as threshold evaluation.
-
-\subsection{Inherent Problems}
-\label{pattern-problems}
-
-These patterns can now help in identifing the problem areas that tend to result
-in suboptimal code when combining multiple ID algorithms. Two issues arise from
-the nested loops that originate from the combination of the ID algorithm
-pattern and the application pattern: loops and common data. A third issue is
-the use of the network.
+The pattern calls the different, linked implementations in sequence in two
+situations: first, when the device observes new messages on the network and,
+secondly, at a regular interval to allow for internal housekeeping, such as
+threshold evaluation.
+
+\section{Inherent Problems}
+\label{problems}
+
+These patterns can now help in identifying the problem areas that tend to
+result in suboptimal code when combining multiple ID algorithms. Besides these
+code-level problems, operational problems also hinder the implementation of an
+IDS for IoT devices.
+
+Two issues arise from the nesting of loops that originate from the combination
+of the ID algorithm pattern and the application pattern: loops and common data.
+A third issue is the use of the network.
 
@@
   \item[Common data] cross-cuts the different algorithms, that have to store
-  information about nodes within its own scope. This results in a lot of
+  information about devices within its own scope. This results in a lot of
   duplicated information in memory.
@@
 
-\bigskip
 Two operational aspects are orthogonal to these problems and amplify them:
@@
   \node (code) [resource,right=of cg]                              {Code};
-
-  \draw [->] (dsl) -> (cg)   node [midway,above=25pt]              {formal};
+  \node (lib)  [resource,below=of code,align=center,yshift=0.5cm]  {Software\\Library};
+
+  \draw [->] (dsl) -> (cg)   node [midway,above=25pt]              {intent};
   \draw [->] (cg)  -> (code) node [midway,above=15pt,align=center] {platform\\ + language};
+
+  \draw [rounded corners,->]
+        ($ (lib.west) $)
+        -- ++(-0.5cm, 0cm)
+        -- ++( 0cm,   1cm)
+        -- ($ (code.west) + (0cm,-0.22cm) $);
 
@@
 The DSL enables researchers to formally describe ID algorithms, without
-focusing on any specific language or platform. A code generator can combine
-such formal descriptions, avoiding the identified problems concerning loops,
-common data and network access. It can also target different languages and
-platforms, while reusing a software library that provides services including
-parsing, network access wrapping,\dots
+focusing on any specific language or platform. A code generator can avoid the
+identified problems concerning loops, common data and network access, as the
+formal description provides the underlying intent, not the technical
+implementation. It can also target different languages and platforms, while
+reusing a software library that provides services including parsing, network
+access wrapping,\dots
 
@@
 
-\begin{figure}[ht]
+\begin{figure}[hb]
   \centering
@@
 The \emph{ideal} situation is that were one ends up with a maximal flexibility
-and no update nor development cost.
-
-If we consider the trade off between flexibility and update cost, we can
-distinguish two extremes: \emph{virtual machines} and \emph{full native
-images}. Virtual machines allow for small updates due to their concise byte
-code, but their instruction set and predefined capabilities restrict them at
-the same time. A full native image on the other hand has unlimited flexibility,
-but requires transfer of a larger image over the network.
-
-Two intermediate alternatives are possible: \emph{modular native image} and
-\emph{reconfigurable virtual machines}. Modular native images allow changes on
-module level, limiting the update size, but the framework to support it is
-static and limits the possibilities. Reconfigurable virtual machines take this
-same approach, but apply it to virtual machines, allowing to \emph{upgrade} the
-instruction set of the virtual machine.
-
-\NAME starts from the full native image, allowing full flexibility, and aims to
-reduce both the update and development cost. Firstly, by applying functional
-code fusion, \NAME optimizes the code organization, which has a positive effect
-on the code size and hence also on the update cost. Secondly, by enabling reuse
-on a functional level and automating the writing phase, the development cost is
-also lowered.
+and no update nor development cost. \NAME starts from the full native image,
+allowing full flexibility, and aims to reduce both the update and development
+cost: first, by applying functional code fusion, \NAME optimizes the code
+organization, which has a positive effect on the code size and hence also on
+the update cost. Second, by enabling reuse on a functional level and automating
+the writing phase, the development cost is also lowered.
 
@@
 pattern that identifies what the algorithm is looking for. This is in fact a
-technical implementation choice. The underlying functional goal is to react
+technical implementation choice. The underlying functional intent is to react
 when such a pattern is present in a message. The loop has \emph{activated} this
@@
 
-All algorithms deal with information about neighboring devices. To centralize
-this information \NAME provides a unified \ttt{nodes} concept as part of its
-DSL and exposes it as an object to the implementer.
-
-Algorithms can extend nodes with their own properties. This way only one
+All algorithms deal with information about other devices in the network. To
+centralize this information \NAME provides a unified \ttt{nodes} concept as
+part of its DSL and exposes it as an object to the implementer.
+
+Algorithms can extend nodes with their own properties. This way, only one
 instance for every known device will be present in memory. Finally, nodes can
-define a scope to execute a function in. This way, nodes also offer a way to
-iterate all known devices, again without the need for a loop.
+define a scope for a function to execute in. This way, nodes also offer a way
+to i.e.\ iterate all known devices, again without the need for a loop.
 
@@
 Sending and receiving of network messages is tightly integrated with the
-previously mentioned concepts of event handling and the nodes domain. An
-underlying software framework will handle all interaction with the network,
+previously mentioned concepts of event handling and the Nodes domain. An
+underlying software library can handle all interaction with the network,
 triggering functions by means of events when parsing new messages and exposing
-an API to send messages through the nodes domain. This approach gives the
+an API to send messages through the Nodes domain. This approach gives the
 framework complete control over communication, allowing optimized marshaling,
@@
 A semantic model \cite{fowler2010domain}, tied closely to the DSL itself, can
-fulfill this role. Semantic transformations evolve the model into a
+fulfill this role. Semantic transformations evolve the model into a more
 syntax-oriented model, comparable to the idea of CodeDom. From there on,
-syntactic transformations an emit source code. Figure \ref{fig:code-generation}
-shows the process that \NAME implements.
+syntactic transformations can further evolve the model into source code. Figure
+\ref{fig:code-generation} shows the process that \NAME implements.
 
@@
   \draw [->] (dsl) -> (sm)   node [midway,above=15pt]              {parse};
-  \draw [->] (sm)  -> (cm)   node [midway,above=15pt,align=center] {semantic\\transform};
+  \draw [->] (sm)  -> (cm)   node [midway,below=10pt,align=center] {semantic\\transform}
+                             node [midway,above=15pt,align=center] {FCF};
   \draw [->] (cm)  -> (code) node [midway,above=15pt]              {emit code};
@@
         -| ($ (sm.west) + (5mm,-5mm) $)
-           node [near start, below=1pt] {type inference};
+           node [near start, below=5pt] {type inference};
 
@@
 
-After parsing the DSL, it's imported into a semantic model. The minimal type
-information present in the initial model allows inferring of the remaining
-types. Through a series of semantic transformations, the model evolves from the
-functional to a more technical syntactic model. This model is comparable to a
-rich abstract syntax tree (AST) with an extensive set of language constructs.
-Syntactic transformations now lower this level of expressiveness according to
-the desired platform and language. In the end the syntactic model can be
-directly emitted to the target language.
-
 \subsubsection{Semantic Model}
@@
 The semantic model is the core of the solution. It offers a way to describe ID
-algorithms in a way that allows semantic reinterpretation. This way,
-reorganizing the functionality of different algorithms allows for more optimal
-execution and memory usage. Hence the name \NAME of the framework: intrusion
-detection functionality organization optimization.
+algorithms in a way that allows semantic reinterpretation through functional
+code fusion. This way, reorganizing the functionality of different algorithms
+allows for more optimal execution and memory usage. Hence the name \NAME of the
+framework: intrusion detection functionality organization optimization.
 
@@
 
+One example of a semantic transformation is type inference, which supports the
+platform independence of the framework.
+
 \subsubsection{Syntactic Model}
@@
 
-Section \ref{pattern-problems} introduced the option to use a software library
-to offload common implementation aspects, thus avoiding small discrepancies in
-code and suboptimal use of shared resources. Section \ref{dsl-unified-msg}
-illustrated this approach with a case for unified messaging: parsing of
-incoming messages and grouping of outgoing messages.
-
-Other common functionality also isn't fully generated. The execution of
+A software library allows offloading of common implementation aspects, thus
+avoiding small discrepancies in code and suboptimal use of shared resources.
+Section \ref{dsl-unified-msg} illustrated this approach with a case for unified
+messaging: parsing of incoming messages and grouping of outgoing messages.
+
+Other common functionality also isn't generated as-is. The execution of
 functions based on a schedule or the implementation of higher level conceptual
@@
 
-Optimizing such algorithms is not within the scope of \NAME. It also gives
+Optimization of such algorithms is not within the scope of \NAME. It also gives
 freedom of choice to the users of the framework to select appropriate
@@
 
-\NAME reduces inefficiencies in execution, memory usage and network access, due
-to the selection of separate ID algorithms. To achieve this, it avoids
+\NAME reduces inefficiencies in network access, execution and memory usage,
+that occur due to a selection of ID algorithms. To achieve this, it avoids
 constructions that are hard to identify and reorganize, such as loops and
@@
 \label{evaluation}
-
-This section discusses a prototype implementation and evaluates its
-performance. We first present the architecture of this implementation and
-consider the impact it has from a development point of view. Next we use this
-implementation to quantify its merits using a side-by-side comparison of
-generated code versus a manual implementation.
 
@@
 
+This section discusses a prototype implementation and evaluates its
+performance. We first present the architecture of this implementation and
+consider the impact it has from a development point of view. Next we use this
+implementation to quantify its merits using a side-by-side comparison of
+generated code versus a manual implementation. We consider experimental results
+regarding the image size, usage of the network and execution time of the
+event-loop. We also take a more analytical approach and look at the handling of
+common data and the size of the source code.
+
 \subsection{Architecture}
 
-Figure \ref{fig:architecture} shows the architecture of the implementation. It
-illustrates the generation of the DSL into compilable code, using the Nodes
+Figure \ref{fig:architecture} shows the architecture of the implementation.
+
+\begin{figure}[ht]
+  \centering
+  \scalebox{.85}{
+\begin{tikzpicture}[
+  node distance=10mm and 20mm,
+  ,>=stealth,thick,black!50,text=black,
+  every new ->/.style={shorten >=1pt},
+]
+  \node (dsl)      [user,align=center]                                     {DSL};
+  \node (code)     [generated,right=of dsl]                                {code};
+  \node (image)    [generated,right=of code]                               {image};
+
+  \node (nodes)    [dom,above right=of dsl,xshift=-20mm]                   {Nodes};
+  \node (platform) [platform,below right=of dsl,align=center,xshift=-20mm] {Platform \&\\Language};
+  \node (lib)      [native,below right=of code,align=center,xshift=-20mm]  {Software\\Library};
+
+  \draw [->] (dsl)  -> (code)  node [midway,above=2pt] {generation};
+  \draw [->] (code) -> (image) node [midway,above=2pt] {compilation};
+
+  \draw [->,dashed] (nodes.south)    -> ++(0,-15mm);
+  \draw [->,dashed] (platform.north) -> ++(0,15mm);
+  \draw [->,dashed] (lib.north)      -> ++(0,15mm);
+\end{tikzpicture}
+}
+  \caption{Architecture of Prototype Implementation}
+  \label{fig:architecture}
+\end{figure}
+
+It illustrates the generation of the DSL into compilable code, using the Nodes
 domain information, a platform and language description and the native software
@@
 
-\begin{figure}[ht]
-  \centering
-  \scalebox{.85}{
-\begin{tikzpicture}[
-  node distance=10mm and 20mm,
-  ,>=stealth,thick,black!50,text=black,
-  every new ->/.style={shorten >=1pt},
-]
-  \node (dsl)      [user,align=center]                                     {DSL};
-  \node (code)     [generated,right=of dsl]                                {code};
-  \node (image)    [generated,right=of code]                               {image};
-
-  \node (nodes)    [dom,above right=of dsl,xshift=-20mm]                   {Nodes};
-  \node (platform) [platform,below right=of dsl,align=center,xshift=-20mm] {Platform \&\\Language};
-  \node (lib)      [native,below right=of code,align=center,xshift=-20mm]  {Software\\Library};
-
-  \draw [->] (dsl)  -> (code)  node [midway,above=2pt] {generation};
-  \draw [->] (code) -> (image) node [midway,above=2pt] {compilation};
-
-  \draw [->,dashed] (nodes.south)    -> ++(0,-15mm);
-  \draw [->,dashed] (platform.north) -> ++(0,15mm);
-  \draw [->,dashed] (lib.north)      -> ++(0,15mm);
-\end{tikzpicture}
-}
-  \caption{Architecture of Prototype Implementation}
-  \label{fig:architecture}
-\end{figure}
-
 This hierarchy is important from a development effort point of view. The level
@@
 To evaluate the prototype we chose to use a system based on the Atmel
-ATMEGA1284p micro-controller \cite{datasheet:atmega1284p} and the Digi XBee S2
-ZigBee module \cite{manual:xbee}. A basic library wrapped the technical calls
-to the hardware components in more easy to use functions.
+ATMEGA1284p micro-controller and the Digi XBee S2 ZigBee module. A basic
+library wrapped the technical calls to the hardware components in more easy to
+use functions.
 
@@
 We evaluated three criteria: the image size of the resulting compiled code, the
-network usage in number of frames and bytes and the time required to perform
+network usage in number of packets and bytes and the time required to perform
 once cycle of the event loop. To do this, we collected these metrics for four
@@
 
-The manual implementation implemented both algorithms as standalone modules and
+The manual implementation realized both algorithms as standalone modules and
 sequentially called them from the base-application's event loop. Given the
@@
 Thirdly, when turning to the goal to reduce the network usage, we see that both
-the number of frames as well as the amount of bytes sent have dropped by 13 and
-7\%.
-
-Finally we see that both algorithms by itself add an equal amount of time to
-the processing of a single event loop cycle, but when combined, the event loop
-is about 7\% faster compared to the manual case.
+the number of packets as well as the amount of bytes sent have dropped by 13
+and 7\%.
+
+Finally, focusing on the execution time, both algorithms by themselves add an
+equal amount of time to the processing of a single event loop cycle. But when
+combined, the event loop is about 7\% faster compared to the manual case.
 
@@
 Looking at the generated source code is another way to validate the framework.
-Reusing common data allows elimination of for example the network address
-within the scope of each individual algorithm. Analysis of the source code with
-both algorithms implemented, shows that this reduces the memory footprint for
-one node with 2 bytes from 27 bytes to 25 bytes. A ZigBee network address
-indeed consists of 16 bits. So this results in a reduction of 7.4\%. Using
-ZigBee's 64 bits unique addresses, this reduction can easily become 20.5\%.
-
-Finally comparing the amount of code versus the size of the formal description
-of IDS algorithms, can be an indication of the development effort. Implementing
-both algorithms manually results in 637 lines of clean and documented C code.
-The corresponding \NAME implementation results in 156 lines of DSL code.
+A simple metric is comparing the amount of code versus the size of the formal
+description of IDS algorithms. This can be an indication of the development
+effort and complexity. In this experiment, implementing both algorithms
+manually results in 637 lines of C code. The corresponding \NAME implementation
+results in 156 lines of DSL code.
+
+We can also look at data structures. Sharing data eliminates redundant
+information, such as a network address, from the scope of each individual
+algorithm. Analysis of the source code with both algorithms implemented, shows
+that this reduces the memory footprint for one node entry by 2 bytes, from 27
+bytes to 25 bytes. A ZigBee network address indeed consists of 16 bits. So this
+results in a reduction of 7.4\%. Using ZigBee's 64 bits unique addresses, this
+reduction can easily become 20.5\%.
 
@@
 
-Section \ref{classification} introduced some ID
-algorithms \cite{ganeriwal2008reputation,mishra2004intrusion,krontiris2009cooperative}
-and presented a classification \cite{mishra2004intrusion,ioannis2007towards,alrajeh2013intrusion}
-based on their detection approach and interaction style.
+Section \ref{classification} introduced some ID algorithms
+\cite{ganeriwal2008reputation,mishra2004intrusion,krontiris2009cooperative} and
+presented a classification
+\cite{mishra2004intrusion,ioannis2007towards,alrajeh2013intrusion} based on
+their detection approach and interaction style.
 
 This section focuses on related work in the field of domain specific languages
-and code generation, both in relation to wireless networks of resource
-constraint devices. These two components are the core of the \NAME framework
-and the solution it proposes. We also look at the more general scope: the
-application in other disciplines and their relevance in the scope of intrusion
-detection in the IoT and the proposed \NAME framework.
+and code generation, both in relation to wireless networks of
+resource-constraint devices. These two components are the core of the \NAME
+framework and the solution it proposes. We also look at their application in
+other disciplines and how this can relate to our context.
 
@@
 
-Within the scope of wireless networks of resource constraint devices, we see
+Within the scope of wireless networks of resource-constraint devices, we see
 the introduction of DSLs \enquote{to shorten the development cycles
@@
 software engineers, and to lead to a wider adoption outside the computer
-science field} \cite{sadilek2008domain,naumowicz2009prototyping}.
+science field} \cite{sadilek2008domain}.
 
@@
 
-The STAT framework with its STATL
-language \cite{eckmann2002statl,vigna2003designing} are not intrusion detection
+The STAT framework with its STATL language
+\cite{eckmann2002statl,vigna2003designing} are not intrusion detection
 specific, but through extensions they allow to model attack scenarios using
-state machines. Our event-driven approach shares this same ideas.
+state machines. Our event-driven approach shares these same ideas.
 
@@
 dealing with limited storage and processing power. Compilers implement
-different techniques to address this at the lowest level \cite{leupers2000code,
-marwedel2002code}.
+different techniques to address this at the lowest level
+\cite{marwedel2002code}.
 
@@
 
-DSLs are a solution here, but other paradigms are possible. A mixed framework
-consisting of a virtual machine (VM) and native code
-extension \cite{sadilek2007energy}, combined with energy-aware compilation,
-seems a promising approach. By offloading the optimization to a VM and only
-compiling parts and not the whole, the solution allows updating at a granular
-scale, thus reducing the update cost, while maintaining flexibility and
-lowering the development effort. As explained in \ref{positioning}, \NAME
-focuses on the reduction of the development effort and update cost, while
-preserving flexibility.
+DSLs offer a solution here, but other paradigms are possible. A mixed framework
+consisting of a virtual machine (VM) and native code extension
+\cite{sadilek2007energy}, combined with energy-aware compilation, seems a
+promising approach. By offloading the optimization to a VM and only compiling
+parts and not the whole, the solution allows updating at a granular scale, thus
+reducing the update cost, while maintaining flexibility and lowering the
+development effort. As explained in \ref{positioning}, \NAME focuses on the
+reduction of the development effort and update cost, while preserving maximal
+flexibility.
 
@@
 
-Wireless resource constraint devices, such as IoT devices, require that
-software respects their restrictions. Efficiently adding multiple intrusion
-detection algorithm requires fusion of these algorithms on a functional level,
-to reduce execution inefficiencies, redundant memory usage and superfluous
-network access.
+Wireless resource-constraint devices, such as IoT devices, require that
+software respects their restrictions. Adding multiple intrusion detection
+algorithm requires fusion of these algorithms on a functional level, to reduce
+execution inefficiencies, redundant memory usage and superfluous network access.
 
@@
 who detect intrusion, to actively and dynamically respond and avoid malicious
-payloads to harm them.
-
-A continued effort is targeting the reduction of non-platform independent parts
-in the implementation. This way it is possible to add more platforms and
-languages with less effort.
+payloads.
+
+A continued effort is targeting the reduction of platform-dependent parts in
+the implementation. This way it is possible to add more platforms and languages
+with less effort.
 

### 2014-11-21 paper.tex
fixed more typos and bad sentences

@@
 attacks on flaws in services can pass unnoticed. This is where intrusion
-detection (ID) comes into play: an intrusion detection systems (IDS) monitors
+detection (ID) comes into play: an intrusion detection system (IDS) monitors
 all traffic that passes through the firewall, looking for patterns of malicious
@@
   \node (decision)              [align=center,right=of detection] {Decision\\Making}
-    child {node (broadcasting)  [align=center] {Broadcasting}}
+    child {node (broadcasting)  [align=center] {Broadcast}}
     child {node (interactive)   [align=center] {Interactive}};
@@
     \Let{$devices_x$}{value}  \Comment{optionally update device info}
-    \State \Call{send}{$device_y$, ``info''} \Comment{optionally exchange info} \label{alg:id-algo-pattern-send1}
+    \State \Call{send}{$devices_y$, ``info''} \Comment{optionally exchange info} \label{alg:id-algo-pattern-send1}
   \EndFunction
@@
 \label{positioning}
-
-Figure \ref{fig:positioning} shows a classification of possible development
-strategies for projects involving wireless devices.
 
@@
 
-Three properties can classify development strategies \cite{sadilek2007energy}:
-flexibility, update cost and development effort.
+Figure \ref{fig:positioning} shows a classification of possible development
+strategies for projects involving wireless devices. Three properties can
+classify development strategies \cite{sadilek2007energy}: flexibility, update
+cost and development effort.
 
@@
 
-To convert the DSL-based ID algorithms into production-ready code, we apply
-code generation techniques. Although different approaches here are applicable,
-like template-based text expansion, XSLT transformations or CodeDom
+To convert the DSL-based ID algorithms into production-ready code, \NAME
+applies code generation techniques. Although different approaches here are
+applicable, like template-based text expansion, XSLT transformations or CodeDom
 \cite{dollard2004code}, we believe that the functional transformations require
@@
 framework: intrusion detection functionality organization optimization.
-
-Figure \ref{fig:meta-model} shows the central part of the semantic model as
-implemented for \NAME.
 
@@
 
-The concept driving this model is the structure starting at the \emph{Model} up
-to the \emph{Execution Strategy}. This structure encompasses the
-\emph{Function}s and underlying \emph{Statement}s and allows for interpretation
-and fusion of the algorithms.
+Figure \ref{fig:meta-model} shows the central part of the semantic model as
+implemented for \NAME. The concept driving this model is the structure starting
+at the \emph{Model} up to the \emph{Execution Strategy}. This structure
+encompasses the \emph{Function}s and underlying \emph{Statement}s and allows
+for interpretation and fusion of the algorithms.
 
@@
 them, the code generator is able to identify the same actions across different
-algorithms. This way it is possible to extract the parsing of incoming
-messages, group actions on the same sets of data or scheduled function
-executions.
+algorithms. This way it is possible to extract the parsing of incoming messages
+and group actions on the same sets of data or scheduled function executions.
 
@@
 transformations to change these patterns to a more optimized organization. It
-first performs semantic transformations on the semantic model itself and ends
-then transforms the semantic model into a syntactic model, encoding the
+first performs semantic transformations on the semantic model itself. A second
+phase transforms the semantic model into a syntactic model, encoding the
 semantics into technical syntax constructs.
@@
 
-Other common functionality also isn't generated as-is. The execution of
+Other common functionality also isn't generated as is. The execution of
 functions based on a schedule or the implementation of higher level conceptual
@@
 
-To evaluate the prototype we chose to use a system based on the Atmel
-ATMEGA1284p micro-controller and the Digi XBee S2 ZigBee module. A basic
-library wrapped the technical calls to the hardware components in more easy to
-use functions.
+To evaluate the prototype we used a system based on the Atmel ATMEGA1284p
+micro-controller and the Digi XBee S2 ZigBee module. A basic library wrapped
+the technical calls to the hardware components in more easy to use functions.
 
 Relying on basic hardware and software allows us to validate that the generator
-is capable of generating code even for the most basic environment available.
-Any step up, both in hardware or software, would offer better and higher
-abstractions that would make it easier to generate code for that situation. It
-shows that the requirements of the generator towards the platform are minimal,
-don't rely on advanced frameworks or operating systems and therefore are
-applicable in any environment, targeting any platform.
+is capable of generating code even for the most basic environment. Any step up,
+both in hardware or software, would offer better and higher abstractions that
+would make it easier to generate code for that situation. It shows that the
+requirements of the generator towards the platform are minimal, don't rely on
+advanced frameworks or operating systems and therefore are applicable in any
+environment, targeting any platform.
 
@@
 a cumulative effect. The impact of both algorithms is simply the sum of the
-impact of each algorithm by itself. In case of the execution time of one cycle
-of the event loop, the total is even a bit higher.
+impact of each algorithm by itself. The execution time of one cycle of the
+event loop with both algorithms enabled is even higher than simply the sum.
 
@@
 reducing the update cost, while maintaining flexibility and lowering the
-development effort. As explained in \ref{positioning}, \NAME focuses on the
-reduction of the development effort and update cost, while preserving maximal
-flexibility.
+development effort. As explained in section \ref{positioning}, \NAME focuses on
+the reduction of the development effort and update cost, while preserving
+maximal flexibility.
 

### 2014-11-21 paper.tex
processed Danny’s remarks on abstract

@@
 of defense. It looks for patterns \& anomalies that can indicate malicious
-behavior. But supporting an adequate set of intrusion detection algorithms
-imposes significant overhead on resource-constraint devices such as those in
-the Internet of Things. Fusion of the algorithms, to optimize sharing of
-resources, can alleviate this, yet, proves to be time-consuming, repetitive and
-error-prone. To address this problem \NAME provides a framework for the
+behavior. However, supporting an adequate set of intrusion detection algorithms
+imposes significant overhead on resource-constrained devices such as those in
+the Internet of Things. Fusion of the algorithms reduces resource consumption
+by optimizing resource sharing, yet, proves to be time-consuming, repetitive
+and error-prone. To address this problem we propose \NAME, a framework for the
 development of efficient intrusion detection systems. It consists of a domain
@@
 functional level, and a code generator to produce more optimally organized
-source code. A side-by-side comparison shows that generated code is smaller and
-reduces message passing overhead, execution time and memory footprint in
-comparison to a straightforward sequential implementation of the same
-algorithms.
+source code. A side-by-side comparison shows that \NAME-generated code is
+smaller and reduces message passing overhead, execution time and memory
+footprint in comparison to a straightforward sequential implementation of the
+same algorithms.
 

### 2014-11-21 paper.tex
processed Danny’s remarks on body

@@
 
-The internet of things (IoT) holds a great potential to positively influence
-our daily work and life through domotics, assisted living, e-health and
-enhanced learning \cite{atzori2010internet}. But with this potential the IoT
-also presents a significant threat: by opening our smart homes and personal
-data to all these interconnected devices, we also open them to everyone who is
-able to break the virtual locks that protect them (i.e.\ passwords, encryption
-keys,\dots). IoT devices typically have limited batteries, processing power and
+The internet of things (IoT) holds great potential to positively influence our
+daily work and life through domotics, assisted living, e-health and enhanced
+learning \cite{atzori2010internet}. But with this potential the IoT also
+presents a significant threat: by opening our smart homes and personal data to
+all of these interconnected devices, we also open them to everyone who is able
+to break the virtual locks that protect them (i.e.\ passwords, encryption
+keys). IoT devices typically have limited batteries, processing power and
 memory. These properties make it hard to add any additional service, including
@@
 all traffic that passes through the firewall, looking for patterns of malicious
-activity, and optionally, after detecting such pattern, alerting the firewall,
+activity, and optionally, after detecting such pattern, alerts the firewall,
 allowing it to take corrective actions \cite{denning1987intrusion}.
@@
 
-In wireless networks of resource-constraint devices, which make up a
-significant part of the IoT, it is not possible to have this single point in
-the network that can oversee all traffic \cite{mishra2004intrusion}. Every
-device has to implement its own lines of defense. But making the IDS a local
-service on each device requires local resources, which are a scarce commodity
-for IoT devices.
+In wireless networks of resource-constrained devices, which make up a
+significant part of the IoT, it is not possible for a single point in the
+network to oversee all traffic \cite{mishra2004intrusion}. Every device has to
+implement its own lines of defense. But making the IDS a local service on each
+device requires local resources, which are a scarce commodity for IoT devices.
 
 Attackers have access to a large set of possible attack vectors
-\cite{aschenbruck2012security}, ranging from the wireless network, over the
+\cite{aschenbruck2012security}, ranging from the wireless network, through the
 medium access control and routing, up to the device itself. Each level presents
@@
 
-Current developments in ID focus on programming frameworks to structure the
+Current developments in ID focus on programming frameworks that structure the
 implementation of algorithms \cite{valero2012di} and offer the required basic
-functional components to implement them \cite{krontiris2008lidea}. The first
-aspect that is missing, is that they don't offer a way to optimize the usage of
-resources. A solution that supports the integration of multiple algorithms on
-IoT devices needs to avoid accumulating their impact on the available resources.
+functional components to implement them \cite{krontiris2008lidea}. A key
+shortcoming of prior ID frameworks, is that they don't offer a way to optimize
+the usage of resources. A solution that supports the integration of multiple
+algorithms on IoT devices needs to avoid accumulating their impact on the
+available resources.
 
 A second omission is support for systematic reuse to create different
-configurations for different devices, both in hard- and software. Addressing
-this in a transparent and automated way is important for any kind of
-development in this area.
+configurations heterogeneous devices. Addressing this in a transparent and
+automated way offers the potential to significantly reduce the effort of
+developing an IDS.
 
@@
 eliminate redundant iterations, tests and computations. Results show improved
-image size, usage of the wireless radio, execution time and memory usage.
+image size, reduced usage of the wireless radio, reduced execution time and
+lower memory usage.
 
@@
 range, wireless devices typically only have a partial view on the entire
-network. This prohibits them to observe potentially crucial events in distant
-parts of the network. The fact that resource-constraint devices typically apply
-duty cycling to lower their energy consumption, further increases this problem,
-introducing periods of time where they even don't observe local events.
+network. This prohibits them from observing potentially crucial events in
+distant parts of the network. The fact that resource-constrained devices
+typically apply duty cycling to lower their energy consumption, further
+increases this problem, introducing periods of time where they cannot observe
+local events.
 
@@
   \item[Broadcast communication] can allow implementing a distributed,
-  cooperative algorithm locally. The reputation-based detection algorithm
-  introduced in \cite{ganeriwal2008reputation} illustrates this. It uses the
-  same principle of a watchdog also found in \cite{mishra2004intrusion}. Using
-  this algorithm, devices can exchange information about the reputation of
-  other devices, solely using one-way broadcast messages. Combining this
-  information allows them to decide about their trust in a given device, based
-  on more than their own, often partial, observations.
+  cooperative algorithm locally: devices can exchange information about the
+  reputation of other devices, solely using one-way broadcast messages.
+  Combining this information allows them to decide about their trust in a given
+  device, based on more than their own, often partial, observations. The
+  reputation-based detection algorithm introduced in
+  \cite{ganeriwal2008reputation} illustrates this. It uses the same principle
+  of a watchdog also found in \cite{mishra2004intrusion}.
 
@@
 \label{scope}
-
-Algorithms that fit this classification are typically platform independent.
-They analyze network messages, looking for patterns that can reveal intrusions
-and all deal with the same information: other devices in the network. To be
-able to formally describe ID algorithms and target the heterogeneous IoT, \NAME
-requires ID algorithms to expose these properties.
-
-A class of ID algorithms that does not have these properties is software
-attestation \cite{seshadri2008sake}. It enables exchanging information about
-the software that is running on a device and identifying devices with modified
-content. These algorithms are highly platform dependent and have no data in
-common with other algorithms, as they look for the effects of an intrusion, not
-the intruder itself.
 
@@
 
+Algorithms that fit this classification are typically platform independent.
+They analyze network messages, looking for patterns that can reveal intrusions
+and all deal with the same information: other devices in the network. To be
+able to formally describe ID algorithms and target the heterogeneous IoT, \NAME
+requires ID algorithms to expose these properties.
+
+A class of ID algorithms that does not have these properties is software
+attestation \cite{seshadri2008sake}. It enables exchanging information about
+the software that is running on a device and identifying devices with modified
+content. These algorithms are highly platform dependent and have no data in
+common with other algorithms, as they look for the effects of an intrusion, not
+the intruder itself.
+
 \section{A Design Pattern for ID Algorithms}
@@
 
-Given this classification, we now introduce a design pattern that identifies
-common functional aspects of these algorithms. It will allow us to define a
-strategy for optimizing the combination of these algorithms. From the
-classification and a survey of algorithms, we extract two activities common to
-most algorithms: the inspection of network traffic and the evaluation of
-thresholds.
+Based upon the classification presented in section \ref{classification}, we now
+introduce a design pattern that identifies common functional aspects of ID
+algorithms. It will allow us to define a strategy for optimizing the
+combination of these algorithms. We extract two activities common to most
+algorithms: the inspection of network traffic and the evaluation of thresholds.
 
@@
 
-But simply checking a threshold after updating some counter, is not a good
-strategy. A simple example is that of a watchdog \cite{mishra2004intrusion},
-observing the actions of other devices. There are valid reasons why a device
-can miss a theoretical window in which one expects activity, i.e.\ the wireless
-network being the most predominantly unreliable component. Directly reacting on
-a basic threshold can result in false positives. A way to overcome this is to
-decouple the detection of problems and the decision what to do about them.
+Updating such statistics and evaluating the corresponding thresholds should be
+distinct operations. The example of a watchdog \cite{mishra2004intrusion}
+illustrates this: a device monitors the actions of other devices and checks if
+they correctly forward messages to distant nodes in the network. Missing
+forwards can indicate a malicious intension to silently drop messages and
+obstruct the normal operation of the network. There are valid reasons why a
+device can miss the theoretical window in which one expects activity, i.e.\ the
+wireless network being the most predominantly unreliable component. Directly
+reacting upon a basic threshold can result in false positives. A way to
+overcome this is to decouple the detection of problems and the decision what to
+do about them.
 
@@
 \begin{algorithmic}[1]
-  \Require{devices: global storage for device information}
+  \Require{devices: global variable for device information}
   \Function{process\_message}{$msg$}
@@
   \item[Loops] by themselves are not a problem, but different loops, that
-  perform the same operations in sequence, can incur an overhead. An example
-  from the ID algorithm pattern is the handling of messages: each algorithm
-  parses each observed message on the network, processing the same byte stream,
+  perform the same operations in sequence, can incur overhead. An example from
+  the ID algorithm pattern is the handling of messages: each algorithm parses
+  each observed message on the network, processing the same byte stream,
   performing comparable operations to find similar patterns.
@@
   each with for example their own network API\@. This also explains why there
-  are no implementations readily available to developers: there is no single
+  are few implementations readily available to developers: there is no single
   platform to target for researchers.
@@
 an API to send messages through the Nodes domain. This approach gives the
-framework complete control over communication, allowing optimized marshaling,
-grouping,\dots
+framework complete control over communication and for example allows for
+optimized marshaling and grouping of messages.
 
@@
 Listing \ref{lst:watchdog} illustrates some of the features mentioned above,
-using a small but representative example of an algorithm: a
-watchdog \cite{mishra2004intrusion}. Like a heartbeat, nodes can broadcast a
-message at a regular interval, allowing other nodes to monitor its
-availability. Missing heartbeats could indicate problems, such as physical
-capturing of a device.
+using a small but representative example of an algorithm: a watchdog
+\cite{mishra2004intrusion}. Like a heartbeat, nodes can broadcast a message at
+a regular interval, allowing other nodes to monitor its availability. Missing
+heartbeats could indicate problems, such as physical capture of a device.
 
@@
 Relying on basic hardware and software allows us to validate that the generator
-is capable of generating code even for the most basic environment. Any step up,
-both in hardware or software, would offer better and higher abstractions that
-would make it easier to generate code for that situation. It shows that the
-requirements of the generator towards the platform are minimal, don't rely on
-advanced frameworks or operating systems and therefore are applicable in any
-environment, targeting any platform.
+is capable of generating code even for the most basic environment. It shows
+that the requirements of the generator towards the platform are minimal, don't
+rely on advanced frameworks or operating systems and therefore are applicable
+in any environment, targeting any platform.
 
@@
 impact to 188\%. We see the effect of the generic handling of common
-functionality, like message parsing, event scheduling,\dots It comes with an
+functionality, like message parsing and event scheduling. It comes with an
 overhead for a single algorithm, but pays for itself when adding more
@@
 or 15\% in this simple case, the impact is small. Taking into account the
-available memory on the ATMega1284p, this is merely 2.3\% of the available
+available memory on the ATMega1284p, this is only 2.3\% of the available
 resources.
@@
 Thirdly, when turning to the goal to reduce the network usage, we see that both
-the number of packets as well as the amount of bytes sent have dropped by 13
-and 7\%.
+the number of packets as well as the amount of bytes sent have dropped by 13\%
+and 7\% respectively.
 
@@
 on novice programmers. It is a clear fit for \emph{sense-and-send} type of
-applications, and doesn't pretend to be the best choice for any type of
-application. It does however share the idea of event to overcome the technical
+applications. It does however share the idea of events to overcome the technical
 nature of loops.
@@
 
-Finally, applying code generation to IDS, typically tries to generate native
-code to avoid the interpretation of the description of patterns.
-I2S \cite{charitakis2003code} is an example that generates code for
-SNORT \cite{roesch1999snort} rules. To optimize this highly efficient code, it
-often targets a specific platform.
+Applying code generation to the context of ID, often focusses on the generation
+of native code to avoid the interpretation of descriptions of patterns to look
+for. I2S \cite{charitakis2003code} is an example that generates code for SNORT
+\cite{roesch1999snort} rules. To optimize this highly efficient code, it often
+targets a specific platform.
 
@@
 operate at the same time. Recent literature describes different algorithms, but
-hardly any are readily available for direct application when implementing an
-IDS for IoT devices. Reselection of algorithms, due to changing requirements or
+few are readily available for direct application when implementing an IDS for
+IoT devices. Reselection of algorithms, due to changing requirements or
 security threats, requires a flexible way to incorporate the different
@@
 software respects their restrictions. Adding multiple intrusion detection
-algorithm requires fusion of these algorithms on a functional level, to reduce
+algorithms requires fusion of these algorithms on a functional level, to reduce
 execution inefficiencies, redundant memory usage and superfluous network access.
@@
 library hosts common functionality, such as network message parsing, sent
-message aggregation, event scheduling,\dots This allows the source code
-generator to focus on the optimization of the organization of the functionality.
+message aggregation and event scheduling. This allows the source code generator
+to focus on the optimization of the organization of the functionality.
 

### 2014-11-23 paper.tex
incorporated comments of review meeting

@@
 to break the virtual locks that protect them (i.e.\ passwords, encryption
-keys). IoT devices typically have limited batteries, processing power and
-memory. These properties make it hard to add any additional service, including
-security.
+keys). Implementation of security measures by itself is hindered by the fact
+that IoT devices typically have limited batteries, processing power and memory.
 
@@
 different opportunities to manipulate data, eavesdrop, perform a form of denial
-of service or even access the network in a legitimate way. For each of these
-attacks an algorithm needs to look for patterns or anomalies. An IDS for IoT
-devices therefore requires a large number of implemented algorithms for an
-adequate coverage.
+of service. For each of these attacks an algorithm needs to look for patterns
+or anomalies. An IDS for IoT devices therefore requires a large number of
+implemented algorithms for an adequate coverage.
 
@@
 algorithms on IoT devices needs to avoid accumulating their impact on the
-available resources.
-
-A second omission is support for systematic reuse to create different
-configurations heterogeneous devices. Addressing this in a transparent and
-automated way offers the potential to significantly reduce the effort of
-developing an IDS.
+available resources. A second omission is support for systematic reuse to
+create different configurations heterogeneous devices. Addressing this in a
+transparent and automated way offers the potential to significantly reduce the
+effort of developing an IDS.
 
@@
 common themes do appear. According to \cite{mishra2004intrusion} and
-\cite{ioannis2007towards} there are three major categories: anomaly detection,
-signature or misuse detection, and specification-based detection. In
-\cite{alrajeh2013intrusion} the authors agree with this typology and add the
-notion of hybrid intrusion detection systems and cross layer intrusion
-detection systems as recent advances in research. Although these additional
-categories seem promising, the authors admit that the impact on
+\cite{ioannis2007towards} there are three major categories: 1) anomaly
+detection, 2) signature or misuse detection, and 3) specification-based
+detection. In \cite{alrajeh2013intrusion} the authors agree with this typology
+and add the notion of hybrid intrusion detection systems and cross layer
+intrusion detection systems as recent advances in research. Although these
+additional categories seem promising, the authors admit that the impact on
 resource-constrained devices will be too much. We therefore focus on the three
@@
   \item[Specification-based detection] uses constraints that describe the
-  correct behavior of the system. Although that this approach can detect
-  previously unknown attacks, it can sometimes not detect specific attacks that
-  anomaly or misuse detection can.
+  correct behavior of the system. Although this approach can detect previously
+  unknown attacks, it can sometimes not detect specific attacks that anomaly or
+  misuse detection can.
 
@@
 Algorithms that fit this classification are typically platform independent.
-They analyze network messages, looking for patterns that can reveal intrusions
-and all deal with the same information: other devices in the network. To be
-able to formally describe ID algorithms and target the heterogeneous IoT, \NAME
-requires ID algorithms to expose these properties.
+They look for patterns in network messages that can reveal intrusions and, in
+doing so, all deal with the same information: other devices in the network. To
+be able to formally describe ID algorithms and target the heterogeneous IoT,
+\NAME requires ID algorithms to expose these properties.
 
@@
   \item[Reimplementing] the algorithms manually to eliminate these issues is
-  not a one-time cost. For example due to the release of a new version of an
-  algorithm, or simply a reselection of algorithms due to a changing risk
-  analysis. Also, changing, or simply implementing the algorithms from scratch,
-  holds the risk of making mistakes.
+  not a one-time cost. A release of a new version of an algorithm, or simply a
+  reselection of algorithms, due to a changing risk analysis, can trigger it.
+  Also, changing, or simply implementing the algorithms from scratch, holds the
+  risk of making mistakes.
 
@@
 
+\begin{figure*}[b]
+  \centering
+  \scalebox{1}{
+\begin{tikzpicture}[
+  node distance=9mm,
+  ,>=stealth,very thick,black!50,text=black,
+  every new ->/.style={shorten >=1pt}
+]
+  \node (dsl)   [user]                                      {DSL};
+  \node (cg)    [representation,right=of dsl,align=center]  {Code\\Generator};
+  \node (code)  [generated,right=of cg]                     {code};
+  \node (comp)  [representation,right=of code]              {Compiler};
+  \node (image) [generated,right=of comp]                   {image};
+
+  \node (nodes) [dom,above=of cg, align=center]             {Nodes\\Domain};
+  \node (plat)  [platform,below=of cg,align=center]         {Platform \&\\Language};          
+  \node (lib)   [native,below=of comp,align=center]         {Software\\Library};
+
+  \draw [->] (dsl)   -> (cg);
+  \draw [->] (nodes) -> (cg);
+  \draw [->] (plat)  -> (cg);
+  \draw [->] (cg)    -> (code);
+  \draw [->] (code)  -> (comp);
+  \draw [->] (lib)   -> (comp);
+  \draw [->] (comp)  -> (image);
+
+\end{tikzpicture}
+}
+
+\caption{The high level design of \NAME, showing the resources, processes and
+artifacts. A grayscale background represents the level of platform dependency,
+with darker shades indicating a higher dependency. Both the source code and the
+installable image are artifacts.}
+
+\label{fig:design}
+\end{figure*}
+
 \section{Design}
@@
 \NAME provides a framework to efficiently implement an IDS for the IoT and
-offers a solution for the identified problems. It consists of a DSL, source
-code generator and software library. Figure \ref{fig:design} visualises the
-overall design.
-
-\begin{figure}[ht]
-  \centering
-  \scalebox{.85}{
-\begin{tikzpicture}[
-  node distance=7mm,
-  ,>=stealth,thick,black!50,text=black,
-  every new ->/.style={shorten >=1pt}
-]
-  \node (dsl)  [resource]                                          {DSL};
-  \node (cg)   [representation,right=of dsl,align=center]          {Code\\Generator};
-  \node (code) [resource,right=of cg]                              {Code};
-  \node (lib)  [resource,below=of code,align=center,yshift=0.5cm]  {Software\\Library};
-
-  \draw [->] (dsl) -> (cg)   node [midway,above=25pt]              {intent};
-  \draw [->] (cg)  -> (code) node [midway,above=15pt,align=center] {platform\\ + language};
-
-  \draw [rounded corners,->]
-        ($ (lib.west) $)
-        -- ++(-0.5cm, 0cm)
-        -- ++( 0cm,   1cm)
-        -- ($ (code.west) + (0cm,-0.22cm) $);
-
-\end{tikzpicture}
-}
-\caption{Overall Design}
-\label{fig:design}
-\end{figure}
+offers a solution for the identified problems. It consists of a 1) DSL, 2)
+source code generator and 3) software library. Figure \ref{fig:design}
+visualizes the overall design.
+
+% figure design is moved up to end up on this page
 
@@
 
-A DSL and code generator defines a development strategy. This implies choices
-with respect to flexibility, update cost and development effort. We will first
-position \NAME and motivate its design with respect to these trade offs.
-Second, we will look at the principal design components of the DSL and how it
-deals with loops and common data. Third, we discuss the structure of the code
-generator and introduce its software library, that offers a way to deal with
-the network access issue.
+A DSL and code generator defines a development strategy. This has implications,
+both on what is possible and how it can be achieved. We will first position
+\NAME and motivate its design with respect to these trade offs. Second, we will
+look at the principal design components of the DSL. Third, we discuss the
+structure of the code generator and introduce its software library, that offers
+a way to deal with for example the network access issue.
 
@@
 
-\newcommand{\ball}[5][above,yshift=+5pt]{
-\tdplottransformmainscreen{#2}{#3}{#4}
-\shadedraw[tdplot_screen_coords, ball color=lightgray,draw=lightgray,shading angle=-90]
-  (\tdplotresx,\tdplotresy) circle (0.20)
-  node[text=gray,font=\footnotesize,#1]{#5};
-}
-
-\begin{figure}[hb]
-  \centering
-  \scalebox{.85}{
-\tdplotsetmaincoords{110}{-10}
-\begin{tikzpicture}[scale=1,tdplot_main_coords]
-  % axis
-  \draw[thick,->] (0,0,0) -- (7,0,0) node[anchor=north east,xshift=10mm]{$flexibility$};
-  \draw[thick,->] (0,0,0) -- (0,7,0) node[anchor=north west]{$update\ cost$};
-  \draw[thick,->] (0,0,0) -- (0,0,4) node[anchor=south]{$development$};
-  % points
-  \coordinate (O)        at (0,0,0);
-  \coordinate (ideal)    at (6,0,0);
-  \coordinate (vm)       at (2,2,2);
-  \coordinate (modular)  at (4,4,4);
-  \coordinate (reconfig) at (4,2,4);
-  \coordinate (full)     at (6,6,4);
-  \coordinate (cg)       at (6,4,2.5);
-  % "lines"
-  \draw[dashed] (2,0,0)    -- (2,2,0) -- (0,2,0);
-  \draw[dashed] (4,0,0)    -- (4,4,0) -- (0,4,0);
-  \draw[dashed] (6,0,0)    -- (6,6,0) -- (0,6,0);
-  \draw[dashed] (vm)       -- (2,2,0);
-  \draw[dashed] (modular)  -- (4,4,0);
-  \draw[dashed] (full)     -- (6,6,0);
-  \draw[dashed] (reconfig) -- (4,2,0);
-  \draw[dashed] (cg)       -- (6,4,0);
-  % "balls"
-  \ball{6}{0}{0}{ideal}
-  \ball{2}{2}{2}{virtual machine}
-  \ball[left,xshift=-5pt]{4}{4}{4}{\shortstack{modular\\native image}}
-  \ball{4}{2}{4}{\shortstack{reconfigurable\\virtual machine}}
-  \ball[right,xshift=+5pt,yshift=+5pt]{6}{6}{4}{full native image}
-  %\ball[right,xshift=+5pt,text=black]{6}{4.5}{3}{code generation};
-  % 5. cg
-  \tdplottransformmainscreen{6}{4}{2.5}
-  \shadedraw[tdplot_screen_coords, ball color=gray,draw=black,shading angle=-90]
-    (\tdplotresx,\tdplotresy) circle (0.20)
-    node[right,xshift=+5pt,text=black,align=center,font=\footnotesize]{code generation\\+ functional code fusion};
-  % "box around our solution" ;-)
-  \tdplottransformmainscreen{0}{0}{0}
-  \draw[tdplot_screen_coords] (4.95,1) rectangle (8.5,0.25);
-
-\end{tikzpicture}
-}
-\caption{Classification of Development Strategies for Wireless Devices (based on \cite{sadilek2007energy})}
-\label{fig:positioning}
-\end{figure}
-
-Figure \ref{fig:positioning} shows a classification of possible development
-strategies for projects involving wireless devices. Three properties can
-classify development strategies \cite{sadilek2007energy}: flexibility, update
-cost and development effort.
-
-\begin{LaTeXdescription}
-
-  \item[Flexibility] is a measure for the number of applications one can deploy
-  on a device without physical access.
-
-  \item[Update cost] is equivalent to the energy required to deploy a new
-  version of an application on a device.
-
-  \item[Development effort] covers the time to write, test and deploy an
-  application.
-
-\end{LaTeXdescription}
-
-The \emph{ideal} situation is that were one ends up with a maximal flexibility
-and no update nor development cost. \NAME starts from the full native image,
-allowing full flexibility, and aims to reduce both the update and development
-cost: first, by applying functional code fusion, \NAME optimizes the code
-organization, which has a positive effect on the code size and hence also on
-the update cost. Second, by enabling reuse on a functional level and automating
-the writing phase, the development cost is also lowered.
+The introduction of a code generator in the development process changes the
+traditional way of working. Most importantly because the actual source code is
+no longer the driving force that can be controlled by developers. It is
+literally an intermediate representation between code generator and compiler.
+
+A code generator constructs this representation using three components, as
+shown in figure \ref{fig:design}: 1) the description of the algorithm in the
+DSL, 2) information about the domain, in this case the Nodes domain and 3)
+information about the platform and target language. At the level of the
+compiler, a software library is introduced to offload specialized
+implementations and allow custom extensions.
+
+The applied grayscale indicates the components' platform dependency: the DSL is
+platform and language independent and deals with subsystems, such as the
+network, in an abstract way. The Nodes domain implements some of these
+abstracted subsystems, but still in a platform and language independent way.
+The platform and language descriptions define the target platform and language
+and convert the abstract constructions in actually supported code. The software
+library is standard native code, targeting the platform and language. Finally,
+the code and image are artifacts of the automated generation and compilation
+process. They aren't a product of manual work and therefore aren't directly
+dependent on the platform.
+
+This hierarchy is important from a development effort point of view. The level
+of dependency on the targeted platform also dictates the level of effort to
+implement changes to that platform. For example: to generate a different
+language will require a complete rewrite of the software library and a partial
+rewrite of the language description of the generator. Although this is a
+reusable, one-time effort, it is important in relation to the scalability of
+the framework itself.
 
@@
 functionality.
+
+Besides avoiding these problems, \NAME's DSL also provides a higher level of
+abstraction with respect to memory management and data access.
+
+This section therefore discusses \NAME's approach to 1) loops, 2) nodes domain,
+3) unified messaging, 4) language support and illustrates these with a complete
+example.
 
@@
 
-All algorithms deal with information about other devices in the network. To
-centralize this information \NAME provides a unified \ttt{nodes} concept as
-part of its DSL and exposes it as an object to the implementer.
+Most, if not all, ID algorithms deal with information about other devices in
+the network. To centralize this information \NAME provides a unified
+\ttt{nodes} concept as part of its DSL and exposes it as an object to the
+implementer.
 
@@
 define a scope for a function to execute in. This way, nodes also offer a way
-to i.e.\ iterate all known devices, again without the need for a loop.
+to i.e.\ iterate all known devices, again without the need for an explicit loop.
 
@@
 optimized marshaling and grouping of messages.
+
+\subsubsection{Language Support}
+\label{language-support}
+
+Using a platform independent language also offers the possibility to relief the
+implementer of explicit memory management and low-level data structures. From a
+language perspective, \NAME offers several programming constructs, typically
+found in modern programming languages with higher levels of abstraction:
+optional typing, object orientation, pattern matching, aspect orientation.
 
@@
 }
-\caption{Overview of \NAME's code generation process}
+\caption{\NAME's code generation process}
 \label{fig:code-generation}
@@
 }
-\caption{Partial Semantic Model for \NAME. Shaded classes indicate the \emph{event/scheduling} construction around functions to allow interpretation.}
+
+\caption{Partial Semantic Model for \NAME. Shaded classes indicate the
+\emph{event/scheduling} construction around functions to allow interpretation
+of the intended functionality.}
+
 \label{fig:meta-model}
@@
 library for the specific platform and target language is possible.
-
-\section{Implementation and Evaluation}
-\label{evaluation}
 
@@
               &            & base  & \multicolumn{2}{c}{watchdog} & \multicolumn{2}{c}{reputation} & \multicolumn{2}{c}{both} \\
-\midrule
-image
-size (bytes)  & manual     & 10466 & 15530 &  +48\% & 13306 &  +27\% & 18334 &  +75\%\\
-              & generated  & 10466 & 18352 &  +75\% & 16376 &  +56\% & 20998 & +101\%\\
-\cmidrule{2-9}
-              & difference &     0 &  2822 &  +18\% &  3070 &  +23\% &  2664 &  +15\%\\
 \midrule
@@
               & difference &     0 &    27 &  +29\% &    33 &  +38\% &   -11 &   -7\%\\
+\midrule
+image
+size (bytes)  & manual     & 10466 & 15530 &  +48\% & 13306 &  +27\% & 18334 &  +75\%\\
+              & generated  & 10466 & 18352 &  +75\% & 16376 &  +56\% & 20998 & +101\%\\
+\cmidrule{2-9}
+              & difference &     0 &  2822 &  +18\% &  3070 &  +23\% &  2664 &  +15\%\\
 \bottomrule
@@
 
-This section discusses a prototype implementation and evaluates its
-performance. We first present the architecture of this implementation and
-consider the impact it has from a development point of view. Next we use this
-implementation to quantify its merits using a side-by-side comparison of
-generated code versus a manual implementation. We consider experimental results
-regarding the image size, usage of the network and execution time of the
-event-loop. We also take a more analytical approach and look at the handling of
-common data and the size of the source code.
-
-\subsection{Architecture}
-
-Figure \ref{fig:architecture} shows the architecture of the implementation.
-
-\begin{figure}[ht]
-  \centering
-  \scalebox{.85}{
-\begin{tikzpicture}[
-  node distance=10mm and 20mm,
-  ,>=stealth,thick,black!50,text=black,
-  every new ->/.style={shorten >=1pt},
-]
-  \node (dsl)      [user,align=center]                                     {DSL};
-  \node (code)     [generated,right=of dsl]                                {code};
-  \node (image)    [generated,right=of code]                               {image};
-
-  \node (nodes)    [dom,above right=of dsl,xshift=-20mm]                   {Nodes};
-  \node (platform) [platform,below right=of dsl,align=center,xshift=-20mm] {Platform \&\\Language};
-  \node (lib)      [native,below right=of code,align=center,xshift=-20mm]  {Software\\Library};
-
-  \draw [->] (dsl)  -> (code)  node [midway,above=2pt] {generation};
-  \draw [->] (code) -> (image) node [midway,above=2pt] {compilation};
-
-  \draw [->,dashed] (nodes.south)    -> ++(0,-15mm);
-  \draw [->,dashed] (platform.north) -> ++(0,15mm);
-  \draw [->,dashed] (lib.north)      -> ++(0,15mm);
-\end{tikzpicture}
-}
-  \caption{Architecture of Prototype Implementation}
-  \label{fig:architecture}
-\end{figure}
-
-It illustrates the generation of the DSL into compilable code, using the Nodes
-domain information, a platform and language description and the native software
-library. The applied grayscale indicates the components' platform dependency:
-the DSL is platform and language independent and deals with subsystems, such as
-the network, in an abstract way. The Nodes domain implements some of these
-abstracted subsystems, but still in a platform and language independent way.
-The platform and language descriptions define the target platform and language
-and convert the abstract constructions in actually supported code. The software
-library is standard native code, targeting the platform and language. Finally,
-the code and image are artifacts of the automated generation and compilation
-process. They aren't a product of manual work and therefore aren't directly
-dependent on the platform.
-
-This hierarchy is important from a development effort point of view. The level
-of dependency on the targeted platform also dictates the level of effort to
-implement changes to that platform. For example: to generate a different
-language will require a complete rewrite of the software library and a partial
-rewrite of the language description of the generator. Although this is a
-reusable, one-time effort, it is important in relation to the scalability of
-the framework itself.
+\section{Implementation and Evaluation}
+\label{evaluation}
+
+In this section we present a prototype implementation and evaluates its
+performance. We first present the implementation consisting of the hardware
+platform, network and software stack. Next we use this implementation to
+quantify its merits using a side-by-side comparison of generated code versus a
+manual implementation. We consider experimental results regarding the image
+size, usage of the network and execution time of the event-loop. We also take a
+more analytical approach and look at the handling of common data and the size
+of the source code.
 
@@
 To evaluate the prototype we used a system based on the Atmel ATMEGA1284p
-micro-controller and the Digi XBee S2 ZigBee module. A basic library wrapped
-the technical calls to the hardware components in more easy to use functions.
+micro-controller and the Digi XBee S2 ZigBee module. A basic software library
+wraps the technical calls to the hardware components in more easy to use
+functions.
 
@@
 
-We evaluated three criteria: the image size of the resulting compiled code, the
-network usage in number of packets and bytes and the time required to perform
-once cycle of the event loop. To do this, we collected these metrics for four
+We evaluated three criteria: the network usage in number of packets and bytes,
+the time required to perform once cycle of the event loop and the image size of
+the resulting compiled code. To do this, we collected these metrics for four
 situations: without ID algorithms, with a watchdog, with reputation-tracking
@@
 A relative value for this difference, to the manual case, is also included.
+Four observations allow us to evaluate the overall performance of \NAME's
+approach.
 
@@
 
-Secondly, we see that the software library that comes with the generated code
+Secondly, when turning to the goal to reduce the network usage, we see that both
+the number of packets as well as the amount of bytes sent have dropped by 13\%
+and 7\% respectively.
+
+Thirdly, focusing on the execution time, both algorithms by themselves add an
+equal amount of time to the processing of a single event loop cycle. But when
+combined, the event loop is about 7\% faster compared to the manual case.
+
+Finally, we see that the software library that comes with the generated code
 adds to the size of the resulting image. But the cost is close to constant and
@@
 
-Thirdly, when turning to the goal to reduce the network usage, we see that both
-the number of packets as well as the amount of bytes sent have dropped by 13\%
-and 7\% respectively.
-
-Finally, focusing on the execution time, both algorithms by themselves add an
-equal amount of time to the processing of a single event loop cycle. But when
-combined, the event loop is about 7\% faster compared to the manual case.
-
 \subsection{Analytical Results}
@@
 manually results in 637 lines of C code. The corresponding \NAME implementation
-results in 156 lines of DSL code.
+results in 156 lines of DSL code. This is largely thanks to the higher level
+coding constructs, as introduced in section \ref{language-support} dealing with
+\NAME's language support.
 
@@
 state machines. Our event-driven approach shares these same ideas.
+
+Although not a real DSL, nesC \cite{gay2003nesc} is a general purpose language
+that shares some typical concepts with \NAME, such as the high degree of
+analysis during code generation to optimize execution. The biggest difference
+is that \NAME starts from the intended functionality and can optimize certain
+constructions that can't be interpreted from more technical source code.
 
@@
 
-Applying code generation to the context of ID, often focusses on the generation
-of native code to avoid the interpretation of descriptions of patterns to look
-for. I2S \cite{charitakis2003code} is an example that generates code for SNORT
+Applying code generation to ID, often focusses on the generation of native code
+to avoid the interpretation of descriptions of patterns to look for. I2S
+\cite{charitakis2003code} is an example that generates code for SNORT
 \cite{roesch1999snort} rules. To optimize this highly efficient code, it often
@@
 Adequate intrusion detection for IoT devices requires multiple algorithms to
-operate at the same time. Recent literature describes different algorithms, but
-few are readily available for direct application when implementing an IDS for
-IoT devices. Reselection of algorithms, due to changing requirements or
-security threats, requires a flexible way to incorporate the different
-algorithms. This eliminates manual implementation and optimization.
-
-Wireless resource-constraint devices, such as IoT devices, require that
-software respects their restrictions. Adding multiple intrusion detection
-algorithms requires fusion of these algorithms on a functional level, to reduce
-execution inefficiencies, redundant memory usage and superfluous network access.
-
-\NAME is a framework consisting of a DSL, a source code generator and
-accompanying software library. Using the DSL, intrusion detection algorithms
-can be formally described in terms of functionality. During source code
-generation, functional code fusion interprets the functional meaning of the
-algorithms and merges common functionality and shared data. The software
-library hosts common functionality, such as network message parsing, sent
-message aggregation and event scheduling. This allows the source code generator
-to focus on the optimization of the organization of the functionality.
-
-Initial experiments confirm the theoretical positive impact on resource usage.
+operate at the same time. Frequent reselection, e.g. due to changing
+requirements or security threats, requires a flexible way to incorporate the
+different algorithms.
+
+Adding multiple intrusion detection algorithms to wireless resource-constraint
+devices requires fusion of these algorithms on a functional level. This can
+reduce execution inefficiencies, redundant memory usage and superfluous network
+access.
+
+\NAME is a framework consisting of a DSL, a source code generator and software
+library. Using the DSL, intrusion detection algorithms can be formally
+described. The code generator applies functional code fusion to interpret the
+intent of algorithms and merges common functionality and shared data. The
+software library hosts common functionality, such as network message parsing
+and event scheduling.
+
+Initial experiments confirm the theoretical positive impact on resource usage:
+grouping of messages lowers the number of packets by 13\%, while the total
+number of bytes sent over the network is reduced by 7\%. Finally, the execution
+time is also improved by 7\%.
+
 The platform independent DSL enables generation of source code targeting
-different platforms and languages, which allows high degrees of reuse in
-heterogenous environments and different configurations.
-
-Future research will focus on the integration of applications with the
-generated intrusion detection systems. We see opportunities for applications
-who detect intrusion, to actively and dynamically respond and avoid malicious
-payloads.
+different platforms and languages, which allows high degrees of reuse with
+heterogenous devices.
+
+As a next step, we want to investigate the relation of \NAME's DSL to reactive
+programming. Both clearly share common concepts with respect to
+event-drivenness and the absence of e.g. explicit loops.
+
+Future research will focus on the integration of applications with a generated
+IDS. We see opportunities for applications who detect intrusion, to actively
+and dynamically respond and avoid malicious payloads.
 

### 2014-11-23 tikz.tex
incorporated comments of review meeting

@@
   native/.style         = { support,  fill=black!70!white, text=white },
-  generated/.style      = { artifact, border },
+  generated/.style      = { artifact, border, dashed },
   resource/.style       = { artifact, border },

### 2014-11-23 paper.tex
added missing memory reduction (conclusion) + typo

@@
 grouping of messages lowers the number of packets by 13\%, while the total
-number of bytes sent over the network is reduced by 7\%. Finally, the execution
-time is also improved by 7\%.
+number of bytes sent over the network is reduced by 7\%. The execution time is
+improved by 7\% and memory usage can be reduced by even 20\%.
 
@@
 Future research will focus on the integration of applications with a generated
-IDS. We see opportunities for applications who detect intrusion, to actively
+IDS. We see opportunities for applications who detect intrusions, to actively
 and dynamically respond and avoid malicious payloads.

### 2014-11-23 paper.tex
improved future work on relation to FRP

@@
 
-As a next step, we want to investigate the relation of \NAME's DSL to reactive
-programming. Both clearly share common concepts with respect to
-event-drivenness and the absence of e.g. explicit loops.
+As a next step, we want to investigate the relation of \NAME to other
+programming paradigms such as functional reactive programming. They share
+common concepts with respect to event-drivenness and the preference to avoid
+explicit loops.
 

### 2014-11-24 paper.tex
fixed various typos, bad sentences,…

@@
 When intrusion prevention fails, intrusion detection serves as a second layer
-of defense. It looks for patterns \& anomalies that can indicate malicious
+of defense. It looks for patterns and anomalies that can indicate malicious
 behavior. However, supporting an adequate set of intrusion detection algorithms
@@
 
-Current developments in ID focus on programming frameworks that structure the
-implementation of algorithms \cite{valero2012di} and offer the required basic
-functional components to implement them \cite{krontiris2008lidea}. A key
-shortcoming of prior ID frameworks, is that they don't offer a way to optimize
-the usage of resources. A solution that supports the integration of multiple
-algorithms on IoT devices needs to avoid accumulating their impact on the
-available resources. A second omission is support for systematic reuse to
-create different configurations heterogeneous devices. Addressing this in a
-transparent and automated way offers the potential to significantly reduce the
-effort of developing an IDS.
+Current developments in ID on resource-constrained devices focus on
+programming frameworks that structure the implementation of algorithms
+\cite{valero2012di} and offer the required basic functional components to
+implement them \cite{krontiris2008lidea}. A key shortcoming of prior ID
+frameworks, is that they don't offer a way to optimize the usage of resources.
+A solution that supports the integration of multiple algorithms on IoT devices
+needs to avoid accumulating their impact on the available resources. A second
+omission is support for systematic reuse to create different configurations
+heterogeneous devices. Addressing this in a transparent and automated way
+offers the potential to significantly reduce the effort of developing an IDS.
 
@@
 the formal description of ID algorithms on a functional level and provides a
-code generator to produces corresponding source code for a given platform and
+code generator to produce corresponding source code for a given platform and
 configuration. This way, \NAME addresses the heterogeneous nature of both the
@@
 algorithms. FCF identifies common data and functions, and organizes code to
-eliminate redundant iterations, tests and computations. Results show improved
-image size, reduced usage of the wireless radio, reduced execution time and
-lower memory usage.
+eliminate redundant iterations, tests and computations. Results show reduced
+usage of the wireless radio, reduced execution time, lower memory usage and
+improved image size.
 
@@
 
-Using a platform independent language also offers the possibility to relief the
-implementer of explicit memory management and low-level data structures. From a
-language perspective, \NAME offers several programming constructs, typically
-found in modern programming languages with higher levels of abstraction:
-optional typing, object orientation, pattern matching, aspect orientation.
+Using a platform independent language also offers the possibility to relieve
+the implementer of explicit memory management and low-level data structures.
+From a language perspective, \NAME offers several programming constructs,
+typically found in modern programming languages with higher levels of
+abstraction: optional typing, object orientation, pattern matching, aspect
+orientation.
 
@@
 applicable, like template-based text expansion, XSLT transformations or CodeDom
-\cite{dollard2004code}, we believe that the functional transformations require
-a richer model.
+\cite{dollard2004code}, we believe that the required transformations demand a
+richer model.
 
@@
 
-In this section we present a prototype implementation and evaluates its
+In this section we present a prototype implementation and evaluate its
 performance. We first present the implementation consisting of the hardware
@@
 quantify its merits using a side-by-side comparison of generated code versus a
-manual implementation. We consider experimental results regarding the image
-size, usage of the network and execution time of the event-loop. We also take a
+manual implementation. We consider experimental results regarding usage of the
+network, execution time of the event-loop and the image size. We also take a
 more analytical approach and look at the handling of common data and the size
@@
 
-We evaluated three criteria: the network usage in number of packets and bytes,
+We evaluated four criteria: the network usage in number of packets and bytes,
 the time required to perform once cycle of the event loop and the image size of
@@
 
-Secondly, when turning to the goal to reduce the network usage, we see that both
-the number of packets as well as the amount of bytes sent have dropped by 13\%
-and 7\% respectively.
+Secondly, when turning to the goal to reduce the network usage, we see that
+both the number of packets as well as the amount of bytes sent have dropped by
+13\% and 7\% respectively.
 
@@
 A simple metric is comparing the amount of code versus the size of the formal
-description of IDS algorithms. This can be an indication of the development
+description of ID algorithms. This can be an indication of the development
 effort and complexity. In this experiment, implementing both algorithms
@@
 and code generation, both in relation to wireless networks of
-resource-constraint devices. These two components are the core of the \NAME
+resource-constrained devices. These two components are the core of the \NAME
 framework and the solution it proposes. We also look at their application in
@@
 
-Within the scope of wireless networks of resource-constraint devices, we see
+Within the scope of wireless networks of resource-constrained devices, we see
 the introduction of DSLs \enquote{to shorten the development cycles
@@
 reducing the update cost, while maintaining flexibility and lowering the
-development effort. As explained in section \ref{positioning}, \NAME focuses on
-the reduction of the development effort and update cost, while preserving
-maximal flexibility.
+development effort. In this terminology, \NAME focuses on the reduction of the
+development effort and update cost, and maximizes flexibility.
 
@@
 
-Adding multiple intrusion detection algorithms to wireless resource-constraint
-devices requires fusion of these algorithms on a functional level. This can
-reduce execution inefficiencies, redundant memory usage and superfluous network
-access.
+Adding multiple intrusion detection algorithms to wireless resource-constrained
+devices requires fusion of algorithms on a functional level. This can reduce
+execution inefficiencies, redundant memory usage and superfluous network access.
 

### 2014-12-07 paper.tex
improved relation to FRP
processed final review meeting’s remarks

@@
 
-\author{\IEEEauthorblockN{Christophe Van Ginneken, Jef Maerien, Danny Hughes, 
-Christophe Huygens, Wouter Joosen}%
+\author{\IEEEauthorblockN{Christophe Van Ginneken, Jef Maerien,
+Christophe Huygens, Wouter Joosen, Danny Hughes}%
 \IEEEauthorblockA{iMinds-DistriNet, KU Leuven\\
@@
 imposes significant overhead on resource-constrained devices such as those in
-the Internet of Things. Fusion of the algorithms reduces resource consumption
-by optimizing resource sharing, yet, proves to be time-consuming, repetitive
-and error-prone. To address this problem we propose \NAME, a framework for the
-development of efficient intrusion detection systems. It consists of a domain
-specific language, to formally describe intrusion detection algorithms on a
-functional level, and a code generator to produce more optimally organized
-source code. A side-by-side comparison shows that \NAME-generated code is
-smaller and reduces message passing overhead, execution time and memory
-footprint in comparison to a straightforward sequential implementation of the
-same algorithms.
+the Internet of Things. Manual fusion of the algorithms reduces resource
+consumption by optimizing resource sharing, yet, proves to be time-consuming,
+repetitive and error-prone. To address this problem we propose \NAME, a
+framework for the development of efficient intrusion detection systems. It
+consists of a domain specific language and a code generator. The language
+allows formally describing the intent of intrusion detection algorithms and
+supports the generator in organizing the source code. A side-by-side comparison
+shows that \NAME-generated code reduces message passing overhead, execution
+time and memory footprint in comparison to sequential calls to individual
+implementations of the algorithms.
 
@@
 
-% IoT + threat
-
-The internet of things (IoT) holds great potential to positively influence our
-daily work and life through domotics, assisted living, e-health and enhanced
-learning \cite{atzori2010internet}. But with this potential the IoT also
-presents a significant threat: by opening our smart homes and personal data to
-all of these interconnected devices, we also open them to everyone who is able
-to break the virtual locks that protect them (i.e.\ passwords, encryption
-keys). Implementation of security measures by itself is hindered by the fact
-that IoT devices typically have limited batteries, processing power and memory.
-
-% classic networks: fw and central ids
-
-In classic networks, firewalls focus on the outer perimeter of the network,
+% wired networks: fw and central ids
+
+In wired networks, firewalls focus on the outer perimeter of the network,
 filtering unwanted packets and protecting the entire internal network. But
@@
 
+% IoT + threat
+
+The internet of things (IoT) holds great potential to positively influence our
+daily work and life through domotics, assisted living, e-health and enhanced
+learning \cite{atzori2010internet}. But with this potential the IoT also
+presents a significant threat: by opening our smart homes and personal data to
+all of these interconnected devices, we also open them to everyone who is able
+to break the virtual locks that protect them (i.e.\ passwords \& encryption
+keys). Implementation of security measures by itself is hindered by the fact
+that IoT devices typically have limited batteries, processing power and memory.
+
 % local ids -> localize -> limited resources
@@
 network to oversee all traffic \cite{mishra2004intrusion}. Every device has to
-implement its own lines of defense. But making the IDS a local service on each
-device requires local resources, which are a scarce commodity for IoT devices.
+implement its own lines of defense. However making the IDS a local service on
+each device requires local resources, which are a scarce commodity for IoT
+devices.
 
 Attackers have access to a large set of possible attack vectors
-\cite{aschenbruck2012security}, ranging from the wireless network, through the
-medium access control and routing, up to the device itself. Each level presents
-different opportunities to manipulate data, eavesdrop, perform a form of denial
-of service. For each of these attacks an algorithm needs to look for patterns
-or anomalies. An IDS for IoT devices therefore requires a large number of
-implemented algorithms for an adequate coverage.
+\cite{aschenbruck2012security}, ranging from the physical layer, through the
+access and routing layers, up to the application layer. Each layer presents
+different opportunities to manipulate data, eavesdrop or perform a form of
+denial of service. For each of these attacks an algorithm needs to look for
+patterns or anomalies. An IDS for IoT devices therefore requires a large number
+of algorithms for an adequate coverage.
 
@@
 
-Current developments in ID on resource-constrained devices focus on
-programming frameworks that structure the implementation of algorithms
-\cite{valero2012di} and offer the required basic functional components to
-implement them \cite{krontiris2008lidea}. A key shortcoming of prior ID
-frameworks, is that they don't offer a way to optimize the usage of resources.
-A solution that supports the integration of multiple algorithms on IoT devices
-needs to avoid accumulating their impact on the available resources. A second
-omission is support for systematic reuse to create different configurations
-heterogeneous devices. Addressing this in a transparent and automated way
-offers the potential to significantly reduce the effort of developing an IDS.
+Current developments in ID on resource-constrained devices focus on programming
+frameworks that structure the implementation of algorithms \cite{valero2012di}
+and offer the required basic functional components to implement them
+\cite{krontiris2008lidea}. A key shortcoming of prior ID frameworks, is that
+they don't offer a way to optimize the usage of resources. A solution that
+supports the integration of multiple algorithms on IoT devices needs to avoid
+accumulating their impact on the available resources. A second omission is
+support for systematic reuse to create different configurations. Addressing
+this in a transparent and automated way offers the potential to significantly
+reduce the effort of developing an IDS.
 
@@
 
-% 1. framework: language + code generator
-
-The first contribution of this paper is a framework called \NAME. It enables
-the formal description of ID algorithms on a functional level and provides a
-code generator to produce corresponding source code for a given platform and
-configuration. This way, \NAME addresses the heterogeneous nature of both the
-IoT and ID algorithms.
-
-% 2. FCF to optimise
-
-The second contribution introduces functional code fusion (FCF) as a source
-code generation paradigm to solve the problem of hard to reuse and combine ID
-algorithms. FCF identifies common data and functions, and organizes code to
-eliminate redundant iterations, tests and computations. Results show reduced
-usage of the wireless radio, reduced execution time, lower memory usage and
-improved image size.
+% 1. classification of ID and design pattern for ID algorithms
+
+The first contribution of this paper consists of the formulation of a design
+pattern for ID algorithms, based on properties identified from a classification.
+
+% 2. framework: language + code generator
+
+The second contribution is a framework called \NAME. It enables the formal
+description of ID algorithms on a functional level and provides a code
+generator producing source code for a given platform and configuration. This
+way, \NAME addresses the heterogeneous nature of both IoT and ID algorithms.
+
+% 3. FCF to optimise
+
+The third contribution introduces functional code fusion (FCF) as a source code
+generation paradigm to address hard to combine ID algorithms. FCF identifies
+common data and functions, and organizes code to eliminate redundant
+iterations, tests and computations. Results show reduced usage of the wireless
+radio, execution time and memory usage.
 
@@
 
-The remainder of this paper proceeds as follows: section \ref{classification}
-analyses and classifies ID algorithms. Section \ref{pattern} looks for patterns
-in the implementation of these classes. Section \ref{problems} identifies the
-problems arising from these patterns. Section \ref{design} describes the design
-we applied to construct \NAME, its domain specific language (DSL) and code
-generator. Section \ref{evaluation} discusses and evaluates an implementation.
-Section \ref{related} explores related work in the field of DSLs and code
-generation. Finally, section \ref{conclusion} summarizes our findings, draws
-conclusions and identifies topics for future work.
+The remainder of this paper proceeds as follows: section \ref{problems}
+describes the inherent problems with the combination of ID algorithms. Section
+\ref{classification} analyses and classifies ID algorithms. Section
+\ref{pattern} looks for patterns in the implementation of these classes.
+Section \ref{design} describes the design we applied to construct \NAME, its
+domain specific language (DSL) and code generator. Section \ref{evaluation}
+discusses and evaluates an implementation. Section \ref{related} explores
+related work in the field of DSLs and code generation. Finally, section
+\ref{conclusion} summarizes our findings, draws conclusions and identifies
+topics for future work.
+
+\section{Inherent Problems}
+\label{problems}
+
+The combination of multiple ID algorithms tends to result in suboptimal code.
+Besides these code-level problems, operational problems also hinder the
+implementation of an IDS for IoT devices.
+
+Two issues arise from the way ID algorithms are implemented and deal with loop
+constructs and common data. A third issue is the use of the network.
+
+\begin{LaTeXdescription}
+
+  \item[Loops] by themselves are not a problem, but different loops, that
+  perform the same operations in sequence, can incur overhead. An example is
+  the handling of messages: each algorithm parses each observed message on the
+  network, processing the same byte stream, performing comparable operations to
+  find similar patterns.
+
+  \item[Common data] cross-cuts the different algorithms, that have to store
+  information about devices within its own scope. This results in a lot of
+  duplicated information in memory.
+
+  \item[The network] allows algorithms to operate on a distributed scale. But
+  scattered calls to the a message sending function cause multiple accesses to
+  the wireless radio, keeping it active and energy consuming more than strictly
+  needed.
+
+\end{LaTeXdescription}
+
+Two operational aspects are orthogonal to these problems and amplify them:
+
+\begin{LaTeXdescription}
+
+  \item[Reimplementing] the algorithms manually to eliminate these issues is
+  not a one-time cost. A release of a new version of an algorithm, or simply a
+  reselection of algorithms, due to a changing risk analysis, can trigger it.
+  Also, changing, or simply implementing the algorithms from scratch, holds the
+  risk of making mistakes.
+
+  \item[The heterogeneity of IoT devices] results in different software stacks,
+  each with for example their own network API\@. This partially explains why
+  there are few implementations readily available to developers: there is no
+  single platform to target for researchers.
+
+\end{LaTeXdescription}
 
@@
 
-Two aspects, currently missing from ID frameworks, that \NAME wants to address,
-are optimization of resources and support for systematic reuse. This section
-uses a classification to define the scope of \NAME, detailing which types of
-algorithms are good candidates for the proposed approach.
+Two aspects currently missing from ID frameworks are optimization of resources
+and support for systematic reuse. This section uses a classification to define
+the scope of \NAME, detailing which types of algorithms are good candidates for
+the proposed approach.
 
@@
 hardly impossible for a single device to detect a complex attack by itself.
-Therefore a second important research topic consists of cooperative algorithms
-to combine information from single identities into a group-based decision about
-alleged intrusions. We identify two ways to construct cooperative decision
-making and focus on the way participants exchange information: broadcast and
-interactive communication.
+Therefore a second important class of ID algorithms consists of cooperative
+algorithms that combine information from single identities into a group-based
+decision about alleged intrusions. We identify two ways to construct
+cooperative decision making and focus on the way participants exchange
+information: broadcast and interactive communication.
 
@@
   \cite{ganeriwal2008reputation} illustrates this. It uses the same principle
-  of a watchdog also found in \cite{mishra2004intrusion}.
+  of a watchdog found in \cite{mishra2004intrusion}.
 
@@
 event. Signature and misuse detection can in some cases, but anomaly detection
-typically deals with some level of uncertainty. Also, in distributed situations
-it is impossible to fully trust any neighboring device. In these situations one
+typically deals with some level of uncertainty. Also, in distributed setups it
+is impossible to fully trust any neighboring device. In these situations one
 has to rely on statistics and thresholds to decide when to drop trust in a
@@
 
-The pattern calls the different, linked implementations in sequence in two
+The pattern calls the individual, linked implementations in sequence in two
 situations: first, when the device observes new messages on the network and,
@@
 threshold evaluation.
-
-\section{Inherent Problems}
-\label{problems}
-
-These patterns can now help in identifying the problem areas that tend to
-result in suboptimal code when combining multiple ID algorithms. Besides these
-code-level problems, operational problems also hinder the implementation of an
-IDS for IoT devices.
-
-Two issues arise from the nesting of loops that originate from the combination
-of the ID algorithm pattern and the application pattern: loops and common data.
-A third issue is the use of the network.
-
-\begin{LaTeXdescription}
-
-  \item[Loops] by themselves are not a problem, but different loops, that
-  perform the same operations in sequence, can incur overhead. An example from
-  the ID algorithm pattern is the handling of messages: each algorithm parses
-  each observed message on the network, processing the same byte stream,
-  performing comparable operations to find similar patterns.
-
-  \item[Common data] cross-cuts the different algorithms, that have to store
-  information about devices within its own scope. This results in a lot of
-  duplicated information in memory.
-
-  \item[The network] allows algorithms to operate on a distributed scale. But
-  scattered calls to the \ttt{SEND} function cause multiple accesses to the
-  wireless radio, keeping it active and energy consuming more than strictly
-  needed.
-
-\end{LaTeXdescription}
-
-Two operational aspects are orthogonal to these problems and amplify them:
-
-\begin{LaTeXdescription}
-
-  \item[Reimplementing] the algorithms manually to eliminate these issues is
-  not a one-time cost. A release of a new version of an algorithm, or simply a
-  reselection of algorithms, due to a changing risk analysis, can trigger it.
-  Also, changing, or simply implementing the algorithms from scratch, holds the
-  risk of making mistakes.
-
-  \item[The heterogeneity of IoT devices] results in different software stacks,
-  each with for example their own network API\@. This also explains why there
-  are few implementations readily available to developers: there is no single
-  platform to target for researchers.
-
-\end{LaTeXdescription}
 
@@
 important code-level problems. These problems are inherent to ID algorithms.
-But at the same time they conflict with their target environment, the IoT\@.
+But at the same time they conflict with the target environment, the IoT\@.
 \NAME provides a framework to efficiently implement an IDS for the IoT and
@@
 
-A DSL and code generator defines a development strategy. This has implications,
-both on what is possible and how it can be achieved. We will first position
-\NAME and motivate its design with respect to these trade offs. Second, we will
-look at the principal design components of the DSL. Third, we discuss the
-structure of the code generator and introduce its software library, that offers
-a way to deal with for example the network access issue.
+Using a DSL and code generator defines a development strategy. This has
+implications, both on what is possible and how it can be achieved. We will
+first position \NAME and motivate its design with respect to these trade offs.
+Second, we will look at the principal design components of the DSL. Third, we
+discuss the structure of the code generator and introduce its software library,
+that offers a way to deal with for example the network access issue.
 
@@
 abstracted subsystems, but still in a platform and language independent way.
-The platform and language descriptions define the target platform and language
-and convert the abstract constructions in actually supported code. The software
-library is standard native code, targeting the platform and language. Finally,
-the code and image are artifacts of the automated generation and compilation
-process. They aren't a product of manual work and therefore aren't directly
-dependent on the platform.
+Platform and language descriptions convert the abstract constructions in
+actually supported code. The software library is standard native code,
+targeting the platform and language. Finally, the code and image are artifacts
+of the automated generation and compilation process. They aren't a product of
+manual work and therefore aren't directly dependent on the platform.
 
 This hierarchy is important from a development effort point of view. The level
-of dependency on the targeted platform also dictates the level of effort to
+of dependency on the targeted platform dictates the level of effort to
 implement changes to that platform. For example: to generate a different
@@
 (semantic) model} \cite{fowler2010domain}. Its sole purpose should be
-populating the model. \NAME's DSL and model focuses on avoiding constructions
+populating the model. \NAME's DSL and model focus on avoiding constructions
 that result in problems concerning loops, common data and the use of the
@@
 Most, if not all, ID algorithms deal with information about other devices in
-the network. To centralize this information \NAME provides a unified
-\ttt{nodes} concept as part of its DSL and exposes it as an object to the
-implementer.
+the network. To share this information \NAME provides a unified \ttt{nodes}
+concept as part of its DSL and exposes it as an object to the implementer.
 
@@
 instance for every known device will be present in memory. Finally, nodes can
-define a scope for a function to execute in. This way, nodes also offer a way
-to i.e.\ iterate all known devices, again without the need for an explicit loop.
+define a scope for a function to execute in. This way, nodes offer a way to
+i.e.\ iterate all known devices, again without the need for an explicit loop.
 
@@
 
-Using a platform independent language also offers the possibility to relieve
-the implementer of explicit memory management and low-level data structures.
-From a language perspective, \NAME offers several programming constructs,
-typically found in modern programming languages with higher levels of
-abstraction: optional typing, object orientation, pattern matching, aspect
-orientation.
+Using a platform independent language offers the possibility to relieve the
+implementer of explicit memory management and low-level data structures. From a
+language perspective, \NAME offers several programming constructs, typically
+found in modern programming languages with higher levels of abstraction:
+optional typing, object orientation, pattern matching, aspect orientation.
 
@@
 To convert the DSL-based ID algorithms into production-ready code, \NAME
-applies code generation techniques. Although different approaches here are
+applies code generation techniques. Although different approaches are
 applicable, like template-based text expansion, XSLT transformations or CodeDom
@@
 
-The syntactic model is an AST for a virtual, expressive language. It contains
-constructs and ideas found in different languages and programming paradigms. No
-single existing programming language supports all features as is. Syntactic
-transformations therefore morph the unsupported constructs into comparable
-constructs supported by the targeted platform and language.
+The syntactic model is like an abstract syntax tree (AST) for a virtual,
+expressive language. It contains constructs and ideas found in different
+languages and programming paradigms. No single existing programming language
+supports all features as is. Syntactic transformations therefore translate the
+unsupported constructs into comparable constructs supported by the targeted
+platform and language.
 
@@
 
-Other common functionality also isn't generated as is. The execution of
+Other common functionality isn't generated as is either. The execution of
 functions based on a schedule or the implementation of higher level conceptual
 functions, such as \ttt{now} and \ttt{sha1}, are calls into the standard
-software library that supports the code generator.
-
-Optimization of such algorithms is not within the scope of \NAME. It also gives
+software library.
+
+Optimization of such algorithms is not within the scope of \NAME. This gives
 freedom of choice to the users of the framework to select appropriate
@@
 
-The DSL is also platform and language independent. This allows the source code
+The DSL is platform and language independent. This allows the source code
 generator to target different configurations without any change to the formal
@@
 performance. We first present the implementation consisting of the hardware
-platform, network and software stack. Next we use this implementation to
-quantify its merits using a side-by-side comparison of generated code versus a
-manual implementation. We consider experimental results regarding usage of the
-network, execution time of the event-loop and the image size. We also take a
-more analytical approach and look at the handling of common data and the size
-of the source code.
+platform, network and software stack. We first take an analytical approach and
+look at the handling of common data and the size of the source code. Next, we
+use the implementation to quantify its merits using a side-by-side comparison
+of generated code versus a manual implementation. We consider experimental
+results regarding usage of the network, execution time of the event-loop and
+the image size.
 
@@
 
+\subsection{Analytical Results}
+
+Looking at the generated source code is a straightforward way to validate the
+framework. A simple metric is comparing the amount of code versus the size of
+the formal description of ID algorithms. This can be an indication of the
+development effort and complexity. In this experiment, implementing both
+algorithms manually results in 637 lines of C code. The corresponding \NAME
+implementation results in 156 lines of DSL code. This is largely thanks to the
+higher level coding constructs, as introduced in section \ref{language-support}
+dealing with \NAME's language support.
+
+Next we look at data structures. Sharing data eliminates redundant information,
+such as a network address, from the scope of each individual algorithm.
+Analysis of the source code with both algorithms implemented, shows that this
+reduces the memory footprint for one node entry by 2 bytes, from 27 bytes to 25
+bytes. In our implementation we used the ZigBee network address, which indeed
+consists of 16 bits. This results in a reduction of 7.4\%. Using ZigBee's 64
+bits unique addresses, this reduction can easily become 20.5\%.
+
 \subsection{Experimental Results}
 
-We evaluated four criteria: the network usage in number of packets and bytes,
-the time required to perform once cycle of the event loop and the image size of
-the resulting compiled code. To do this, we collected these metrics for four
-situations: without ID algorithms, with a watchdog, with reputation-tracking
-and with both algorithms implemented.
+Besides this analysis we also instrumented the code and evaluated four
+criteria: the network usage in number of packets and bytes, the time required
+to perform once cycle of the event loop and the image size of the resulting
+compiled code. To do this, we collected these metrics for four situations:
+without ID algorithms, with a watchdog, with reputation-tracking and with both
+algorithms implemented.
 
@@
 both the number of packets as well as the amount of bytes sent have dropped by
-13\% and 7\% respectively.
+13\% and 7\%.
 
@@
 resources.
-
-\subsection{Analytical Results}
-
-Looking at the generated source code is another way to validate the framework.
-A simple metric is comparing the amount of code versus the size of the formal
-description of ID algorithms. This can be an indication of the development
-effort and complexity. In this experiment, implementing both algorithms
-manually results in 637 lines of C code. The corresponding \NAME implementation
-results in 156 lines of DSL code. This is largely thanks to the higher level
-coding constructs, as introduced in section \ref{language-support} dealing with
-\NAME's language support.
-
-We can also look at data structures. Sharing data eliminates redundant
-information, such as a network address, from the scope of each individual
-algorithm. Analysis of the source code with both algorithms implemented, shows
-that this reduces the memory footprint for one node entry by 2 bytes, from 27
-bytes to 25 bytes. A ZigBee network address indeed consists of 16 bits. So this
-results in a reduction of 7.4\%. Using ZigBee's 64 bits unique addresses, this
-reduction can easily become 20.5\%.
 
@@
 devices requires fusion of algorithms on a functional level. This can reduce
-execution inefficiencies, redundant memory usage and superfluous network access.
+execution inefficiencies as well as redundant memory usage and network access.
 
@@
 
-As a next step, we want to investigate the relation of \NAME to other
-programming paradigms such as functional reactive programming. They share
-common concepts with respect to event-drivenness and the preference to avoid
-explicit loops.
+As a next step, we will implement more ID algorithms using \NAME and continue
+the experiments on a larger scale. This way we will also be able to verify the
+initially observed inclination in the evolution of the image size with growing
+numbers of algorithms.
 
@@
 
-This research is partially funded by the Interuniversity Attraction Poles
-Programme Belgian State, Belgian Science Policy, and by the Research Fund KU
-Leuven. This research is partially funded by the EU FP7 project NESSoS\@. We
-would like to thank the reviewers for their thoughtful and helpful comments
-that enhanced the readability of this paper.
+This research is partially funded by the Research Fund KU Leuven and partially
+by the EU FP7 project NESSoS\@.
 

### 2014-12-08 paper.tex
fixed typo’s, sentences, semi-colons,…

@@
 shows that \NAME-generated code reduces message passing overhead, execution
-time and memory footprint in comparison to sequential calls to individual
+time and memory footprint in comparison to sequential calls into individual
 implementations of the algorithms.
@@
 In wired networks, firewalls focus on the outer perimeter of the network,
-filtering unwanted packets and protecting the entire internal network. But
+filtering unwanted packets and protecting the entire internal network. Still,
 attacks on flaws in services can pass unnoticed. This is where intrusion
@@
 daily work and life through domotics, assisted living, e-health and enhanced
-learning \cite{atzori2010internet}. But with this potential the IoT also
-presents a significant threat: by opening our smart homes and personal data to
-all of these interconnected devices, we also open them to everyone who is able
-to break the virtual locks that protect them (i.e.\ passwords \& encryption
-keys). Implementation of security measures by itself is hindered by the fact
-that IoT devices typically have limited batteries, processing power and memory.
+learning \cite{atzori2010internet}. With this potential the IoT also presents a
+significant threat: by opening our smart homes and personal data to all of
+these interconnected devices, we also open them to everyone who is able to
+break the virtual locks that protect them (i.e.\ passwords \& encryption keys).
+Implementation of security measures by itself is hindered by the fact that IoT
+devices typically have limited batteries, processing power and memory.
 
@@
 and offer the required basic functional components to implement them
-\cite{krontiris2008lidea}. A key shortcoming of prior ID frameworks, is that
+\cite{krontiris2008lidea}. A key shortcoming of prior ID frameworks is that
 they don't offer a way to optimize the usage of resources. A solution that
@@
 
-The first contribution of this paper consists of the formulation of a design
+The first contribution of this paper consists in the formulation of a design
 pattern for ID algorithms, based on properties identified from a classification.
@@
 The remainder of this paper proceeds as follows: section \ref{problems}
-describes the inherent problems with the combination of ID algorithms. Section
+describes the inherent problems in the combination of ID algorithms. Section
 \ref{classification} analyses and classifies ID algorithms. Section
@@
 
-The combination of multiple ID algorithms tends to result in suboptimal code.
-Besides these code-level problems, operational problems also hinder the
-implementation of an IDS for IoT devices.
+The combination of ID algorithms tends to result in suboptimal code. Besides
+these code-level problems, operational problems also hinder the implementation.
 
@@
   \item[Loops] by themselves are not a problem, but different loops, that
-  perform the same operations in sequence, can incur overhead. An example is
+  perform the same operations in sequence, can cause overhead. An example is
   the handling of messages: each algorithm parses each observed message on the
@@
 
-  \item[The network] allows algorithms to operate on a distributed scale. But
-  scattered calls to the a message sending function cause multiple accesses to
-  the wireless radio, keeping it active and energy consuming more than strictly
-  needed.
+  \item[The network] allows algorithms to operate on a distributed scale.
+  However, scattered calls to a message sending function cause multiple
+  accesses to the wireless radio, keeping it active and energy consuming more
+  than strictly needed.
 
@@
 typically apply duty cycling to lower their energy consumption, further
-increases this problem, introducing periods of time where they cannot observe
+increases this problem, introducing periods of time when they cannot observe
 local events.
@@
 additional categories seem promising, the authors admit that the impact on
-resource-constrained devices will be too much. We therefore focus on the three
+resource-constrained devices will be too high. We therefore focus on the three
 categories agreed upon by most sources.
@@
   \item[Signature or misuse detection] starts with known patterns of attacks.
-  It tries to find evidence for intrusions in the actions of and communication
+  It tries to find evidence of intrusions in the actions of and communication
   with a device. Because attacks often use valid ways to communicate with
@@
 Not all algorithms can decide if an attack is real or not, based on a single
-event. Signature and misuse detection can in some cases, but anomaly detection
-typically deals with some level of uncertainty. Also, in distributed setups it
-is impossible to fully trust any neighboring device. In these situations one
-has to rely on statistics and thresholds to decide when to drop trust in a
-given device.
+event. Signature and misuse detection can do that in some cases, but anomaly
+detection typically deals with some level of uncertainty. Also, in distributed
+setups it is impossible to fully trust any neighboring device. In these
+situations one has to rely on statistics and thresholds to decide when to drop
+trust in a given device.
 
@@
 The previous sections identified loops, common data and the network as three
-important code-level problems. These problems are inherent to ID algorithms.
-But at the same time they conflict with the target environment, the IoT\@.
-\NAME provides a framework to efficiently implement an IDS for the IoT and
-offers a solution for the identified problems. It consists of a 1) DSL, 2)
-source code generator and 3) software library. Figure \ref{fig:design}
-visualizes the overall design.
+important code-level problems. These problems are inherent to ID algorithms,
+but conflict with the target environment, the IoT\@. \NAME provides a framework
+to efficiently implement an IDS for the IoT and offers a solution for the
+identified problems. It consists of a 1) DSL, 2) source code generator and 3)
+software library. Figure \ref{fig:design} visualizes the overall design.
 
@@
 The introduction of a code generator in the development process changes the
-traditional way of working. Most importantly because the actual source code is
-no longer the driving force that can be controlled by developers. It is
-literally an intermediate representation between code generator and compiler.
+traditional way of working. This is most importantly due to the fact that the
+actual source code is no longer the driving force that can be controlled by
+developers. It is literally an intermediate representation between code
+generator and compiler.
 
@@
 shown in figure \ref{fig:design}: 1) the description of the algorithm in the
-DSL, 2) information about the domain, in this case the Nodes domain and 3)
+DSL, 2) information about the domain, in this case the Nodes domain, and 3)
 information about the platform and target language. At the level of the
@@
 of dependency on the targeted platform dictates the level of effort to
-implement changes to that platform. For example: to generate a different
+implement changes in that platform. For example: to generate a different
 language will require a complete rewrite of the software library and a partial
@@
 This section therefore discusses \NAME's approach to 1) loops, 2) nodes domain,
-3) unified messaging, 4) language support and illustrates these with a complete
-example.
+3) unified messaging, 4) language support, and illustrates these with a
+complete example.
 
@@
 Most, if not all, ID algorithms deal with information about other devices in
-the network. To share this information \NAME provides a unified \ttt{nodes}
+the network. To share this information, \NAME provides a unified \ttt{nodes}
 concept as part of its DSL and exposes it as an object to the implementer.
 
-Algorithms can extend nodes with their own properties. This way, only one
-instance for every known device will be present in memory. Finally, nodes can
-define a scope for a function to execute in. This way, nodes offer a way to
-i.e.\ iterate all known devices, again without the need for an explicit loop.
+Algorithms can extend nodes with their own properties. This allows for only one
+instance per known device to be present in memory. Finally, nodes can define a
+scope for a function to execute in. This way, nodes offer a way to i.e.\
+iterate all known devices, again without the need for an explicit loop.
 
@@
 to be more active than needed. Delaying the sending of messages and grouping
-them, can lower this overhead.
+them can lower this overhead.
 
@@
 The semantic model is the core of the solution. It offers a way to describe ID
-algorithms in a way that allows semantic reinterpretation through functional
-code fusion. This way, reorganizing the functionality of different algorithms
+algorithms so that semantic reinterpretation through functional code fusion is
+possible. This reorganization of the functionality of different algorithms
 allows for more optimal execution and memory usage. Hence the name \NAME of the
@@
 avoiding small discrepancies in code and suboptimal use of shared resources.
-Section \ref{dsl-unified-msg} illustrated this approach with a case for unified
-messaging: parsing of incoming messages and grouping of outgoing messages.
-
-Other common functionality isn't generated as is either. The execution of
+Previously we illustrated this approach with a case for unified messaging:
+parsing of incoming messages and grouping of outgoing messages.
+
+More common functionality isn't generated verbatim either. The execution of
 functions based on a schedule or the implementation of higher level conceptual
@@
 constructions that are hard to identify and reorganize, such as loops and
-common data. A DSL, focussing on event-handling, enforces these restrictions
-and enables formally describing ID algorithms.
+common data. A DSL, focusing on event-handling, enforces these restrictions and
+enables formally describing ID algorithms.
 
@@
 In this section we present a prototype implementation and evaluate its
-performance. We first present the implementation consisting of the hardware
-platform, network and software stack. We first take an analytical approach and
-look at the handling of common data and the size of the source code. Next, we
-use the implementation to quantify its merits using a side-by-side comparison
-of generated code versus a manual implementation. We consider experimental
-results regarding usage of the network, execution time of the event-loop and
-the image size.
+performance. We start by introducing the implementation, consisting of the
+hardware platform, network and software stack. We then continue with an
+analytical approach, looking at the handling of common data and the size of the
+source code. Next, we use the implementation to quantify its merits using a
+side-by-side comparison of generated code versus a manual implementation. We
+consider experimental results regarding usage of the network, execution time of
+the event-loop and image size.
 
@@
 that the requirements of the generator towards the platform are minimal, don't
-rely on advanced frameworks or operating systems and therefore are applicable
+rely on advanced frameworks or operating systems, and therefore are applicable
 in any environment, targeting any platform.
@@
 application was an includable piece of code, which allowed integration in a
-manual implementation and simple merging when generation. On top of this
-baseline we added two intrusion detection related algorithms: a
-watchdog \cite{mishra2004intrusion}, that allows nodes to validate each other's
+manual implementation and easy merging when generating. On top of this baseline
+we added two intrusion detection related algorithms: a watchdog
+\cite{mishra2004intrusion}, that allows nodes to validate each other's
 presence, and a reputation-based algorithm \cite{ganeriwal2008reputation}, that
@@
 the formal description of ID algorithms. This can be an indication of the
-development effort and complexity. In this experiment, implementing both
-algorithms manually results in 637 lines of C code. The corresponding \NAME
-implementation results in 156 lines of DSL code. This is largely thanks to the
-higher level coding constructs, as introduced in section \ref{language-support}
-dealing with \NAME's language support.
+development effort and complexity. Implementing both algorithms manually
+results in 637 lines of C code. The corresponding \NAME implementation results
+in 156 lines of DSL code. This is largely thanks to the higher level coding
+constructs, as introduced in section \ref{design}, dealing with \NAME's
+language support.
 
@@
 bytes. In our implementation we used the ZigBee network address, which indeed
-consists of 16 bits. This results in a reduction of 7.4\%. Using ZigBee's 64
-bits unique addresses, this reduction can easily become 20.5\%.
+consists of 16 bits. This results in a reduction of 7.4\% of dynamic memory per
+known device. Using ZigBee's 64 bits unique addresses, this reduction can
+easily become 20.5\%.
 
@@
 
-Besides this analysis we also instrumented the code and evaluated four
-criteria: the network usage in number of packets and bytes, the time required
-to perform once cycle of the event loop and the image size of the resulting
-compiled code. To do this, we collected these metrics for four situations:
-without ID algorithms, with a watchdog, with reputation-tracking and with both
-algorithms implemented.
+Besides this analytical approach, we also instrumented the code and evaluated
+four criteria: the network usage in number of packets and bytes, the time
+required to perform once cycle of the event loop and the image size of the
+resulting compiled code. To do this, we collected these metrics for four
+situations: without ID algorithms, with a watchdog, with reputation-tracking
+and with both algorithms implemented.
 
@@
 Table \ref{tbl:results} shows the collected data for the manual and the
-generated implementation, as well as a comparison in absolute values and
-relative.
+generated implementation, as well as a comparison in absolute values and in
+relative terms.
 
@@
 impact of each algorithm by itself. The execution time of one cycle of the
-event loop with both algorithms enabled is even higher than simply the sum.
+event loop with both algorithms enabled is even higher than simply the sum of
+their individual times.
 
@@
 The first observation deals with the base case: both the manual and the
-generated implementations perform exactly the same. There is no fundamental
-structural difference in the overall architecture of the code.
-
-Secondly, when turning to the goal to reduce the network usage, we see that
+generated implementations perform in exactly the same way. There is no
+fundamental structural difference in the overall architecture of the code.
+
+Secondly, when turning to the goal of reducing the network usage, we see that
 both the number of packets as well as the amount of bytes sent have dropped by
@@
 Thirdly, focusing on the execution time, both algorithms by themselves add an
-equal amount of time to the processing of a single event loop cycle. But when
+equal amount of time to the processing of a single event loop cycle. When
 combined, the event loop is about 7\% faster compared to the manual case.
@@
 Finally, we see that the software library that comes with the generated code
-adds to the size of the resulting image. But the cost is close to constant and
-is lower for both algorithms together. With roughly 3KB of additional overhead,
-or 15\% in this simple case, the impact is small. Taking into account the
-available memory on the ATMega1284p, this is only 2.3\% of the available
+adds to the size of the resulting image. However, the cost is close to constant
+and is lower for both algorithms together. With roughly 3KB of additional
+overhead, or 15\% in this simple case, the impact is small. Taking into account
+the available memory on the ATMega1284p, this is only 2.3\% of the available
 resources.
@@
 description of algorithms in general. The Basic-like syntax does enable more
-productive development. But this approach doesn't address the underlying
+productive development. Still, this approach doesn't address the underlying
 resource problems inherently connected to the context.
@@
 on novice programmers. It is a clear fit for \emph{sense-and-send} type of
-applications. It does however share the idea of events to overcome the technical
-nature of loops.
+applications. It does however share the idea of using events to overcome the
+technical nature of loops.
 
@@
 optimizations \cite{panda2001data} can result in improvements on a lower level.
-But really preserving energy requires higher-level strategies, targeting
+To really preserve energy one requires higher-level strategies, targeting
 architecture, radio handling, efficient software,\dots \cite{naik2001software}.
@@
 
-Applying code generation to ID, often focusses on the generation of native code
+Applying code generation to ID often focusses on the generation of native code
 to avoid the interpretation of descriptions of patterns to look for. I2S
@@
 requirements or security threats, requires a flexible way to incorporate the
-different algorithms.
+algorithms.
 
@@
 described. The code generator applies functional code fusion to interpret the
-intent of algorithms and merges common functionality and shared data. The
+intent of algorithms and merges common functionality \& shared data. The
 software library hosts common functionality, such as network message parsing
@@
 
-Initial experiments confirm the theoretical positive impact on resource usage:
-grouping of messages lowers the number of packets by 13\%, while the total
-number of bytes sent over the network is reduced by 7\%. The execution time is
-improved by 7\% and memory usage can be reduced by even 20\%.
+Initial experiments confirm the theoretically positive impact on resource
+usage: grouping of messages lowers the number of packets by 13\%, while the
+total number of bytes sent over the network is reduced by 7\%. The execution
+time is improved by 7\% and memory usage can be reduced by even 20\%.
 
@@
 As a next step, we will implement more ID algorithms using \NAME and continue
-the experiments on a larger scale. This way we will also be able to verify the
+experiments on a larger scale. This way we will also be able to verify the
 initially observed inclination in the evolution of the image size with growing
@@
 Future research will focus on the integration of applications with a generated
-IDS. We see opportunities for applications who detect intrusions, to actively
+IDS. We see opportunities for applications that detect intrusions, to actively
 and dynamically respond and avoid malicious payloads.

### 2014-12-08 paper.tex
applied some more fixes by Wouter

@@
 The internet of things (IoT) holds great potential to positively influence our
-daily work and life through domotics, assisted living, e-health and enhanced
-learning \cite{atzori2010internet}. With this potential the IoT also presents a
-significant threat: by opening our smart homes and personal data to all of
-these interconnected devices, we also open them to everyone who is able to
-break the virtual locks that protect them (i.e.\ passwords \& encryption keys).
-Implementation of security measures by itself is hindered by the fact that IoT
-devices typically have limited batteries, processing power and memory.
+daily work and life through domotics, assisted living, e-health, enhanced
+learning, etc. \cite{atzori2010internet}. With this potential the IoT also
+presents a significant threat: by opening our smart homes and personal data to
+all of these interconnected devices, we also open them to everyone who is able
+to break the virtual locks that protect them (i.e.\ passwords \& encryption
+keys). Implementation of security measures by itself is hindered by the fact
+that IoT devices typically have limited batteries, processing power and memory.
 
@@
 patterns or anomalies. An IDS for IoT devices therefore requires a large number
-of algorithms for an adequate coverage.
+of algorithms to achieve adequate coverage.
 
@@
 \cite{krontiris2008lidea}. A key shortcoming of prior ID frameworks is that
-they don't offer a way to optimize the usage of resources. A solution that
+they do not offer a way to optimize the usage of resources. A solution that
 supports the integration of multiple algorithms on IoT devices needs to avoid
@@
 is capable of generating code even for the most basic environment. It shows
-that the requirements of the generator towards the platform are minimal, don't
+that the requirements of the generator towards the platform are minimal, do not
 rely on advanced frameworks or operating systems, and therefore are applicable
@@
 \cite{mishra2004intrusion,ioannis2007towards,alrajeh2013intrusion} based on
-their detection approach and interaction style.
+detection approach and interaction style.
 
@@
 framework and the solution it proposes. We also look at their application in
-other disciplines and how this can relate to our context.
+other disciplines and analyze how this can relate to our context.
 

### 2015-04-15 ecrc-procs/procs-template-generic.tex
added Elsevier Procedia style + converted paper

% Template for Elsevier CRC journal article
% version 1.2 dated 09 May 2011

% This file (c) 2009-2011 Elsevier Ltd.  Modifications may be freely made,
% provided the edited file is saved under a different name

% This file contains modifications for Procedia Computer Science

% Changes since version 1.1
% - added "procedia" option compliant with ecrc.sty version 1.2a
%   (makes the layout approximately the same as the Word CRC template)
% - added example for generating copyright line in abstract

%-----------------------------------------------------------------------------------

%% This template uses the elsarticle.cls document class and the extension package ecrc.sty
%% For full documentation on usage of elsarticle.cls, consult the documentation "elsdoc.pdf"
%% Further resources available at http://www.elsevier.com/latex

%-----------------------------------------------------------------------------------

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                          %%
%% Important note on usage                                  %%
%% -----------------------                                  %%
%% This file should normally be compiled with PDFLaTeX      %%
%% Using standard LaTeX should work but may produce clashes %%
%%                                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% The '3p' and 'times' class options of elsarticle are used for Elsevier CRC
%% The 'procedia' option causes ecrc to approximate to the Word template
\documentclass[3p,times,procedia]{elsarticle}
\flushbottom

%% The `ecrc' package must be called to make the CRC functionality available
\usepackage{ecrc}
%\usepackage{amsmath}

%% The ecrc package defines commands needed for running heads and logos.
%% For running heads, you can set the journal name, the volume, the starting page and the authors

%% set the volume if you know. Otherwise `00'
\volume{00}

%% set the starting page if not 1
\firstpage{1}

%% Give the name of the journal
\journalname{Procedia Computer Science}

%% Give the author list to appear in the running head
%% Example \runauth{C.V. Radhakrishnan et al.}
\runauth{Author name}

%% The choice of journal logo is determined by the \jid and \jnltitlelogo commands.
%% A user-supplied logo with the name <\jid>logo.pdf will be inserted if present.
%% e.g. if \jid{yspmi} the system will look for a file yspmilogo.pdf
%% Otherwise the content of \jnltitlelogo will be set between horizontal lines as a default logo

%% Give the abbreviation of the Journal.
\jid{procs}

%% Give a short journal name for the dummy logo (if needed)
%\jnltitlelogo{Procedia Computer Science}

%% Hereafter the template follows `elsarticle'.
%% For more details see the existing template files elsarticle-template-harv.tex and elsarticle-template-num.tex.

%% Elsevier CRC generally uses a numbered reference style
%% For this, the conventions of elsarticle-template-num.tex should be followed (included below)
%% If using BibTeX, use the style file elsarticle-num.bst

%% End of ecrc-specific commands
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% The amssymb package provides various useful mathematical symbols

\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers after \end{frontmatter}.
%% \usepackage{lineno}

%% natbib.sty is loaded by default. However, natbib options can be
%% provided with \biboptions{...} command. Following options are
%% valid:

%%   round  -  round parentheses are used (default)
%%   square -  square brackets are used   [option]
%%   curly  -  curly braces are used      {option}
%%   angle  -  angle brackets are used    <option>
%%   semicolon  -  multiple citations separated by semi-colon
%%   colon  - same as semicolon, an earlier confusion
%%   comma  -  separated by comma
%%   numbers-  selects numerical citations
%%   super  -  numerical citations as superscripts
%%   sort   -  sorts multiple citations according to order in ref. list
%%   sort&compress   -  like sort, but also compresses numerical citations
%%   compress - compresses without sorting
%%
%\biboptions{sort&compress}

% \biboptions{}

% if you have landscape tables
\usepackage[figuresright]{rotating}
%\usepackage{harvard}
% put your own definitions here:x
%   \newcommand{\cZ}{\cal{Z}}
%   \newtheorem{def}{Definition}[section]
%   ...

% add words to TeX's hyphenation exception list
%\hyphenation{author another created financial paper re-commend-ed Post-Script}

% declarations for front matter

\begin{document}

\begin{frontmatter}

%% Title, authors and addresses

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for the associated footnote;
%% use the fnref command within \author or \address for footnotes;
%% use the fntext command for the associated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for the associated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%%
%% \title{Title\tnoteref{label1}}
%% \tnotetext[label1]{}
%% \author{Name\corref{cor1}\fnref{label2}}
%% \ead{email address}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
%% \address{Address\fnref{label3}}
%% \fntext[label3]{}

\dochead{6th International Conference on Emerging Ubiquitous Systems and Pervasive Networks, EUSPN-2015 and the 5th International Conference on Current and Future Trends of Information and Communication Technologies in Healthcare, ICTH 2015,}
%% Use \dochead if there is an article header, e.g. \dochead{Short communication}
%% \dochead can also be used to include a conference title, if directed by the editors
%% e.g. \dochead{17th International Conference on Dynamical Processes in Excited States of Solids}

\title{Click here, type the title of your paper, Capitalize first letter}

%% use optional labels to link authors explicitly to addresses:
%% \author[label1,label2]{<author name>}
%% \address[label1]{<address>}
%% \address[label2]{<address>}

\author[a]{First Author}
\author[b]{Second Author}
\author[a,b]{Third Author\corref{cor1}}

\address[a]{First affiliation, Address, City and Postcode, Country}
\address[b]{Second affiliation, Address, City and Postcode, Country}

\begin{abstract}
%% Text of abstract
Click here and insert your abstract text.
\end{abstract}

\begin{keyword}
Type your keywords here, separated by semicolons ;

%% keywords here, in the form: keyword \sep keyword

%% PACS codes here, in the form: \PACS code \sep code

%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)

\end{keyword}
\cortext[cor1]{Corresponding author. Tel.: +0-000-000-0000 ; fax: +0-000-000-0000.}
\end{frontmatter}

%\correspondingauthor[*]{Corresponding author. Tel.: +0-000-000-0000 ; fax: +0-000-000-0000.}
\email{author@institute.xxx}

%%
%% Start line numbering here if you want
%%
% \linenumbers

%% main text

\vspace*{-10pt}

%\enlargethispage{-7mm}
\section{Main Text}
\label{main}

Here introduce the paper, and put a nome¬nclature if necessary, in a box with the same font size as the rest of the paper. The paragraphs continue from here and are only separated by headings, subheadings, images and formulae. The section headings are arranged by numbers, bold and 10 pt. Here follows further instructions for authors.

\begin{nomenclature}
\begin{deflist}[A]
\defitem{A}\defterm{radius of}
\defitem{B}\defterm{position of}
\defitem{C}\defterm{further nomenclature continues down the page inside the text box\vspace*{-8pt}}
\end{deflist}
\end{nomenclature}
\vspace*{-2pt}

\subsection{ Structure}
Files must be in LaTeX format only and should be formatted for direct printing, using the CRC LaTeX template provided. Figures and tables should be embedded and not supplied separately.

Please make sure that you use as much as possible normal fonts in your documents. Special fonts, such as fonts used in the Far East (Japanese, Chinese, Korean, etc.) may cause problems during processing. To avoid unnecessary errors you are strongly advised to use the `spellchecker' function of TeX Editor. Follow this order when typing manuscripts: Title, Authors, Affiliations, Abstract, Keywords, Main text (including figures and tables), Acknowledgements, References, Appendix. Collate acknowledgements in a separate section at the end of the article and do not include them on the title page, as a footnote to the title or otherwise.

Bulleted lists may be included and should look like this:
\begin{itemize}[]
\item First point
\item Second point
\item And so on
\end{itemize}

Ensure that you return to the `Els-body-text' style, the style that you will mainly be using for large blocks of text, when you have completed your bulleted list.

Please do not alter the formatting and style layouts which have been set up in this template document. As indicated in the template, papers should be prepared in single column format suitable for direct printing onto paper with trim size $192 \times 262$ mm. Do not number pages on the front, as page numbers will be added separately for the preprints and the Proceedings. Leave a line clear between paragraphs. All the required style templates are provided in the file ``LaTeX Template'' with the appropriate name supplied, e.g. choose 1. Els1st-order-head for your first order heading text, els-abstract-text for the abstract text etc.

\subsection{ Tables}

All tables should be numbered with Arabic numerals. Every table should have a
caption. Headings should be placed above tables, left justified. Only
horizontal lines should be used within a table, to distinguish the column
headings from the body of the table, and immediately above and below the table.
Tables must be embedded into the text and not supplied separately. Below is an
example which the authors may find useful.

\begin{table}[h]
\caption{An example of a table.}
\begin{tabular*}{\hsize}{@{\extracolsep{\fill}}lll@{}}
\toprule
An example of a column heading & Column A ({\it{t}}) & Column B ({\it{t}})\\
\colrule
And an entry &   1 &  2\\
And another entry  & 3 &  4\\
And another entry &  5 &  6\\
\botrule
\end{tabular*}
\end{table}

%\enlargethispage{12pt}

\subsection{Construction of references}

References must be listed at the end of the paper. Do not begin them on a new page unless this is absolutely necessary. Authors should ensure that every reference in the text appears in the list of references and vice versa. Indicate references by\kern-1.5pt\cite{Vander} or \cite{Strunk,Mettam} in the text.

Some examples of how your references should be listed are given at the end of this template in the `References' section, which will allow you to assemble your reference list according to the correct format and font size.

Reference generation by using bibliography style commands for LaTeX template only.

The author may use ``model3a-num-names.bst'' as per the style required in document. The sample bib file could be referred.
If the author may using bibstyle for providing references author must comment the bibliography section in TeX file, Bibtex will generate the reference automatically.

If the author may not able to view the references in output same could be done by copying the bibliography section from ``filename.bbl'' file and paste in TeX file.

\subsection{Section headings}
Section headings should be left justified, bold, with the first letter capitalized and numbered consecutively, starting with the Introduction. Sub-section headings should be in capital and lower-case italic letters, numbered 1.1, 1.2, etc,~and left justified, with second~and subsequent lines indented. All headings should have a minimum of two text lines after them before a page or column break.
Ensure the text area is not blank except for the last page.

\subsection{General guidelines for the preparation of your text}
Avoid hyphenation at the end of a line. Symbols denoting vectors and matrices should be indicated in bold type. Scalar variable names should normally be expressed using italics. Weights and measures should be expressed in SI units. All non-standard abbreviations or symbols must be defined when first mentioned, or a glossary provided.

\subsection{File naming and delivery}
Please title your files in this order `procedia acronym\_conference acronym\_authorslastname'.~Submit both the source file and the PDF to the Guest Editor.

Artwork filenames should comply with the syntax ``aabbbbbb.ccc'', where:\vspace*{-12pt}
\begin{itemize}
\item a = artwork component type
\item b = manuscript reference code
\item c = standard file extension

Component types:
\item gr = figure
\item pl = plate
\item sc = scheme
\item fx = fixed graphic
\end{itemize}

\subsection{Footnotes}
Footnotes should be avoided if possible. Necessary footnotes should be denoted in the text by consecutive superscript letters\footnote{Footnote text.}. The footnotes should be typed single spaced, and in smaller type size (8 pt), at the foot of the page in which they are mentioned, and separated from the main text by a one line space extending at the foot of the column. The `Els-footnote' style is available in the ``TeX Template'' for the text of the footnote.

Please do not change the margins of the template as this can result in the footnote falling outside printing range.

\section{Illustrations}
All figures should be numbered with Arabic numerals (1,2,3,\,$\ldots.$). Every figure should have a caption. All\break photographs, schemas, graphs and diagrams are to be referred to as figures. Line drawings should be good quality\break scans or true electronic output. Low-quality scans are not acceptable. Figures must be embedded into the text and not supplied separately. In MS word input the figures must be properly coded. Preferred format of figures are PNG, JPEG, GIF etc. Lettering and symbols should be clearly defined either in the caption or in a legend provided as part of the figure. Figures should be placed at the top or bottom of a page wherever possible, as close as possible to the first reference to them in the paper. Please ensure that all the figures are of 300 DPI resolutions as this will facilitate good output.
\begin{figure}[t]\vspace*{4pt}
%\centerline{\includegraphics{fx1}\hspace*{5mm}\includegraphics{fx1}}
\centerline{\includegraphics{gr1}}
\caption{(a) first picture; (b) second picture.}\vspace*{-6pt}
\end{figure}

The figure number and caption should be typed below the illustration in 8 pt and left justified [{{\bfseries\itshape Note:}} one-line captions of length less than column width (or full typesetting width or oblong) centered]. For more guidelines and information to help you submit high quality artwork please visit: http://www.elsevier.com/artworkinstructions\break Artwork has no text along the side of it in the main body of the text. However, if two images fit next to each other, these may be placed next to each other to save space. For example, see Fig.~1.

\section{Equations}
Equations and formulae should be typed in Mathtype, and numbered consecutively with Arabic numerals in parentheses on the right hand side of the page (if referred to explicitly in the text). They should also be separated from the surrounding text by one space
\begin{equation}
\begin{array}{lcl}
\displaystyle X_r &=& \displaystyle\dot{Q}^{''}_{rad}\left/\left(\dot{Q}^{''}_{rad} + \dot{Q}^{''}_{conv}\right)\right.\\[6pt]
\displaystyle \rho &=& \displaystyle\frac{\vec{E}}{J_c(T={\rm const.})\cdot\left(P\cdot\left(\displaystyle\frac{\vec{E}}{E_c}\right)^m+(1-P)\right)}
\end{array}
\end{equation}

\vspace*{-6pt}
\section{Online license transfer}
All authors are required to complete the Procedia exclusive license transfer agreement before the article can be published, which they can do online. This transfer agreement enables Elsevier to protect the copyrighted material for the authors, but does not relinquish the authors' proprietary rights. The copyright transfer covers the exclusive rights to reproduce and distribute the article, including reprints, photographic reproductions, microfilm or any other reproductions of similar nature and translations. Authors are responsible for obtaining from the copyright holder, the permission to reproduce any figures for which copyright exists.

\section*{Acknowledgements}

Acknowledgements and Reference heading should be left justified, bold, with the first letter capitalized but have no numbers. Text below continues as normal.

%% The Appendices part is started with the command \appendix;
%% appendix sections are then done as normal sections
%% \appendix

%% \section{}
%% \label{}

\appendix
\section{An example appendix}
Authors including an appendix section should do so before References section. Multiple appendices should all have headings in the style used above. They will automatically be ordered A, B, C etc.

\subsection{Example of a sub-heading within an appendix}
There is also the option to include a subheading within the Appendix if you wish.
%% References
%%
%% Following citation commands can be used in the body text:
%% Usage of \cite is as follows:
%%   \cite{key}         ==>>  [#]
%%   \cite[chap. 2]{key} ==>> [#, chap. 2]
%%

%The citation must be used in following style: \cite{article-minimal} \cite{article-full} \cite{article-crossref} \cite{whole-journal}.
%% References with BibTeX database:

%\bibliography{xampl}
%\bibliographystyle{model3a-num-names}

%% Authors are advised to use a BibTeX database file for their reference list.
%% The provided style file elsarticle-num.bst formats references in the required Procedia style

%% For references without a BibTeX database:

\begin{thebibliography}{}

%% \bibitem must have the following form:
%%   \bibitem{key}...
%%
\bibitem{Vander}
Van der Geer J, Hanraads JAJ, Lupton RA. The art of writing a
scientific article. {\it J Sci Commun} 2000;{\bf 163}:51-9.
\bibitem{Strunk}
Strunk Jr W, White EB. {\it The elements of style}. 3rd ed. New York:
Macmillan; 1979.
\bibitem{Mettam}
Mettam GR, Adams LB. How to prepare an electronic version of your article. In: Jones BS, Smith RZ, editors. {\it Introduction to the electronic age}. New York: E-Publishing Inc; 1999. p. 281-304.

 \end{thebibliography}

\clearpage

%%%% This page is for instructions only, once the article is finalize please omit the below text before creating the final PDF
\normalMode

\section*{Instructions to Authors for LaTeX template:}

\section{ZIP mode for LaTeX template:}

The zip package is created as per the guide lines present on the URL http://www.elsevier.com/author-schemas/ preparing-crc-journal-articles-with-latex for creating the LaTeX zip file of Procedia LaTeX template.  The zip generally contains the following files:
\begin{Itemize}[]\leftskip-17.7pt\labelsep3.3pt
\item ecrc.sty
\item  elsarticle.cls
\item elsdoc.pdf
\item .bst file
\item Manuscript templates for use with these bibliographic styles
\item  Generic and journal specific logos, etc.
\end{Itemize}

The LaTeX package is the main LaTeX template. All LaTeX support files are required for LaTeX pdf generation from the LaTeX template package.

{\bf Reference style .bst file used for collaboration support:} In the LaTeX template packages of all Procedia titles a new ``.bst'' file is used which supports collaborations downloaded from the path http://www.elsevier.com/author-schemas/the-elsarticle-latex-document-class

\section{Reference styles used in  Procedia master templates:}
\let\footnotesize\normalsize
\hspace*{-10pt}\begin{tabular*}{\hsize}{@{}ll@{}}
{\bf Title}&{\bf Reference style} \\[6pt]
AASPRO  & 2 Harvard\\
AASRI Procedia  & 3 Vancouver Numbered\\
APCBEE Procedia  & 3 Vancouver Numbered\\
EGYPRO  & 3 Vancouver Numbered\\
FINE    & 2 Harvard\\
IERI Procedia  & 3 Vancouver Numbered\\
MATPR  & 1a Numbered without article titles\\
MSPRO  & 2 Harvard\\
PHPRO  & 2 Harvard\\
PIUTAM  & 3a Embellished Vancouver \\
Procedia CIRP  & 3 Vancouver Numbered\\
PROCHE  & 3a Embellished Vancouver \\
PROCS  & 3a Embellished Vancouver \\
PROENG  & 1 Numbered\\
PROENV  & 3a Embellished Vancouver \\
PROEPS  & 3a Embellished Vancouver \\
PROFOO    & 3a Embellished Vancouver \\
PROMFG  & 1a Numbered without article titles\\
PROTCY  & 3 Vancouver Numbered\\
PROVAC  & 3a Embellished Vancouver \\
SBSPRO  & 5 APA\\
SEPRO  & 3a Embellished Vancouver \\
AQPRO & 2 Harvard\\
UMKPRO & 5 APA\\
\end{tabular*}

\end{document}

%%
%% End of file `procs-template.tex'.

### 2015-04-15 paper-euspn15.tex
added Elsevier Procedia style + converted paper

--- 2014-12-08 paper.tex
@@
-\documentclass[conference]{IEEEtran}
-\usepackage[pdftex]{graphicx}
-\usepackage[cmex10]{amsmath}
-\interdisplaylinepenalty=2500
-\usepackage{array}
-\usepackage{mdwmath}
-\usepackage{mdwtab}
-\usepackage{eqparbox}
-\usepackage[font=footnotesize]{subfig}
-\usepackage{fixltx2e}
-\usepackage{url}
-\hyphenation{}
+% This paper will be submitted to EUSPN15.
+
+\documentclass[3p,times,procedia]{elsarticle}
+\flushbottom
+\usepackage{ecrc}
+\volume{00}
+\firstpage{1}
+\journalname{Procedia Computer Science}
+\runauth{Christophe Van Ginneken, Jef Maerien,
+Christophe Huygens, Wouter Joosen, Danny Hughes}
+\jid{procs}
+\usepackage{amssymb}
+\usepackage[figuresright]{rotating}
 
@@
 
+\begin{frontmatter}
+
+\dochead{6th International Conference on Emerging Ubiquitous Systems and
+Pervasive Networks, EUSPN-2015 and the 5th International Conference on Current
+and Future Trends of Information and Communication Technologies in Healthcare,
+ICTH 2015,}
+
 \title{
-\NAME: A framework for \\
-Efficient Intrusion Detection\\
-in the Internet of Things
+\NAME: A framework for Efficient Intrusion Detection in the Internet of Things
 }
 
-\author{\IEEEauthorblockN{Christophe Van Ginneken, Jef Maerien,
-Christophe Huygens, Wouter Joosen, Danny Hughes}%
-\IEEEauthorblockA{iMinds-DistriNet, KU Leuven\\
-3001 Leuven, Belgium\\
-\{firstname.lastname\}@cs.kuleuven.be}}
-
-\maketitle
+\author[a]{Christophe Van Ginneken}
+\author[a]{Jef Maerien}
+\author[a]{Christophe Huygens}
+\author[a]{\\Wouter Joosen}
+\author[a]{Danny Hughes}
+
+\address[a]{
+iMinds-DistriNet, KU Leuven, 3001 Leuven, Belgium\\
+\{firstname.lastname\}@cs.kuleuven.be
+}
 
@@
 
+\begin{keyword}
+
+IoT, Domain Specific Languages, Code Generation, Source Code Optimization,
+Intrusion Detection
+
+\end{keyword}
+
+\end{frontmatter}
+
+\vspace*{-10pt}
+
 \section{Introduction}
@@
 
-\begin{LaTeXdescription}
+\begin{description}
 
@@
 
-\end{LaTeXdescription}
+\end{description}
 
@@
 
-\begin{figure}[ht]
+\begin{figure}[t]
   \centering
@@
 
-\begin{figure}[ht]
-\captionof{lstlisting}{ID Algorithm Pattern\label{alg:id-algo-pattern}}
-\begin{algorithmic}[1]
+\begin{algorithm}[t]
+  \caption{ID Algorithm Pattern}
+  \label{alg:id-algo-pattern}
+  \begin{algorithmic}
     \Require{devices: global variable for device information}
@@
     \EndFunction
-  \State
   \end{algorithmic}
-\end{figure}
+\end{algorithm}
 
@@
 
-\begin{figure}[ht]
-\captionof{lstlisting}{Application pattern of multiple ID algorithms\label{alg:id-algo-application}}
-\begin{algorithmic}[1]
+\begin{algorithm}[t]
+  \caption{Application pattern of multiple ID algorithms}
+  \label{alg:id-algo-application}
+  \begin{algorithmic}
     \Let{$msg$}{$network$} \Comment{all observed messages}
@@
   \end{algorithmic}
-\end{figure}
+\end{algorithm}
 
@@
 
-\begin{figure*}[b]
+\begin{figure}[b]
   \centering
@@
 }
-
 \caption{The high level design of \NAME, showing the resources, processes and
@@
 \label{fig:design}
-\end{figure*}
+\end{figure}
 
@@
 }
-
 \caption{Partial Semantic Model for \NAME. Shaded classes indicate the
@@
 of the intended functionality.}
-
 \label{fig:meta-model}
@@
 
-% table goes here, because a spanning table is placed on the next page, relative
-% to where it's defined.
-\begin{table*}[b]
-  \centering
-  \begin{tabular}{llrrrrrrr}
+\section{Implementation and Evaluation}
+\label{evaluation}
+
+In this section we present a prototype implementation and evaluate its
+performance. We start by introducing the implementation, consisting of the
+hardware platform, network and software stack. We then continue with an
+analytical approach, looking at the handling of common data and the size of the
+source code. Next, we use the implementation to quantify its merits using a
+side-by-side comparison of generated code versus a manual implementation. We
+consider experimental results regarding usage of the network, execution time of
+the event-loop and image size.
+
+\subsection{Implementation}
+
+To evaluate the prototype we used a system based on the Atmel ATMEGA1284p
+micro-controller and the Digi XBee S2 ZigBee module. A basic software library
+wraps the technical calls to the hardware components in more easy to use
+functions.
+
+Relying on basic hardware and software allows us to validate that the generator
+is capable of generating code even for the most basic environment. It shows
+that the requirements of the generator towards the platform are minimal, do not
+rely on advanced frameworks or operating systems, and therefore are applicable
+in any environment, targeting any platform.
+
+For the evaluation we constructed a small meshed network. In this network, not
+all devices have direct access to one another, requiring routing through shared
+neighboring devices.
+
+We started from a basic application that measures light intensity. The
+application was an includable piece of code, which allowed integration in a
+manual implementation and easy merging when generating. On top of this baseline
+we added two intrusion detection related algorithms: a watchdog
+\cite{mishra2004intrusion}, that allows nodes to validate each other's
+presence, and a reputation-based algorithm \cite{ganeriwal2008reputation}, that
+checks if a parent node is cooperative and actually forwards messages further
+upstream.
+
+\subsection{Analytical Results}
+
+Looking at the generated source code is a straightforward way to validate the
+framework. A simple metric is comparing the amount of code versus the size of
+the formal description of ID algorithms. This can be an indication of the
+development effort and complexity. Implementing both algorithms manually
+results in 637 lines of C code. The corresponding \NAME implementation results
+in 156 lines of DSL code. This is largely thanks to the higher level coding
+constructs, as introduced in section \ref{design}, dealing with \NAME's
+language support.
+
+Next we look at data structures. Sharing data eliminates redundant information,
+such as a network address, from the scope of each individual algorithm.
+Analysis of the source code with both algorithms implemented, shows that this
+reduces the memory footprint for one node entry by 2 bytes, from 27 bytes to 25
+bytes. In our implementation we used the ZigBee network address, which indeed
+consists of 16 bits. This results in a reduction of 7.4\% of dynamic memory per
+known device. Using ZigBee's 64 bits unique addresses, this reduction can
+easily become 20.5\%.
+
+\subsection{Experimental Results}
+
+Besides this analytical approach, we also instrumented the code and evaluated
+four criteria: the network usage in number of packets and bytes, the time
+required to perform once cycle of the event loop and the image size of the
+resulting compiled code. To do this, we collected these metrics for four
+situations: without ID algorithms, with a watchdog, with reputation-tracking
+and with both algorithms implemented.
+
+The manual implementation realized both algorithms as standalone modules and
+sequentially called them from the base-application's event loop. Given the
+\NAME descriptions of the algorithms, the code generator generated all four
+cases.
+
+Table \ref{tbl:results} shows the collected data for the manual and the
+generated implementation, as well as a comparison in absolute values and in
+relative terms.
+
+\begin{table}[b]
+  \caption{Results and Side-by-Side Comparison.}
+  \label{tbl:results}
+\begin{tabular*}{\hsize}{@{\extracolsep{\fill}}llrrrrrrr@{}}
 \toprule
@@
 \bottomrule
-  \end{tabular}
-  \caption{Results and Side-by-Side Comparison.}
-  \label{tbl:results}
-\end{table*}
-
-\section{Implementation and Evaluation}
-\label{evaluation}
-
-In this section we present a prototype implementation and evaluate its
-performance. We start by introducing the implementation, consisting of the
-hardware platform, network and software stack. We then continue with an
-analytical approach, looking at the handling of common data and the size of the
-source code. Next, we use the implementation to quantify its merits using a
-side-by-side comparison of generated code versus a manual implementation. We
-consider experimental results regarding usage of the network, execution time of
-the event-loop and image size.
-
-\subsection{Implementation}
-
-To evaluate the prototype we used a system based on the Atmel ATMEGA1284p
-micro-controller and the Digi XBee S2 ZigBee module. A basic software library
-wraps the technical calls to the hardware components in more easy to use
-functions.
-
-Relying on basic hardware and software allows us to validate that the generator
-is capable of generating code even for the most basic environment. It shows
-that the requirements of the generator towards the platform are minimal, do not
-rely on advanced frameworks or operating systems, and therefore are applicable
-in any environment, targeting any platform.
-
-For the evaluation we constructed a small meshed network. In this network, not
-all devices have direct access to one another, requiring routing through shared
-neighboring devices.
-
-We started from a basic application that measures light intensity. The
-application was an includable piece of code, which allowed integration in a
-manual implementation and easy merging when generating. On top of this baseline
-we added two intrusion detection related algorithms: a watchdog
-\cite{mishra2004intrusion}, that allows nodes to validate each other's
-presence, and a reputation-based algorithm \cite{ganeriwal2008reputation}, that
-checks if a parent node is cooperative and actually forwards messages further
-upstream.
-
-\subsection{Analytical Results}
-
-Looking at the generated source code is a straightforward way to validate the
-framework. A simple metric is comparing the amount of code versus the size of
-the formal description of ID algorithms. This can be an indication of the
-development effort and complexity. Implementing both algorithms manually
-results in 637 lines of C code. The corresponding \NAME implementation results
-in 156 lines of DSL code. This is largely thanks to the higher level coding
-constructs, as introduced in section \ref{design}, dealing with \NAME's
-language support.
-
-Next we look at data structures. Sharing data eliminates redundant information,
-such as a network address, from the scope of each individual algorithm.
-Analysis of the source code with both algorithms implemented, shows that this
-reduces the memory footprint for one node entry by 2 bytes, from 27 bytes to 25
-bytes. In our implementation we used the ZigBee network address, which indeed
-consists of 16 bits. This results in a reduction of 7.4\% of dynamic memory per
-known device. Using ZigBee's 64 bits unique addresses, this reduction can
-easily become 20.5\%.
-
-\subsection{Experimental Results}
-
-Besides this analytical approach, we also instrumented the code and evaluated
-four criteria: the network usage in number of packets and bytes, the time
-required to perform once cycle of the event loop and the image size of the
-resulting compiled code. To do this, we collected these metrics for four
-situations: without ID algorithms, with a watchdog, with reputation-tracking
-and with both algorithms implemented.
-
-The manual implementation realized both algorithms as standalone modules and
-sequentially called them from the base-application's event loop. Given the
-\NAME descriptions of the algorithms, the code generator generated all four
-cases.
-
-Table \ref{tbl:results} shows the collected data for the manual and the
-generated implementation, as well as a comparison in absolute values and in
-relative terms.
+  \end{tabular*}
+\end{table}
 
@@
 
-\bibliographystyle{IEEEtran}
+\bibliographystyle{elsarticle-num}
 \bibliography{literature/referenties}

### 2015-04-16 paper-euspn15.tex
cleaned up the paper up to 9 pages (1 to go)

@@
 \title{
-\NAME: A framework for Efficient Intrusion Detection in the Internet of Things
+\FOO: A framework for efficient source code generation\\
+of detection algorithms for the Internet of Things
 }
 
-\author[a]{Christophe Van Ginneken}
-\author[a]{Jef Maerien}
-\author[a]{Christophe Huygens}
-\author[a]{\\Wouter Joosen}
-\author[a]{Danny Hughes}
+\author[]{Christophe Van Ginneken}
+\author[]{Jef Maerien}
+\author[]{Christophe Huygens}
+\author[]{\\Wouter Joosen}
+\author[]{Danny Hughes}
 
-\address[a]{
+\address{
 iMinds-DistriNet, KU Leuven, 3001 Leuven, Belgium\\
@@
 
-When intrusion prevention fails, intrusion detection serves as a second layer
-of defense. It looks for patterns and anomalies that can indicate malicious
-behavior. However, supporting an adequate set of intrusion detection algorithms
-imposes significant overhead on resource-constrained devices such as those in
-the Internet of Things. Manual fusion of the algorithms reduces resource
-consumption by optimizing resource sharing, yet, proves to be time-consuming,
-repetitive and error-prone. To address this problem we propose \NAME, a
-framework for the development of efficient intrusion detection systems. It
-consists of a domain specific language and a code generator. The language
-allows formally describing the intent of intrusion detection algorithms and
-supports the generator in organizing the source code. A side-by-side comparison
-shows that \NAME-generated code reduces message passing overhead, execution
-time and memory footprint in comparison to sequential calls into individual
-implementations of the algorithms.
+The ability to detect patterns in its usage is what makes \emph{things}
+\emph{smart}. The more patterns it can detect, the more accurate its
+operational results will be. However, supporting an adequate set of pattern
+detection algorithms imposes significant overhead on resource-constrained
+devices such as those in the Internet of Things. Manual fusion of the
+algorithms reduces resource consumption by optimizing resource sharing, yet,
+proves to be time-consuming, repetitive and error-prone. To address this
+problem we propose \FOO, a framework for the generation of efficient detection
+algorithms. It consists of a domain specific language and a code generator. The
+language allows formally describing the intent of detection algorithms and
+supports the generator in organizing the source code. As a case study, we apply
+\FOO to the generation of an Intrusion Detection System for wireless sensor
+nodes. A side-by-side comparison shows that \FOO-generated source code reduces
+message passing overhead, execution time and memory footprint in comparison to
+sequential calls into stand-alone implementations of the algorithms.
 
@@
 
-IoT, Domain Specific Languages, Code Generation, Source Code Optimization,
-Intrusion Detection
+IoT, Domain Specific Languages, Source Code Generation, Source Code
+Optimization, Intrusion Detection, Context Awareness
 
@@
 
-% context
+The Internet of Things (IoT), the mother of all ubiquitous and pervasive
+networks, populates our daily lives with millions of internet-connected
+\emph{smart things}: from smart watches that assist visually impaired
+persons\cite{porzi2013smart} to monitoring systems for flood hazardous
+areas\cite{hughes2006gridstix}. Their applications cannot be more diverse,
+however they share one common concern: each of these \emph{things} is
+resource-constrained, relying on a battery and operating with limited
+processing power and memory.
 
-% wired networks: fw and central ids
+Although their applications at first seem unrelated, under the hood they share
+a common paradigm: pattern recognition. With the introduction of \emph{smart
+things}, we want to augment our lives with information that is hard, or at
+least time and resource consuming, to obtain otherwise. The ability to detect
+patterns in its usage or its applied context, is what makes these \emph{things}
+\emph{smart}. A watch that recognizes gestures can perform certain actions on
+behalf of its wearer. That same watch can also detect a wet floor and notify
+this hazard.
 
-In wired networks, firewalls focus on the outer perimeter of the network,
-filtering unwanted packets and protecting the entire internal network. Still,
-attacks on flaws in services can pass unnoticed. This is where intrusion
-detection (ID) comes into play: an intrusion detection system (IDS) monitors
-all traffic that passes through the firewall, looking for patterns of malicious
-activity, and optionally, after detecting such pattern, alerts the firewall,
-allowing it to take corrective actions \cite{denning1987intrusion}.
+Monitoring human activities is an active research field for more than a
+decade\cite{lara2013survey}, but with the advent of the IoT it now moves into
+the real world. Equipped with wearables with numerous sensors, the
+possibilities seem endless, but so are the number of detection and recognition
+algorithms that have to run simultaneously on these small \emph{things}, to
+support our augmented reality.
 
-% IoT + threat
-
-The internet of things (IoT) holds great potential to positively influence our
-daily work and life through domotics, assisted living, e-health, enhanced
-learning, etc. \cite{atzori2010internet}. With this potential the IoT also
-presents a significant threat: by opening our smart homes and personal data to
-all of these interconnected devices, we also open them to everyone who is able
-to break the virtual locks that protect them (i.e.\ passwords \& encryption
-keys). Implementation of security measures by itself is hindered by the fact
-that IoT devices typically have limited batteries, processing power and memory.
-
-% local ids -> localize -> limited resources
-
-In wireless networks of resource-constrained devices, which make up a
-significant part of the IoT, it is not possible for a single point in the
-network to oversee all traffic \cite{mishra2004intrusion}. Every device has to
-implement its own lines of defense. However making the IDS a local service on
-each device requires local resources, which are a scarce commodity for IoT
-devices.
-
-Attackers have access to a large set of possible attack vectors
-\cite{aschenbruck2012security}, ranging from the physical layer, through the
-access and routing layers, up to the application layer. Each layer presents
-different opportunities to manipulate data, eavesdrop or perform a form of
-denial of service. For each of these attacks an algorithm needs to look for
-patterns or anomalies. An IDS for IoT devices therefore requires a large number
-of algorithms to achieve adequate coverage.
-
-% gap analysis
-
-Current developments in ID on resource-constrained devices focus on programming
-frameworks that structure the implementation of algorithms \cite{valero2012di}
-and offer the required basic functional components to implement them
-\cite{krontiris2008lidea}. A key shortcoming of prior ID frameworks is that
-they do not offer a way to optimize the usage of resources. A solution that
-supports the integration of multiple algorithms on IoT devices needs to avoid
-accumulating their impact on the available resources. A second omission is
-support for systematic reuse to create different configurations. Addressing
-this in a transparent and automated way offers the potential to significantly
-reduce the effort of developing an IDS.
+Another prime example of a class of algorithms that inhibit the same properties
+related to pattern recognition is Intrusion Detection (ID). The vast number of
+security related opportunities for attackers makes way for an equally sized
+amount of ID algorithms, all with a similar structure, but largely independent
+operation.
 
@@
 
-% 1. classification of ID and design pattern for ID algorithms
-
-The first contribution of this paper consists in the formulation of a design
-pattern for ID algorithms, based on properties identified from a classification.
-
-% 2. framework: language + code generator
-
-The second contribution is a framework called \NAME. It enables the formal
-description of ID algorithms on a functional level and provides a code
+The contributions of this paper are threefold: The first contribution consists
+in the formulation of a design pattern for detection/recognition algorithms for
+IoT devices. This defines the target algorithms that fit our proposed
+framework. The second contribution is a framework called \FOO.\@ It enables the
+formal description of algorithms on a functional level and provides a code
 generator producing source code for a given platform and configuration. This
-way, \NAME addresses the heterogeneous nature of both IoT and ID algorithms.
-
-% 3. FCF to optimise
-
-The third contribution introduces functional code fusion (FCF) as a source code
-generation paradigm to address hard to combine ID algorithms. FCF identifies
-common data and functions, and organizes code to eliminate redundant
+way, \FOO also addresses the heterogeneous nature of both IoT and the
+algorithms. The third contribution introduces functional code fusion (FCF) as a
+source code generation paradigm to address the combination of algorithms. FCF
+identifies common data and functions, and organizes code to eliminate redundant
 iterations, tests and computations. Results show reduced usage of the wireless
@@
 The remainder of this paper proceeds as follows: section \ref{problems}
-describes the inherent problems in the combination of ID algorithms. Section
-\ref{classification} analyses and classifies ID algorithms. Section
-\ref{pattern} looks for patterns in the implementation of these classes.
-Section \ref{design} describes the design we applied to construct \NAME, its
-domain specific language (DSL) and code generator. Section \ref{evaluation}
-discusses and evaluates an implementation. Section \ref{related} explores
-related work in the field of DSLs and code generation. Finally, section
-\ref{conclusion} summarizes our findings, draws conclusions and identifies
-topics for future work.
+identifies problems that arise when combining multiple algorithms in the
+context of resource-constrained devices. Section \ref{pattern} formulates a
+pattern common to the detection/recognition algorithms in the IoT context.
+Section \ref{design} introduces our framework, its design and components. In
+section \ref{evaluation} we apply our solution to case study related to ID,
+showing an actual implementation and evaluation of the experimental results.
+Section \ref{related} explores related work in the field of DSLs and code
+generation. Finally, section \ref{conclusion} summarizes our findings, draws
+conclusions and identifies topics for future work.
 
-\section{Inherent Problems}
+\section{Problems when combining multiple algorithms}
 \label{problems}
 
-The combination of ID algorithms tends to result in suboptimal code. Besides
-these code-level problems, operational problems also hinder the implementation.
+Simply taking existing implementations of algorithms and calling them in
+sequence, to allow them to perform their functionality, will in most cases
+result in suboptimal code. Three prime examples on the source code level
+illustrate this:
 
-Two issues arise from the way ID algorithms are implemented and deal with loop
-constructs and common data. A third issue is the use of the network.
+\vspace{-2mm}
 
@@
   network, processing the same byte stream, performing comparable operations to
-  find similar patterns.
+  find similar patterns. These loops can also hold calls to poll sensor for
+  their values. This results in an execution overhead, putting a strain on the
+  power consumption.
+
+\vspace{-2mm}
 
   \item[Common data] cross-cuts the different algorithms, that have to store
-  information about devices within its own scope. This results in a lot of
-  duplicated information in memory.
+  the same information over and over again within their own scope. This results
+  in a lot of duplicated information in memory, raising the hardware
+  requirements.
+
+\vspace{-2mm}
 
   \item[The network] allows algorithms to operate on a distributed scale.
-  However, scattered calls to a message sending function cause multiple
-  accesses to the wireless radio, keeping it active and energy consuming more
-  than strictly needed.
+  Scattered calls to a message sending function cause multiple accesses to the
+  wireless radio, keeping it active and energy consuming more than strictly
+  needed.
 
@@
 
-Two operational aspects are orthogonal to these problems and amplify them:
+Two operational aspects are orthogonal to these problems and further amplify
+them:
+
+\vspace{-2mm}
 
@@
 
-  \item[Reimplementing] the algorithms manually to eliminate these issues is
+  \item[Reimplementing] algorithms manually to eliminate these issues is
   not a one-time cost. A release of a new version of an algorithm, or simply a
-  reselection of algorithms, due to a changing risk analysis, can trigger it.
+  reselection of algorithms, due to a changing requirements, can trigger it.
   Also, changing, or simply implementing the algorithms from scratch, holds the
@@
 
+\vspace{-2mm}
+
   \item[The heterogeneity of IoT devices] results in different software stacks,
   each with for example their own network API\@. This partially explains why
-  there are few implementations readily available to developers: there is no
-  single platform to target for researchers.
+  there are few implementations of algorithms readily available to developers:
+  there is no single platform to target for researchers, which lowers the
+  opportunity to reuse prior work.
 
@@
 
-\section{A Classification of ID Algorithms}
-\label{classification}
-
-Two aspects currently missing from ID frameworks are optimization of resources
-and support for systematic reuse. This section uses a classification to define
-the scope of \NAME, detailing which types of algorithms are good candidates for
-the proposed approach.
-
-Research into ID in wireless networks typically focuses on two major topics:
-the devices as single entities and the network as a group of such devices. Both
-are important: a group cannot make a decision without members that detect
-malicious behavior and a group-based decision is often needed because devices
-in the network can miss certain events that could reveal intrusions. In a
-wireless network, where not all participants are within each others' radio
-range, wireless devices typically only have a partial view on the entire
-network. This prohibits them from observing potentially crucial events in
-distant parts of the network. The fact that resource-constrained devices
-typically apply duty cycling to lower their energy consumption, further
-increases this problem, introducing periods of time when they cannot observe
-local events.
-
-This leads to two complementary classes of ID activities: local intrusion
-detection and cooperative decision making.
-
-\subsection{Local Intrusion Detection}
-\label{detection}
-
-There are different ways to construct a taxonomy for intrusion detection, but
-common themes do appear. According to \cite{mishra2004intrusion} and
-\cite{ioannis2007towards} there are three major categories: 1) anomaly
-detection, 2) signature or misuse detection, and 3) specification-based
-detection. In \cite{alrajeh2013intrusion} the authors agree with this typology
-and add the notion of hybrid intrusion detection systems and cross layer
-intrusion detection systems as recent advances in research. Although these
-additional categories seem promising, the authors admit that the impact on
-resource-constrained devices will be too high. We therefore focus on the three
-categories agreed upon by most sources.
-
-\begin{description}
-
-  \item[Anomaly detection] considers a baseline profile and evaluates all
-  activity according to this profile. It tries to flag aberrant behavior as a
-  possible intrusion. The baseline profile can evolve over time, especially in
-  wireless and mobile environments. Single events should therefore not result
-  in absolute decisions.
-
-  \item[Signature or misuse detection] starts with known patterns of attacks.
-  It tries to find evidence of intrusions in the actions of and communication
-  with a device. Because attacks often use valid ways to communicate with
-  services on the devices, this approach can produce false positives.
-
-  \item[Specification-based detection] uses constraints that describe the
-  correct behavior of the system. Although this approach can detect previously
-  unknown attacks, it can sometimes not detect specific attacks that anomaly or
-  misuse detection can.
-
-\end{description}
-
-\subsection{Cooperative Decision Making}
-\label{coorperative}
-
-These taxonomies focus on the act of detecting intrusions themselves. In a
-distributed environment, communication between participants may be even more
-important.
-
-Because not all devices are always actively participating in the network, they
-can miss events that would otherwise lead them to detect intrusions. It is
-hardly impossible for a single device to detect a complex attack by itself.
-Therefore a second important class of ID algorithms consists of cooperative
-algorithms that combine information from single identities into a group-based
-decision about alleged intrusions. We identify two ways to construct
-cooperative decision making and focus on the way participants exchange
-information: broadcast and interactive communication.
-
-\begin{description}
-
-  \item[Broadcast communication] can allow implementing a distributed,
-  cooperative algorithm locally: devices can exchange information about the
-  reputation of other devices, solely using one-way broadcast messages.
-  Combining this information allows them to decide about their trust in a given
-  device, based on more than their own, often partial, observations. The
-  reputation-based detection algorithm introduced in
-  \cite{ganeriwal2008reputation} illustrates this. It uses the same principle
-  of a watchdog found in \cite{mishra2004intrusion}.
-
-  \item[Interactive communication] allows algorithms to exchange information to
-  reach a consensus. In \cite{krontiris2009cooperative}, the authors first
-  present a theoretical foundation to analyze cooperative algorithms. Given
-  these foundations they continue to present an algorithm consisting of five
-  phases, which essentially implement a voting system based on the Guy Fawkes
-  protocol \cite{anderson1998new}, to identify an intruder. It enables the
-  exchange of suspected intruder information and allows for distributed
-  authentication of those \emph{votes}. Based on these authenticated votes, the
-  participants can make a distributed decision about the jointly identified
-  intruder.
-
-\end{description}
-
-\subsection{Properties and Scope of \NAME}
-\label{scope}
-
-Figure \ref{fig:classification} visually summarizes the dual classification and
-scope of \NAME.
-
-\begin{figure}[t]
-  \centering
-  \scalebox{.85}{
-\begin{tikzpicture}[
-  sibling distance=20mm,node distance=40mm
-  ]
-  \node (detection)             [align=center] {Detecting\\Intrusions}
-    child {node (anomaly)       [align=center] {Anomaly}}
-    child {node (signature)     [align=center] {Signature\\Misuse}}
-    child {node (specification) [align=center] {Specification}};
-  \node (decision)              [align=center,right=of detection] {Decision\\Making}
-    child {node (broadcasting)  [align=center] {Broadcast}}
-    child {node (interactive)   [align=center] {Interactive}};
-\end{tikzpicture}
-}
-  \caption{Classification of ID Algorithms}
-  \label{fig:classification}
-\end{figure}
-
-Algorithms that fit this classification are typically platform independent.
-They look for patterns in network messages that can reveal intrusions and, in
-doing so, all deal with the same information: other devices in the network. To
-be able to formally describe ID algorithms and target the heterogeneous IoT,
-\NAME requires ID algorithms to expose these properties.
-
-A class of ID algorithms that does not have these properties is software
-attestation \cite{seshadri2008sake}. It enables exchanging information about
-the software that is running on a device and identifying devices with modified
-content. These algorithms are highly platform dependent and have no data in
-common with other algorithms, as they look for the effects of an intrusion, not
-the intruder itself.
-
-\section{A Design Pattern for ID Algorithms}
+\section{A Design Pattern for Detection/Recognition Algorithms}
 \label{pattern}
 
-Based upon the classification presented in section \ref{classification}, we now
-introduce a design pattern that identifies common functional aspects of ID
-algorithms. It will allow us to define a strategy for optimizing the
-combination of these algorithms. We extract two activities common to most
-algorithms: the inspection of network traffic and the evaluation of thresholds.
+The source code level problems that arise when combining multiple algorithms,
+as presented in section \ref{problems}, are in fact inherent to the design
+pattern that is the foundation for the type of algorithms we identify in
+typical detection/recognition applications in IoT applications. The formulation
+of such a design pattern will allow us to define a strategy for the
+optimization of the combination of these algorithms. We start by extracting two
+common activities of these algorithms: the collection of sensor values and the
+evaluation of thresholds. Periodically collecting sensor values requires
+preparing the hardware for value extraction, waiting for the value to become
+available, reading the actual bytes, interpreting them into a meaningful value
+and recording them for further analysis. This can be as simple as reading a
+single byte, but can also be as complex as parsing a network message.
 
-Network traffic is the carrier along which malicious payloads reach a device
-directly. Indirectly a device can also detect aberrant patterns of
-communication by other devices, as a result of a successful attack.
-
-Not all algorithms can decide if an attack is real or not, based on a single
-event. Signature and misuse detection can do that in some cases, but anomaly
-detection typically deals with some level of uncertainty. Also, in distributed
-setups it is impossible to fully trust any neighboring device. In these
-situations one has to rely on statistics and thresholds to decide when to drop
-trust in a given device.
-
-Updating such statistics and evaluating the corresponding thresholds should be
-distinct operations. The example of a watchdog \cite{mishra2004intrusion}
-illustrates this: a device monitors the actions of other devices and checks if
-they correctly forward messages to distant nodes in the network. Missing
-forwards can indicate a malicious intension to silently drop messages and
-obstruct the normal operation of the network. There are valid reasons why a
-device can miss the theoretical window in which one expects activity, i.e.\ the
-wireless network being the most predominantly unreliable component. Directly
-reacting upon a basic threshold can result in false positives. A way to
-overcome this is to decouple the detection of problems and the decision what to
-do about them.
+A single recorded sensor value hardly ever results in a decision. The
+combination, often over time, of several of these values and the evaluation of
+thresholds will be key to decide if a change in the environment requires an
+action. Updating such statistics and evaluating the corresponding thresholds
+should be distinct operations. Therefore this is a separate function that is
+also periodically executed.
 
 This decoupling of detection and decision is the core of the design pattern for
-ID algorithms. Listing \ref{alg:id-algo-pattern} presents the design pattern in
-pseudo-code.
+ID algorithms. Listing \ref{alg:pattern} presents the design pattern in
+pseudo-code. Applying the pattern in the context of multiple algorithms,
+extends its scope and a pattern such as described in listing
+\ref{alg:application} will typically appear. The pattern calls the individual,
+implementations in sequence in two situations: first, to collect sensor
+readings, and secondly, at a regular interval to allow for internal
+housekeeping, such as threshold evaluation.
 
-\begin{algorithm}[t]
-  \caption{ID Algorithm Pattern}
-  \label{alg:id-algo-pattern}
+We clearly see the three aforementioned problems manifest itself: first, the
+nesting of loops causes redundant executions of potentially the same sensor
+readings inside the individual algorithms' $collect\_data$ functions. Second,
+each individual algorithm stores the same sensor values in memory. Third,
+scattered calls to e.g. network functions (in case of distributed algorithms)
+lead to suboptimal use of the wireless radio.
+
+Implementing the solution from scratch, it would be possible to identify these
+problems and reorganize the code in such a way that sensor values are only read
+once, stored in memory once and network messages parsed once. In section
+\ref{problems} we identified two operational problems that show that this is
+not feasible or at least suboptimal. We therefore present \FOO, a framework
+that enables formally describing such algorithms and fusing common data and
+functions in an automated way, to overcome both these problems.
+
+\vspace{-2mm}
+
+\begin{algorithm}[b]
+  \caption{Detection/Recognition Algorithm Pattern}
+  \label{alg:pattern}
   \begin{algorithmic}
-    \Require{devices: global variable for device information}
-    \Function{process\_message}{$msg$}
-      \ForEach{$byte \in msg$} \label{alg:id-algo-pattern-loop1}
-       \State \dots \Comment{analyze byte-sequences}
+    \Require{readings: global variables to hold collected sensor data for the given algorithm}
+    \Require{devices: global variable to information about neighbor devices} \Comment{optional in case of distributed algorithms}
+    \Function{collect\_data}{}
+      \ForEach{$sensor \in required\_sensors$}
+        \Let{values[]}{$sensor$}   \Comment{collect sensor data}
+        \Let{$reading_{sensor}$}{\Call{interpret}{values[]}}  \Comment{update interpreted sensor info}
       \EndFor
-      \Let{$devices_x$}{value}  \Comment{optionally update device info}
-      \State \Call{send}{$devices_y$, ``info''} \Comment{optionally exchange info} \label{alg:id-algo-pattern-send1}
+      \State \Call{send}{$devices_x$, ``info''} \Comment{optionally exchange info} \label{alg:id-algo-pattern-send1}
     \EndFunction
-    \State
     \Function{do\_housekeeping}{}
-      \ForEach{$device \in devices$} \label{alg:id-algo-pattern-loop2} \label{alg:id-algo-pattern-common-data}
-        \If{$device > \dots$} \Comment{validate recorded value}
+      \ForEach{$reading \in readings$} \label{alg:id-algo-pattern-loop2} \label{alg:id-algo-pattern-common-data}
+        \If{$reading > \dots$} \Comment{validate recorded value}
           \State \dots \Comment{take actions}
-          \State \Call{send}{$device$, ``info''} \Comment{communicate} \label{alg:id-algo-pattern-send2}
+          \State \Call{send}{$device_x$, ``info''} \Comment{optionally communicate} \label{alg:id-algo-pattern-send2}
         \EndIf
@@
 
-Applying the pattern in the context of multiple algorithms, extends its scope
-and a pattern such as described in listing \ref{alg:id-algo-application} will
-typically appear.
-
-\begin{algorithm}[t]
-  \caption{Application pattern of multiple ID algorithms}
-  \label{alg:id-algo-application}
+\begin{algorithm}[b]
+  \caption{Application pattern of multiple algorithms}
+  \label{alg:application}
   \begin{algorithmic}
-    \Let{$msg$}{$network$} \Comment{all observed messages}
     \ForEach{$algorithm \in algorithms$}
-      \State \Call{algorithm.progress\_message}{msg}
+      \State \Call{algorithm.collect\_data}{}
     \EndFor
-    \State \dots
-    \Comment{at a given interval}
-    \ForEach{$algorithm \in algorithms$}
+    \ForEach{$algorithm \in algorithms$}        \Comment{at a given interval}
       \State \Call{algorithm.do\_housekeeping}{}
@@
 
-The pattern calls the individual, linked implementations in sequence in two
-situations: first, when the device observes new messages on the network and,
-secondly, at a regular interval to allow for internal housekeeping, such as
-threshold evaluation.
+\section{Design}
+\label{design}
+
+The previous sections identified loops, common data and the network as three
+important code-level problems. These problems are inherent to detection
+algorithms, but conflict with the target environment, the IoT\@. \FOO provides
+a framework to efficiently implement these algorithms for the IoT and offers a
+solution for all identified problems. It consists of a 1) DSL, 2) source code
+generator and 3) software library. Figure \ref{fig:design} visualizes the
+design.
 
@@
   \centering
-  \scalebox{1}{
+  \scalebox{0.75}{
 \begin{tikzpicture}[
@@
 
-  \node (nodes) [dom,above=of cg, align=center]             {Nodes\\Domain};
+  \node (domains) [dom,above=of cg, align=center]             {Functional\\Domains};
   \node (plat)    [platform,below=of cg,align=center]         {Platform \&\\Language};          
@@
   \draw [->] (dsl)     -> (cg);
-  \draw [->] (nodes) -> (cg);
+  \draw [->] (domains) -> (cg);
   \draw [->] (plat)    -> (cg);
@@
   \draw [->] (comp)    -> (image);
-
 \end{tikzpicture}
 }
-\caption{The high level design of \NAME, showing the resources, processes and
+\caption{The high level design of \FOO, showing the resources, processes and
 artifacts. A grayscale background represents the level of platform dependency,
@@
 installable image are artifacts.}
-
 \label{fig:design}
@@
 
-\section{Design}
-\label{design}
+The DSL enables researchers to formally describe algorithms, without focusing
+on any specific language or platform. A code generator can avoid the identified
+problems concerning loops, common data and network access, as the formal
+description provides the underlying intent, not the technical implementation.
+It can also target different languages and platforms, while reusing a software
+library that provides services including parsing, network access wrapping,\dots
+It applies the Functional Code Fusion (FCF) paradigm to identify the hotspots
+for optimization. A key component of FCF is the introduction of
+\emph{functional domains} that harbor common functionality.
 
-The previous sections identified loops, common data and the network as three
-important code-level problems. These problems are inherent to ID algorithms,
-but conflict with the target environment, the IoT\@. \NAME provides a framework
-to efficiently implement an IDS for the IoT and offers a solution for the
-identified problems. It consists of a 1) DSL, 2) source code generator and 3)
-software library. Figure \ref{fig:design} visualizes the overall design.
+Using a DSL and code generator defines a development strategy. We will first
+position \FOO and motivate its design. Second, we will look at the principal
+design components of the DSL.\@ Third, we discuss the structure of the code
+generator and introduce its software library, that offers a way to deal with
+for example the network access issue.
 
-% figure design is moved up to end up on this page
-
-The DSL enables researchers to formally describe ID algorithms, without
-focusing on any specific language or platform. A code generator can avoid the
-identified problems concerning loops, common data and network access, as the
-formal description provides the underlying intent, not the technical
-implementation. It can also target different languages and platforms, while
-reusing a software library that provides services including parsing, network
-access wrapping,\dots
-
-Using a DSL and code generator defines a development strategy. This has
-implications, both on what is possible and how it can be achieved. We will
-first position \NAME and motivate its design with respect to these trade offs.
-Second, we will look at the principal design components of the DSL. Third, we
-discuss the structure of the code generator and introduce its software library,
-that offers a way to deal with for example the network access issue.
-
-\subsection{\NAME as a Development Strategy}
+\vspace{-1mm}
+\subsection{\FOO as a Development Strategy}
 \label{positioning}
 
-The introduction of a code generator in the development process changes the
-traditional way of working. This is most importantly due to the fact that the
-actual source code is no longer the driving force that can be controlled by
-developers. It is literally an intermediate representation between code
-generator and compiler.
+The introduction of a code generator in the development process changes the way
+of working. The actual source code is no longer the driving force, controlled
+by developers. It is literally an intermediate representation between code
+generator and compiler. A code generator constructs this representation using
+three components, as shown in figure \ref{fig:design}: 1) the description of
+the algorithm in the DSL, 2) information about the functional domains, and 3)
+information about the platform and target language. At the level of the
+compiler, a software library offloads specialized implementations and allow
+custom extensions. The applied grayscale indicates the components' platform
+dependency. This hierarchy is important from a development effort point of
+view. The level of dependency on the target platform dictates the level of
+effort to implement changes. This is important in relation to the scalability
+of the framework itself.
 
-A code generator constructs this representation using three components, as
-shown in figure \ref{fig:design}: 1) the description of the algorithm in the
-DSL, 2) information about the domain, in this case the Nodes domain, and 3)
-information about the platform and target language. At the level of the
-compiler, a software library is introduced to offload specialized
-implementations and allow custom extensions.
-
-The applied grayscale indicates the components' platform dependency: the DSL is
-platform and language independent and deals with subsystems, such as the
-network, in an abstract way. The Nodes domain implements some of these
-abstracted subsystems, but still in a platform and language independent way.
-Platform and language descriptions convert the abstract constructions in
-actually supported code. The software library is standard native code,
-targeting the platform and language. Finally, the code and image are artifacts
-of the automated generation and compilation process. They aren't a product of
-manual work and therefore aren't directly dependent on the platform.
-
-This hierarchy is important from a development effort point of view. The level
-of dependency on the targeted platform dictates the level of effort to
-implement changes in that platform. For example: to generate a different
-language will require a complete rewrite of the software library and a partial
-rewrite of the language description of the generator. Although this is a
-reusable, one-time effort, it is important in relation to the scalability of
-the framework itself.
-
-\subsection{\NAME as a Domain Specific Language}
+\vspace{-1mm}
+\subsection{\FOO as a Domain Specific Language}
 \label{dsl-design}
@@
 (semantic) model} \cite{fowler2010domain}. Its sole purpose should be
-populating the model. \NAME's DSL and model focus on avoiding constructions
-that result in problems concerning loops, common data and the use of the
-network. To do this, \NAME uses an event-based approach to defining
-functionality.
+populating the model. \FOO's DSL and model focus on avoiding constructions that
+result in problems concerning loops, common data and the use of the network. To
+do this, \FOO uses an event-based approach to defining functionality.
+Describing algorithms in function of events that occur will remove the
+technical descriptions and allows FCF to identify the actual intent of the
+algorithms with respect to the identified source code problems.
 
-Besides avoiding these problems, \NAME's DSL also provides a higher level of
-abstraction with respect to memory management and data access.
-
-This section therefore discusses \NAME's approach to 1) loops, 2) nodes domain,
-3) unified messaging, 4) language support, and illustrates these with a
-complete example.
-
+\vspace{-1mm}
 \subsubsection{Loops}
@@
 program code are technical constructs and in fact often hide the intended
-functionality.
+functionality. The example of parsing network messages illustrates this. Each
+algorithm deals with parsing in the same way: loop over the bytes in the
+payload and look for a pattern that identifies what the algorithm is looking
+for. This is in fact a technical implementation choice. The underlying
+functional intent is to react when such a pattern is present in a message. The
+loop has \emph{activated} this functionality, but the real goal is to
+\emph{passively} react on an event, not \emph{actively} look for it. Event
+handling therefore can transform constructions with loops to their functional
+meaning, allowing identification and fusion of common functionality.
 
-The example of parsing network messages illustrates this. Each algorithm deals
-with parsing in the same way: loop over the bytes in the payload and look for a
-pattern that identifies what the algorithm is looking for. This is in fact a
-technical implementation choice. The underlying functional intent is to react
-when such a pattern is present in a message. The loop has \emph{activated} this
-functionality, but the real goal is to \emph{passively} react on an event, not
-\emph{actively} look for it. Event handling therefore can transform
-constructions with loops to their functional meaning, allowing identification
-and fusion of common functionality.
+\vspace{-1mm}
+\subsubsection{Common Data}
 
-The same goes for actively polling some variable until it reaches a certain
-threshold. When transformed to a reaction to an event, it eliminates the need
-for a technical loop-construction.
+Different detection algorithms typically deal with the same classes of
+information, be it sensor data or network messages. Functional domains
+consolidate these and store and manage this data in a unified way. \FOO
+provides the concept of such domains as plug-ins to the DSL, extending it with
+domain specific storage and logic. For example in the case of distributed
+algorithms, such as ID algorithms, a \emph{Nodes} domain will manage
+information about neighbor nodes, exposing them as global objects to the
+algorithms, relieving them from managing these themselves. Functional domains
+are accessible as collections, allowing execution of functions in their scope,
+eliminating the need for explicit loops.
 
-\subsubsection{The Nodes Domain}
-
-Most, if not all, ID algorithms deal with information about other devices in
-the network. To share this information, \NAME provides a unified \ttt{nodes}
-concept as part of its DSL and exposes it as an object to the implementer.
-
-Algorithms can extend nodes with their own properties. This allows for only one
-instance per known device to be present in memory. Finally, nodes can define a
-scope for a function to execute in. This way, nodes offer a way to i.e.\
-iterate all known devices, again without the need for an explicit loop.
-
+\vspace{-1mm}
 \subsubsection{Unified Messaging}
@@
 to be more active than needed. Delaying the sending of messages and grouping
-them can lower this overhead.
+them can lower this overhead. Sending and receiving of network messages is
+tightly integrated with the previously mentioned concepts of event handling and
+for example the Nodes domain. An underlying software library can handle all
+interaction with the network, triggering functions by means of events when
+parsing new messages and exposing an API to send messages through the Nodes
+domain. This approach gives the framework complete control over communication
+and allows for optimized marshaling and grouping of messages.
 
-Sending and receiving of network messages is tightly integrated with the
-previously mentioned concepts of event handling and the Nodes domain. An
-underlying software library can handle all interaction with the network,
-triggering functions by means of events when parsing new messages and exposing
-an API to send messages through the Nodes domain. This approach gives the
-framework complete control over communication and for example allows for
-optimized marshaling and grouping of messages.
+\vspace{-1mm}
+\subsection{\FOO as a Code Generator}
 
-\subsubsection{Language Support}
-\label{language-support}
+To convert the DSL-based ID algorithms into production-ready code, \FOO applies
+code generation techniques. Although different approaches are applicable, like
+template-based text expansion, XSLT transformations or CodeDom
+\cite{dollard2004code}, we believe that the required transformations demand a
+richer model. A semantic model \cite{fowler2010domain}, tied closely to the DSL
+itself, can fulfill this role. Semantic transformations evolve the model into a
+more syntax-oriented model, comparable to the idea of CodeDom. From there on,
+syntactic transformations can further evolve the model into source code. Figure
+\ref{fig:code-generation} shows the process that \FOO implements.
 
-Using a platform independent language offers the possibility to relieve the
-implementer of explicit memory management and low-level data structures. From a
-language perspective, \NAME offers several programming constructs, typically
-found in modern programming languages with higher levels of abstraction:
-optional typing, object orientation, pattern matching, aspect orientation.
-
-\subsubsection{An Example}
-
-Listing \ref{lst:watchdog} illustrates some of the features mentioned above,
-using a small but representative example of an algorithm: a watchdog
-\cite{mishra2004intrusion}. Like a heartbeat, nodes can broadcast a message at
-a regular interval, allowing other nodes to monitor its availability. Missing
-heartbeats could indicate problems, such as physical capture of a device.
-
-\lstinputlisting[
-  language=foo-lang,
-  float,
-  basicstyle=\footnotesize\ttfamily\color{black},
-  label=lst:watchdog,
-  caption=Watchdog in \NAME
-]{src/watchdog.foo}
-
-\subsection{\NAME as a Code Generator}
-\label{code-generator-design}
-
-To convert the DSL-based ID algorithms into production-ready code, \NAME
-applies code generation techniques. Although different approaches are
-applicable, like template-based text expansion, XSLT transformations or CodeDom
-\cite{dollard2004code}, we believe that the required transformations demand a
-richer model.
-
-A semantic model \cite{fowler2010domain}, tied closely to the DSL itself, can
-fulfill this role. Semantic transformations evolve the model into a more
-syntax-oriented model, comparable to the idea of CodeDom. From there on,
-syntactic transformations can further evolve the model into source code. Figure
-\ref{fig:code-generation} shows the process that \NAME implements.
-
-\begin{figure}[t]
+\begin{figure}[b]
   \centering
-  \scalebox{.85}{
+  \scalebox{.75}{
 \begin{tikzpicture}[
@@
 }
-\caption{\NAME's code generation process}
+\caption{\FOO's code generation process}
 \label{fig:code-generation}
@@
 
+\vspace{-1mm}
 \subsubsection{Semantic Model}
 
-The semantic model is the core of the solution. It offers a way to describe ID
-algorithms so that semantic reinterpretation through functional code fusion is
-possible. This reorganization of the functionality of different algorithms
-allows for more optimal execution and memory usage. Hence the name \NAME of the
-framework: intrusion detection functionality organization optimization.
+The semantic model is the core of the generator. It offers a way to describe
+algorithms so that semantic reinterpretation through FCF is possible. This
+reorganization of the functionality of different algorithms allows for more
+optimal execution and memory usage. Hence the name \FOO of the framework:
+functionality organization optimization. Figure \ref{fig:meta-model} shows the
+central part of the semantic model as implemented for \FOO.\@ The concept driving
+this model is the structure starting at the \emph{Model} up to the
+\emph{Execution Strategy}. This structure encompasses the \emph{Function}s and
+underlying \emph{Statement}s and allows for interpretation and fusion of the
+algorithms.
 
-\begin{figure}[t]
+\begin{figure}[b]
   \centering
-  \scalebox{.85}{
+  \scalebox{.75}{
 \begin{tikzpicture}[
@@
         child {node (strategy) [class,align=center,xshift=-10.5mm,fill=black!10!white] {Execution\\Strategy} edge from parent [composite]
-          child {node(every) [class,xshift=-9mm,fill=black!10!white] {Every} edge from parent [subclass]}
-          child {node (when) [class,xshift=9mm,fill=black!10!white] {When} edge from parent [subclass]}
+          child {node(every) [class,xshift=-9mm,fill=black!10!white] {Interval} edge from parent [subclass]}
+          child {node (when) [class,xshift=9mm,fill=black!10!white] {Event} edge from parent [subclass]}
         }
@@
   \node (statement) [class,right=of function]              {Statement};
-  \node (nodes)     [class,left=of module,xshift=-13mm]    {Nodes\\Domain};
+  \node (domain)    [class,left=of module,xshift=-13mm]    {Functional\\Domain};
 
-  \draw [aggregate]  (module)   -- (nodes);
+  \draw [aggregate]  (module)   -- (domain);
   \draw [composite]  (module)   -- (constant);
@@
 
-  \draw [composite]  (nodes)    -- (scope);
+  \draw [composite]  (domain)   -- (scope);
   \draw [dependency] (strategy) -- (function);
@@
 }
-\caption{Partial Semantic Model for \NAME. Shaded classes indicate the
+\caption{Partial Semantic Model for \FOO.\@ Shaded classes indicate the
 \emph{event/scheduling} construction around functions to allow interpretation
@@
 
-Figure \ref{fig:meta-model} shows the central part of the semantic model as
-implemented for \NAME. The concept driving this model is the structure starting
-at the \emph{Model} up to the \emph{Execution Strategy}. This structure
-encompasses the \emph{Function}s and underlying \emph{Statement}s and allows
-for interpretation and fusion of the algorithms.
-
+\vspace{-1mm}
 \subsubsection{Functional Code Fusion (FCF)}
@@
 and group actions on the same sets of data or scheduled function executions.
-
 FCF typically looks for patterns in the semantic model and performs model
@@
 phase transforms the semantic model into a syntactic model, encoding the
-semantics into technical syntax constructs.
+semantics into technical syntax constructs. One example of a semantic
+transformation is type inference, which supports the platform independence of
+the framework.
 
-One example of a semantic transformation is type inference, which supports the
-platform independence of the framework.
-
+\vspace{-1mm}
 \subsubsection{Syntactic Model}
@@
 unsupported constructs into comparable constructs supported by the targeted
-platform and language.
+platform and language. Depending on the target platform, some constructions
+aren't transformed into actually corresponding code. In selected cases, the
+generator uses a software library. For example when it needs to parse incoming
+messages or schedule tasks. During the generation process, it looks for
+patterns that indicate any of these actions and transforms them into calls to
+the \FOO software library.
 
-Depending on the target platform, some constructions aren't transformed into
-actually corresponding code. In selected cases, the generator uses a software
-library. For example when it needs to parse incoming messages or schedule
-tasks. During the generation process, it looks for patterns that indicate any
-of these actions and transforms them into calls to the \NAME software library.
-
-\subsection{\NAME as a Software Library}
+\vspace{-1mm}
+\subsection{\FOO as a Software Library}
 \label{software-lib-design}
@@
 Previously we illustrated this approach with a case for unified messaging:
-parsing of incoming messages and grouping of outgoing messages.
+parsing of incoming messages and grouping of outgoing messages. More common
+functionality isn't generated verbatim either. The execution of functions based
+on a schedule or the implementation of higher level conceptual functions are
+calls into the standard software library. Optimization of such algorithms is
+not within the scope of \FOO.\@ This gives freedom of choice to the users of
+the framework to select appropriate implementations for this standard
+functionality.
 
-More common functionality isn't generated verbatim either. The execution of
-functions based on a schedule or the implementation of higher level conceptual
-functions, such as \ttt{now} and \ttt{sha1}, are calls into the standard
-software library.
-
-Optimization of such algorithms is not within the scope of \NAME. This gives
-freedom of choice to the users of the framework to select appropriate
-implementations for this standard functionality.
-
-\subsection{\NAME as a Framework}
-
-\NAME reduces inefficiencies in network access, execution and memory usage,
-that occur due to a selection of ID algorithms. To achieve this, it avoids
-constructions that are hard to identify and reorganize, such as loops and
-common data. A DSL, focusing on event-handling, enforces these restrictions and
-enables formally describing ID algorithms.
-
-The DSL is platform and language independent. This allows the source code
-generator to target different configurations without any change to the formal
-descriptions of the algorithms.
-
-A supporting software library allows separation of standard functionality that
-is out of scope for the generator. Optimization of the implementation of this
-library for the specific platform and target language is possible.
-
-\section{Implementation and Evaluation}
+\section{Implementation and Evaluation: Case Study on Intrusion Detection in Wireless Sensor Networks}
 \label{evaluation}
 
-In this section we present a prototype implementation and evaluate its
-performance. We start by introducing the implementation, consisting of the
-hardware platform, network and software stack. We then continue with an
-analytical approach, looking at the handling of common data and the size of the
-source code. Next, we use the implementation to quantify its merits using a
-side-by-side comparison of generated code versus a manual implementation. We
-consider experimental results regarding usage of the network, execution time of
-the event-loop and image size.
+In wired networks, firewalls focus on the outer perimeter of the network,
+filtering unwanted packets. Still, attacks on services can pass unnoticed. This
+is where an Intrusion Detection System (IDS) comes into play: monitoring all
+traffic that passes through the firewall, looking for patterns of malicious
+activity\cite{denning1987intrusion}. In meshed networks it is not possible for
+a single point in the network to oversee all traffic
+\cite{mishra2004intrusion}. Every device has to implement its own IDS.
 
+Attackers have access to a large set of possible attack vectors
+\cite{aschenbruck2012security}, ranging from the physical layer up to the
+application layer. Each layer presents different opportunities for launching an
+attack. For each of these attacks an algorithm needs to look for patterns or
+anomalies. An IDS therefore requires a many algorithms to achieve adequate
+coverage.
+
+\vspace{-1mm}
 \subsection{Implementation}
@@
 To evaluate the prototype we used a system based on the Atmel ATMEGA1284p
-micro-controller and the Digi XBee S2 ZigBee module. A basic software library
-wraps the technical calls to the hardware components in more easy to use
-functions.
-
-Relying on basic hardware and software allows us to validate that the generator
-is capable of generating code even for the most basic environment. It shows
-that the requirements of the generator towards the platform are minimal, do not
-rely on advanced frameworks or operating systems, and therefore are applicable
-in any environment, targeting any platform.
-
-For the evaluation we constructed a small meshed network. In this network, not
-all devices have direct access to one another, requiring routing through shared
-neighboring devices.
+micro-controller and the Digi XBee S2 ZigBee module, running no specialized
+operating system. Relying on basic hardware and software allows us to validate
+that the generator is capable of generating code even for the most basic
+environment. For the evaluation we constructed a small meshed network. In this
+network, not all devices have direct access to one another, requiring routing
+through shared neighboring devices.
 
 We started from a basic application that measures light intensity. The
-application was an includable piece of code, which allowed integration in a
-manual implementation and easy merging when generating. On top of this baseline
-we added two intrusion detection related algorithms: a watchdog
+application was an includable piece of code, allowing integration in a both a
+manual implementation and generated code base. On top of this baseline we added
+two intrusion detection related algorithms: a watchdog
 \cite{mishra2004intrusion}, that allows nodes to validate each other's
 presence, and a reputation-based algorithm \cite{ganeriwal2008reputation}, that
-checks if a parent node is cooperative and actually forwards messages further
-upstream.
+checks if a parent node is cooperative and actually forwards messages.
 
+\vspace{-1mm}
 \subsection{Analytical Results}
 
-Looking at the generated source code is a straightforward way to validate the
-framework. A simple metric is comparing the amount of code versus the size of
-the formal description of ID algorithms. This can be an indication of the
-development effort and complexity. Implementing both algorithms manually
-results in 637 lines of C code. The corresponding \NAME implementation results
-in 156 lines of DSL code. This is largely thanks to the higher level coding
-constructs, as introduced in section \ref{design}, dealing with \NAME's
-language support.
+Looking at the generated source code, a simple metric is comparing the amount
+of code versus the size of the formal description of ID algorithms.
+Implementing both algorithms manually results in 637 lines of C code. The
+corresponding \FOO implementation requires only 156 lines of DSL code. Looking
+at data structures, we see that sharing data eliminates redundant information,
+such as a network address. This reduces the memory footprint for one node entry
+by 2 bytes, from 27 bytes to 25 bytes. In our implementation we used the ZigBee
+network address, which consists of 16 bits. This results in a reduction of
+7.4\% of dynamic memory per known device. Using ZigBee's 64 bits unique
+addresses, this reduction can easily become 20.5\%.
 
-Next we look at data structures. Sharing data eliminates redundant information,
-such as a network address, from the scope of each individual algorithm.
-Analysis of the source code with both algorithms implemented, shows that this
-reduces the memory footprint for one node entry by 2 bytes, from 27 bytes to 25
-bytes. In our implementation we used the ZigBee network address, which indeed
-consists of 16 bits. This results in a reduction of 7.4\% of dynamic memory per
-known device. Using ZigBee's 64 bits unique addresses, this reduction can
-easily become 20.5\%.
-
+\vspace{-1mm}
 \subsection{Experimental Results}
 
-Besides this analytical approach, we also instrumented the code and evaluated
-four criteria: the network usage in number of packets and bytes, the time
-required to perform once cycle of the event loop and the image size of the
-resulting compiled code. To do this, we collected these metrics for four
+We also instrumented the code and evaluated four criteria: network usage
+(number of packets and bytes), the time required to perform once cycle of the
+event loop and the resulting image size. We collected these metrics for four
 situations: without ID algorithms, with a watchdog, with reputation-tracking
-and with both algorithms implemented.
+and with both algorithms. The manual implementation realized both algorithms as
+standalone modules and sequentially called them from the base-application's
+event loop. Given the \FOO descriptions of the algorithms, the code generator
+generated all four cases. Table \ref{tbl:results} shows the collected data.
 
-The manual implementation realized both algorithms as standalone modules and
-sequentially called them from the base-application's event loop. Given the
-\NAME descriptions of the algorithms, the code generator generated all four
-cases.
-
-Table \ref{tbl:results} shows the collected data for the manual and the
-generated implementation, as well as a comparison in absolute values and in
-relative terms.
-
-\begin{table}[b]
-  \caption{Results and Side-by-Side Comparison.}
+\begin{table}[h]
+  \caption{Experimental Results and Side-by-Side Comparison.}
   \label{tbl:results}
@@
 These results show that adding algorithms manually to the base-application has
-a cumulative effect. The impact of both algorithms is simply the sum of the
-impact of each algorithm by itself. The execution time of one cycle of the
-event loop with both algorithms enabled is even higher than simply the sum of
-their individual times.
+a cumulative effect. In case of the generated implementation, this effect is no
+longer observed: individual algorithms do have a larger impact by themselves,
+but the combination of both algorithms is less than the sum of both. Most
+remarkable is the effect on the time of one event loop cycle: both algorithms
+by itself add 152\%. Adding a second algorithm only augments the impact to
+188\%. We see the effect of the generic handling of common functionality. It
+comes with an overhead for a single algorithm, but pays for itself when adding
+more algorithms.
 
-In case of the generated implementation, this effect is no longer observed:
-individual algorithms do have a larger impact by themselves, but the
-combination of both algorithms is less than the sum of both.
-
-Most remarkable is the effect on the time of one event loop cycle: both
-algorithms by itself add 152\%. Adding a second algorithm only augments the
-impact to 188\%. We see the effect of the generic handling of common
-functionality, like message parsing and event scheduling. It comes with an
-overhead for a single algorithm, but pays for itself when adding more
-algorithms.
-
-Table \ref{tbl:results} also compares the two situations by subtracting the
-manual case from the generated case, showing the impact of the code generation.
-A relative value for this difference, to the manual case, is also included.
-Four observations allow us to evaluate the overall performance of \NAME's
-approach.
-
-The first observation deals with the base case: both the manual and the
-generated implementations perform in exactly the same way. There is no
-fundamental structural difference in the overall architecture of the code.
-
-Secondly, when turning to the goal of reducing the network usage, we see that
-both the number of packets as well as the amount of bytes sent have dropped by
-13\% and 7\%.
-
-Thirdly, focusing on the execution time, both algorithms by themselves add an
-equal amount of time to the processing of a single event loop cycle. When
-combined, the event loop is about 7\% faster compared to the manual case.
-
-Finally, we see that the software library that comes with the generated code
-adds to the size of the resulting image. However, the cost is close to constant
-and is lower for both algorithms together. With roughly 3KB of additional
-overhead, or 15\% in this simple case, the impact is small. Taking into account
-the available memory on the ATMega1284p, this is only 2.3\% of the available
-resources.
+Table \ref{tbl:results} also compares the two situations, showing the impact of
+the code generation: Both the number of packets as well as the amount of bytes
+sent have dropped by 13\% and 7\%. In case of the execution time, both
+algorithms by themselves add an equal amount of time to a single event loop
+cycle. When combined, the event loop is about 7\% faster compared to the manual
+case. Finally, we see that the software library adds to the size of the
+resulting image. However, the cost is close to constant and is lower for both
+algorithms together. With roughly 3KB of additional overhead, or 15\% in this
+simple case, the impact is small. Taking into account the available memory on
+the ATMega1284p, this is only 2.3\% of the available resources.
 
@@
 
-Section \ref{classification} introduced some ID algorithms
-\cite{ganeriwal2008reputation,mishra2004intrusion,krontiris2009cooperative} and
-presented a classification
-\cite{mishra2004intrusion,ioannis2007towards,alrajeh2013intrusion} based on
-detection approach and interaction style.
-
 This section focuses on related work in the field of domain specific languages
-and code generation, both in relation to wireless networks of
-resource-constrained devices. These two components are the core of the \NAME
-framework and the solution it proposes. We also look at their application in
-other disciplines and analyze how this can relate to our context.
-
-\subsection{Domain Specific Languages}
+and code generation, in relation to wireless networks of resource-constrained
+devices. These two components are the core of the \FOO framework.
 
@@
 software engineers, and to lead to a wider adoption outside the computer
-science field} \cite{sadilek2008domain}.
-
-TinyScript \cite{levis2004tinyscript} is an example of a language for wireless
-sensor networks. It's a complete programing language that aims to ease the
-description of algorithms in general. The Basic-like syntax does enable more
-productive development. Still, this approach doesn't address the underlying
-resource problems inherently connected to the context.
-
-SEAL \cite{elsts2013seal} is a novel complete programming language that focuses
-on novice programmers. It is a clear fit for \emph{sense-and-send} type of
-applications. It does however share the idea of using events to overcome the
-technical nature of loops.
-
-Shifting the scope to intrusion detection, we learn that DSLs are predominantly
-used to describe patterns of events found in network
-packets \cite{sekar1999high,roesch1999snort} in a concise way.
-
-BMSL \cite{uppuluri2001experiences} is a DSL that allows the specification of
-patterns and actions. Patterns can contain events extracted from log files,
-system calls and network packets.
-
-The STAT framework with its STATL language
-\cite{eckmann2002statl,vigna2003designing} are not intrusion detection
-specific, but through extensions they allow to model attack scenarios using
-state machines. Our event-driven approach shares these same ideas.
-
-Although not a real DSL, nesC \cite{gay2003nesc} is a general purpose language
-that shares some typical concepts with \NAME, such as the high degree of
-analysis during code generation to optimize execution. The biggest difference
-is that \NAME starts from the intended functionality and can optimize certain
-constructions that can't be interpreted from more technical source code.
-
-\subsection{Code Generation}
+science field} \cite{sadilek2008domain}. TinyScript \cite{levis2004tinyscript}
+is an example of a language for wireless sensor networks. It's a complete
+programing language that aims to ease the description of algorithms in general.
+The Basic-like syntax does enable more productive development. Still, this
+approach doesn't address the underlying resource problems inherently connected
+to the context. SEAL \cite{elsts2013seal} is a novel complete programming
+language that focuses on novice programmers. It is a clear fit for
+\emph{sense-and-send} type of applications. It does however share the idea of
+using events to overcome the technical nature of loops. In the context of
+intrusion detection, we learn that DSLs are predominantly used to describe
+patterns of events found in network packets
+\cite{sekar1999high,roesch1999snort} in a concise way. The STAT framework with
+its STATL language \cite{eckmann2002statl,vigna2003designing} allow to model
+scenarios using state machines. Our event-driven approach shares these same
+ideas. Although not a real DSL, nesC \cite{gay2003nesc} is a general purpose
+language that shares some typical concepts with \FOO, such as the high degree
+of analysis during code generation to optimize execution. The biggest
+difference is that \FOO starts from the intended functionality and can optimize
+certain constructions that can't be interpreted from more technical source code.
 
@@
 by reducing the size of the code base or systematically improving the code
-quality.
-
-Embedded systems are a prime example of an environment where code needs to be
-of a high quality. Focus on size and performance are a logical choice when
-dealing with limited storage and processing power. Compilers implement
-different techniques to address this at the lowest level
-\cite{marwedel2002code}.
-
-Optimizing for power efficiency however proves to be concurrent to size and
-performance optimization, and requires different strategies. Data and memory
-optimizations \cite{panda2001data} can result in improvements on a lower level.
-To really preserve energy one requires higher-level strategies, targeting
-architecture, radio handling, efficient software,\dots \cite{naik2001software}.
-Most of these are beyond the control of the compiler and require a higher level
-of abstraction to address them.
-
-DSLs offer a solution here, but other paradigms are possible. A mixed framework
+quality. Embedded systems are a prime example of an environment where code
+needs to be of a high quality. Focus on size and performance are a logical
+choice when dealing with limited storage and processing power. Compilers
+implement different techniques to address this at the lowest level
+\cite{marwedel2002code}. Optimizing for power efficiency however proves to be
+concurrent to size and performance optimization, and requires different
+strategies. Data and memory optimizations \cite{panda2001data} can result in
+improvements on a lower level. To really preserve energy one requires
+higher-level strategies, targeting architecture, radio handling, efficient
+software,\dots \cite{naik2001software}. Most of these are beyond the control of
+the compiler and require a higher level of abstraction to address them. DSLs
+offer a solution here, but other paradigms are possible. A mixed framework
 consisting of a virtual machine (VM) and native code extension
@@
 reducing the update cost, while maintaining flexibility and lowering the
-development effort. In this terminology, \NAME focuses on the reduction of the
+development effort. In this terminology, \FOO focuses on the reduction of the
 development effort and update cost, and maximizes flexibility.
-
-Applying code generation to ID often focusses on the generation of native code
-to avoid the interpretation of descriptions of patterns to look for. I2S
-\cite{charitakis2003code} is an example that generates code for SNORT
-\cite{roesch1999snort} rules. To optimize this highly efficient code, it often
-targets a specific platform.
 
@@
 
-Adequate intrusion detection for IoT devices requires multiple algorithms to
-operate at the same time. Frequent reselection, e.g. due to changing
-requirements or security threats, requires a flexible way to incorporate the
-algorithms.
+Many IoT applications typically are constructed of similar algorithms that
+adhere to a pattern of collecting sensor data and periodically checking
+recorded values for patterns that require action. Fusing these algorithms is
+hard but needed to deal with the resource-constrained devices they are
+targeting. Frequent reselection, e.g. due to changing requirements or changing
+security threats, requires a flexible way to incorporate these algorithms in
+subsequent iterations of the software.
 
-Adding multiple intrusion detection algorithms to wireless resource-constrained
-devices requires fusion of algorithms on a functional level. This can reduce
-execution inefficiencies as well as redundant memory usage and network access.
-
-\NAME is a framework consisting of a DSL, a source code generator and software
-library. Using the DSL, intrusion detection algorithms can be formally
-described. The code generator applies functional code fusion to interpret the
-intent of algorithms and merges common functionality \& shared data. The
-software library hosts common functionality, such as network message parsing
-and event scheduling.
+\FOO is a framework consisting of a DSL, a source code generator and software
+library. The DSL allows formally describing algorithms, enabling the code
+generator to apply functional code fusion to interpret the intent of algorithms
+and merge common functionality \& shared data. The software library hosts
+common functionality, such as sensor access, message parsing and event
+scheduling. The platform independent DSL enables generation of source code
+targeting different platforms and languages, which allows high degrees of reuse
+with heterogenous devices.
 
@@
 
-The platform independent DSL enables generation of source code targeting
-different platforms and languages, which allows high degrees of reuse with
-heterogenous devices.
-
-As a next step, we will implement more ID algorithms using \NAME and continue
-experiments on a larger scale. This way we will also be able to verify the
-initially observed inclination in the evolution of the image size with growing
-numbers of algorithms.
-
-Future research will focus on the integration of applications with a generated
-IDS. We see opportunities for applications that detect intrusions, to actively
-and dynamically respond and avoid malicious payloads.
-
-A continued effort is targeting the reduction of platform-dependent parts in
-the implementation. This way it is possible to add more platforms and languages
-with less effort.
-
-In the longer run, we want to investigate the possibility to extend the
-functional code fusion paradigm to other domains beyond intrusion detection.
+Future research will focus on the integration of applications with the
+generated functionality. We see opportunities for applications to adopt the
+event-driven approach and hope to expand the capabilities of \FOO to this
+level. A continued effort is targeting the reduction of platform-dependent
+parts in the implementation. This way it is possible to add more platforms and
+languages with less effort. In the longer run, we want to investigate the
+possibility to extend the functional code fusion paradigm to more domains.
 

### 2015-04-17 paper-euspn15.tex
completed text clean-up, squeezed whitespace

@@
 \input{packages}
+\input{squeeze}
 
@@
 
-\author[]{Christophe Van Ginneken}
-\author[]{Jef Maerien}
-\author[]{Christophe Huygens}
-\author[]{\\Wouter Joosen}
-\author[]{Danny Hughes}
+\author{Christophe Van Ginneken}
+\author{Jef Maerien}
+\author{Christophe Huygens}
+\author{Wouter Joosen}
+\author{Danny Hughes}
 
@@
 proves to be time-consuming, repetitive and error-prone. To address this
-problem we propose \FOO, a framework for the generation of efficient detection
-algorithms. It consists of a domain specific language and a code generator. The
-language allows formally describing the intent of detection algorithms and
-supports the generator in organizing the source code. As a case study, we apply
-\FOO to the generation of an Intrusion Detection System for wireless sensor
-nodes. A side-by-side comparison shows that \FOO-generated source code reduces
-message passing overhead, execution time and memory footprint in comparison to
-sequential calls into stand-alone implementations of the algorithms.
+problem we propose \FOO, a framework for the generation of efficient
+combinations of detection algorithms. It consists of a domain specific language
+and a code generator. The language allows formally describing the intent of
+detection algorithms and supports the generator in organizing the source code.
+As a case study, we apply \FOO to the generation of an intrusion detection
+system for wireless sensor nodes. A side-by-side comparison shows that
+\FOO-generated source code reduces message passing overhead, execution time and
+memory footprint in comparison to sequential calls into standalone
+implementations of the algorithms.
 
@@
 
-IoT, Domain Specific Languages, Source Code Generation, Source Code
-Optimization, Intrusion Detection, Context Awareness
+Internet of Things, Domain Specific Languages, Source Code Generation, Source
+Code Optimization, Intrusion Detection, Context Awareness
 
@@
 
-\vspace*{-10pt}
-
 \section{Introduction}
+\label{introduction}
 
@@
 \emph{smart things}: from smart watches that assist visually impaired
-persons\cite{porzi2013smart} to monitoring systems for flood hazardous
-areas\cite{hughes2006gridstix}. Their applications cannot be more diverse,
-however they share one common concern: each of these \emph{things} is
-resource-constrained, relying on a battery and operating with limited
-processing power and memory.
+persons\cite{porzi2013smart} to monitoring systems for waste management in our
+\emph{smart cities}\cite{zanella2014internet}. Their applications cannot be
+more diverse, however they share one common concern: each of these
+\emph{things} is resource-constrained, relying on a battery and operating with
+limited processing power and memory.
 
@@
 behalf of its wearer. That same watch can also detect a wet floor and notify
-this hazard.
+this hazard\cite{porzi2013smart}.
 
@@
 decade\cite{lara2013survey}, but with the advent of the IoT it now moves into
-the real world. Equipped with wearables with numerous sensors, the
-possibilities seem endless, but so are the number of detection and recognition
-algorithms that have to run simultaneously on these small \emph{things}, to
-support our augmented reality.
+the real world at an increased pace. Equipped with wearables, hosting numerous
+sensors, the possibilities seem endless, but so are the number of detection and
+recognition algorithms that have to run simultaneously on these small
+\emph{things}, to support our augmented reality.
 
 Another prime example of a class of algorithms that inhibit the same properties
-related to pattern recognition is Intrusion Detection (ID). The vast number of
-security related opportunities for attackers makes way for an equally sized
-amount of ID algorithms, all with a similar structure, but largely independent
-operation.
+is Intrusion Detection (ID). The large number of security related opportunities
+for attackers requires an equally sized amount of ID algorithms, all with a
+similar structure, still independent in their operation.
 
@@
 
-The contributions of this paper are threefold: The first contribution consists
+The contributions of this work are threefold: the first contribution consists
 in the formulation of a design pattern for detection/recognition algorithms for
@@
 context of resource-constrained devices. Section \ref{pattern} formulates a
-pattern common to the detection/recognition algorithms in the IoT context.
-Section \ref{design} introduces our framework, its design and components. In
-section \ref{evaluation} we apply our solution to case study related to ID,
-showing an actual implementation and evaluation of the experimental results.
-Section \ref{related} explores related work in the field of DSLs and code
-generation. Finally, section \ref{conclusion} summarizes our findings, draws
-conclusions and identifies topics for future work.
+pattern common to detection/recognition algorithms in the IoT context. Section
+\ref{design} introduces our framework, its design and components. In section
+\ref{evaluation} we apply it to a case study related to ID and evaluate the
+results. Section \ref{related} explores related work in the field of DSLs and
+code generation. Finally, section \ref{conclusion} summarizes our findings,
+draws conclusions and identifies topics for future work.
 
@@
 Simply taking existing implementations of algorithms and calling them in
-sequence, to allow them to perform their functionality, will in most cases
-result in suboptimal code. Three prime examples on the source code level
-illustrate this:
+sequence, will in most cases result in suboptimal code. Three prime examples on
+the source code level illustrate this:
 
-\vspace{-2mm}
-
-\begin{description}
+\begin{description}[style=unboxed,leftmargin=0cm]
 
@@
   network, processing the same byte stream, performing comparable operations to
-  find similar patterns. These loops can also hold calls to poll sensor for
-  their values. This results in an execution overhead, putting a strain on the
+  find similar patterns. Polling sensors for their value is another example.
+  This redundancy results in an execution overhead, putting a strain on the
   power consumption.
 
-\vspace{-2mm}
-
   \item[Common data] cross-cuts the different algorithms, that have to store
-  the same information over and over again within their own scope. This results
-  in a lot of duplicated information in memory, raising the hardware
-  requirements.
-
-\vspace{-2mm}
+  the same information over and over again within their own scope. Typical
+  examples here include information about neighboring nodes, sensor readings,
+  etc.\ This results in a lot of duplicated information in memory, raising the
+  hardware requirements.
 
@@
 
-\vspace{-2mm}
-
-\begin{description}
+\begin{description}[style=unboxed,leftmargin=0cm]
 
@@
 
-\vspace{-2mm}
-
   \item[The heterogeneity of IoT devices] results in different software stacks,
   each with for example their own network API\@. This partially explains why
-  there are few implementations of algorithms readily available to developers:
-  there is no single platform to target for researchers, which lowers the
-  opportunity to reuse prior work.
+  there are hardly any implementations of algorithms readily available to
+  developers: there is no single platform to target for researchers, which
+  lowers the opportunity to reuse prior work.
 
@@
 
-\section{A Design Pattern for Detection/Recognition Algorithms}
+\section{A design pattern for detection/recognition algorithms}
 \label{pattern}
@@
 as presented in section \ref{problems}, are in fact inherent to the design
-pattern that is the foundation for the type of algorithms we identify in
-typical detection/recognition applications in IoT applications. The formulation
+pattern that forms the foundation for the type of algorithms we identify in
+typical detection/recognition applications in an IoT context. The formulation
 of such a design pattern will allow us to define a strategy for the
 optimization of the combination of these algorithms. We start by extracting two
-common activities of these algorithms: the collection of sensor values and the
-evaluation of thresholds. Periodically collecting sensor values requires
-preparing the hardware for value extraction, waiting for the value to become
-available, reading the actual bytes, interpreting them into a meaningful value
-and recording them for further analysis. This can be as simple as reading a
-single byte, but can also be as complex as parsing a network message.
+common activities: the collection of sensor values and the evaluation of
+thresholds. Periodically collecting sensor values requires preparing the
+hardware for value extraction, waiting for the value to become available,
+reading the bytes from the sensor, interpreting these bytes and recording this
+interpretation for further analysis. This can be as simple as reading a single
+byte, but can also be as complex as parsing a network message.
 
-A single recorded sensor value hardly ever results in a decision. The
-combination, often over time, of several of these values and the evaluation of
-thresholds will be key to decide if a change in the environment requires an
-action. Updating such statistics and evaluating the corresponding thresholds
-should be distinct operations. Therefore this is a separate function that is
-also periodically executed.
+A single sensor value hardly ever results in a decision. The combination, often
+over time, of these values and the evaluation of thresholds will be key to
+decide if a change in the environment requires an action. Updating such
+statistics and evaluating the corresponding thresholds should be distinct
+operations. This decoupling of detection and decision is the core of the design
+pattern for detection algorithms. Listing \ref{alg:pattern} presents the design
+pattern in pseudo-code. Applying the pattern in the context of multiple
+algorithms, extends its scope and an application pattern such as described in
+listing \ref{alg:application} appears.
 
-This decoupling of detection and decision is the core of the design pattern for
-ID algorithms. Listing \ref{alg:pattern} presents the design pattern in
-pseudo-code. Applying the pattern in the context of multiple algorithms,
-extends its scope and a pattern such as described in listing
-\ref{alg:application} will typically appear. The pattern calls the individual,
-implementations in sequence in two situations: first, to collect sensor
-readings, and secondly, at a regular interval to allow for internal
-housekeeping, such as threshold evaluation.
+We see the three aforementioned problems manifest themselves: first, nesting of
+loops causes redundant executions of potentially the same sensor readings
+inside individual algorithms' $collect\_data$ functions. Second, each
+individual algorithm stores the same sensor values in memory. Third, scattered
+calls to network functions (in case of distributed algorithms) lead to
+suboptimal use of the wireless radio.
 
-We clearly see the three aforementioned problems manifest itself: first, the
-nesting of loops causes redundant executions of potentially the same sensor
-readings inside the individual algorithms' $collect\_data$ functions. Second,
-each individual algorithm stores the same sensor values in memory. Third,
-scattered calls to e.g. network functions (in case of distributed algorithms)
-lead to suboptimal use of the wireless radio.
-
-Implementing the solution from scratch, it would be possible to identify these
-problems and reorganize the code in such a way that sensor values are only read
-once, stored in memory once and network messages parsed once. In section
+When implementing the solution from scratch, it would be possible to identify
+these problems and reorganize the code in such a way that it reads and stores
+sensor values in memory once or centrally parses network messages. In section
 \ref{problems} we identified two operational problems that show that this is
-not feasible or at least suboptimal. We therefore present \FOO, a framework
+not feasible, or at least suboptimal. We therefore present \FOO, a framework
 that enables formally describing such algorithms and fusing common data and
@@
 
-\vspace{-2mm}
-
-\begin{algorithm}[b]
+\begin{algorithm}[t]
+  \small
   \caption{Detection/Recognition Algorithm Pattern}
@@
 
-\begin{algorithm}[b]
+\begin{algorithm}[t]
+  \small
   \caption{Application pattern of multiple algorithms}
@@
 The previous sections identified loops, common data and the network as three
-important code-level problems. These problems are inherent to detection
+important code level problems. These problems are inherent to detection
 algorithms, but conflict with the target environment, the IoT\@. \FOO provides
-a framework to efficiently implement these algorithms for the IoT and offers a
-solution for all identified problems. It consists of a 1) DSL, 2) source code
-generator and 3) software library. Figure \ref{fig:design} visualizes the
-design.
+a framework to efficiently implement these algorithms and offers a solution for
+all identified problems. It consists of a 1) DSL, 2) source code generator and
+3) software library.
 
-\begin{figure}[b]
+The DSL enables researchers to formally describe algorithms, without focusing
+on a specific language or platform. The code generator can avoid the identified
+problems concerning loops, common data and network access, because the formal
+description provides the underlying intent, not the technical implementation.
+It applies Functional Code Fusion (FCF) to identify hotspots for optimization.
+A key component of FCF are \emph{functional domains} that harbor common
+functionality.
+
+We will first position \FOO as a development strategy and motivate its design.
+Second, we will look at the principal design components of the DSL.\@ Third, we
+discuss the structure of the code generator and introduce its software library.
+
+\subsection{\FOO as a development strategy}
+\label{positioning}
+
+The introduction of a code generator in the development process changes the way
+of working: the source code is no longer the driving force, controlled by
+developers. It is literally an intermediate representation between code
+generator and compiler. A code generator constructs this representation using
+three components, as shown in figure \ref{fig:design}: 1) the description of
+the algorithm in the DSL, 2) information about the functional domains, and 3)
+information about the platform and target language. At the level of the
+compiler, a software library offloads specialized implementations and allows
+custom extensions. The applied grayscale indicates the components' platform
+dependency. This hierarchy is important from a development effort point of
+view. The level of dependency on the target platform dictates the level of
+effort to implement changes. This is important in relation to the scalability
+of the framework itself.
+
+\begin{figure}[t]
   \centering
-  \scalebox{0.75}{
+  \scalebox{0.70}{
 \begin{tikzpicture}[
@@
 
-The DSL enables researchers to formally describe algorithms, without focusing
-on any specific language or platform. A code generator can avoid the identified
-problems concerning loops, common data and network access, as the formal
-description provides the underlying intent, not the technical implementation.
-It can also target different languages and platforms, while reusing a software
-library that provides services including parsing, network access wrapping,\dots
-It applies the Functional Code Fusion (FCF) paradigm to identify the hotspots
-for optimization. A key component of FCF is the introduction of
-\emph{functional domains} that harbor common functionality.
-
-Using a DSL and code generator defines a development strategy. We will first
-position \FOO and motivate its design. Second, we will look at the principal
-design components of the DSL.\@ Third, we discuss the structure of the code
-generator and introduce its software library, that offers a way to deal with
-for example the network access issue.
-
-\vspace{-1mm}
-\subsection{\FOO as a Development Strategy}
-\label{positioning}
-
-The introduction of a code generator in the development process changes the way
-of working. The actual source code is no longer the driving force, controlled
-by developers. It is literally an intermediate representation between code
-generator and compiler. A code generator constructs this representation using
-three components, as shown in figure \ref{fig:design}: 1) the description of
-the algorithm in the DSL, 2) information about the functional domains, and 3)
-information about the platform and target language. At the level of the
-compiler, a software library offloads specialized implementations and allow
-custom extensions. The applied grayscale indicates the components' platform
-dependency. This hierarchy is important from a development effort point of
-view. The level of dependency on the target platform dictates the level of
-effort to implement changes. This is important in relation to the scalability
-of the framework itself.
-
-\vspace{-1mm}
-\subsection{\FOO as a Domain Specific Language}
+\subsection{\FOO as a domain specific language}
 \label{dsl-design}
@@
 do this, \FOO uses an event-based approach to defining functionality.
-Describing algorithms in function of events that occur will remove the
-technical descriptions and allows FCF to identify the actual intent of the
-algorithms with respect to the identified source code problems.
+Describing algorithms based on events removes the technical interpretations and
+allows FCF to identify the intent of algorithms and avoid the identified source
+code problems.
 
-\vspace{-1mm}
 \subsubsection{Loops}
@@
 
-\vspace{-1mm}
-\subsubsection{Common Data}
+\subsubsection{Common data}
 
@@
 domain specific storage and logic. For example in the case of distributed
-algorithms, such as ID algorithms, a \emph{Nodes} domain will manage
-information about neighbor nodes, exposing them as global objects to the
-algorithms, relieving them from managing these themselves. Functional domains
-are accessible as collections, allowing execution of functions in their scope,
-eliminating the need for explicit loops.
+algorithms a \emph{Nodes} domain will manage information about neighbor nodes,
+exposing them as global objects to the algorithms, relieving them from managing
+these themselves. Functional domains are also accessible as collections,
+allowing execution of functions in their scope, again eliminating the need for
+explicit loops.
 
-\vspace{-1mm}
-\subsubsection{Unified Messaging}
+\subsubsection{Unified messaging}
 \label{dsl-unified-msg}
@@
 tightly integrated with the previously mentioned concepts of event handling and
-for example the Nodes domain. An underlying software library can handle all
-interaction with the network, triggering functions by means of events when
-parsing new messages and exposing an API to send messages through the Nodes
-domain. This approach gives the framework complete control over communication
-and allows for optimized marshaling and grouping of messages.
+for example the \emph{Nodes} domain. An underlying software library can handle
+all interaction with the network, triggering functions by means of events when
+parsing messages and exposing an API to send messages. This approach gives the
+framework complete control over communication and allows for optimized
+marshaling and grouping of messages.
 
-\vspace{-1mm}
-\subsection{\FOO as a Code Generator}
+\subsection{\FOO as a code generator}
 
@@
 itself, can fulfill this role. Semantic transformations evolve the model into a
-more syntax-oriented model, comparable to the idea of CodeDom. From there on,
-syntactic transformations can further evolve the model into source code. Figure
-\ref{fig:code-generation} shows the process that \FOO implements.
+more syntax-oriented model, comparable to the idea of CodeDom. Syntactic
+transformations can further evolve the model into source code. Figure
+\ref{fig:code-generation} shows this process.
 
-\begin{figure}[b]
+\begin{figure}[t]
   \centering
-  \scalebox{.75}{
+  \scalebox{.70}{
 \begin{tikzpicture}[
@@
   \node (cm)   [representation,right=of sm,align=center]  {Syntactic\\Model};
-  \node (code) [resource,right=of cm]        {Code};
+  \node (code) [resource,right=of cm]                     {Source Code};
 
@@
                              node [midway,above=15pt,align=center] {FCF};
-  \draw [->] (cm)  -> (code) node [midway,above=15pt]              {emit code};
+  \draw [->] (cm)  -> (code) node [midway,above=15pt]              {emit};
 
@@
 
-\vspace{-1mm}
-\subsubsection{Semantic Model}
+\subsubsection{Semantic model}
 
@@
   \centering
-  \scalebox{.75}{
+  \scalebox{.70}{
 \begin{tikzpicture}[
@@
 \caption{Partial Semantic Model for \FOO.\@ Shaded classes indicate the
-\emph{event/scheduling} construction around functions to allow interpretation
-of the intended functionality.}
+\emph{event/scheduling} construction around functions.}
 \label{fig:meta-model}
@@
 
-\vspace{-1mm}
-\subsubsection{Functional Code Fusion (FCF)}
+\subsubsection{Functional code fusion (FCF)}
 
@@
 
-\vspace{-1mm}
-\subsubsection{Syntactic Model}
+\subsubsection{Syntactic model}
 
-The syntactic model is like an abstract syntax tree (AST) for a virtual,
-expressive language. It contains constructs and ideas found in different
-languages and programming paradigms. No single existing programming language
-supports all features as is. Syntactic transformations therefore translate the
-unsupported constructs into comparable constructs supported by the targeted
-platform and language. Depending on the target platform, some constructions
-aren't transformed into actually corresponding code. In selected cases, the
-generator uses a software library. For example when it needs to parse incoming
-messages or schedule tasks. During the generation process, it looks for
-patterns that indicate any of these actions and transforms them into calls to
-the \FOO software library.
+The syntactic model is like an abstract syntax tree for a virtual, expressive
+language. It contains constructs and ideas found in different languages and
+programming paradigms. No single existing programming language supports all
+features as is. Syntactic transformations therefore translate the unsupported
+constructs into comparable constructs supported by the targeted platform and
+language. Depending on the target platform, some constructions aren't
+transformed into actually corresponding code. In selected cases, the generator
+uses a software library. For example when it needs to parse incoming messages
+or schedule tasks. During the generation process, it looks for patterns that
+indicate any of these actions and transforms them into calls to the \FOO
+software library.
 
-\vspace{-1mm}
-\subsection{\FOO as a Software Library}
+\subsection{\FOO as a software library}
 \label{software-lib-design}
@@
 parsing of incoming messages and grouping of outgoing messages. More common
-functionality isn't generated verbatim either. The execution of functions based
-on a schedule or the implementation of higher level conceptual functions are
-calls into the standard software library. Optimization of such algorithms is
-not within the scope of \FOO.\@ This gives freedom of choice to the users of
-the framework to select appropriate implementations for this standard
-functionality.
+functionality isn't generated verbatim either. A standard software library
+provides functionality for the execution of functions based on a schedule or
+the implementation of higher level conceptual functions. Optimization of these
+type of algorithms is not within the scope of \FOO.\@ This further gives
+freedom of choice to the users of the framework to select appropriate
+implementations for this standard functionality.
 
-\section{Implementation and Evaluation: Case Study on Intrusion Detection in Wireless Sensor Networks}
+\section{
+Implementation and evaluation:
+a case study on intrusion detection in wireless sensor networks
+}
 \label{evaluation}
@@
 a single point in the network to oversee all traffic
-\cite{mishra2004intrusion}. Every device has to implement its own IDS.
+\cite{mishra2004intrusion}. So every device has to implement its own IDS.\@
 
@@
 attack. For each of these attacks an algorithm needs to look for patterns or
-anomalies. An IDS therefore requires a many algorithms to achieve adequate
+anomalies. An IDS therefore requires a lot of algorithms to achieve adequate
 coverage.
 
-\vspace{-1mm}
 \subsection{Implementation}
@@
 We started from a basic application that measures light intensity. The
-application was an includable piece of code, allowing integration in a both a
+application was an includable piece of code, allowing integration in both a
 manual implementation and generated code base. On top of this baseline we added
@@
 
-\vspace{-1mm}
-\subsection{Analytical Results}
+\subsection{Analytical results}
 
@@
 
-\vspace{-1mm}
-\subsection{Experimental Results}
+\subsection{Experimental results}
 
@@
 
-\begin{table}[h]
-  \caption{Experimental Results and Side-by-Side Comparison.}
+\begin{table}[ht]
+  \caption{Experimental results and side-by-side comparison.}
   \label{tbl:results}
@@
 
-These results show that adding algorithms manually to the base-application has
-a cumulative effect. In case of the generated implementation, this effect is no
+It is clear that adding algorithms manually to the base-application has a
+cumulative effect. In case of the generated implementation, this effect is no
 longer observed: individual algorithms do have a larger impact by themselves,
@@
 
-Table \ref{tbl:results} also compares the two situations, showing the impact of
-the code generation: Both the number of packets as well as the amount of bytes
-sent have dropped by 13\% and 7\%. In case of the execution time, both
-algorithms by themselves add an equal amount of time to a single event loop
-cycle. When combined, the event loop is about 7\% faster compared to the manual
-case. Finally, we see that the software library adds to the size of the
-resulting image. However, the cost is close to constant and is lower for both
-algorithms together. With roughly 3KB of additional overhead, or 15\% in this
-simple case, the impact is small. Taking into account the available memory on
-the ATMega1284p, this is only 2.3\% of the available resources.
+Table \ref{tbl:results} also compares the manual versus de generated versions,
+showing the impact of the code generation: both the number of packets as well
+as the amount of bytes sent have dropped by 13\% and 7\%. In case of the
+execution time, both algorithms by themselves add an equal amount of time to a
+single event loop cycle. When combined, the event loop is about 7\% faster
+compared to the manual case. Finally, we see that the software library adds to
+the size of the resulting image. However, the cost is close to constant and is
+lower for both algorithms together. With roughly 3KB of additional overhead, or
+15\% in this simple case, the impact is small. Taking into account the
+available memory on the ATMega1284p, this is only 2.3\% of the available
+resources.
 
-\section{Related Work}
+\section{Related work}
 \label{related}
@@
 \cite{sekar1999high,roesch1999snort} in a concise way. The STAT framework with
-its STATL language \cite{eckmann2002statl,vigna2003designing} allow to model
-scenarios using state machines. Our event-driven approach shares these same
-ideas. Although not a real DSL, nesC \cite{gay2003nesc} is a general purpose
-language that shares some typical concepts with \FOO, such as the high degree
-of analysis during code generation to optimize execution. The biggest
-difference is that \FOO starts from the intended functionality and can optimize
-certain constructions that can't be interpreted from more technical source code.
+its STATL language \cite{eckmann2002statl,vigna2003designing} allows modeling
+scenarios using state machines. Our event-driven approach shares this idea.
+Although not a real DSL, nesC \cite{gay2003nesc} is a general purpose language
+that shares some typical concepts with \FOO, such as the high degree of
+analysis during code generation to optimize execution. The biggest difference
+is that \FOO starts from the intended functionality and can optimize certain
+constructions that can't be interpreted from more technical source code.
 
@@
 consisting of a virtual machine (VM) and native code extension
-\cite{sadilek2007energy}, combined with energy-aware compilation, seems a
-promising approach. By offloading the optimization to a VM and only compiling
-parts and not the whole, the solution allows updating at a granular scale, thus
-reducing the update cost, while maintaining flexibility and lowering the
-development effort. In this terminology, \FOO focuses on the reduction of the
-development effort and update cost, and maximizes flexibility.
+\cite{sadilek2007energy}, combined with energy-aware compilation, seems a valid
+approach. By offloading the optimization to a VM and only compiling parts and
+not the whole, the solution allows updating at a granular scale, thus reducing
+the update cost, while maintaining flexibility and lowering the development
+effort. In this terminology, \FOO focuses on the reduction of the development
+effort and update cost, and maximizes flexibility.
 
-\section{Conclusion and Future Work}
+\section{Conclusion and future work}
 \label{conclusion}
@@
 total number of bytes sent over the network is reduced by 7\%. The execution
-time is improved by 7\% and memory usage can be reduced by even 20\%.
+time is improved by 7\% and memory usage can be reduced even by 20\%.
 

### 2015-04-17 squeeze.tex
completed text clean-up, squeezed whitespace

% white space sqeezing ;-)

\setlength\floatsep{0.5\baselineskip plus 3pt minus 2pt}
\setlength\textfloatsep{0.5\baselineskip plus 3pt minus 2pt}
\setlength\intextsep{0.5\baselineskip plus 3pt minus 2pt}

\usepackage{titlesec}
\titlespacing\section{0pt}{5pt plus 4pt minus 2pt}{5pt plus 2pt minus 2pt}
\titlespacing\subsection{0pt}{5pt plus 4pt minus 2pt}{3pt plus 2pt minus 2pt}
\titlespacing\subsubsection{0pt}{5pt plus 4pt minus 2pt}{2pt plus 2pt minus 2pt}
\titlespacing\item{0pt}{5pt plus 4pt minus 2pt}{5pt plus 2pt minus 2pt}

\usepackage{enumitem}
\setlist{nolistsep}

### 2015-04-17 paper-euspn15.tex
final reading before draft release

@@
 
-The ability to detect patterns in its usage is what makes \emph{things}
-\emph{smart}. The more patterns it can detect, the more accurate its
-operational results will be. However, supporting an adequate set of pattern
+The ability to detect patterns in its application context is what makes
+\emph{things} \emph{smart}. The more patterns it can detect, the more accurate
+its operational results will be. However, supporting an adequate set of pattern
 detection algorithms imposes significant overhead on resource-constrained
@@
 \emph{smart things}: from smart watches that assist visually impaired
-persons\cite{porzi2013smart} to monitoring systems for waste management in our
+persons\cite{porzi2013smart} to monitoring systems for waste management in
 \emph{smart cities}\cite{zanella2014internet}. Their applications cannot be
@@
 recognition algorithms that have to run simultaneously on these small
-\emph{things}, to support our augmented reality.
+\emph{things}.
 
 Another prime example of a class of algorithms that inhibit the same properties
-is Intrusion Detection (ID). The large number of security related opportunities
+is intrusion detection (ID). The large number of security related opportunities
 for attackers requires an equally sized amount of ID algorithms, all with a
@@
 generator producing source code for a given platform and configuration. This
-way, \FOO also addresses the heterogeneous nature of both IoT and the
+way, \FOO also addresses the heterogeneous nature of both the IoT and the
 algorithms. The third contribution introduces functional code fusion (FCF) as a
@@
 
-  \item[Reimplementing] algorithms manually to eliminate these issues is
-  not a one-time cost. A release of a new version of an algorithm, or simply a
-  reselection of algorithms, due to a changing requirements, can trigger it.
+  \item[Reimplementing] algorithms manually to eliminate these issues is not a
+  one-time cost. A release of a new version of an algorithm, or simply a
+  reselection of algorithms, due to changing requirements, can trigger it.
   Also, changing, or simply implementing the algorithms from scratch, holds the
@@
   \small
-  \caption{Detection/Recognition Algorithm Pattern}
+  \caption{Detection/recognition algorithm pattern}
   \label{alg:pattern}
   \begin{algorithmic}
-    \Require{readings: global variables to hold collected sensor data for the given algorithm}
-    \Require{devices: global variable to information about neighbor devices} \Comment{optional in case of distributed algorithms}
+    \Require{readings: global variable to hold collected sensor data for the given algorithm}
+    \Require{devices: global variable to store information about neighbor devices} \Comment{optional in case of distributed algorithms}
     \Function{collect\_data}{}
@@
 compiler, a software library offloads specialized implementations and allows
-custom extensions. The applied grayscale indicates the components' platform
+for custom extensions. The applied grayscale indicates the components' platform
 dependency. This hierarchy is important from a development effort point of
@@
 result in problems concerning loops, common data and the use of the network. To
-do this, \FOO uses an event-based approach to defining functionality.
-Describing algorithms based on events removes the technical interpretations and
-allows FCF to identify the intent of algorithms and avoid the identified source
-code problems.
+do this, \FOO uses an event-based approach to define functionality. Describing
+algorithms based on events removes the technical interpretations and allows FCF
+to identify the intent of algorithms and avoid the identified source code
+problems.
 
@@
 algorithms a \emph{Nodes} domain will manage information about neighbor nodes,
-exposing them as global objects to the algorithms, relieving them from managing
-these themselves. Functional domains are also accessible as collections,
-allowing execution of functions in their scope, again eliminating the need for
-explicit loops.
+exposing them as objects to the algorithms, relieving them from managing these
+themselves. Functional domains are also accessible as collections, allowing
+execution of functions in their scope, again eliminating the need for explicit
+loops.
 
@@
 
-To convert the DSL-based ID algorithms into production-ready code, \FOO applies
+To convert the DSL-based algorithms into production-ready code, \FOO applies
 code generation techniques. Although different approaches are applicable, like
@@
 
-\subsubsection{Functional code fusion (FCF)}
+\subsubsection{Functional code fusion}
 
@@
 provides functionality for the execution of functions based on a schedule or
-the implementation of higher level conceptual functions. Optimization of these
+the implementation of higher level conceptual functions. Optimization of this
 type of algorithms is not within the scope of \FOO.\@ This further gives
@@
 We also instrumented the code and evaluated four criteria: network usage
-(number of packets and bytes), the time required to perform once cycle of the
+(number of packets and bytes), the time required to perform one cycle of the
 event loop and the resulting image size. We collected these metrics for four
@@
 
-Many IoT applications typically are constructed of similar algorithms that
-adhere to a pattern of collecting sensor data and periodically checking
-recorded values for patterns that require action. Fusing these algorithms is
-hard but needed to deal with the resource-constrained devices they are
-targeting. Frequent reselection, e.g. due to changing requirements or changing
-security threats, requires a flexible way to incorporate these algorithms in
-subsequent iterations of the software.
+IoT applications are typically constructed of similar algorithms that adhere to
+a pattern of collecting sensor data and periodically checking recorded values
+for patterns that require action. Fusing these algorithms is hard but needed to
+deal with the resource-constrained devices they are targeting. Frequent
+reselection, e.g. due to changing requirements or changing security threats,
+requires a flexible way to incorporate these algorithms in subsequent
+iterations of the software.
 
@@
 
-Initial experiments confirm the theoretically positive impact on resource
-usage: grouping of messages lowers the number of packets by 13\%, while the
-total number of bytes sent over the network is reduced by 7\%. The execution
-time is improved by 7\% and memory usage can be reduced even by 20\%.
+Initial experiments confirm the expected positive impact on resource usage:
+grouping of messages lowers the number of packets by 13\%, while the total
+number of bytes sent over the network is reduced by 7\%. The execution time is
+improved by 7\% and memory usage can be reduced by 20\%.
 
